### 2011-09-20 survey.tex
first import, sections + initial bib

\documentclass[journal]{IEEEtran}

\usepackage{cite}

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi

\usepackage{graphicx}
\usepackage[cmex10]{amsmath}
\usepackage{algorithmic}
\usepackage{array}
\usepackage{mdwmath}
\usepackage{mdwtab}
\usepackage{eqparbox}
%\usepackage[tight,footnotesize]{subfigure}
%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
%\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{fixltx2e}
\usepackage{stfloats}
\usepackage{url}

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{TODO: A Survey of StarCraft AI Techniques}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{FirstName~LastName,~\IEEEmembership{Member,~IEEE,}
        Jim~Raynor,~\IEEEmembership{Fellow,~RR,}
        and~Sarah~Kerrigan,~\IEEEmembership{Life~Fellow,~ZS}% <-this % stops a space
\thanks{FirstName~LastName is with the Department of Names
GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
\thanks{J. Raynor and S. Kerrigane are with the Romeo\&Juliet Inc.}% <-this % stops a space
\thanks{Manuscript received April 19, 2499; revised January 11, 2500.}}

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

% The paper headers
\markboth{TCIAIG ~Vol.~X, No.~Y, Month~Year}%
{Shell \MakeLowercase{\textit{et al.}}: TODO: Title here}

\maketitle

\begin{abstract}
TODO
\end{abstract}

\begin{IEEEkeywords}
review, RTS, StarCraft, machine learning, planning, TODO ...

\end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}
\IEEEPARstart{S}{tarcraft} AI competitions have caused many AI techniques to be
applied to RTS AI. We will list and classify these approaches, explain their 
power and their downsides and conclude on what is left to achieve human-level 
RTS AI. TODO (test \cite{WeberCig10})

\section{Challenges, why is it hard to do a good RTS AI?}

\section{AI Techniques review}
\subsection{Overview}
\subsection{Case study 1: EISBot}
\subsection{Case study 2: NOVA}
\subsection{Case study 3: BroodwarBotQ}

%\begin{figure}[!t]
%\centering
%\includegraphics[width=4cm]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results}
%\label{fig_sim}
%\end{figure}

% Note that IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command, the
% \label for the overall figure must come after \caption.
% \hfil must be used as a separator to get equal spacing.
% The subfigure.sty package works much the same way, except \subfigure is
% used instead of \subfloat.
%
%\begin{figure*}[!t]
%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
%\label{fig_second_case}}}
%\caption{Simulation results}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

\section{Discussion, what is left to do / open challenges}

\section{Conclusion}

\section*{Acknowledgment}

% Can use something like this to put references on a page
% by themselves when using endfloat and the captionsoff option.
\ifCLASSOPTIONcaptionsoff
  \newpage
\fi

%\begin{thebibliography}{1}
%\end{thebibliography}

\bibliographystyle{IEEEtran}                                                    
\bibliography{survey}

% biography section
% 
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{biography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

\begin{IEEEbiographynophoto}{FirstName LastName}
Biography text here.
\end{IEEEbiographynophoto}

\begin{IEEEbiography}[{\includegraphics[width=2cm, keepaspectratio]{jim.jpg}}]{Jim Raynor}
Biography text here.
\end{IEEEbiography}

% if you will not have a photo at all:

% insert where needed to balance the two columns on the last page with
% biographies
%\newpage

% You can push biographies down or up by placing
% a \vfill before or after them. The appropriate
% use of \vfill depends on what kind of text is
% on the last page and whether or not the columns
% are being equalized.

%\vfill

% Can be used to pull up biographies so that the bottom of the last one
% is flush with the other column.
%\enlargethispage{-5in}

% that's all folks
\end{document}

### 2011-09-20 survey.tex
biography test

@@
 %\end{thebibliography}
-
 \bibliographystyle{IEEEtran}                                                    
@@
 
-% biography section
-% 
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{biography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
-\begin{IEEEbiographynophoto}{FirstName LastName}
-Biography text here.
-\end{IEEEbiographynophoto}
+%\begin{IEEEbiographynophoto}{FirstName LastName}
+%Biography text here.
+%\end{IEEEbiographynophoto}
 
 \begin{IEEEbiography}[{\includegraphics[width=2cm, keepaspectratio]{jim.jpg}}]{Jim Raynor}
-Biography text here.
+Jim Raynor was a Confederate marshal on Mar Sara at the time of the first zerg incursions on that world. He is now with Raynor's Raiders Inc.
 \end{IEEEbiography}
 
-% if you will not have a photo at all:
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-% You can push biographies down or up by placing
-% a \vfill before or after them. The appropriate
-% use of \vfill depends on what kind of text is
-% on the last page and whether or not the columns
-% are being equalized.
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
-
-% that's all folks
 \end{document}

### 2011-10-09 survey.tex
Started writing a section on overall architecture of bots (still WIP)

@@
 
-\usepackage{cite}
+\usepackage{cite,color,graphicx}
 
@@
 
+{\color{red} Here we can use Buro's 2003 paper as a starting point. Much has changed since, so we should update, and put his predictions in perspective}
+
 \section{AI Techniques review}
@@
 
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=4cm]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation Results}
-%\label{fig_sim}
-%\end{figure}
+\section{AI Architectures for RTS AI}\label{sec:architecture}
 
-% Note that IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
+{\color{red}[SANTI: This section is WIP, so, don't read it yet :)]}
 
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command, the
-% \label for the overall figure must come after \caption.
-% \hfil must be used as a separator to get equal spacing.
-% The subfigure.sty package works much the same way, except \subfigure is
-% used instead of \subfloat.
-%
-%\begin{figure*}[!t]
-%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-%\label{fig_second_case}}}
-%\caption{Simulation results}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
+Playing an RTS game involves dealing with all the problems described above.
+A few approaches, like CAT \cite{aha2005cat}, Darmok \cite{ontanon2010darmok} or XXX \cite{some-rl-system} ({\color{red}Which was that system by Stuart Russell that played Warcraft?}) try to deal with the problem in a monolithic manner, by using a single AI technique.
+This resembles approaches to solve other games, such as Chess or Go, where a single game-tree search approach is enough to play the game at human level.
+However, none of those systems aims at achieving near human performance.
+In order to achieve human-level performance, RTS AI designers use a lot of domain knowledge in order to divide the task of playing the game into a collection of sub-problems, which can be dealt-with using individual AI techniques (as discussed in the previous section).
+Thus, an integration architecture is required to put together all of those techniques into a single coherent system, which is the focus of this section.
 
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
+...
+
+\begin{figure}[ta]
+%    \centering
+    \includegraphics[width=\columnwidth]{figures/figure-bot-architectures.pdf}
+    \caption{General architecture of 5 Starcraft AI bots. {\color{red} (This figure is unreadable as of now, but we'll figure out a way to better present this info)}}
+    \label{fig:bot-architecture}
+\end{figure}
+
+{\color{red} Present some representative examples (Figure \ref{fig:bot-architecture}), which have slightly different architectures.
+I could mention THANATOS as something different using learning, and Darmok as a monolithic architecture that aims at learning how to play the game, without hoping to play at human level.}
+
+{\color{red} Then, draw some general conclusions: 
+\begin{itemize}
+\item two types of aggregation, hierarchical, and decentralized (bidding?). Hierarchical is multiscale, with actions at different levels of abstraction.
+\item high level strategy is hard-coded (unlike in games like Chess or Go): different expressivity of the high-level strategy of the different systems, compare them.
+It's like a very high level programming language for RTS bots.
+\end{itemize}}
 

### 2011-10-18 survey.tex
Completed a first pass on the architecture section

@@
 \section{AI Techniques review}
+
 \subsection{Overview}
@@
 
-{\color{red}[SANTI: This section is WIP, so, don't read it yet :)]}
-
 Playing an RTS game involves dealing with all the problems described above.
-A few approaches, like CAT \cite{aha2005cat}, Darmok \cite{ontanon2010darmok} or XXX \cite{some-rl-system} ({\color{red}Which was that system by Stuart Russell that played Warcraft?}) try to deal with the problem in a monolithic manner, by using a single AI technique.
+A few approaches, like CAT \cite{aha2005cat}, Darmok \cite{ontanon2010darmok} or ALisp \cite{marthi2005wargus} try to deal with the problem in a monolithic manner, by using a single AI technique.
 This resembles approaches to solve other games, such as Chess or Go, where a single game-tree search approach is enough to play the game at human level.
@@
 Thus, an integration architecture is required to put together all of those techniques into a single coherent system, which is the focus of this section.
-
-...
 
@@
 
-{\color{red} Present some representative examples (Figure \ref{fig:bot-architecture}), which have slightly different architectures.
-I could mention THANATOS as something different using learning, and Darmok as a monolithic architecture that aims at learning how to play the game, without hoping to play at human level.}
+Figure \ref{fig:bot-architecture} shows some representative examples of the integration architectures used by different bots in the AIIDE 2011 Starcraft AI competition \cite{???}.
+Each box represents an individual module with a clearly defined task (only modules with a black background can send actions directly to Starcraft).
+Dashed arrows represent data flow, and solid arrows represent control (when a module can command another module to perform some task).
+For example, we can see how SPAR is divided in two sets of modules: {\em situation analysis} and {\em decision making}, the first with three modules dedicated to analyze the current situation of the game, and the later with 4 modules dedicated to exploit that information to decide what to do.
+We can see how the decision making aspect of SPAR is organized hierarchically, with the higher-level module ({\em strategic decision}) issuing commands to the next module ({\em tactical decision}), which sends commands to the next module ({\em action implementation}), and so on.
+Only the lower-level modules can send actions directly to Starcraft.
 
-{\color{red} Then, draw some general conclusions: 
+On the other hand, bots such as NOVA or BroodwarBotQ (BBQ) only use a hierarchical organization for {\em micro-management} (controlling the attack units), but use a decentralized organization for the rest of the bot.
+In Nova and BBQ, there is a collection of modules that control different aspects of the game (workers, production, construction, etc.).
+These modules can all send actions directly to Starcraft.
+In Nova those modules coordinate only through writing data in a shared blackboard, and in BBQ they coordinate only when they have to use a shared resource (unit) by means of an arbitrator.
+
+By analyzing the structure of these bots, we can see that there are two main tools that can be used when designing an integration architecture:
+
 \begin{itemize}
-\item two types of aggregation, hierarchical, and decentralized (bidding?). Hierarchical is multiscale, with actions at different levels of abstraction.
-\item high level strategy is hard-coded (unlike in games like Chess or Go): different expressivity of the high-level strategy of the different systems, compare them.
-It's like a very high level programming language for RTS bots.
-\end{itemize}}
+\item {\em Abstraction}: complex tasks can be formulated at different levels of abstraction.
+For example, playing an RTS game can be seen as issuing individual low-level actions to each of the units in the game, or at a higher level, it can be seen as deploying a specific strategy (e.g.
+a ``BBS strategy'', or a ``Reaver Drop'' strategy).
+Some bots, reason at multiple levels of abstraction at the same time, making the task of playing Starcraft simpler.
+Assuming that each module in the architecture of a bot has a goal and determines some actions to achieve that goal, the actions determined by higher-level modules are considered as the goals of the lower level modules.
+In this way, each module can focus on reasoning at only one level of abstraction, thus, making the problem easier.
+
+\item {\em Divide-and-conquer}: playing a complex RTS, such as Starcraft, requires performing many conceptually different tasks, such as gathering resources, attacking, placing buildings, etc.
+Assuming each of these tasks can be performed relatively independently and without interference, we can have one module focusing on each of the tasks independently, thus making the problem easier.
+\end{itemize}
+
+If we imagine the different tasks to perform in a complex RTS game in a two-dimensional plane, where the vertical axis represents abstraction, and the horizontal axis represents the different aspects of the game (micro-management, resource gathering, etc.), abstraction can be seen as dividing the space with horizontal lines, whereas divide-and-conquer divides the space using vertical lines.
+
+Different bots, use different combinations of these two tools. Looking back at Figure \ref{fig:bot-architecture}, we can see the following use of abstraction and divide-in-conquer in the bots:
+
+\begin{itemize}
+\item BTHAI: uses a two-tier abstraction hierarchy, where a collection of high-level modules command a collection of lower-level agents in charge of each of the units.
+At the high-level, BTHAI uses divide-and-conquer, having multiple high-level modules issuing commands to the lower-level units.
+\item BBQ: uses abstraction for micro-management, and divide-and-conquer for macro-management and intelligence gathering.
+To avoid conflicts between modules (since the individual tasks of each of the modules are not completely independent), BBQ uses an arbitrator.
+\item Nova: is similar in design as BBQ, and uses abstraction for micro-management, and divide-and-conquer for macro-management.
+The differences are that Nova does not have an arbitrator to resolve conflicts, but has a higher-level module ({\em strategy manager}), which posts information to the blackboard that the rest of modules follow (thus, making use of abstraction).
+\item SPAR: only uses abstraction.
+Its high-level module determines the strategy to use, and the tactical decision module divides it into a collection of {\em abstract actions}, that are executed by the lower-level modules.
+\item Skynet: makes extensive use of both abstraction and divide-and-conquer.
+Considering the decision making component of Skynet, we can see a high level module that issues commands to a series of tactics modules.
+The collection of tactic modules queue {\em tasks} (that are analogous to the abstract actions used in SPAR).
+Each different task has a specific low level module that knows how to execute it.
+Thus, Skynet uses a 3 layered abstraction hierarchy, and uses divide-and-conquer in all levels except the highest.
+\end{itemize}
+
+Additionally, except for BTHAI, all other agents use divide-and-conquer at a higher-level bot design and divide all the modules into two or three categories: {\em information gathering} and {\em decision}, or {\em information gathering}, {\em micro-management} and {\em macro-management}.
+
+Notice that most bots using divide-and-conquer (except for BBQ), assume that each of the modules can act independently and that their actions can be executed without interference.
+BBQ is the exception, including an arbitrator that makes sure that modules do not send contradictory orders to the same unit.
+However, very little bots handle the problem of how to coordinate resource usage amongst modules, for instance BTHAI uses a first-come-first-serve policy for spending resources, the first module that requests resources is the one that gets them.
+The exceptions are Nova and Skynet implement some rudimentary prioritization based on the high level strategy.
+
+One interesting aspect of the five bots described above is that, while all of them are reactive at the lower level (unit control), most if not all of them, are scripted at the highest level of abstraction.
+BTHAI reads build and squad formations from a predefined script, Nova's {\em Strategy Manager} is a predefined finite-state machine, BBQ's construction manager reads the build order from a predefined script, and Skynet's {\em BuildOrder Manager} is basically a predefined script.
+Such scripts describe the strategy that the bots will use, however, such strategy is always fixed.
+One could see this pre-scripting as if each bot defined a ``high-level programming language'' to describe Starcraft strategies, and the bots themselves are just interpreters of such strategy.
+Compared to current approaches for Chess or Go, this scripting seems a bit rigid and inflexible, but responds to the much higher complexity of the Starcraft game.
+
+In conclusion, we can see that there are two basic tools that can be used in an integration architecture: abstraction and divide-and-conquer, which are widely used by the existing Starcraft bots.
+However, several open questions remain:
+
+{\color{red}
+\begin{itemize}
+\item What is the best task decomposition for RTS games?
+\item Resource coordination in a divide-and-conquer approaches.
+\item Reactive high-level strategies that do not depend on scripting.
+\item ...
+\end{itemize}
+}
 

### 2011-11-17 survey.tex
Tentative structure of the paper
I've added a tentative structure of the paper with sections and
subsections. Let's discuss it.

@@
 \begin{abstract}
-TODO
+TODO: Idea of the paper is: ``one-stop guide on what is the
+state of the art in Starcraft AI''. It should help people participating in the competition focus
+their efforts, and also should help people implementing AI for RTS games in
+general (e.g. industry).  In Gabriel's words ``RTS AI problems, Solutions, State-of-the-art, conclude on what's "solved" (since Buro 2004) and what's not.''
+
+For example, if someone wants to implement a bot, and wonders "how should I do scouting", our paper should provide a summary of the existing techniques, and pointers to know more. 
 \end{abstract}
@@
 \begin{IEEEkeywords}
-review, RTS, StarCraft, machine learning, planning, TODO ...
+review, RTS, games, StarCraft, machine learning, planning, TODO ...
 
@@
 
-\section{Introduction}
-\IEEEPARstart{S}{tarcraft} AI competitions have caused many AI techniques to be
+\section{Introduction}\label{sec:intro}
+\IEEEPARstart{S}{ince} Michael Buro's call for research in RTS games \cite{Buro03rts}, many researchers have answered the call.
+Specially, AI competitions like the Starcraft one have caused many AI techniques to be
 applied to RTS AI. We will list and classify these approaches, explain their 
 power and their downsides and conclude on what is left to achieve human-level 
-RTS AI. TODO (test \cite{WeberCig10})
+RTS AI.
 
-\section{Challenges, why is it hard to do a good RTS AI?}
+{\color{blue}
+Motivate the paper, and provide an outline.
 
-{\color{red} Here we can use Buro's 2003 paper as a starting point. Much has changed since, so we should update, and put his predictions in perspective}
+Some arguments to use in the motivation could be that games are a good application to motivate novel AI research (as has been happening throughout the history of AI), and that techniques and algorithms developed for RTS games, in addition to be useful and relevant to the game industry, have broader application to other areas.
 
-\section{AI Techniques review}
+Reiterate that the goal of this paper is to provide a one-stop guide on what is the
+state of the art in Starcraft AI
+}
 
-\subsection{Overview}
-\subsection{Case study 1: EISBot}
-\subsection{Case study 2: NOVA}
-\subsection{Case study 3: BroodwarBotQ}
+\section{Real-Time Strategy Games}\label{sec:rts}
 
-\section{AI Architectures for RTS AI}\label{sec:architecture}
+\subsection{Starcraft}\label{subsec:starcraft}
+
+{\color{blue}
+Introduce the game plus screenshot. Not too lengthy
+}
+
+\subsection{Complexity of RTS Game AI}\label{subsec:complexity}
+
+{\color{blue}
+Here discuss the complexity of RTS games in general, but focusing on Starcraft.
+The most important thing of this section is to demonstrate that RTS games are complex, and this is the motivation for requiring all the challenges and research presented afterwards.
+}
+
+\subsection{Challenges in RTS Game AI}\label{subsec:challenges}
+
+{\color{blue}
+Many years have passed since Buro's call for research (8!!). he identified 6 challenges:
+\begin{itemize}
+\item Resource management
+\item Decision making under uncertainty
+\item Spatial and temporal reasoning
+\item Collaboration
+\item Opponent modeling, Learning
+\item Adversarial real-time planning
+\end{itemize}
+
+There has been a lot of work in many, but others have been untouched (e.g.
+Collaboration).
+Additionally, other challenges that are not in the list appeared, and have been worked on, like: how to exploit the existing domain knowledge (strategies, build-orders, replays, etc.), or how to design an architecture that integrates all the modules required for an agent to play an RTS games.
+This section should provide an updated view of challenges, and set the ground for the rest of the paper.
+The remainder of the paper should be structured around the challenges that we list here.
+
+I think that a good list could be something like this (I'm open to all kind of criticism here :)):
+
+\begin{itemize}
+\item Task Decomposition (or ``Architecture'')
+\item Integration of Domain Knowledge
+\item Reasoning with Uncertainty (including information gathering)
+\item Opponent Modeling and Adaptation
+\item Group and Individual Control (``micro'')
+\item Planning and Resource Allocation (``macro'')
+\end{itemize}
+
+I've now listed them as challenges, but maybe we can list them as questions.
+In each of the subsequent sections, we should use the bots in the competitions (BBQ, EISBot, Nova, UAlbertaBot, etc.) to illustrate the state-of-the-art.
+Eventually, we could refer to other work though if need be.
+We could also compare with what is being used in industry, and use the real AI of Starcraft as an example (Bob Fitch's talk in AIIDE provides enough material for that).
+}
+
+\section{Challenge 1: Task Decomposition}\label{sec:architecture}
+
+{\color{blue}
+Here we should say that complex RTS games require multi-scale reasoning \cite{WeberCig10}, and thus complex architectures that integrate many AI modules have been devised.
+Discuss how different researchers have divided up the task of playing Starcraft into a collection of subtasks and integrated them (see the end of this paper for a preliminary write-up on this)
+}
+
+\section{Challenge 2: Integration of Domain Knowledge}\label{sec:domainknowledge}
+
+{\color{blue}
+All bots attempt at integrating human domain knowledge in a way or another, be it in the form of build-orders, heuristics, strategies, or by automatically learning from human replays.
+This has been shown to be key to good performance in bots (for now, as Dave was saying: ``Starcraft AI is like Chess was in the 60s'').
+
+Talk about learning, scripting, and other techniques to incorporate such knowledge.
+}
+
+\section{Challenge 3: Reasoning with Uncertainty}\label{sec:uncertainty}
+
+{\color{blue}
+I think there are two important aspects of uncertainty in RTS games.
+
+First is trying to model uncertainty and try to reduce it (information gathering, and representation of knowledge), this includes scouting, and maintaining good estimates of locations and number of units of the enemy, etc.
+
+Second is actually using the uncertain information. Taking decisions like army strength estimation (like to attack or retreat).
+}
+
+\section{Challenge 4: Opponent modeling and Adaptation}\label{sec:adaptation}
+
+{\color{blue}
+What has been done in modeling the strategy of the opponent and adapt accordingly? which bots do adapt themselves to what they learn from the opponent and how?
+}
+
+{\color{red} I'm a bit worried that Challenge 3 and 4 will overlap significantly, since a large source of uncertainty is the opponent, so, part of opponent modeling has a lot to do with uncertainty.
+What do you think?}
+
+\section{Challenge 5: Group and Individual Control}\label{sec:micro}
+
+{\color{blue}
+One of the big areas where we have seen a lot of research is in micro-management, where participants in the competitions have used reinforcement-learning (Overmind), potential-fields (BTHAI, Nova, and others), or many other techniques for coordinating both individual unit and group behavior.
+}
+
+\section{Challenge 6: Planning and Resource Allocation}\label{sec:planning}
+
+{\color{blue}
+This section is intended to refer to all those techniques concerning ``macro'', since resource allocation and planning is considered together in most state-of-the-art bots (do you agree?).
+Here we can talk about things like build-order-optimization, resource allocation to different aspects of the game (army, research, etc.)
+}
+
+\section{Empirical Comparison of RTS Bots}\label{sec:experiments}
+
+{\color{blue}
+What are the similarities and differences of the current bots in the competition?
+which techniques are widespread, and which are experimental?
+which features decide over winning?
+Which features would have a high impact if we work on them?
+}
+
+{\color{red} Given that we have Ben, Mike (Preuss), Dave and possibly Michael (Buro) on board, we could include results from the competitions.
+But, given that this is not a standard ``competition paper'', maybe you guys want to reserve that data to put together a separate paper.
+It's your call, since it's your data :)}
+
+\section{Discussion and Outlook}\label{sec:conclusions}
+
+{\color{blue}	
+what is solved, what is not, what needs more work, which challenges have not even addressed.
+This section should go back to the list of challenges from Section \ref{subsec:challenges} and discuss them in the light of what has been said afterwards.
+I think this is a VERY IMPORTANT section, since it should provide our vision on the field.
+
+For example, notice that a big challenge missing from the above sections is {\em Spatial and Temporal Reasoning}.
+While I think (and I'm sure you'll agree) that this should be a huge topic in RTS game AI, I haven't seen much of this beyond potential fields.
+Because of this (as Michael Buro mentioned), many bots fall prey of easy tricks due to lack of common sense reasoning.
+
+It should also present an outlook of what is to come and what is needed.
+}
+
+\section*{Acknowledgments} {\color{blue} This research is partially funded by projects ... and ... }
+
+\appendix
+
+\section*{This is the old task-division section}
+
+{\color{red}
+I've included here the section we had for not losing it :)
+}
 
 Playing an RTS game involves dealing with all the problems described above.
-A few approaches, like CAT \cite{aha2005cat}, Darmok \cite{ontanon2010darmok} or ALisp \cite{marthi2005wargus} try to deal with the problem in a monolithic manner, by using a single AI technique.
+A few approaches, like CAT \cite{LTW}, Darmok \cite{OntanonMSR10} or ALisp \cite{Marthi05} try to deal with the problem in a monolithic manner, by using a single AI technique.
 This resembles approaches to solve other games, such as Chess or Go, where a single game-tree search approach is enough to play the game at human level.
@@
 
-\begin{figure}[ta]
-%    \centering
-    \includegraphics[width=\columnwidth]{figures/figure-bot-architectures.pdf}
-    \caption{General architecture of 5 Starcraft AI bots. {\color{red} (This figure is unreadable as of now, but we'll figure out a way to better present this info)}}
+\begin{figure*}[ta]
+    \centering
+    \includegraphics[width=\textwidth]{figures/figure-bot-architectures-wide.pdf}
+    \caption{General architecture of 5 Starcraft AI bots.}
     \label{fig:bot-architecture}
-\end{figure}
+\end{figure*}
 
@@
 
-{\color{red}
+{\color{blue}
 \begin{itemize}
-\item What is the best task decomposition for RTS games?
-\item Resource coordination in a divide-and-conquer approaches.
-\item Reactive high-level strategies that do not depend on scripting.
-\item ...
+\item Strengths and weaknesses?
+\item Resource coordination in a divide-and-conquer approaches?
+\item High-level strategies that do not depend on scripting? (not here, this should be discussed in Section \ref{sec:planning})
 \end{itemize}
 }
-
-\section{Discussion, what is left to do / open challenges}
-
-\section{Conclusion}
-
-\section*{Acknowledgment}
 

### 2011-11-24 survey.tex
added the spatial reasoning section

@@
 \item Planning and Resource Allocation (``macro'')
+\item Spatial reasoning
 \end{itemize}
@@
 
+\section{Challenge 7: Spatial Reasoning}\label{sec:spatial}
+
+{\color{blue}
+I think (and I'm sure you'll agree) that this should be a huge topic in RTS game AI, I haven't seen much of this beyond potential fields.
+Because of this (as Michael Buro mentioned), many bots fall prey of easy tricks due to lack of common sense reasoning.
+}
+
 \section{Empirical Comparison of RTS Bots}\label{sec:experiments}
@@
 I think this is a VERY IMPORTANT section, since it should provide our vision on the field.
-
-For example, notice that a big challenge missing from the above sections is {\em Spatial and Temporal Reasoning}.
-While I think (and I'm sure you'll agree) that this should be a huge topic in RTS game AI, I haven't seen much of this beyond potential fields.
-Because of this (as Michael Buro mentioned), many bots fall prey of easy tricks due to lack of common sense reasoning.
 

### 2011-12-12 survey.tex
just some minor edits
drafting about learning types and uncertainty

@@
 their efforts, and also should help people implementing AI for RTS games in
-general (e.g. industry).  In Gabriel's words ``RTS AI problems, Solutions, State-of-the-art, conclude on what's "solved" (since Buro 2004) and what's not.''
+general (e.g. industry).  In Gabriel's words ``RTS AI problems, Solutions, State-of-the-art, conclude on what's "solved" since (Buro 2004) and what's not.''
 
@@
 
+When talking about learning in RTS games, with regard to a particular match against an opponent, one has to differentiate between:
+\begin{itemize}
+\item prior learning: done before the match, like when a bot learns what is likely to happen on a given match-up and map from replays, or when a bot optimizes its micro-management parameters through simulated annealing or genetic evolution.
+\item in-match learning: done between the different games of match, ``the opponent rushed me last game so I will play this one safe''.
+\item in-game learning: done between the different actions of the game, ``I tried to attack here and it was a failure, let us try somewhere else''.
+\end{itemize}
+All these learning can be supervised (labeled replays, scripted situations) or unsupervised (unlabeled data, exploratory situations).
+The last two kinds of situations for learning are more inclined towards reinforcement learning because the AI has to deal with the exploration-exploitation duality.
+
 \section{Challenge 3: Reasoning with Uncertainty}\label{sec:uncertainty}
@@
 }
+
+There are two main kinds of uncertainty, existing for two different reasons, in RTS games.
+First, there is the incompleteness of information, due to the fog of war, which leads to some kind of ``extensional''
+%``extensive'' 
+uncertainty, that can be lowered by good scouting, and knowledge representation and reasoning (to infer what is possible given what has been seen).
+Also, there is the fact that we will never be able to read the opponent's mind, the ``intentional''
+%or ``intensive'' 
+uncertainty that we can't really be sure to scratch off, even with the best scouting possible.
+For intentional uncertainty, the AI, as the human, can only build a sensible model of what the opponent is likely to do, to think, or to think knowing what we think he thinks.
+Both these kinds of uncertainty can be studied through probabilities, but $P(Hidden | Seen)$ can be \textit{directly} linked to data only for extensional uncertainty.
+%now for the clumsy part:
+One could argue that the extensional data is the only thing that matters because it is the only thing that has an influence on the game, but in reality, for very highly skilled players, discovering the intensions of the enemy is the key to winning.
+It allows for a compact understanding of the game state and development, while also giving additional information about the future.
+Whereas a player is doing action A to follow it up by action B or C depends on its intension, while the information about action A is visible and information about B or C is hidden, only intension allows for a choice.
 

### 2011-12-12 survey.tex
typos + bib

@@
 \begin{itemize}
-\item prior learning: done before the match, like when a bot learns what is likely to happen on a given match-up and map from replays, or when a bot optimizes its micro-management parameters through simulated annealing or genetic evolution.
+\item prior learning: done before the match, like when a bot learns what is likely to happen on a given match-up (and map) from replays, or when a bot optimizes its micro-management parameters through simulated annealing or genetic evolution.
 \item in-match learning: done between the different games of match, ``the opponent rushed me last game so I will play this one safe''.
@@
 All these learning can be supervised (labeled replays, scripted situations) or unsupervised (unlabeled data, exploratory situations).
-The last two kinds of situations for learning are more inclined towards reinforcement learning because the AI has to deal with the exploration-exploitation duality.
+The last two kinds of situations of learning are more inclined towards reinforcement learning because the AI has to deal with the exploration-exploitation duality.
+%prior learning: supervised \cite{WeberCIG09} \cite{SynnaeveCIG11}, unsupervised: \cite{SynnaeveAIIDE11}
 
@@
 uncertainty that we can't really be sure to scratch off, even with the best scouting possible.
-For intentional uncertainty, the AI, as the human, can only build a sensible model of what the opponent is likely to do, to think, or to think knowing what we think he thinks.
-Both these kinds of uncertainty can be studied through probabilities, but $P(Hidden | Seen)$ can be \textit{directly} linked to data only for extensional uncertainty.
+For intentional uncertainty, the AI, as the human player, can only build a sensible model of what the opponent is likely to do, to think, or to think knowing what we think he thinks.
+Both these kinds of uncertainty can be studied through probabilities, but only $P(Hidden | Seen)$ can be \textit{directly} linked to data only for extensional uncertainty.
 %now for the clumsy part:
-One could argue that the extensional data is the only thing that matters because it is the only thing that has an influence on the game, but in reality, for very highly skilled players, discovering the intensions of the enemy is the key to winning.
+One could argue that the extensional data is the only thing that matters because it is the only thing that has an influence on the game, but in reality, for very highly skilled players, discovering the intentions of the enemy is the key to winning.
 It allows for a compact understanding of the game state and development, while also giving additional information about the future.
-Whereas a player is doing action A to follow it up by action B or C depends on its intension, while the information about action A is visible and information about B or C is hidden, only intension allows for a choice.
+Whereas a player is doing action A to follow it up by action B or C depends on its intention, while the information about action A is visible and information about B or C is hidden, only intention allows for a choice.
 

### 2012-02-18 survey.tex
Reorganized the content of the paper into the new structure

@@
 \begin{IEEEkeywords}
-review, RTS, games, StarCraft, machine learning, planning, TODO ...
+Game AI, Real-Time Strategy, Starcraft, Review1
 
@@
 {\color{blue}
-Introduce the game plus screenshot. Not too lengthy
-}
-
-\subsection{Complexity of RTS Game AI}\label{subsec:complexity}
-
-{\color{blue}
-Here discuss the complexity of RTS games in general, but focusing on Starcraft.
-The most important thing of this section is to demonstrate that RTS games are complex, and this is the motivation for requiring all the challenges and research presented afterwards.
+Introduce the game plus screenshot. Not too lengthy.
 }
@@
 {\color{blue}
+This section should be brief, and just give us some basic ideas of which are the challenges in RTS Game AI, so that when we refer to related work, we can use this as a reference point.
+The most important thing of this section is to demonstrate that RTS games are complex.
+
 Many years have passed since Buro's call for research (8!!). he identified 6 challenges:
@@
 Additionally, other challenges that are not in the list appeared, and have been worked on, like: how to exploit the existing domain knowledge (strategies, build-orders, replays, etc.), or how to design an architecture that integrates all the modules required for an agent to play an RTS games.
-This section should provide an updated view of challenges, and set the ground for the rest of the paper.
-The remainder of the paper should be structured around the challenges that we list here.
-
-I think that a good list could be something like this (I'm open to all kind of criticism here :)):
 
@@
 
-I've now listed them as challenges, but maybe we can list them as questions.
-In each of the subsequent sections, we should use the bots in the competitions (BBQ, EISBot, Nova, UAlbertaBot, etc.) to illustrate the state-of-the-art.
-Eventually, we could refer to other work though if need be.
-We could also compare with what is being used in industry, and use the real AI of Starcraft as an example (Bob Fitch's talk in AIIDE provides enough material for that).
+Maybe the list above goes better in Section \ref{sec:questions}?
 }
 
-\section{Challenge 1: Task Decomposition}\label{sec:architecture}
+\section{Existing work on RTS Game AI}\label{sec:review}
 
 {\color{blue}
-Here we should say that complex RTS games require multi-scale reasoning \cite{WeberCig10}, and thus complex architectures that integrate many AI modules have been devised.
-Discuss how different researchers have divided up the task of playing Starcraft into a collection of subtasks and integrated them (see the end of this paper for a preliminary write-up on this)
+Include here the literature review of existing work on RTS Game AI.
+We could also compare with what is being used in industry, and use the real AI of Starcraft as an example (Bob Fitch's talk in AIIDE provides enough material for that).
+I have included here as subsections the two blurbs of text we had for integration of domain knowledge and reasoning with uncertainty, since I guess they correspond here.
 }
 
-\section{Challenge 2: Integration of Domain Knowledge}\label{sec:domainknowledge}
+%\subsection{Integration of Domain Knowledge}\label{subsec:domainknowledge}
 
@@
 Talk about learning, scripting, and other techniques to incorporate such knowledge.
-}
 
@@
 %prior learning: supervised \cite{WeberCIG09} \cite{SynnaeveCIG11}, unsupervised: \cite{SynnaeveAIIDE11}
+}
 
-\section{Challenge 3: Reasoning with Uncertainty}\label{sec:uncertainty}
+%\subsection{Reasoning with Uncertainty}\label{subsec:uncertainty}
 
 {\color{blue}
-I think there are two important aspects of uncertainty in RTS games.
-
-First is trying to model uncertainty and try to reduce it (information gathering, and representation of knowledge), this includes scouting, and maintaining good estimates of locations and number of units of the enemy, etc.
-
-Second is actually using the uncertain information. Taking decisions like army strength estimation (like to attack or retreat).
-}
+I think there are two important aspects of uncertainty in RTS games:
+\begin{itemize}
+\item First is trying to model uncertainty and try to reduce it (information gathering, and representation of knowledge), this includes scouting, and maintaining good estimates of locations and number of units of the enemy, etc.
+\item Second is actually using the uncertain information. Taking decisions like army strength estimation (like to attack or retreat).
+\end{itemize}
 
@@
 Whereas a player is doing action A to follow it up by action B or C depends on its intention, while the information about action A is visible and information about B or C is hidden, only intention allows for a choice.
+}
 
-\section{Challenge 4: Opponent modeling and Adaptation}\label{sec:adaptation}
+\section{State of the Art Bots for Starcraft}\label{sec:bot}
 
 {\color{blue}
-What has been done in modeling the strategy of the opponent and adapt accordingly? which bots do adapt themselves to what they learn from the opponent and how?
+Here, describe the existing top bots for Starcraft.
+The difference between this section and the previous, is that the previous focuses on existing ideas, techniques, etc.
+whereas this section focuses on how some of those have been put together to create specific bots that play Starcraft.
 }
 
-{\color{red} I'm a bit worried that Challenge 3 and 4 will overlap significantly, since a large source of uncertainty is the opponent, so, part of opponent modeling has a lot to do with uncertainty.
-What do you think?}
-
-\section{Challenge 5: Group and Individual Control}\label{sec:micro}
+\section{The Starcraft AI Competition}\label{sec:competition}
 
 {\color{blue}
-One of the big areas where we have seen a lot of research is in micro-management, where participants in the competitions have used reinforcement-learning (Overmind), potential-fields (BTHAI, Nova, and others), or many other techniques for coordinating both individual unit and group behavior.
+Here describe the results obtained in the competition comparing the bots.
 }
 
-\section{Challenge 6: Planning and Resource Allocation}\label{sec:planning}
+\section{Open Questions in RTS Game AI}\label{sec:questions}
 
 {\color{blue}
-This section is intended to refer to all those techniques concerning ``macro'', since resource allocation and planning is considered together in most state-of-the-art bots (do you agree?).
-Here we can talk about things like build-order-optimization, resource allocation to different aspects of the game (army, research, etc.)
+Here, describe what is left to be done.
 }
 
-\section{Challenge 7: Spatial Reasoning}\label{sec:spatial}
+\section{Empirical Evidence of What is Missing in Existing Bots}\label{sec:experiments}
 
 {\color{blue}
-I think (and I'm sure you'll agree) that this should be a huge topic in RTS game AI, I haven't seen much of this beyond potential fields.
-Because of this (as Michael Buro mentioned), many bots fall prey of easy tricks due to lack of common sense reasoning.
+Here, whatever experiments we want to include to support which aspects of RTS Game AI require more work.
 }
 
-\section{Empirical Comparison of RTS Bots}\label{sec:experiments}
-
-{\color{blue}
-What are the similarities and differences of the current bots in the competition?
-which techniques are widespread, and which are experimental?
-which features decide over winning?
-Which features would have a high impact if we work on them?
-}
-
-{\color{red} Given that we have Ben, Mike (Preuss), Dave and possibly Michael (Buro) on board, we could include results from the competitions.
-But, given that this is not a standard ``competition paper'', maybe you guys want to reserve that data to put together a separate paper.
-It's your call, since it's your data :)}
-
-\section{Discussion and Outlook}\label{sec:conclusions}
-
-{\color{blue}	
-what is solved, what is not, what needs more work, which challenges have not even addressed.
-This section should go back to the list of challenges from Section \ref{subsec:challenges} and discuss them in the light of what has been said afterwards.
-I think this is a VERY IMPORTANT section, since it should provide our vision on the field.
-
-It should also present an outlook of what is to come and what is needed.
-}
+\section{Discussion and Conclusions}\label{sec:conclusions}
 
@@
 I've included here the section we had for not losing it :)
+}
+
+{\color{blue}
+Here we should say that complex RTS games require multi-scale reasoning \cite{WeberCig10}, and thus complex architectures that integrate many AI modules have been devised.
+Discuss how different researchers have divided up the task of playing Starcraft into a collection of subtasks and integrated them (see the end of this paper for a preliminary write-up on this)
 }

### 2012-02-26 survey.tex
first paragraph on micro approaches: CBR and RL

@@
 their efforts, and also should help people implementing AI for RTS games in
-general (e.g. industry).  In Gabriel's words ``RTS AI problems, Solutions, State-of-the-art, conclude on what's "solved" since (Buro 2004) and what's not.''
+general (e.g. industry).  In Gabriel's words ``RTS AI problems, Solutions, State-of-the-art, conclude on what's "solved" since \cite{Buro03rts} and what's not.''
 
@@
 \item Planning and Resource Allocation (``macro'')
-\item Spatial reasoning
+\item Spatial reasoning (``tactics'')
 \end{itemize}
@@
 }
+
+\subsection{Micro-management}
+
+Some interesting uses of reinforcement learning (RL) \cite{Sutton} to RTS research are in the domain of \textit{micro-management} (group and individual units control): \cite{Marthi05} employs concurrent hierarchical Q-learning (units Q-functions are combined at the group level) RL to efficiently control units in a ``one robot with multiple effectors'' fashion.
+\cite{Madeira06} advocates the use of prior domain knowledge to allow faster RL learning and applied their work on a turn-based strategy game.
+In real game setups, the actions spaces to explore is gigantic, it requires to use the structure of the game in a partial program (or a partial Markov decision process) and a shape function (or a heuristic) \cite{Marthi05}.
+There are other was to learn parameters of an underlying model, for instance \cite{GA} used evolutionary learning techniques, but face the same problem of dimensionality.
+The difficulty to work with multi-scale goals and plans is handled directly by case-based reasoning (CBR), which has been adapted for micro-management with continuous action models \cite{Molineaux08}, an integrated RL/CBR algorithm using continous models, or with hybrid CBR/RL transfer learning \cite{CBR-RL}.
+Reactive planning (for behavior languages) \cite{WeberCig10}, a decompositional planning similar to hierarchical task networks \cite{HTNPlanning}, allows for plans to be changed at different granularity levels and so for multi-scale (hierarchical) goals integration.
 

### 2012-02-26 survey.tex
micro: state of the art with potential fields

@@
 
+Potential fields \cite{Hagelback08,Hagelback09} are a powerful way to specify complex (with the combination of several fields, requiring a parameter-tuning method) behavior policies and are used in modern commercial games and RTS AI are using derivative of it (Supreme Commander flowfield).
+Self-organizing-maps (SOM) learned with evolutionary algorithms provide some way to learn (realistic) parameters for RTS units behavior \cite{teamCompositionRTS}.
+Bayesian programming has been applied to inverse fusion of the sensory inputs of the units \cite{SynnaeveMicroCig11}, which subsumes potential fields, allowing for integration of tactical goals directly in micro-management.
+There are also some cognitive approaches to RTS AI \cite{SORTS}, in which there is a human-like attention model for micro-management.
+
 %\subsection{Integration of Domain Knowledge}\label{subsec:domainknowledge}

### 2012-02-26 survey.tex
minimal intro on micro

@@
 \item Opponent Modeling and Adaptation
-\item Group and Individual Control (``micro'')
+\item Group and Individual Control (``micro''): the task of controlling units efficiently (we can not speak of optimality here due to the huge state space, and of the enemy behavior entails numerous Nash equilibriums), focusing fire to diminish enemy's firepower and keeping our units alive the longest, casting defensive and offensive spells and abilities.
 \item Planning and Resource Allocation (``macro'')

### 2012-02-26 survey.tex
added some biblio on opponent modeling

--- 2012-02-26 survey.tex
@@
 \item Reasoning with Uncertainty (including information gathering)
-\item Opponent Modeling and Adaptation
-\item Group and Individual Control (``micro'')
+\item Opponent Modeling and Adaptation: opponent modeling is key if we have to adapt our strategy.
+As different strategies are dominating each others, forming multiple Nash equilibria, the AI has to be able to infer the intend of its opponent.
+\item Group and Individual Control (``micro''): the task of controlling units efficiently (we can not speak of optimality here due to the huge state space, and of the enemy behavior entails numerous Nash equilibria), focusing fire to diminish enemy's firepower and keeping our units alive the longest, casting defensive and offensive spells and abilities.
 \item Planning and Resource Allocation (``macro'')
@@
 There are also some cognitive approaches to RTS AI \cite{SORTS}, in which there is a human-like attention model for micro-management.
+
+\subsection{Opponent modeling, strategy prediction}
+
+There are related works in the domains of opponent modeling \cite{HsiehS08,SchaddBS07,OBRecog}.
+The main methods used to these ends are case-based reasoning (CBR) and planning or plan recognition \cite{LTW,CBR_Planning,OntanonCBR,HTNPlanning,Ramirez}.
+There are precedent works of Bayesian plan recognition \cite{BMPR}, even in games with Albrecht \textit{et al.} \cite{BayesianRecog} using dynamic Bayesian networks to recognize a user's plan in a multi-player dungeon adventure.
+
+\cite{LTW} used CBR to perform dynamic plan retrieval extracted from domain knowledge in Wargus (Warcraft II clone).
+\cite{CBR_Planning} base their real-time case-based planning (CBP) system on a plan dependency graph which is learned from human demonstration.
+In \cite{OntanonCBR,PlanRetrieval}, they use CBR and expert demonstrations on Wargus.
+They update their previous CPB approach by using ``situation assessment'' for plan retrieval.
+They improve the speed of CPB by using a decision tree to select relevant features.
+Hsieh and Sun \cite{HsiehS08} based their work on \cite{LTW} CBR model and used StarCraft replays to construct states and building sequences.
+Strategies are choices of building construction order in their model.
+
+\cite{SchaddBS07} describe opponent modeling through hierarchically structured models of the opponent behaviour and they applied their work to the Spring RTS (Total Annihilation clone).
+\cite{HTNPlanning} use hierarchical task networks (HTN) to model strategies in a first person shooter with the goal to use HTN planners.
+\cite{OBRecog} improve the probabilistic hostile agent task tracker (PHATT \cite{PHATT}, a simulated HMM for plan recognition) by encoding strategies as HTN.
+%\cite{Madeira06}
+\cite{WeberCig09} proposed ``a data mining approach to strategy prediction'' and performed supervised learning on StarCraft replays.
+\cite{HMMstrat_RTS_AIIDE11} used HMM to learn the transition probabilities of sequences of constructions and kept the most probables to produce probabilistic behavior models (in StarCraft).
+\cite{SynnaeveOpeningCig11} continued on \cite{WeberCig09} and used a Bayesian semi-supervised (labeling based on clustering) model to learn from replays and predict openings (early game strategies) from StarCraft replays, while \cite{SynnaeveAIIDE11} presents a whole game Bayesian tech-tree predictor based on unserpervised learning from replays.
 

### 2012-02-27 survey.tex
adding missing sty files

--- 2012-02-26 survey.tex
@@
 their efforts, and also should help people implementing AI for RTS games in
-general (e.g. industry).  In Gabriel's words ``RTS AI problems, Solutions, State-of-the-art, conclude on what's "solved" since \cite{Buro03rts} and what's not.''
+general (e.g. industry).  In Gabriel's words ``RTS AI problems, Solutions, State-of-the-art, conclude on what's "solved" since (Buro 2004) and what's not.''
 
@@
 \item Planning and Resource Allocation (``macro'')
-\item Spatial reasoning (``tactics'')
+\item Spatial reasoning
 \end{itemize}
@@
 }
-
-\subsection{Micro-management}
-
-Some interesting uses of reinforcement learning (RL) \cite{Sutton} to RTS research are in the domain of \textit{micro-management} (group and individual units control): \cite{Marthi05} employs concurrent hierarchical Q-learning (units Q-functions are combined at the group level) RL to efficiently control units in a ``one robot with multiple effectors'' fashion.
-\cite{Madeira06} advocates the use of prior domain knowledge to allow faster RL learning and applied their work on a turn-based strategy game.
-In real game setups, the actions spaces to explore is gigantic, it requires to use the structure of the game in a partial program (or a partial Markov decision process) and a shape function (or a heuristic) \cite{Marthi05}.
-There are other was to learn parameters of an underlying model, for instance \cite{GA} used evolutionary learning techniques, but face the same problem of dimensionality.
-The difficulty to work with multi-scale goals and plans is handled directly by case-based reasoning (CBR), which has been adapted for micro-management with continuous action models \cite{Molineaux08}, an integrated RL/CBR algorithm using continous models, or with hybrid CBR/RL transfer learning \cite{CBR-RL}.
-Reactive planning (for behavior languages) \cite{WeberCig10}, a decompositional planning similar to hierarchical task networks \cite{HTNPlanning}, allows for plans to be changed at different granularity levels and so for multi-scale (hierarchical) goals integration.
-
-Potential fields \cite{Hagelback08,Hagelback09} are a powerful way to specify complex (with the combination of several fields, requiring a parameter-tuning method) behavior policies and are used in modern commercial games and RTS AI are using derivative of it (Supreme Commander flowfield).
-Self-organizing-maps (SOM) learned with evolutionary algorithms provide some way to learn (realistic) parameters for RTS units behavior \cite{teamCompositionRTS}.
-Bayesian programming has been applied to inverse fusion of the sensory inputs of the units \cite{SynnaeveMicroCig11}, which subsumes potential fields, allowing for integration of tactical goals directly in micro-management.
-There are also some cognitive approaches to RTS AI \cite{SORTS}, in which there is a human-like attention model for micro-management.
 
@@
 
-Figure \ref{fig:bot-architecture} shows some representative examples of the integration architectures used by different bots in the AIIDE 2011 Starcraft AI competition \cite{???}.
+Figure \ref{fig:bot-architecture} shows some representative examples of the integration architectures used by different bots in the AIIDE 2011 Starcraft AI competition \cite{url1}.
 Each box represents an individual module with a clearly defined task (only modules with a black background can send actions directly to Starcraft).
@@
 \item Resource coordination in a divide-and-conquer approaches?
-\item High-level strategies that do not depend on scripting? (not here, this should be discussed in Section \ref{sec:planning})
+\item High-level strategies that do not depend on scripting? (not here, this should be discussed in Section) % \ref{sec:planning})
 \end{itemize}

### 2012-02-27 survey.tex
literature review

@@
 
-\usepackage{cite,color,graphicx}
+\usepackage{cite,graphicx}
+\usepackage[usenames,dvipsnames]{xcolor}
 
@@
 }
+{\color{ForestGreen}
+Here it is a brief description, I don't think we need a screenshot.
+}
+Starcraft: Brood War is a popular RTS game released in 1998 by Blizzard Entertainment.
+Starcraft is set in a science-fiction based world where the player must choose one of the three races: Terran, Protoss or Zerg.
+The good work done by the people of Blizzard makes this game one of the most extremely well-balanced RTS games ever created.
+\begin{itemize}
+	\item Terrans, human exiled from Earth, provide units that are versatile and flexible giving a balanced option between Protoss and Zergs.
+	\item Protoss units have lengthy and expensive manufacturing process, but they are strong and resistant. These conditions makes players follow a strategy of quality over quantity.
+	\item Zergs, the insectoid race, units are cheap and weak. They can be produced fast, encouraging players to overwhelm their opponents with sheer numbers.
+\end{itemize}
 
@@
 Whereas a player is doing action A to follow it up by action B or C depends on its intention, while the information about action A is visible and information about B or C is hidden, only intention allows for a choice.
+}
+
+\subsection{Terrain Analysis}
+{\color{ForestGreen}
+Terrain analysis supplies the AI with chunks of abstract information about the map in order to help in making decisions.
+Terrain analysis can provide automatic detection of regions, choke points and other areas of interest.
+This analysis is usually performed off-line, in order to save processing time during gameplay.
+
+The final goal of terrain analysis is having all the useful tactical information of the map.
+To do this we can use different algorithms, one of the most used, coming from computer vision, being \textit{skeletons by influence zones}, also known as \textit{Voronoi diagrams}\cite{Perkins_2010}.
+Through this kind of diagrams we will be able to generate a graph with the centroid of the regions and the choke points of each region.
+
+Another common technique is the utilization of an influence map to detect and mark relevant zones of the map as resources location.
+}
+
+\subsection{Pathfinding}
+{\color{ForestGreen}
+The most common pathfinding algorithm is A*, but the big problem of this algorithm is the time and  memory consumption, a show-stopper in a real-time environment.
+
+To solve this the search space needs to be reduced as much as possible.
+One common technique is reducing the tile map search into a navigation graph map.
+To do this some polygonal triangulations are calculated and abstracted on a grid-based map and this is the TRA* (Triangulation Reduction A*) algorithm\cite{Demyen_2006}.
+
+** There are a lot of new pathfinding algorithms. Do we need to review all of them? **
+}
+
+\subsection{Group behavior}
+{\color{ForestGreen}
+Craig Reynols introduced algorithmic steering behaviors to simulate the aggregate motion of swarms using procedural animation\cite{Reynolds_1999}.
+His model describes how a system of multiple agents , each steering according to simple local rules, can produce the collective movement found in flocks of birds.
+With this behaviours we are able to simulate intelligent squad movements and do tactical moves to flank the opponent in a RTS game\cite{Danielsiek_2008}, improving the survival of our units.
+
+A good alternative to Flocking is using Reciprocal Velocity Obstacles (RVO)\cite{Berg_2008}, since it is easier to control the behavior and they guarantee to prevent collisions.With this algorithm we can move a large squad crossing another squad and avoid all possible collisions between units.
+This type of techniques are already applied in commercial RTS games like Starcraft 2 or Supreme Commander 2 with great success.
+
+Another popular technique is Potential Fields.
+basically, Potential Fields work like a charged particle moving through a magnetic field.
+The process of tuning and optimizing the parameters of potential field charges can be time reduced applying learning algorithms such as Q-Learning\cite{Liu_2008}.
+Another common issue we have to deal with when using potential fields is the local optima problem that can stuck the unit away from the goal.
+Some examples of the utilities of potential files in RTS games are avoiding obstacles (navigation), avoiding opponent fire, or staying at maximum shooting distance\cite{Hagelback08}\cite{Hagelback09}.
+}
+
+\subsection{Decision Making}
+{\color{ForestGreen}
+The most common decision making are Decision Trees.
+Some of the advantages of the decision trees are: simple to understand and interpret acting like a white box where we can explain why a decision is taken, and they are compatible with other decision techniques.
+For complex decisions we have many algorithms to generate an optimum decision tree like ID3 or C4.5, depends on the problem we will use one or another.
+
+A good alternative are Finite State Machine.
+The idea behind FSM is to decompose an object's behavior into easily manageable "chunks" or states, such as "attacking an enemy", "gathering resources" or "repairing" and establish the conditions that trigger transitions between them.
+}
+
+\subsection{Opponent modelling}
+{\color{ForestGreen}
+Opponent modelling is the process of keeping track of what the enemy is doing during a game in order to estimate the probability of it using a specific strategy.
+The key point here is to predict what units the enemy will build, when he will build them, when and where he will attack and so on.
+This will give us information about what kind of opponent it is allowing us to adapt our strategy and select the best counter.
+
+Science we are dealing with imperfect information, the best solution is to build a Bayesian model which can deal perfectly with the uncertainty.
+The difficult part is building that model.
+Some approaches use traces of game logs to analyze the state of the world, but those traces have all the information of the game, so we have to add some noise to emulate the real use of the model in a non-deterministic world.
+
+A Threat Map or Influence Map is a way to store relevant information about the opponent and keep track of the conflict areas or the secure ones.
+Statical threat maps can be designed where the threat is always "real" or concepts like heat map over time can be applied to make predictions.
 }

### 2012-02-27 survey.tex

@@
 \item Reasoning with Uncertainty (including information gathering)
-\item Opponent Modeling and Adaptation
-\item Group and Individual Control (``micro'')
+\item Opponent Modeling and Adaptation: opponent modeling is key if we have to adapt our strategy.
+As different strategies are dominating each others, forming multiple Nash equilibria, the AI has to be able to infer the intend of its opponent.
+\item Group and Individual Control (``micro''): the task of controlling units efficiently (we can not speak of optimality here due to the huge state space, and of the enemy behavior entails numerous Nash equilibria), focusing fire to diminish enemy's firepower and keeping our units alive the longest, casting defensive and offensive spells and abilities.
 \item Planning and Resource Allocation (``macro'')
-\item Spatial reasoning
+\item Spatial reasoning (``tactics'')
 \end{itemize}
@@
 }
+
+\subsection{Micro-management}
+
+Some interesting uses of reinforcement learning (RL) \cite{Sutton} to RTS research are in the domain of \textit{micro-management} (group and individual units control): \cite{Marthi05} employs concurrent hierarchical Q-learning (units Q-functions are combined at the group level) RL to efficiently control units in a ``one robot with multiple effectors'' fashion.
+\cite{Madeira06} advocates the use of prior domain knowledge to allow faster RL learning and applied their work on a turn-based strategy game.
+In real game setups, the actions spaces to explore is gigantic, it requires to use the structure of the game in a partial program (or a partial Markov decision process) and a shape function (or a heuristic) \cite{Marthi05}.
+There are other was to learn parameters of an underlying model, for instance \cite{GA} used evolutionary learning techniques, but face the same problem of dimensionality.
+The difficulty to work with multi-scale goals and plans is handled directly by case-based reasoning (CBR), which has been adapted for micro-management with continuous action models \cite{Molineaux08}, an integrated RL/CBR algorithm using continous models, or with hybrid CBR/RL transfer learning \cite{CBR-RL}.
+Reactive planning (for behavior languages) \cite{WeberCig10}, a decompositional planning similar to hierarchical task networks \cite{HTNPlanning}, allows for plans to be changed at different granularity levels and so for multi-scale (hierarchical) goals integration.
+
+Potential fields \cite{Hagelback08,Hagelback09} are a powerful way to specify complex (with the combination of several fields, requiring a parameter-tuning method) behavior policies and are used in modern commercial games and RTS AI are using derivative of it (Supreme Commander flowfield).
+Self-organizing-maps (SOM) learned with evolutionary algorithms provide some way to learn (realistic) parameters for RTS units behavior \cite{teamCompositionRTS}.
+Bayesian programming has been applied to inverse fusion of the sensory inputs of the units \cite{SynnaeveMicroCig11}, which subsumes potential fields, allowing for integration of tactical goals directly in micro-management.
+There are also some cognitive approaches to RTS AI \cite{SORTS}, in which there is a human-like attention model for micro-management.
+
+\subsection{Opponent modeling, strategy prediction}
+
+There are related works in the domains of opponent modeling \cite{HsiehS08,SchaddBS07,OBRecog}.
+The main methods used to these ends are case-based reasoning (CBR) and planning or plan recognition \cite{LTW,CBR_Planning,OntanonCBR,HTNPlanning,Ramirez}.
+There are precedent works of Bayesian plan recognition \cite{BMPR}, even in games with Albrecht \textit{et al.} \cite{BayesianRecog} using dynamic Bayesian networks to recognize a user's plan in a multi-player dungeon adventure.
+
+\cite{LTW} used CBR to perform dynamic plan retrieval extracted from domain knowledge in Wargus (Warcraft II clone).
+\cite{CBR_Planning} base their real-time case-based planning (CBP) system on a plan dependency graph which is learned from human demonstration.
+In \cite{OntanonCBR,PlanRetrieval}, they use CBR and expert demonstrations on Wargus.
+They update their previous CPB approach by using ``situation assessment'' for plan retrieval.
+They improve the speed of CPB by using a decision tree to select relevant features.
+Hsieh and Sun \cite{HsiehS08} based their work on \cite{LTW} CBR model and used StarCraft replays to construct states and building sequences.
+Strategies are choices of building construction order in their model.
+
+\cite{SchaddBS07} describe opponent modeling through hierarchically structured models of the opponent behaviour and they applied their work to the Spring RTS (Total Annihilation clone).
+\cite{HTNPlanning} use hierarchical task networks (HTN) to model strategies in a first person shooter with the goal to use HTN planners.
+\cite{OBRecog} improve the probabilistic hostile agent task tracker (PHATT \cite{PHATT}, a simulated HMM for plan recognition) by encoding strategies as HTN.
+%\cite{Madeira06}
+\cite{WeberCig09} proposed ``a data mining approach to strategy prediction'' and performed supervised learning on StarCraft replays.
+\cite{HMMstrat_RTS_AIIDE11} used HMM to learn the transition probabilities of sequences of constructions and kept the most probables to produce probabilistic behavior models (in StarCraft).
+\cite{SynnaeveOpeningCig11} continued on \cite{WeberCig09} and used a Bayesian semi-supervised (labeling based on clustering) model to learn from replays and predict openings (early game strategies) from StarCraft replays, while \cite{SynnaeveAIIDE11} presents a whole game Bayesian tech-tree predictor based on unserpervised learning from replays.
 

### 2012-02-28 survey.tex
competition part started

@@
 
-\section{The Starcraft AI Competition}\label{sec:competition}
+\section{Recent Starcraft AI Competitions}\label{sec:competition}
 
-{\color{blue}
+{\color{magenta}
+
+\subsection{CIG 2011}
+\label{sec:cig2011}
+
+An initial attempt to run a StarCraft tournament at the Computational
+Intelligence in Games conference 2010 suffered from technical problems.
+These mainly stemmed from the desire to use evolved, largely untested
+maps which proved to look interesting but made the submitted bots crash
+so often that it would have been unjustifiable to announce a winner.
+
+At the CIG 2011, the tournament was therefore run with a (secret) selection
+of maps used in league play. It was organized by Tobias Mahlmann and Mike
+Preuss and attracted 10 bots.
+
+\begin{table*}[htb]
+\caption{Bracket A results at CIG 2011, 
+qualified for the final round: UAlbertaBot and Skynet.}
+\label{tab:correlations}
+\begin{small}
+\begin{center}
+\begin{tabular}{|l|c|c|c|}
+\hline
+Crashes & games & bot	& wins\\ \hline
+0 &	 40 &	 UAlbertaBot &	 33\\
+1 &   40 &	 Skynet	  &  31\\
+2 &	 40 &	 AIUR	  &  24\\
+1 &	 40 &	 Nova	  &  8\\
+0 &	 40 &	 LSAI	  &  4\\
+\hline
+\end{tabular}
+\end{center}
+\end{small}
+\end{table*}
+
+Bracket B
+
+Crashes	 Games	 Bot	 Wins
+12	 40	 Xelnaga	 25
+3	 40	 BroodwarBotQ	 23
+0	 40	 BTHAI	 23
+17	 40	 Protoss Beast Jelly	 17
+0	 40	 EvoBot	 12
+As BroodwarBotQ and BTHAI have the same number of wins, their direct encounter is evaluated which accounted 6 : 4 for the BroodwarBotQ.
+
+Qualified for the final round: Xelnaga and BroodwarBotQ. Note that all qualified bots play Protoss.
+Maps used for the first round: (2)MatchPoint 1.3, (4)Fighting Spirit 1.3, iCCupdestination 1.1, iCCup gaia, iCCup great barrier reef
+
+Final Round
+
+The final round was played in a similar mode as each of the first round brackets, using another set of 5 maps (see below), resulting in 30 games for each bot.
+
+Crashes	 Games	 Bot	 Wins
+0	 30	 Skynet	 26
+0	 30	 UAlbertaBot	 22
+3	 30	 Xelnaga	 11
+2	 30	 BroodwarBotQ	 1
+
 Here describe the results obtained in the competition comparing the bots.

### 2012-03-08 survey.tex

@@
 
-{\color{blue}
-Here, describe the existing top bots for Starcraft.
-The difference between this section and the previous, is that the previous focuses on existing ideas, techniques, etc.
-whereas this section focuses on how some of those have been put together to create specific bots that play Starcraft.
+{\color{red} add UAlbertaBot}
+
+Thanks to the recent organization of international game AI competitions focused around the popular StarCraft game (see Section \ref{sec:competition}), several groups have been working on integrating many of the techniques described in the previous section into complete ``bots'', capable of playing complete StarCraft games.
+In this section we will overview some of the currently available top bots, and focus on the architecture being used to integrate all the techniques being used for the different aspects of the bot.
+
+Playing an RTS game involves dealing with all the problems described above.
+A few approaches, like CAT \cite{LTW}, Darmok \cite{OntanonMSR10} or ALisp \cite{Marthi05} try to deal with the problem in a monolithic manner, by using a single AI technique.
+This resembles approaches to solve other games, such as Chess or Go, where a single game-tree search approach is enough to play the game at human level.
+However, none of those systems aims at achieving near human performance.
+In order to achieve human-level performance, RTS AI designers use a lot of domain knowledge in order to divide the task of playing the game into a collection of sub-problems, which can be dealt-with using individual AI techniques (as discussed in the previous section).
+
+\begin{figure*}[ta]
+    \centering
+    \includegraphics[width=\textwidth]{figures/figure-bot-architectures-wide.pdf}
+    \caption{General architecture of 5 Starcraft AI bots.}
+    \label{fig:bot-architecture}
+\end{figure*}
+
+Figure \ref{fig:bot-architecture} shows some representative examples of the integration architectures used by different bots in the AIIDE and CIG 2011 Starcraft AI competitions \cite{url1,url2}: BTHAI \cite{???}, BroodwarQ \cite{???}, SPAR \cite{???}, Nova \cite{???}, and Skynet \cite{???} (see Section \ref{sec:competition} for a comparison of their performance).
+Each box represents an individual module with a clearly defined task (only modules with a black background can send actions directly to Starcraft).
+Dashed arrows represent data flow, and solid arrows represent control (when a module can command another module to perform some task).
+For example, we can see how SPAR is divided in two sets of modules: {\em situation analysis} and {\em decision making}, the first with three modules dedicated to analyze the current situation of the game, and the later with 4 modules dedicated to exploit that information to decide what to do.
+We can see how the decision making aspect of SPAR is organized hierarchically, with the higher-level module ({\em strategic decision}) issuing commands to the next module ({\em tactical decision}), which sends commands to the next module ({\em action implementation}), and so on.
+Only the lower-level modules can send actions directly to Starcraft.
+
+On the other hand, bots such as NOVA or BroodwarBotQ (BBQ) only use a hierarchical organization for {\em micro-management} (controlling the attack units), but use a decentralized organization for the rest of the bot.
+In Nova and BBQ, there is a collection of modules that control different aspects of the game (workers, production, construction, etc.).
+These modules can all send actions directly to Starcraft.
+In Nova those modules coordinate only through writing data in a shared blackboard, and in BBQ they coordinate only when they have to use a shared resource (unit) by means of an arbitrator.
+
+By analyzing the structure of these bots, we can see that there are two main tools that can be used when designing an integration architecture:
+
+\begin{itemize}
+\item {\em Abstraction}: complex tasks can be formulated at different levels of abstraction.
+For example, playing an RTS game can be seen as issuing individual low-level actions to each of the units in the game, or at a higher level, it can be seen as deploying a specific strategy (e.g.
+a ``BBS strategy'', or a ``Reaver Drop'' strategy).
+Some bots, reason at multiple levels of abstraction at the same time, making the task of playing Starcraft simpler.
+Assuming that each module in the architecture of a bot has a goal and determines some actions to achieve that goal, the actions determined by higher-level modules are considered as the goals of the lower level modules.
+In this way, each module can focus on reasoning at only one level of abstraction, thus, making the problem easier.
+
+\item {\em Divide-and-conquer}: playing a complex RTS, such as Starcraft, requires performing many conceptually different tasks, such as gathering resources, attacking, placing buildings, etc.
+Assuming each of these tasks can be performed relatively independently and without interference, we can have one module focusing on each of the tasks independently, thus making the problem easier.
+\end{itemize}
+
+If we imagine the different tasks to perform in a complex RTS game in a two-dimensional plane, where the vertical axis represents abstraction, and the horizontal axis represents the different aspects of the game (micro-management, resource gathering, etc.), abstraction can be seen as dividing the space with horizontal lines, whereas divide-and-conquer divides the space using vertical lines.
+
+Different bots, use different combinations of these two tools. Looking back at Figure \ref{fig:bot-architecture}, we can see the following use of abstraction and divide-in-conquer in the bots:
+
+\begin{itemize}
+\item BTHAI: uses a two-tier abstraction hierarchy, where a collection of high-level modules command a collection of lower-level agents in charge of each of the units.
+At the high-level, BTHAI uses divide-and-conquer, having multiple high-level modules issuing commands to the lower-level units.
+\item BBQ: uses abstraction for micro-management, and divide-and-conquer for macro-management and intelligence gathering.
+To avoid conflicts between modules (since the individual tasks of each of the modules are not completely independent), BBQ uses an arbitrator.
+\item Nova: is similar in design as BBQ, and uses abstraction for micro-management, and divide-and-conquer for macro-management.
+The differences are that Nova does not have an arbitrator to resolve conflicts, but has a higher-level module ({\em strategy manager}), which posts information to the blackboard that the rest of modules follow (thus, making use of abstraction).
+\item SPAR: only uses abstraction.
+Its high-level module determines the strategy to use, and the tactical decision module divides it into a collection of {\em abstract actions}, that are executed by the lower-level modules.
+\item Skynet: makes extensive use of both abstraction and divide-and-conquer.
+Considering the decision making component of Skynet, we can see a high level module that issues commands to a series of tactics modules.
+The collection of tactic modules queue {\em tasks} (that are analogous to the abstract actions used in SPAR).
+Each different task has a specific low level module that knows how to execute it.
+Thus, Skynet uses a 3 layered abstraction hierarchy, and uses divide-and-conquer in all levels except the highest.
+\end{itemize}
+
+Additionally, except for BTHAI, all other agents use divide-and-conquer at a higher-level bot design and divide all the modules into two or three categories: {\em information gathering} and {\em decision}, or {\em information gathering}, {\em micro-management} and {\em macro-management}.
+
+Notice that most bots using divide-and-conquer (except for BBQ), assume that each of the modules can act independently and that their actions can be executed without interference.
+BBQ is the exception, including an arbitrator that makes sure that modules do not send contradictory orders to the same unit.
+However, very little bots handle the problem of how to coordinate resource usage amongst modules, for instance BTHAI uses a first-come-first-serve policy for spending resources, the first module that requests resources is the one that gets them.
+Nova and Skynet are exceptions, and implement some rudimentary prioritization based on the high level strategy.
+
+One interesting aspect of the five bots described above is that, while all of them are reactive at the lower level (unit control), most if not all of them, are scripted at the highest level of abstraction.
+BTHAI reads build and squad formations from a predefined script, Nova's {\em Strategy Manager} is a predefined finite-state machine, BBQ's construction manager reads the build order from a predefined script, and Skynet's {\em BuildOrder Manager} is basically a predefined script.
+Such scripts describe the strategy that the bots will use, however, such strategy is always fixed.
+One could see this pre-scripting as if each bot defined a ``high-level programming language'' to describe Starcraft strategies, and the bots themselves are just interpreters of such strategy.
+Compared to current approaches for Chess or Go, this scripting seems a rigid and inflexible, but responds to the much higher complexity of the Starcraft game.
+
+In conclusion, we can see that there are two basic tools that can be used in an integration architecture: abstraction and divide-and-conquer, which are widely used by the existing Starcraft bots.
+However, several open questions remain:
+
+{\color{red}
+This section is missing:
+\begin{itemize}
+\item Strengths and weaknesses?
+\item Which AI techniques do each bot use for each task?
+\end{itemize}
 }
@@
 
-\appendix
-
-\section*{This is the old task-division section}
-
-{\color{red}
-I've included here the section we had for not losing it :)
-}
-
-{\color{blue}
-Here we should say that complex RTS games require multi-scale reasoning \cite{WeberCig10}, and thus complex architectures that integrate many AI modules have been devised.
-Discuss how different researchers have divided up the task of playing Starcraft into a collection of subtasks and integrated them (see the end of this paper for a preliminary write-up on this)
-}
-
-Playing an RTS game involves dealing with all the problems described above.
-A few approaches, like CAT \cite{LTW}, Darmok \cite{OntanonMSR10} or ALisp \cite{Marthi05} try to deal with the problem in a monolithic manner, by using a single AI technique.
-This resembles approaches to solve other games, such as Chess or Go, where a single game-tree search approach is enough to play the game at human level.
-However, none of those systems aims at achieving near human performance.
-In order to achieve human-level performance, RTS AI designers use a lot of domain knowledge in order to divide the task of playing the game into a collection of sub-problems, which can be dealt-with using individual AI techniques (as discussed in the previous section).
-Thus, an integration architecture is required to put together all of those techniques into a single coherent system, which is the focus of this section.
-
-\begin{figure*}[ta]
-    \centering
-    \includegraphics[width=\textwidth]{figures/figure-bot-architectures-wide.pdf}
-    \caption{General architecture of 5 Starcraft AI bots.}
-    \label{fig:bot-architecture}
-\end{figure*}
-
-Figure \ref{fig:bot-architecture} shows some representative examples of the integration architectures used by different bots in the AIIDE 2011 Starcraft AI competition \cite{url1}.
-Each box represents an individual module with a clearly defined task (only modules with a black background can send actions directly to Starcraft).
-Dashed arrows represent data flow, and solid arrows represent control (when a module can command another module to perform some task).
-For example, we can see how SPAR is divided in two sets of modules: {\em situation analysis} and {\em decision making}, the first with three modules dedicated to analyze the current situation of the game, and the later with 4 modules dedicated to exploit that information to decide what to do.
-We can see how the decision making aspect of SPAR is organized hierarchically, with the higher-level module ({\em strategic decision}) issuing commands to the next module ({\em tactical decision}), which sends commands to the next module ({\em action implementation}), and so on.
-Only the lower-level modules can send actions directly to Starcraft.
-
-On the other hand, bots such as NOVA or BroodwarBotQ (BBQ) only use a hierarchical organization for {\em micro-management} (controlling the attack units), but use a decentralized organization for the rest of the bot.
-In Nova and BBQ, there is a collection of modules that control different aspects of the game (workers, production, construction, etc.).
-These modules can all send actions directly to Starcraft.
-In Nova those modules coordinate only through writing data in a shared blackboard, and in BBQ they coordinate only when they have to use a shared resource (unit) by means of an arbitrator.
-
-By analyzing the structure of these bots, we can see that there are two main tools that can be used when designing an integration architecture:
-
-\begin{itemize}
-\item {\em Abstraction}: complex tasks can be formulated at different levels of abstraction.
-For example, playing an RTS game can be seen as issuing individual low-level actions to each of the units in the game, or at a higher level, it can be seen as deploying a specific strategy (e.g.
-a ``BBS strategy'', or a ``Reaver Drop'' strategy).
-Some bots, reason at multiple levels of abstraction at the same time, making the task of playing Starcraft simpler.
-Assuming that each module in the architecture of a bot has a goal and determines some actions to achieve that goal, the actions determined by higher-level modules are considered as the goals of the lower level modules.
-In this way, each module can focus on reasoning at only one level of abstraction, thus, making the problem easier.
-
-\item {\em Divide-and-conquer}: playing a complex RTS, such as Starcraft, requires performing many conceptually different tasks, such as gathering resources, attacking, placing buildings, etc.
-Assuming each of these tasks can be performed relatively independently and without interference, we can have one module focusing on each of the tasks independently, thus making the problem easier.
-\end{itemize}
-
-If we imagine the different tasks to perform in a complex RTS game in a two-dimensional plane, where the vertical axis represents abstraction, and the horizontal axis represents the different aspects of the game (micro-management, resource gathering, etc.), abstraction can be seen as dividing the space with horizontal lines, whereas divide-and-conquer divides the space using vertical lines.
-
-Different bots, use different combinations of these two tools. Looking back at Figure \ref{fig:bot-architecture}, we can see the following use of abstraction and divide-in-conquer in the bots:
-
-\begin{itemize}
-\item BTHAI: uses a two-tier abstraction hierarchy, where a collection of high-level modules command a collection of lower-level agents in charge of each of the units.
-At the high-level, BTHAI uses divide-and-conquer, having multiple high-level modules issuing commands to the lower-level units.
-\item BBQ: uses abstraction for micro-management, and divide-and-conquer for macro-management and intelligence gathering.
-To avoid conflicts between modules (since the individual tasks of each of the modules are not completely independent), BBQ uses an arbitrator.
-\item Nova: is similar in design as BBQ, and uses abstraction for micro-management, and divide-and-conquer for macro-management.
-The differences are that Nova does not have an arbitrator to resolve conflicts, but has a higher-level module ({\em strategy manager}), which posts information to the blackboard that the rest of modules follow (thus, making use of abstraction).
-\item SPAR: only uses abstraction.
-Its high-level module determines the strategy to use, and the tactical decision module divides it into a collection of {\em abstract actions}, that are executed by the lower-level modules.
-\item Skynet: makes extensive use of both abstraction and divide-and-conquer.
-Considering the decision making component of Skynet, we can see a high level module that issues commands to a series of tactics modules.
-The collection of tactic modules queue {\em tasks} (that are analogous to the abstract actions used in SPAR).
-Each different task has a specific low level module that knows how to execute it.
-Thus, Skynet uses a 3 layered abstraction hierarchy, and uses divide-and-conquer in all levels except the highest.
-\end{itemize}
-
-Additionally, except for BTHAI, all other agents use divide-and-conquer at a higher-level bot design and divide all the modules into two or three categories: {\em information gathering} and {\em decision}, or {\em information gathering}, {\em micro-management} and {\em macro-management}.
-
-Notice that most bots using divide-and-conquer (except for BBQ), assume that each of the modules can act independently and that their actions can be executed without interference.
-BBQ is the exception, including an arbitrator that makes sure that modules do not send contradictory orders to the same unit.
-However, very little bots handle the problem of how to coordinate resource usage amongst modules, for instance BTHAI uses a first-come-first-serve policy for spending resources, the first module that requests resources is the one that gets them.
-The exceptions are Nova and Skynet implement some rudimentary prioritization based on the high level strategy.
-
-One interesting aspect of the five bots described above is that, while all of them are reactive at the lower level (unit control), most if not all of them, are scripted at the highest level of abstraction.
-BTHAI reads build and squad formations from a predefined script, Nova's {\em Strategy Manager} is a predefined finite-state machine, BBQ's construction manager reads the build order from a predefined script, and Skynet's {\em BuildOrder Manager} is basically a predefined script.
-Such scripts describe the strategy that the bots will use, however, such strategy is always fixed.
-One could see this pre-scripting as if each bot defined a ``high-level programming language'' to describe Starcraft strategies, and the bots themselves are just interpreters of such strategy.
-Compared to current approaches for Chess or Go, this scripting seems a bit rigid and inflexible, but responds to the much higher complexity of the Starcraft game.
-
-In conclusion, we can see that there are two basic tools that can be used in an integration architecture: abstraction and divide-and-conquer, which are widely used by the existing Starcraft bots.
-However, several open questions remain:
-
-{\color{blue}
-\begin{itemize}
-\item Strengths and weaknesses?
-\item Resource coordination in a divide-and-conquer approaches?
-\item High-level strategies that do not depend on scripting? (not here, this should be discussed in Section) % \ref{sec:planning})
-\end{itemize}
-}
-
 % Can use something like this to put references on a page

### 2012-03-08 survey.tex
Updated the bot comparison section to include UAlbertaBot

@@
 
-{\color{red} add UAlbertaBot}
-
 Thanks to the recent organization of international game AI competitions focused around the popular StarCraft game (see Section \ref{sec:competition}), several groups have been working on integrating many of the techniques described in the previous section into complete ``bots'', capable of playing complete StarCraft games.
@@
     \includegraphics[width=\textwidth]{figures/figure-bot-architectures-wide.pdf}
-    \caption{General architecture of 5 Starcraft AI bots.}
+    \caption{Architecture of 6 Starcraft bots. Modules with black background sent commands directly to Starcraft, dashed arrows represent data flow, and solid arrows represent control.}
     \label{fig:bot-architecture}
@@
 
-Figure \ref{fig:bot-architecture} shows some representative examples of the integration architectures used by different bots in the AIIDE and CIG 2011 Starcraft AI competitions \cite{url1,url2}: BTHAI \cite{???}, BroodwarQ \cite{???}, SPAR \cite{???}, Nova \cite{???}, and Skynet \cite{???} (see Section \ref{sec:competition} for a comparison of their performance).
+Figure \ref{fig:bot-architecture} shows some representative examples of the integration architectures used by different bots in the AIIDE and CIG 2011 Starcraft AI competitions \cite{url1,url2}: BroodwarQ \cite{???}, Nova \cite{???}, UAlbertaBot \cite{???}, Skynet \cite{???}, SPAR \cite{???}, and BTHAI \cite{???} (see Section \ref{sec:competition} for a comparison of their performance).
 Each box represents an individual module with a clearly defined task (only modules with a black background can send actions directly to Starcraft).
@@
 \begin{itemize}
-\item BTHAI: uses a two-tier abstraction hierarchy, where a collection of high-level modules command a collection of lower-level agents in charge of each of the units.
-At the high-level, BTHAI uses divide-and-conquer, having multiple high-level modules issuing commands to the lower-level units.
-\item BBQ: uses abstraction for micro-management, and divide-and-conquer for macro-management and intelligence gathering.
+\item BroodwarBotQ: uses abstraction for micro-management, and divide-and-conquer for macro-management and intelligence gathering.
 To avoid conflicts between modules (since the individual tasks of each of the modules are not completely independent), BBQ uses an arbitrator.
-\item Nova: is similar in design as BBQ, and uses abstraction for micro-management, and divide-and-conquer for macro-management.
+\item Nova: is similar in design as BroodwarBotQ, and uses abstraction for micro-management, and divide-and-conquer for macro-management.
 The differences are that Nova does not have an arbitrator to resolve conflicts, but has a higher-level module ({\em strategy manager}), which posts information to the blackboard that the rest of modules follow (thus, making use of abstraction).
-\item SPAR: only uses abstraction.
-Its high-level module determines the strategy to use, and the tactical decision module divides it into a collection of {\em abstract actions}, that are executed by the lower-level modules.
+\item UAlbertaBot: also uses abstraction in micro-management like the previous two bots.
+But it also uses it in macro-management: as can be seen, the production manager sends commands to the building manager, who is in charge of producing the buildings.
+This bot also uses divide-and-conquer, and tasks like scouting and resource gathering are managed by separate, independent modules.
 \item Skynet: makes extensive use of both abstraction and divide-and-conquer.
@@
 Thus, Skynet uses a 3 layered abstraction hierarchy, and uses divide-and-conquer in all levels except the highest.
+\item SPAR: only uses abstraction.
+Its high-level module determines the strategy to use, and the tactical decision module divides it into a collection of {\em abstract actions}, that are executed by the lower-level modules.
+\item BTHAI: uses a two-tier abstraction hierarchy, where a collection of high-level modules command a collection of lower-level agents in charge of each of the units.
+At the high-level, BTHAI uses divide-and-conquer, having multiple high-level modules issuing commands to the lower-level units.
 \end{itemize}
@@
 
-Notice that most bots using divide-and-conquer (except for BBQ), assume that each of the modules can act independently and that their actions can be executed without interference.
-BBQ is the exception, including an arbitrator that makes sure that modules do not send contradictory orders to the same unit.
+Most bots using divide-and-conquer (except for BBQ and UAlbertaBot), assume that each of the modules can act independently and that their actions can be executed without interference.
+BBQ and UAlbertaBot use an arbitrator ({\em Game Commander'} in UAlbertaBot) that makes sure that modules do not send contradictory orders to the same unit.
 However, very little bots handle the problem of how to coordinate resource usage amongst modules, for instance BTHAI uses a first-come-first-serve policy for spending resources, the first module that requests resources is the one that gets them.
@@
 Compared to current approaches for Chess or Go, this scripting seems a rigid and inflexible, but responds to the much higher complexity of the Starcraft game.
+An interesting exception to that is UAlbertaBot, which uses a search algorithm in the {\em Production Manager} to find near-optimal build orders.
 

### 2012-03-09 analyzis/totext.py
fixes for totest.py + features upgrades/units

@@ -18,32 +18,53 @@ header = True
 body = False
 names = {}
 races = {}
-time_built = {}
+time_had = {}
 buildings = {'T' : ['Expansion', 'Comsat_Station', 'Nuclear_Silo', 'Supply_Depot', 'Refinery', 'Barracks', 'Academy', 'Factory', 'Starport', 'Control_Tower', 'Science_Facility', 'Covert_Ops', 'Physics_Lab', 'Machine_Shop', 'Engineering_Bay', 'Armory', 'Missile_Turret', 'Bunker'],
         'P' : ['Expansion', 'Robotics_Facility', 'Pylon', 'Assimilator', 'Observatory', 'Gateway', 'Photon_Cannon', 'Citadel_of_Adun', 'Cybernetics_Core', 'Templar_Archives', 'Forge', 'Stargate', 'Fleet_Beacon', 'Arbiter_Tribunal', 'Robotics_Support_Bay', 'Shield_Battery'], 
         'Z' : ['Overlord', 'Infested_Command_Center', 'Expansion', 'Lair', 'Hive', 'Nydus_Canal', 'Hydralisk_Den', 'Defiler_Mound', 'Greater_Spire', 'Queens_Nest', 'Evolution_Chamber', 'Ultralisk_Cavern', 'Spire', 'Spawning_Pool', 'Creep_Colony', 'Spore_Colony', 'Sunken_Colony', 'Extractor']}
+doubling = {'T' : ['Expansion', 'Refinery', 'Barracks', 'Factory', 'Armory'],
+        'P' : ['Expansion', 'Pylon', 'Assimilator', 'Gateway'],
+        'Z' : ['Expansion', 'Extractor']}
+tripling = {'T' : ['Expansion', 'Barracks', 'Factory'],
+        'P' : ['Expansion', 'Gateway'],
+        'Z' : ['Expansion']}
+quadrupling = {'T' : ['Expansion'],
+        'P' : ['Expansion', 'Gateway'],
+        'Z' : ['Expansion']}
+units = {'T' : ['SCV', 'Dropship', 'Marine', 'Ghost', 'Vulture', 'Goliath', 'Siege Tank Tank Mode', 'Wraith', 'Science Vessel', 'Battlecruiser', 'Siege Tank Siege Mode', 'Firebat', 'Medic', 'Valkyrie' ], 
[…]

### 2012-03-09 analyzis/totext.py
per bot txt output for games from LMR format

@@ -17,6 +17,7 @@ if len(sys.argv) > 1:
 header = True
 body = False
 names = {}
+files_bots = {}
 races = {}
 time_had = {}
 buildings = {'T' : ['Expansion', 'Comsat_Station', 'Nuclear_Silo', 'Supply_Depot', 'Refinery', 'Barracks', 'Academy', 'Factory', 'Starport', 'Control_Tower', 'Science_Facility', 'Covert_Ops', 'Physics_Lab', 'Machine_Shop', 'Engineering_Bay', 'Armory', 'Missile_Turret', 'Bunker'],
@@ -43,22 +44,35 @@ def write(tb, fo):
     for p, t in tb.iteritems():
         for e in buildings[races[p][0]]:
             b = races[p] + '_' + e
-            fo.write(b + ' ' + str(t.get(b, 0)/24) + '; ')
+            twr = b + ' ' + str(t.get(b, 0)/24) + '; '
+            fo.write(twr)
+            files_bots[names[p]].write(twr)
         for e in doubling[races[p][0]]:
             b = races[p] + '_' + e + '2'
-            fo.write(b + ' ' + str(t.get(b, 0)/24) + '; ')
+            twr = b + ' ' + str(t.get(b, 0)/24) + '; '
+            fo.write(twr)
+            files_bots[names[p]].write(twr)
         for e in tripling[races[p][0]]:
             b = races[p] + '_' + e + '3'
-            fo.write(b + ' ' + str(t.get(b, 0)/24) + '; ')
+            twr = b + ' ' + str(t.get(b, 0)/24) + '; '
+            fo.write(twr)
+            files_bots[names[p]].write(twr)
         for e in quadrupling[races[p][0]]:
             b = races[p] + '_' + e + '4'
-            fo.write(b + ' ' + str(t.get(b, 0)/24) + '; ')
+            twr = b + ' ' + str(t.get(b, 0)/24) + '; '
+            fo.write(twr)
+            files_bots[names[p]].write(twr)
         for e in units[races[p][0]]:
             u = races[p] + '_' + e
-            fo.write(u + ' ' + str(t.get(u, 0)/24) + '; ')
+            twr = u + ' ' + str(t.get(u, 0)/24) + '; '
+            fo.write(twr)
+            files_bots[names[p]].write(twr)
         for e in rchupgs[races[p][0]]:
             r = races[p] + '_' + e
-            fo.write(r + ' ' + str(t.get(r, 0)/24) + '; ')
[…]

### 2012-03-09 analyzis/totext.py
correcting some bugs in totest.py

@@ -18,6 +18,7 @@ header = True
 body = False
 names = {}
 files_bots = {}
+files_races = {}
 races = {}
 time_had = {}
 buildings = {'T' : ['Expansion', 'Comsat_Station', 'Nuclear_Silo', 'Supply_Depot', 'Refinery', 'Barracks', 'Academy', 'Factory', 'Starport', 'Control_Tower', 'Science_Facility', 'Covert_Ops', 'Physics_Lab', 'Machine_Shop', 'Engineering_Bay', 'Armory', 'Missile_Turret', 'Bunker'],
@@ -32,47 +33,54 @@ tripling = {'T' : ['Expansion', 'Barracks', 'Factory'],
 quadrupling = {'T' : ['Expansion'],
         'P' : ['Expansion', 'Gateway'],
         'Z' : ['Expansion']}
-units = {'T' : ['SCV', 'Dropship', 'Marine', 'Ghost', 'Vulture', 'Goliath', 'Siege Tank Tank Mode', 'Wraith', 'Science Vessel', 'Battlecruiser', 'Siege Tank Siege Mode', 'Firebat', 'Medic', 'Valkyrie' ], 
-        'P' : ['Probe', 'Shuttle', 'Observer', 'Dragoon', 'Zealot', 'Archon', 'Reaver', 'High Templar', 'Arbiter', 'Carrier', 'Scout', 'Dark Archon', 'Corsair', 'Dark Templar'], 
+units = {'T' : ['SCV', 'Dropship', 'Marine', 'Ghost', 'Vulture', 'Goliath', 'Siege_Tank_Tank_Mode', 'Wraith', 'Science_Vessel', 'Battlecruiser', 'Siege_Tank_Siege_Mode', 'Firebat', 'Medic', 'Valkyrie' ], 
+        'P' : ['Probe', 'Shuttle', 'Observer', 'Dragoon', 'Zealot', 'Archon', 'Reaver', 'High_Templar', 'Arbiter', 'Carrier', 'Scout', 'Dark_Archon', 'Corsair', 'Dark_Templar'], 
         'Z' : ['Drone', 'Overlord', 'Zergling', 'Devourer', 'Guardian', 'Ultralisk', 'Queen', 'Hydralisk', 'Mutalisk', 'Scourge', 'Lurker', 'Defiler']}
 rchupgs = {'T' : ['Spider_Mines', ],
[…]

### 2012-03-22 analyzis/totext.py
added the --names option to totext.py

@@ -12,6 +12,7 @@ import sys
 # python totext.py LMRFILEHERE [--bots] [--races]
 # --bots will also make one .txt file for each bot
 # --races will also make one .txt file for each race
+# --names will also put bots names in the big .txt file
 
 f = sys.stdin
 o = sys.stdout
@@ -24,6 +25,8 @@ if len(sys.argv) > 1:
         do_bots = True
     if "--races" in sys.argv:
         do_races = True
+    if "--names" in sys.argv:
+        do_names = True
 
 
 header = True
@@ -108,6 +111,8 @@ def write(tb, fo):
             files_bots[names[p]].write('\n')
         if do_races:
             files_races[races[p]].write('\n')
+        if do_names:
+            fo.write('Name ' + names[p] + '; ')
         fo.write('\n')
     if loser != "":
         write_for_one(loser, tb.pop(loser))

### 2012-03-22 survey.tex
Unifying section 'Existing work'

--- 2012-03-08 survey.tex
@@
 
-{\color{blue}
-Introduce the game plus screenshot. Not too lengthy.
-}
-{\color{ForestGreen}
-Here it is a brief description, I don't think we need a screenshot.
-}
 Starcraft: Brood War is a popular RTS game released in 1998 by Blizzard Entertainment.
@@
 
-{\color{blue}
-Include here the literature review of existing work on RTS Game AI.
-We could also compare with what is being used in industry, and use the real AI of Starcraft as an example (Bob Fitch's talk in AIIDE provides enough material for that).
-I have included here as subsections the two blurbs of text we had for integration of domain knowledge and reasoning with uncertainty, since I guess they correspond here.
+\subsection{Perception}
+Every AI needs to observe the world in order to analyze it and take decisions. In RTS games this task isn't trivial since we have to deal with imperfect information.
+
+\textbf{Spatial}
+
+Terrain analysis supplies the AI with chunks of abstract information about the map in order to help in making decisions.
+This analysis is usually performed off-line, in order to save processing time during gameplay.
+
+Pottinger \cite{Pottinger00} described \emph{BANG!} engine implemented by Ensemble Studios for the game Age of Empires 2.
+This engine provides terrain analysis functionalities to the game using influence maps and areas with connectivity information.
+Later works \cite{Forbus01} showed the importance to have qualitative spatial information, and some others \cite{Karavelas04}\cite{Hale08} focused on spatial decomposition.
+The most recent work \cite{Perkins10} presented an algorithm using Voronoi diagrams to detect regions and relevant choke points.
+
+\textbf{Temporal}
+
+In \cite{WeberAIIDE11} a particle model is used to tracks opponent units under fog of war over time.
+
+{\color{ForestGreen}
+Anyone did anything here? I think only Skynet uses temporal reasoning...
+Maybe heat maps?? but I can't find any reference...
 }
 
-\subsection{Micro-management}
+\textbf{Enemy}
 
-Some interesting uses of reinforcement learning (RL) \cite{Sutton} to RTS research are in the domain of \textit{micro-management} (group and individual units control): \cite{Marthi05} employs concurrent hierarchical Q-learning (units Q-functions are combined at the group level) RL to efficiently control units in a ``one robot with multiple effectors'' fashion.
-\cite{Madeira06} advocates the use of prior domain knowledge to allow faster RL learning and applied their work on a turn-based strategy game.
-In real game setups, the actions spaces to explore is gigantic, it requires to use the structure of the game in a partial program (or a partial Markov decision process) and a shape function (or a heuristic) \cite{Marthi05}.
-There are other was to learn parameters of an underlying model, for instance \cite{GA} used evolutionary learning techniques, but face the same problem of dimensionality.
-The difficulty to work with multi-scale goals and plans is handled directly by case-based reasoning (CBR), which has been adapted for micro-management with continuous action models \cite{Molineaux08}, an integrated RL/CBR algorithm using continous models, or with hybrid CBR/RL transfer learning \cite{CBR-RL}.
-Reactive planning (for behavior languages) \cite{WeberCig10}, a decompositional planning similar to hierarchical task networks \cite{HTNPlanning}, allows for plans to be changed at different granularity levels and so for multi-scale (hierarchical) goals integration.
+Opponent modeling is the process of keeping track of what the enemy is doing during a game in order to estimate the probability of it using a specific strategy.
 
-Potential fields \cite{Hagelback08,Hagelback09} are a powerful way to specify complex (with the combination of several fields, requiring a parameter-tuning method) behavior policies and are used in modern commercial games and RTS AI are using derivative of it (Supreme Commander flowfield).
-Self-organizing-maps (SOM) learned with evolutionary algorithms provide some way to learn (realistic) parameters for RTS units behavior \cite{teamCompositionRTS}.
-Bayesian programming has been applied to inverse fusion of the sensory inputs of the units \cite{SynnaeveMicroCig11}, which subsumes potential fields, allowing for integration of tactical goals directly in micro-management.
-There are also some cognitive approaches to RTS AI \cite{SORTS}, in which there is a human-like attention model for micro-management.
-
-\subsection{Opponent modeling, strategy prediction}
-
-There are related works in the domains of opponent modeling \cite{HsiehS08,SchaddBS07,OBRecog}.
-The main methods used to these ends are case-based reasoning (CBR) and planning or plan recognition \cite{LTW,CBR_Planning,OntanonCBR,HTNPlanning,Ramirez}.
-There are precedent works of Bayesian plan recognition \cite{BMPR}, even in games with Albrecht \textit{et al.} \cite{BayesianRecog} using dynamic Bayesian networks to recognize a user's plan in a multi-player dungeon adventure.
-
-\cite{LTW} used CBR to perform dynamic plan retrieval extracted from domain knowledge in Wargus (Warcraft II clone).
+The main methods used to these ends are case-based reasoning (CBR) and planning or plan recognition \cite{LTW,CBR_Planning,OntanonCBR,HTNPlanning,Ramirez}.\cite{LTW} used CBR to perform dynamic plan retrieval extracted from domain knowledge in Wargus (Warcraft II clone).
 \cite{CBR_Planning} base their real-time case-based planning (CBP) system on a plan dependency graph which is learned from human demonstration.
@@
 
-\cite{SchaddBS07} describe opponent modeling through hierarchically structured models of the opponent behaviour and they applied their work to the Spring RTS (Total Annihilation clone).
+\cite{SchaddBS07} describe opponent modeling through hierarchically structured models of the opponent behavior and they applied their work to the Spring RTS (Total Annihilation clone).
 \cite{HTNPlanning} use hierarchical task networks (HTN) to model strategies in a first person shooter with the goal to use HTN planners.
@@
 \cite{HMMstrat_RTS_AIIDE11} used HMM to learn the transition probabilities of sequences of constructions and kept the most probables to produce probabilistic behavior models (in StarCraft).
-\cite{SynnaeveOpeningCig11} continued on \cite{WeberCig09} and used a Bayesian semi-supervised (labeling based on clustering) model to learn from replays and predict openings (early game strategies) from StarCraft replays, while \cite{SynnaeveAIIDE11} presents a whole game Bayesian tech-tree predictor based on unserpervised learning from replays.
+\cite{SynnaeveOpeningCig11} continued on \cite{WeberCig09} and used a Bayesian semi-supervised (labeling based on clustering) model to learn from replays and predict openings (early game strategies) from StarCraft replays, while \cite{SynnaeveAIIDE11} presents a whole game Bayesian tech-tree predictor based on unsupervised learning from replays.
+
+Most commercial games avoid gather this information making their AI omniscient and using perfect information all the time.
+But sometimes they simulate some scouting tasks as Bob Fitch described in his AIIDE 2011 keynote for the game series WarCraft and StarCraft I.
+
+{\color{ForestGreen}
+Nobody had described an efficient way to scout the opponent... better opponent information better opponent modeling...
+}
+
+\textbf{Ally}
+
+{\color{ForestGreen}
+Anyone did anything here?
+}
+
+\subsection{Strategy}
+
+The most common decision making are Decision Trees.
+Some of the advantages of the decision trees are: simple to understand and interpret acting like a white box where we can explain why a decision is taken, and they are compatible with other decision techniques.
+For complex decisions we have many algorithms to generate an optimum decision tree like ID3 or C4.5, depends on the problem we will use one or another.
+
+A good alternative are Finite State Machine.
+The idea behind FSM is to decompose an object's behavior into easily manageable "chunks" or states, such as "attacking an enemy", "gathering resources" or "repairing" and establish the conditions that trigger transitions between them.
+
+\subsection{Giving orders}
+
+{\color{ForestGreen}
+Explain centralized models (like blackboard) and distributed/hierarchically models.
+}
+
+\subsection{Executing orders}
+
+\textbf{Pathfinding}
+
+The most common pathfinding algorithm is A*, but the big problem of this algorithm is the time and  memory consumption, a show-stopper in a real-time environment.
+
+To solve this the search space needs to be reduced as much as possible.
+One common technique is reducing the tile map search into a navigation graph map.
+To do this some polygonal triangulations are calculated and abstracted on a grid-based map and this is the TRA* (Triangulation Reduction A*) algorithm\cite{Demyen_2006}.
+
+{\color{ForestGreen}
+There are a lot of new pathfinding algorithms. Do we need to review all of them?
+}
+
+\textbf{Group behavior}
+
+Craig Reynols introduced algorithmic steering behaviors to simulate the aggregate motion of swarms using procedural animation\cite{Reynolds_1999}.
+His model describes how a system of multiple agents , each steering according to simple local rules, can produce the collective movement found in flocks of birds.
+With this behaviours we are able to simulate intelligent squad movements and do tactical moves to flank the opponent in a RTS game\cite{Danielsiek_2008}, improving the survival of our units.
+
+A good alternative to Flocking is using Reciprocal Velocity Obstacles (RVO)\cite{Berg_2008}, since it is easier to control the behavior and they guarantee to prevent collisions.
+With this algorithm we can move a large squad crossing another squad and avoid all possible collisions between units.
+This type of techniques are already applied in commercial RTS games like Starcraft 2 or Supreme Commander 2 with great success.
+
+Another popular technique is Potential Fields.
+basically, Potential Fields work like a charged particle moving through a magnetic field.
+The process of tuning and optimizing the parameters of potential field charges can be time reduced applying learning algorithms such as Q-Learning\cite{Liu_2008}.
+Another common issue we have to deal with when using potential fields is the local optima problem that can stuck the unit away from the goal.
+Some examples of the utilities of potential files in RTS games are avoiding obstacles (navigation), avoiding opponent fire, or staying at maximum shooting distance\cite{Hagelback08,Hagelback09}.
+
+There are some interesting uses of reinforcement learning (RL) \cite{Sutton}: \cite{Marthi05} employs concurrent hierarchical Q-learning (units Q-functions are combined at the group level) RL to efficiently control units in a ``one robot with multiple effectors'' fashion.
+\cite{Madeira06} advocates the use of prior domain knowledge to allow faster RL learning and applied their work on a turn-based strategy game.
+In real game setups, the actions spaces to explore is gigantic, it requires to use the structure of the game in a partial program (or a partial Markov decision process) and a shape function (or a heuristic) \cite{Marthi05}.
+There are other was to learn parameters of an underlying model, for instance \cite{GA} used evolutionary learning techniques, but face the same problem of dimensionality.
+The difficulty to work with multi-scale goals and plans is handled directly by case-based reasoning (CBR), which has been adapted for units behavior with continuous action models \cite{Molineaux08}, an integrated RL/CBR algorithm using continuous models, or with hybrid CBR/RL transfer learning \cite{CBR-RL}.
+Reactive planning (for behavior languages) \cite{WeberCig10}, a decompositional planning similar to hierarchical task networks \cite{HTNPlanning}, allows for plans to be changed at different granularity levels and so for multi-scale (hierarchical) goals integration.
+
+Self-organizing-maps (SOM) learned with evolutionary algorithms provide some way to learn (realistic) parameters for RTS units behavior \cite{teamCompositionRTS}.
+Bayesian programming has been applied to inverse fusion of the sensory inputs of the units \cite{SynnaeveMicroCig11}, which subsumes potential fields, allowing for integration of tactical goals directly in micro-management.
+There are also some cognitive approaches to RTS AI \cite{SORTS}, in which there is a human-like attention model for micro-management.
 
@@
 Whereas a player is doing action A to follow it up by action B or C depends on its intention, while the information about action A is visible and information about B or C is hidden, only intention allows for a choice.
-}
-
-\subsection{Terrain Analysis}
-{\color{ForestGreen}
-Terrain analysis supplies the AI with chunks of abstract information about the map in order to help in making decisions.
-Terrain analysis can provide automatic detection of regions, choke points and other areas of interest.
-This analysis is usually performed off-line, in order to save processing time during gameplay.
-
-The final goal of terrain analysis is having all the useful tactical information of the map.
-To do this we can use different algorithms, one of the most used, coming from computer vision, being \textit{skeletons by influence zones}, also known as \textit{Voronoi diagrams}\cite{Perkins_2010}.
-Through this kind of diagrams we will be able to generate a graph with the centroid of the regions and the choke points of each region.
-
-Another common technique is the utilization of an influence map to detect and mark relevant zones of the map as resources location.
-}
-
-\subsection{Pathfinding}
-{\color{ForestGreen}
-The most common pathfinding algorithm is A*, but the big problem of this algorithm is the time and  memory consumption, a show-stopper in a real-time environment.
-
-To solve this the search space needs to be reduced as much as possible.
-One common technique is reducing the tile map search into a navigation graph map.
-To do this some polygonal triangulations are calculated and abstracted on a grid-based map and this is the TRA* (Triangulation Reduction A*) algorithm\cite{Demyen_2006}.
-
-** There are a lot of new pathfinding algorithms. Do we need to review all of them? **
-}
-
-\subsection{Group behavior}
-{\color{ForestGreen}
-Craig Reynols introduced algorithmic steering behaviors to simulate the aggregate motion of swarms using procedural animation\cite{Reynolds_1999}.
-His model describes how a system of multiple agents , each steering according to simple local rules, can produce the collective movement found in flocks of birds.
-With this behaviours we are able to simulate intelligent squad movements and do tactical moves to flank the opponent in a RTS game\cite{Danielsiek_2008}, improving the survival of our units.
-
-A good alternative to Flocking is using Reciprocal Velocity Obstacles (RVO)\cite{Berg_2008}, since it is easier to control the behavior and they guarantee to prevent collisions.With this algorithm we can move a large squad crossing another squad and avoid all possible collisions between units.
-This type of techniques are already applied in commercial RTS games like Starcraft 2 or Supreme Commander 2 with great success.
-
-Another popular technique is Potential Fields.
-basically, Potential Fields work like a charged particle moving through a magnetic field.
-The process of tuning and optimizing the parameters of potential field charges can be time reduced applying learning algorithms such as Q-Learning\cite{Liu_2008}.
-Another common issue we have to deal with when using potential fields is the local optima problem that can stuck the unit away from the goal.
-Some examples of the utilities of potential files in RTS games are avoiding obstacles (navigation), avoiding opponent fire, or staying at maximum shooting distance\cite{Hagelback08}\cite{Hagelback09}.
-}
-
-\subsection{Decision Making}
-{\color{ForestGreen}
-The most common decision making are Decision Trees.
-Some of the advantages of the decision trees are: simple to understand and interpret acting like a white box where we can explain why a decision is taken, and they are compatible with other decision techniques.
-For complex decisions we have many algorithms to generate an optimum decision tree like ID3 or C4.5, depends on the problem we will use one or another.
-
-A good alternative are Finite State Machine.
-The idea behind FSM is to decompose an object's behavior into easily manageable "chunks" or states, such as "attacking an enemy", "gathering resources" or "repairing" and establish the conditions that trigger transitions between them.
-}
-
-\subsection{Opponent modelling}
-{\color{ForestGreen}
-Opponent modelling is the process of keeping track of what the enemy is doing during a game in order to estimate the probability of it using a specific strategy.
-The key point here is to predict what units the enemy will build, when he will build them, when and where he will attack and so on.
-This will give us information about what kind of opponent it is allowing us to adapt our strategy and select the best counter.
-
-Science we are dealing with imperfect information, the best solution is to build a Bayesian model which can deal perfectly with the uncertainty.
-The difficult part is building that model.
-Some approaches use traces of game logs to analyze the state of the world, but those traces have all the information of the game, so we have to add some noise to emulate the real use of the model in a non-deterministic world.
-
-A Threat Map or Influence Map is a way to store relevant information about the opponent and keep track of the conflict areas or the secure ones.
-Statical threat maps can be designed where the threat is always "real" or concepts like heat map over time can be applied to make predictions.
 }
@@
 \end{document}
+
+In his AIIDE 2010 keynote, Chris Jurnet described the technique XXX, implemented by [COMPANY] in the game YYY \cite{gameurl}

### 2012-05-04 survey.tex

@@
 Thanks to the recent organization of international game AI competitions focused around the popular StarCraft game (see Section \ref{sec:competition}), several groups have been working on integrating many of the techniques described in the previous section into complete ``bots'', capable of playing complete StarCraft games.
-In this section we will overview some of the currently available top bots, and focus on the architecture being used to integrate all the techniques being used for the different aspects of the bot.
+In this section we will overview some of the currently available top bots, and focus, first on the architecture being used to integrate all the techniques being used for the different aspects of the bot, and then on which subsets of techniques are used for each bot.
+
+\subsection{RTS Bot Architectures}\label{sec:integration}
 
@@
 These modules can all send actions directly to Starcraft.
-In Nova those modules coordinate only through writing data in a shared blackboard, and in BBQ they coordinate only when they have to use a shared resource (unit) by means of an arbitrator.
+In Nova those modules coordinate mostly through writing data in a shared blackboard, and in BBQ they coordinate only when they have to use a shared resource (unit) by means of an arbitrator.
 
@@
 In conclusion, we can see that there are two basic tools that can be used in an integration architecture: abstraction and divide-and-conquer, which are widely used by the existing Starcraft bots.
-However, several open questions remain:
 
-{\color{red}
-This section is missing:
-\begin{itemize}
-\item Strengths and weaknesses?
-\item Which AI techniques do each bot use for each task?
-\end{itemize}
-}
+This section has focused on discussing the architecture of existing Starcraft bots.
+Let us now focus on what is the state of the art on the techniques used for each of the individual modules used by them.
+
+\subsection{Individual AI Techniques}\label{sec:techniques}
+
+{\color{red} TODO}
 

### 2012-05-04 survey.tex
added a ref + beginning cleaning

@@
 \subsection{Perception}
-Every AI needs to observe the world in order to analyze it and take decisions. In RTS games this task isn't trivial since we have to deal with imperfect information.
+XXX Every AI needs to observe the world in order to analyze it and take decisions. In RTS games this task isn't trivial since we have to deal with imperfect information.
 
-\textbf{Spatial}
+%\textbf{Spatial}
 
-Terrain analysis supplies the AI with chunks of abstract information about the map in order to help in making decisions.
+XXX Terrain analysis supplies the AI with chunks of abstract information about the map in order to help in making decisions.
 This analysis is usually performed off-line, in order to save processing time during gameplay.
 
-Pottinger \cite{Pottinger00} described \emph{BANG!} engine implemented by Ensemble Studios for the game Age of Empires 2.
+Pottinger \cite{Pottinger00} described \emph{BANG!} engine implemented by Ensemble Studios for the game Age of Empires II.
 This engine provides terrain analysis functionalities to the game using influence maps and areas with connectivity information.
-Later works \cite{Forbus01} showed the importance to have qualitative spatial information, and some others \cite{Karavelas04}\cite{Hale08} focused on spatial decomposition.
-The most recent work \cite{Perkins10} presented an algorithm using Voronoi diagrams to detect regions and relevant choke points.
 
-\textbf{Temporal}
+\cite{Forbus2002} showed the importance to have qualitative spatial informationfor wargames, for which they used geometric and pathfinding analysis.
 
-In \cite{WeberAIIDE11} a particle model is used to tracks opponent units under fog of war over time.
+\cite{Hale08} presented a 2D geometric navigation mesh generation method from expanding convex regions from seeds.
 
-{\color{ForestGreen}
-Anyone did anything here? I think only Skynet uses temporal reasoning...
-Maybe heat maps?? but I can't find any reference...
-}
+\cite{Perkins10} applied Voronoi decomposition (then pruning) %\cite{Karavelas04}
+to detect regions and relevant choke points in RTS maps. This approach is implemented for StarCraft in BWTA\footnote{\url{http://code.google.com/p/bwta/}}.
 
-\textbf{Enemy}
+%\textbf{Temporal}
+% Isla 2006 Game AI Programming Wisdom 3. Charles River Media. chapter Probabilistic Target Tracking and Search Using Occupancy Maps, 379–388. ?
 
-Opponent modeling is the process of keeping track of what the enemy is doing during a game in order to estimate the probability of it using a specific strategy.
+\cite{Hladky2008} benchmarked hidden semi-Markov models (HSMM) and particle filters in FPS units tracking.
+They showed that the accuracy of occupancy maps was improved using movement models (learned from the player behavior) in HSMM.
+% comparable with human experts ?
+
+\cite{HagelbackJ08} presented a multi-agent potential fields based bot able to deal with fog of war in Tankbattle. XXX
+
+\cite{WeberAIIDE11} used a particle model, with a simple linear trajectory update, to track opponent units under fog of war in StarCraft.
+
+% {\color{ForestGreen}
+% Anyone did anything here? I think only Skynet uses temporal reasoning...
+% Maybe heat maps?? but I can't find any reference...
+% }
+
+%\textbf{Enemy}
+
+XXX Opponent modeling is the process of keeping track of what the enemy is doing during a game in order to estimate the probability of it using a specific strategy.
 
@@
 
-\textbf{Ally}
-
-{\color{ForestGreen}
-Anyone did anything here?
-}
+% \textbf{Ally}
+% 
+% {\color{ForestGreen}
+% Anyone did anything here?
+% }
 

### 2012-05-04 survey.tex
re-organizing the previous works sections

@@
 
-\subsection{Perception}
+\begin{figure*}
+    \centering
+    \includegraphics[width=0.9\columnwidth]{figures/levels_abstraction.pdf}
+    \caption{RTS AI levels of abstraction and their properties: uncertainty (from intentions and from partial observation) is going higher as the abstraction levels are raising.
+The timings on the right correspond to an estimate of the duration of a behavior switch in StarCraft.}
+    \label{fig:levels-abstraction}
+\end{figure*}
+
+%\subsection{Integration of Domain Knowledge}\label{subsec:domainknowledge}
+{\color{blue}
+All bots attempt at integrating human domain knowledge in a way or another, be it in the form of build-orders, heuristics, strategies, or by automatically learning from human replays.
+This has been shown to be key to good performance in bots (for now, as Dave was saying: ``Starcraft AI is like Chess was in the 60s'').
+
+Talk about learning, scripting, and other techniques to incorporate such knowledge.
+
+When talking about learning in RTS games, with regard to a particular match against an opponent, one has to differentiate between:
+\begin{itemize}
+\item prior learning: done before the match, like when a bot learns what is likely to happen on a given match-up (and map) from replays, or when a bot optimizes its micro-management parameters through simulated annealing or genetic evolution.
+\item in-match learning: done between the different games of match, ``the opponent rushed me last game so I will play this one safe''.
+\item in-game learning: done between the different actions of the game, ``I tried to attack here and it was a failure, let us try somewhere else''.
+\end{itemize}
+All these learning can be supervised (labeled replays, scripted situations) or unsupervised (unlabeled data, exploratory situations).
+The last two kinds of situations of learning are more inclined towards reinforcement learning because the AI has to deal with the exploration-exploitation duality.
+%prior learning: supervised \cite{WeberCIG09} \cite{SynnaeveCIG11}, unsupervised: \cite{SynnaeveAIIDE11}
+}
+
+%\subsection{Reasoning with Uncertainty}\label{subsec:uncertainty}
+{\color{blue}
+I think there are two important aspects of uncertainty in RTS games:
+\begin{itemize}
+\item First is trying to model uncertainty and try to reduce it (information gathering, and representation of knowledge), this includes scouting, and maintaining good estimates of locations and number of units of the enemy, etc.
+\item Second is actually using the uncertain information. Taking decisions like army strength estimation (like to attack or retreat).
+\end{itemize}
+
+There are two main kinds of uncertainty, existing for two different reasons, in RTS games.
+First, there is the incompleteness of information, due to the fog of war, which leads to some kind of ``extensional''
+%``extensive'' 
+uncertainty, that can be lowered by good scouting, and knowledge representation and reasoning (to infer what is possible given what has been seen).
+Also, there is the fact that we will never be able to read the opponent's mind, the ``intentional''
+%or ``intensive'' 
+uncertainty that we can't really be sure to scratch off, even with the best scouting possible.
+For intentional uncertainty, the AI, as the human player, can only build a sensible model of what the opponent is likely to do, to think, or to think knowing what we think he thinks.
+Both these kinds of uncertainty can be studied through probabilities, but only $P(Hidden | Seen)$ can be \textit{directly} linked to data only for extensional uncertainty.
+%now for the clumsy part:
+One could argue that the extensional data is the only thing that matters because it is the only thing that has an influence on the game, but in reality, for very highly skilled players, discovering the intentions of the enemy is the key to winning.
+It allows for a compact understanding of the game state and development, while also giving additional information about the future.
+Whereas a player is doing action A to follow it up by action B or C depends on its intention, while the information about action A is visible and information about B or C is hidden, only intention allows for a choice.
+}
+
+\subsection{Strategy}
 XXX Every AI needs to observe the world in order to analyze it and take decisions. In RTS games this task isn't trivial since we have to deal with imperfect information.
@@
 
-\cite{WeberAIIDE11} used a particle model, with a simple linear trajectory update, to track opponent units under fog of war in StarCraft.
+\cite{WeberAIIDE11} used a particle model with a simple linear trajectory update to track opponent units under fog of war in StarCraft.
 
@@
 
-\subsection{Strategy}
+\subsection{Tactics}
 
@@
 
-\subsection{Giving orders}
+\subsection{Units Control}
 
-{\color{ForestGreen}
-Explain centralized models (like blackboard) and distributed/hierarchically models.
-}
-
-\subsection{Executing orders}
-
-\textbf{Pathfinding}
-
+% {\color{ForestGreen}
+% Explain centralized models (like blackboard) and distributed/hierarchically models.
+% }
 The most common pathfinding algorithm is A*, but the big problem of this algorithm is the time and  memory consumption, a show-stopper in a real-time environment.
-
 To solve this the search space needs to be reduced as much as possible.
@@
 
-{\color{ForestGreen}
-There are a lot of new pathfinding algorithms. Do we need to review all of them?
-}
-
-\textbf{Group behavior}
-
-Craig Reynols introduced algorithmic steering behaviors to simulate the aggregate motion of swarms using procedural animation\cite{Reynolds_1999}.
-His model describes how a system of multiple agents , each steering according to simple local rules, can produce the collective movement found in flocks of birds.
-With this behaviours we are able to simulate intelligent squad movements and do tactical moves to flank the opponent in a RTS game\cite{Danielsiek_2008}, improving the survival of our units.
+Craig Reynolds introduced algorithmic steering behaviors to simulate the aggregate motion of swarms using procedural animation\cite{Reynolds_1999}.
+His model describes how a system of multiple agents, each steering according to simple local rules, can produce the collective movement found in flocks of birds.
+With this behaviors, we are able to simulate intelligent squad movements and do tactical moves to flank the opponent in a RTS game\cite{Danielsiek_2008,SynnaeveMicroCig11}, improving the survival of our units.
 
 A good alternative to Flocking is using Reciprocal Velocity Obstacles (RVO)\cite{Berg_2008}, since it is easier to control the behavior and they guarantee to prevent collisions.
-With this algorithm we can move a large squad crossing another squad and avoid all possible collisions between units.
+With this algorithm, we can move a large squad crossing another squad and avoid all possible collisions between units.
 This type of techniques are already applied in commercial RTS games like Starcraft 2 or Supreme Commander 2 with great success.
@@
 There are also some cognitive approaches to RTS AI \cite{SORTS}, in which there is a human-like attention model for micro-management.
-
-%\subsection{Integration of Domain Knowledge}\label{subsec:domainknowledge}
-
-{\color{blue}
-All bots attempt at integrating human domain knowledge in a way or another, be it in the form of build-orders, heuristics, strategies, or by automatically learning from human replays.
-This has been shown to be key to good performance in bots (for now, as Dave was saying: ``Starcraft AI is like Chess was in the 60s'').
-
-Talk about learning, scripting, and other techniques to incorporate such knowledge.
-
-When talking about learning in RTS games, with regard to a particular match against an opponent, one has to differentiate between:
-\begin{itemize}
-\item prior learning: done before the match, like when a bot learns what is likely to happen on a given match-up (and map) from replays, or when a bot optimizes its micro-management parameters through simulated annealing or genetic evolution.
-\item in-match learning: done between the different games of match, ``the opponent rushed me last game so I will play this one safe''.
-\item in-game learning: done between the different actions of the game, ``I tried to attack here and it was a failure, let us try somewhere else''.
-\end{itemize}
-All these learning can be supervised (labeled replays, scripted situations) or unsupervised (unlabeled data, exploratory situations).
-The last two kinds of situations of learning are more inclined towards reinforcement learning because the AI has to deal with the exploration-exploitation duality.
-%prior learning: supervised \cite{WeberCIG09} \cite{SynnaeveCIG11}, unsupervised: \cite{SynnaeveAIIDE11}
-}
-
-%\subsection{Reasoning with Uncertainty}\label{subsec:uncertainty}
-
-{\color{blue}
-I think there are two important aspects of uncertainty in RTS games:
-\begin{itemize}
-\item First is trying to model uncertainty and try to reduce it (information gathering, and representation of knowledge), this includes scouting, and maintaining good estimates of locations and number of units of the enemy, etc.
-\item Second is actually using the uncertain information. Taking decisions like army strength estimation (like to attack or retreat).
-\end{itemize}
-
-There are two main kinds of uncertainty, existing for two different reasons, in RTS games.
-First, there is the incompleteness of information, due to the fog of war, which leads to some kind of ``extensional''
-%``extensive'' 
-uncertainty, that can be lowered by good scouting, and knowledge representation and reasoning (to infer what is possible given what has been seen).
-Also, there is the fact that we will never be able to read the opponent's mind, the ``intentional''
-%or ``intensive'' 
-uncertainty that we can't really be sure to scratch off, even with the best scouting possible.
-For intentional uncertainty, the AI, as the human player, can only build a sensible model of what the opponent is likely to do, to think, or to think knowing what we think he thinks.
-Both these kinds of uncertainty can be studied through probabilities, but only $P(Hidden | Seen)$ can be \textit{directly} linked to data only for extensional uncertainty.
-%now for the clumsy part:
-One could argue that the extensional data is the only thing that matters because it is the only thing that has an influence on the game, but in reality, for very highly skilled players, discovering the intentions of the enemy is the key to winning.
-It allows for a compact understanding of the game state and development, while also giving additional information about the future.
-Whereas a player is doing action A to follow it up by action B or C depends on its intention, while the information about action A is visible and information about B or C is hidden, only intention allows for a choice.
-}
 

### 2012-05-11 survey.tex
Clarified that diagrams come from analyzing sourcecode
reworked the first part of the state of the art, intro to problems

@@
 
-\begin{figure*}
-    \centering
-    \includegraphics[width=0.9\columnwidth]{figures/levels_abstraction.pdf}
-    \caption{RTS AI levels of abstraction and their properties: uncertainty (from intentions and from partial observation) is going higher as the abstraction levels are raising.
-The timings on the right correspond to an estimate of the duration of a behavior switch in StarCraft.}
-    \label{fig:levels-abstraction}
-\end{figure*}
+\subsection{Problems}
 
-%\subsection{Integration of Domain Knowledge}\label{subsec:domainknowledge}
 {\color{blue}
-All bots attempt at integrating human domain knowledge in a way or another, be it in the form of build-orders, heuristics, strategies, or by automatically learning from human replays.
-This has been shown to be key to good performance in bots (for now, as Dave was saying: ``Starcraft AI is like Chess was in the 60s'').
+\textbf{Planning}
+TODO (Santi?) just a paragraph on exposing why there is a planning problem in RTS AI.
 
-Talk about learning, scripting, and other techniques to incorporate such knowledge.
+\textbf{Spatial and Temporal Reasoning}
+TODO just a paragraph on exposing why there are temporal and spatial reasoning in RTS AI.
 
+\textbf{Learning}\\
 When talking about learning in RTS games, with regard to a particular match against an opponent, one has to differentiate between:
 \begin{itemize}
-\item prior learning: done before the match, like when a bot learns what is likely to happen on a given match-up (and map) from replays, or when a bot optimizes its micro-management parameters through simulated annealing or genetic evolution.
-\item in-match learning: done between the different games of match, ``the opponent rushed me last game so I will play this one safe''.
-\item in-game learning: done between the different actions of the game, ``I tried to attack here and it was a failure, let us try somewhere else''.
+\item prior learning: done before the match, like when a bot learns what is likely to happen on a given match-up (and map) from replays, or when a bot optimizes its parameters beforehand through simulated annealing or genetic evolution.
+\item in-game learning: done between the different actions (decisions) taken during a game.
+For instance, a bot can take into account the successes and failures of its attacks (location, type, timings).
+\item in-match learning: done between the different games of match, it encompasses opponent modeling across multiple games and adapting the bots' strategy to its successes and failures.
 \end{itemize}
@@
 %prior learning: supervised \cite{WeberCIG09} \cite{SynnaeveCIG11}, unsupervised: \cite{SynnaeveAIIDE11}
-}
 
-%\subsection{Reasoning with Uncertainty}\label{subsec:uncertainty}
-{\color{blue}
-I think there are two important aspects of uncertainty in RTS games:
-\begin{itemize}
-\item First is trying to model uncertainty and try to reduce it (information gathering, and representation of knowledge), this includes scouting, and maintaining good estimates of locations and number of units of the enemy, etc.
-\item Second is actually using the uncertain information. Taking decisions like army strength estimation (like to attack or retreat).
-\end{itemize}
+%%% I think there are two important aspects of uncertainty in RTS games:
+%%% \begin{itemize}
+%%% \item First is trying to model uncertainty and try to reduce it (information gathering, and representation of knowledge), this includes scouting, and maintaining good estimates of locations and number of units of the enemy, etc.
+%%% \item Second is actually using the uncertain information. Taking decisions like army strength estimation (like to attack or retreat).
+%%% \end{itemize}
 
-There are two main kinds of uncertainty, existing for two different reasons, in RTS games.
-First, there is the incompleteness of information, due to the fog of war, which leads to some kind of ``extensional''
+\textbf{Uncertainty}\\
+In RTS games, there are two main kinds of uncertainty, existing for two different reasons.
+First, there is incompleteness of information, due to the fog of war, which leads to some kind of ``extensional''
 %``extensive'' 
-uncertainty, that can be lowered by good scouting, and knowledge representation and reasoning (to infer what is possible given what has been seen).
-Also, there is the fact that we will never be able to read the opponent's mind, the ``intentional''
+uncertainty, that can be lowered by good scouting, and knowledge representation (to infer what is possible given what has been seen).
+Secondly, there is the fact that we will never be able to read the opponent's mind, the ``intentional''
 %or ``intensive'' 
-uncertainty that we can't really be sure to scratch off, even with the best scouting possible.
+uncertainty that we cannot really be sure to scratch off, even with the best scouting possible.
 For intentional uncertainty, the AI, as the human player, can only build a sensible model of what the opponent is likely to do, to think, or to think knowing what we think he thinks.
-Both these kinds of uncertainty can be studied through probabilities, but only $P(Hidden | Seen)$ can be \textit{directly} linked to data only for extensional uncertainty.
+Both these kinds of uncertainty can be studied through probabilities, but only for extensional uncertainty can $P(Hidden | Seen)$ be \textit{directly} linked to data.
 %now for the clumsy part:
@@
 }
+
+\subsection{Decomposition}
+\begin{figure*}
+    \centering
+    \includegraphics[width=0.9\columnwidth]{figures/levels_abstraction.pdf}
+    \caption{RTS AI levels of abstraction and their properties: uncertainty (from intentions and from partial observation) is going higher as the abstraction levels are raising.
+The timings on the right correspond to an estimate of the duration of a behavior switch in StarCraft.
+Spatial and temporal reasoning are indicated for part for which greedy solutions are really not efficient.}
+    \label{fig:levels-abstraction}
+\end{figure*}
+As a work on RTS AI has to address several of these aforementioned problems together, we decided to study decompose RTS AI in three tasks, widely known by RTS players: strategy, tactics and units control (micro-management), see Figure~\ref{fig:levels-abstraction}.
+We consider strategy to be everything related to the \textit{technology tree}, being the buildings and the upgrades, and the army composition.
+It is the most intentional level, as a player performs a strategy with future stances (aggressive, defensive, economy, technology) and tactics in mind.
+To change its strategy, we consider that a player need about 3 minutes: they should need to construct a few new buildings and/or research some upgrades and/or produce units changing the army composition.
+Tactics represent the attacks and defenses.
+For the human player, decisions on tactics are made from spatial (exploiting the terrain) and temporal (army movements) reasoning, constrained on the possible types of attacks by the army composition of the player and their opponent.
+Units control, also called micro-management, is how the player takes action to maximize the efficiency of its units (applying the tactics) in real-time (simultaneous) and adversary conditions.
+As shown in Figure~\ref{fig:levels-abstraction}, strategy constrains possible tactics which in turn constrain units control.
+On a prediction/adaptation point of view, the units observation level gives information about the opponent's tactics and strategy.
+Flaws in opponents tactics can also steer our strategy into exploiting them.\\
+perhaps TODO examples?
 
@@
     \includegraphics[width=\textwidth]{figures/figure-bot-architectures-wide.pdf}
-    \caption{Architecture of 6 Starcraft bots. Modules with black background sent commands directly to Starcraft, dashed arrows represent data flow, and solid arrows represent control.}
+    \caption{Architecture of 6 Starcraft bots obtained by analyzing their source code.
+Modules with black background sent commands directly to Starcraft, dashed arrows represent data flow, and solid arrows represent control.}
     \label{fig:bot-architecture}

### 2012-05-11 survey.tex
moving out blue
correct colors/plan

@@
 
-\subsection{Starcraft}\label{subsec:starcraft}
+\subsection{StarCraft {\color{blue}- Situation}}\label{subsec:starcraft}
 
@@
 
-\subsection{Challenges in RTS Game AI}\label{subsec:challenges}
+{\color{blue}TODO: explain the gameplay a little more, how are ppl not familiar to SC able to understand the problems for AI that we raise below?}
+
+\subsection{Challenges in RTS Game AI {\color{blue}- Problems}}\label{subsec:challenges}
 
@@
 
-\section{Existing work on RTS Game AI}\label{sec:review}
-
-\subsection{Problems}
-
 {\color{blue}
+\begin{LARGE}
+TODO merge above and bottom\\
+\end{LARGE}
 \textbf{Planning}
@@
 TODO just a paragraph on exposing why there are temporal and spatial reasoning in RTS AI.
+}
 
@@
 Whereas a player is doing action A to follow it up by action B or C depends on its intention, while the information about action A is visible and information about B or C is hidden, only intention allows for a choice.
-}
 
-\subsection{Decomposition}
+\section{Existing work on RTS Game AI - \color{blue}{Resolutions}}\label{sec:review}
+
+\subsection{RTS AI Decomposition}
 \begin{figure*}
@@
 Flaws in opponents tactics can also steer our strategy into exploiting them.\\
+{\color{blue}
 perhaps TODO examples?
+}
 
@@
 
-\section{State of the Art Bots for Starcraft}\label{sec:bot}
+\section{State of the Art Bots for Starcraft - \color{blue}{Information/Productions}}\label{sec:bot}
 

### 2012-05-11 survey.tex
review: strategy

@@
 
-\subsection{Strategy}
-XXX Every AI needs to observe the world in order to analyze it and take decisions. In RTS games this task isn't trivial since we have to deal with imperfect information.
-
-%\textbf{Spatial}
-
-XXX Terrain analysis supplies the AI with chunks of abstract information about the map in order to help in making decisions.
-This analysis is usually performed off-line, in order to save processing time during gameplay.
-
-Pottinger \cite{Pottinger00} described \emph{BANG!} engine implemented by Ensemble Studios for the game Age of Empires II.
-This engine provides terrain analysis functionalities to the game using influence maps and areas with connectivity information.
-
-\cite{Forbus2002} showed the importance to have qualitative spatial informationfor wargames, for which they used geometric and pathfinding analysis.
-
-\cite{Hale08} presented a 2D geometric navigation mesh generation method from expanding convex regions from seeds.
-
-\cite{Perkins10} applied Voronoi decomposition (then pruning) %\cite{Karavelas04}
-to detect regions and relevant choke points in RTS maps. This approach is implemented for StarCraft in BWTA\footnote{\url{http://code.google.com/p/bwta/}}.
-
-%\textbf{Temporal}
-% Isla 2006 Game AI Programming Wisdom 3. Charles River Media. chapter Probabilistic Target Tracking and Search Using Occupancy Maps, 379–388. ?
-
-\cite{Hladky2008} benchmarked hidden semi-Markov models (HSMM) and particle filters in FPS units tracking.
-They showed that the accuracy of occupancy maps was improved using movement models (learned from the player behavior) in HSMM.
-% comparable with human experts ?
-
-\cite{HagelbackJ08} presented a multi-agent potential fields based bot able to deal with fog of war in Tankbattle. XXX
-
-\cite{WeberAIIDE11} used a particle model with a simple linear trajectory update to track opponent units under fog of war in StarCraft.
-
-% {\color{ForestGreen}
-% Anyone did anything here? I think only Skynet uses temporal reasoning...
-% Maybe heat maps?? but I can't find any reference...
-% }
-
-%\textbf{Enemy}
-
-XXX Opponent modeling is the process of keeping track of what the enemy is doing during a game in order to estimate the probability of it using a specific strategy.
-
-The main methods used to these ends are case-based reasoning (CBR) and planning or plan recognition \cite{LTW,CBR_Planning,OntanonCBR,HTNPlanning,Ramirez}.\cite{LTW} used CBR to perform dynamic plan retrieval extracted from domain knowledge in Wargus (Warcraft II clone).
-\cite{CBR_Planning} base their real-time case-based planning (CBP) system on a plan dependency graph which is learned from human demonstration.
-In \cite{OntanonCBR,PlanRetrieval}, they use CBR and expert demonstrations on Wargus.
-They update their previous CPB approach by using ``situation assessment'' for plan retrieval.
-They improve the speed of CPB by using a decision tree to select relevant features.
-Hsieh and Sun \cite{HsiehS08} based their work on \cite{LTW} CBR model and used StarCraft replays to construct states and building sequences.
-Strategies are choices of building construction order in their model.
-
-\cite{SchaddBS07} describe opponent modeling through hierarchically structured models of the opponent behavior and they applied their work to the Spring RTS (Total Annihilation clone).
-\cite{HTNPlanning} use hierarchical task networks (HTN) to model strategies in a first person shooter with the goal to use HTN planners.
-\cite{OBRecog} improve the probabilistic hostile agent task tracker (PHATT \cite{PHATT}, a simulated HMM for plan recognition) by encoding strategies as HTN.
-%\cite{Madeira06}
-\cite{WeberCig09} proposed ``a data mining approach to strategy prediction'' and performed supervised learning on StarCraft replays.
-\cite{HMMstrat_RTS_AIIDE11} used HMM to learn the transition probabilities of sequences of constructions and kept the most probables to produce probabilistic behavior models (in StarCraft).
-\cite{SynnaeveOpeningCig11} continued on \cite{WeberCig09} and used a Bayesian semi-supervised (labeling based on clustering) model to learn from replays and predict openings (early game strategies) from StarCraft replays, while \cite{SynnaeveAIIDE11} presents a whole game Bayesian tech-tree predictor based on unsupervised learning from replays.
-
-Most commercial games avoid gather this information making their AI omniscient and using perfect information all the time.
-But sometimes they simulate some scouting tasks as Bob Fitch described in his AIIDE 2011 keynote for the game series WarCraft and StarCraft I.
-
-{\color{ForestGreen}
-Nobody had described an efficient way to scout the opponent... better opponent information better opponent modeling...
+{\color{blue}
+perhaps TODO ``how it's done in the industry'' regrouping the first paragraph of each of the following 3 sections?
 }
-
-% \textbf{Ally}
-% 
-% {\color{ForestGreen}
-% Anyone did anything here?
-% }
-
-\subsection{Tactics}
-
-The most common decision making are Decision Trees.
+%The most common decision making are Decision Trees.
 Some of the advantages of the decision trees are: simple to understand and interpret acting like a white box where we can explain why a decision is taken, and they are compatible with other decision techniques.
@@
 
-A good alternative are Finite State Machine.
-The idea behind FSM is to decompose an object's behavior into easily manageable "chunks" or states, such as "attacking an enemy", "gathering resources" or "repairing" and establish the conditions that trigger transitions between them.
+\subsection{Strategy}
+%XXX Every AI needs to observe the world in order to analyze it and take decisions. In RTS games this task isn't trivial since we have to deal with imperfect information.
+
+In the industry, strategies are often ``hard coded'' with finite state machines (FSM) \cite{FSM_AIGameProgWisdom2003}.
+The idea behind FSM is to decompose an object's behavior into easily manageable states, such as ``attacking'', ``gathering resources'' or ``repairing'' and establish the conditions that trigger transitions between them.
+Hierarchical FSM give higher (strategic) states some control over lower oned (actions).
+However, strategic decisions have to be made in a context, not in a vacuum.
+It is hard to encode dynamic, adaptive behaviors in (H)FSM due to the exponential number of transitions (between states) to account for.
+To this end, several approaches have been developed like hierarchical task networks, behavior trees and the comeback of STRIPS \cite{FikesSTRIPS} planning in the industry \cite{orkinGDC_FEAR}.
+Regardless, one cannot but notice that strategy adaptation is not the strong point of industrial RTS AI.
+Strategy adaptation is part of what is called opponent modeling.
+An adapting robot needs to keep track of what the enemy is doing in order to estimate the probability that they will use a specific strategy and adapt its own strategy to it.
+
+In research, the main methods used for strategic decision-making are case-based reasoning (CBR) and planning or plan recognition \cite{LTW,CBR_Planning,OntanonCBR,HTNPlanning,Ramirez}.
+Aha et al.
+\cite{LTW} used CBR to perform dynamic plan retrieval extracted from domain knowledge in Wargus (Warcraft II clone).
+Onta\~{n}\'{o}n et al.
+\cite{CBR_Planning} base their real-time case-based planning (CBP) system on a plan dependency graph which is learned from human demonstration.
+In \cite{OntanonCBR,PlanRetrieval}, they use CBR learned from expert demonstrations on Wargus.
+They update their previous CBP approach by using situation assessment for retrieving the ``right'' plan, and improve the speed of CBP by using a decision tree to select relevant features.
+Hsieh and Sun \cite{HsiehS08} based their work on \cite{LTW} CBR model and used StarCraft replays to construct states and building sequences (``build orders'').
+Schadd et al.
+\cite{SchaddBS07} describe opponent modeling through hierarchically structured models of the opponent behavior and they applied their work to the Spring RTS (Total Annihilation clone).
+Hoang et al.
+\cite{HTNPlanning} use hierarchical task networks (HTN) to model strategies in a first person shooter with the goal to use HTN planners.
+Kabanza et al.
+\cite{OBRecog} improve the probabilistic hostile agent task tracker (PHATT \cite{PHATT}, a simulated HMM for plan recognition) by encoding strategies as HTN.
+%Madeira et al.
+\cite{Madeira06}
+
+Weber and Mateas \cite{WeberCig09} proposed ``a data mining approach to strategy prediction'' and performed supervised learning (from buildings features) on labeled StarCraft replays.
+Dereszynski et al.
+\cite{HMMstrat_RTS_AIIDE11} used HMM to learn the transition probabilities of sequences of constructions and kept the most probables to produce probabilistic behavior models (in StarCraft).
+Synnaeve and Bessi\`{e}re \cite{SynnaeveOpeningCig11} extended \cite{WeberCig09} and used a Bayesian semi-supervised (labeling based on EM clustering considering appropriate features) model to learn from replays and predict openings (early game strategies) from StarCraft replays.
+Then \cite{SynnaeveAIIDE11}, they presented an unsupervised learning Bayesian model for tech-tree prediction, still using replays.
+
+Most commercial games avoid gathering this information by using perfect information all the time (cheating).
+Sometimes they simulate some scouting tasks as Bob Fitch described in his AIIDE 2011 keynote for the WarCraft game series and StarCraft: Broodwar.
+\{\color{blue}XXX (perhaps move it above, to the start of subsection}
+
+%{\color{ForestGreen}
+%Nobody had described an efficient way to scout the opponent... better opponent information better opponent modeling...
+%}
+
+\subsection{Tactics}
+Terrain analysis supplies the AI with chunks of abstract information about the map in order to help in making decisions.
+This analysis is usually performed off-line, in order to save processing time during gameplay.
+Pottinger \cite{Pottinger00} described the \emph{BANG} engine implemented by Ensemble Studios for the game Age of Empires II.
+This engine provides terrain analysis functionalities to the game using influence maps and areas with connectivity information.
+Forbus et al.
+\cite{Forbus2002} showed the importance to have qualitative spatial information for wargames, for which they used geometric and pathfinding analysis.
+Hale et al.
+\cite{Hale08} presented a 2D geometric navigation mesh generation method from expanding convex regions from seeds.
+Finally, Perkins \cite{Perkins10} applied Voronoi decomposition (then pruning) %\cite{Karavelas04}
+to detect regions and relevant choke points in RTS maps. This approach is implemented for StarCraft in BWTA\footnote{\url{http://code.google.com/p/bwta/}}.
+% Isla 2006 Game AI Programming Wisdom 3. Charles River Media. chapter Probabilistic Target Tracking and Search Using Occupancy Maps, 379–388. ?
+
+Hladky and Bulitko \cite{Hladky2008} benchmarked hidden semi-Markov models (HSMM) and particle filters in FPS units tracking.
+They showed that the accuracy of occupancy maps was improved using movement models (learned from the player behavior) in HSMM.
+% comparable with human experts ?
+Hagelb\"{a}ck and Johansson \cite{HagelbackJ08} presented a multi-agent potential fields based bot able to deal with fog of war in Tankbattle. XXX 
+Weber et al. \cite{WeberAIIDE11} used a particle model with a simple linear trajectory update to track opponent units under fog of war in StarCraft.
 

### 2012-05-11 survey.tex
first pass on state-of-the-art

@@
 
-{\color{blue}
-perhaps TODO ``how it's done in the industry'' regrouping the first paragraph of each of the following 3 sections?
-}
+%{\color{blue}
+%perhaps TODO ``how it's done in the industry'' regrouping the first paragraph of each of the following 3 sections?
+%}
+
+\subsection{Strategy}
+%Every AI needs to observe the world in order to analyze it and take decisions. In RTS games this task isn't trivial since we have to deal with imperfect information.
 %The most common decision making are Decision Trees.
@@
 For complex decisions we have many algorithms to generate an optimum decision tree like ID3 or C4.5, depends on the problem we will use one or another.
-
-\subsection{Strategy}
-%XXX Every AI needs to observe the world in order to analyze it and take decisions. In RTS games this task isn't trivial since we have to deal with imperfect information.
 
@@
 \cite{HTNPlanning} use hierarchical task networks (HTN) to model strategies in a first person shooter with the goal to use HTN planners.
-Kabanza et al.
-\cite{OBRecog} improve the probabilistic hostile agent task tracker (PHATT \cite{PHATT}, a simulated HMM for plan recognition) by encoding strategies as HTN.
 %Madeira et al.
@@
 Sometimes they simulate some scouting tasks as Bob Fitch described in his AIIDE 2011 keynote for the WarCraft game series and StarCraft: Broodwar.
-\{\color{blue}XXX (perhaps move it above, to the start of subsection}
+{\color{blue}XXX (perhaps move it above, to the start of subsection}
 
@@
 \subsection{Tactics}
-Terrain analysis supplies the AI with chunks of abstract information about the map in order to help in making decisions.
-This analysis is usually performed off-line, in order to save processing time during gameplay.
+Units have different abilities, which leads to different possible tactics.
+Most often, fast or invisible or flying units are not cost-effective in head-to-head fights against slower bulky units.
+Taking good tactical decisions is taking advantage of the terrain and positions of the different units (in a timely manner).
+Terrain analysis supplies the AI with structured information about the map in order to help making decisions.
+This analysis is usually performed off-line, in order to save CPU time during the game.
 Pottinger \cite{Pottinger00} described the \emph{BANG} engine implemented by Ensemble Studios for the game Age of Empires II.
@@
 
-Hladky and Bulitko \cite{Hladky2008} benchmarked hidden semi-Markov models (HSMM) and particle filters in FPS units tracking.
+Hladky and Bulitko \cite{Hladky2008} benchmarked hidden semi-Markov models (HSMM) and particle filters in first person shooter games (FPS) units tracking.
 They showed that the accuracy of occupancy maps was improved using movement models (learned from the player behavior) in HSMM.
-% comparable with human experts ?
-Hagelb\"{a}ck and Johansson \cite{HagelbackJ08} presented a multi-agent potential fields based bot able to deal with fog of war in Tankbattle. XXX 
-Weber et al. \cite{WeberAIIDE11} used a particle model with a simple linear trajectory update to track opponent units under fog of war in StarCraft.
+Hagelb\"{a}ck and Johansson \cite{HagelbackJ08} presented a multi-agent potential fields based bot able to deal with fog of war in Tankbattle.
+Weber et al.
+\cite{WeberAIIDE11} used a particle model with a simple linear trajectory update to track opponent units under fog of war in StarCraft.
+They also produced tactical goals through reactive planning and goal-driven autonomy \cite{WeberCig10,Weber10}, finding the more relevant goal(s) to spawn in unforeseen situations.
+Kabanza et al.
+\cite{OBRecog} improve the probabilistic hostile agent task tracker (PHATT \cite{PHATT}, a simulated HMM for plan recognition) by encoding strategies as HTN, used for plan and intent recognition to find tactical opportunities.
+Sharma et al.
+\cite{CBR-RL} combined CBR and reinforcement learning to enable reuse of tactical plan components.
+Cadena and Garrido \cite{CadenaG11} used fuzzy CBR (fuzzy case matching) for strategic and tactical planning.
+\cite{Chung05} adapted Monte-Carlo tree search (MCTS) to planning in RTS games and applied it to a capture-the-flag mod of Open RTS.
+Balla and Fern \cite{UCT} applied upper confidence bounds on trees (UCT: a MCTS algorithm) to tactical assault planning in Wargus.
+On influence maps, \cite{teamCompositionRTS} studied team composition and maneuvering by learning a self-organizing map, while \cite{HagelbackJ08} presented a multi-agent potential field based bot.
+Finally, Avery et al.
+\cite{Avery09} and Smith et al.
+\cite{SmithCIG10} co-evolved influence map trees for spatial (tactical) reasoning in RTS games.
 
 \subsection{Units Control}
+Units control, also called micro-management in RTS games, aims at maximizing the effectiveness of units, it starts with pathfinding and goes up to simultaneous control of units of different types in complex battles on heterogeneous terrain.
+The most common pathfinding algorithm is A*, but its big problem is CPU time and memory consumption, hard to satisfy in a complex, dynamic, real-time environment on multiple units.
+%To solve this the search space needs to be reduced as much as possible.
+A common technique consists in reducing the search into a navigation graph map.
+Among others, Triangulation Reduction A* computes polygonal triangulations on a grid-based map \cite{Demyen_2006}.
+One of the problems of A* related approaches in RTS games is that moving units block/get in the way of/ each others and the path has to be recomputed often.
+Craig Reynolds \cite{Reynolds_1999} introduced algorithmic steering behaviors, which simulate the aggregate motion of swarms.
+His model describes how a system of multiple agents, each steering according to simple local rules, can produce the collective movement found in flocks of birds.
+Danielsiek et al.
+\cite{Danielsiek_2008} studied intelligent squad movements to do tactical moves to flank the opponent in a RTS game this way.
+An alternative to flocking is to use reciprocal velocity obstacles (RVO) \cite{Berg_2008}, since it is easier to control the behavior and they guarantee to prevent collisions.
+With this algorithm, we can move a large squad crossing another squad and avoid all possible collisions between units.
+{\color{blue}Do we put RVO?
+As I don't remember it being used in any game or game AI research.} In recent commercial RTS games like Starcraft 2 or Supreme Commander 2, flocking-like behaviors are inspired of continuum crowds (``flow field'') \cite{Treuille2006}, and a comprehensive review about (grid-based) pathfinding was recently done by Sturtevant \cite{sturtevant2012benchmarks}.
 
-% {\color{ForestGreen}
-% Explain centralized models (like blackboard) and distributed/hierarchically models.
-% }
-The most common pathfinding algorithm is A*, but the big problem of this algorithm is the time and  memory consumption, a show-stopper in a real-time environment.
-To solve this the search space needs to be reduced as much as possible.
-One common technique is reducing the tile map search into a navigation graph map.
-To do this some polygonal triangulations are calculated and abstracted on a grid-based map and this is the TRA* (Triangulation Reduction A*) algorithm\cite{Demyen_2006}.
+Another popular technique is potential fields (PF) and influence maps.
+Basically, PF work like a charged particle moving through a magnetic field.
+The parameters of PF charges can be learned through Q-Learning \cite{Liu_2008}.
+Another common issue of potential fields is the local optima problem that can stuck the unit away from the goal.
+Some uses of potential fields in RTS games are: avoiding obstacles (navigation), avoiding opponent fire, or staying at maximum shooting distance \cite{Hagelback08,Hagelback09}.
+Self-organizing-maps (SOM) learned with evolutionary algorithms provide some way to learn (realistic) parameters for RTS units behavior \cite{teamCompositionRTS}.
+Bayesian modeling has been applied to inverse fusion of the sensory inputs of the units \cite{SynnaeveMicroCig11}, which subsumes potential fields, allowing for integration of tactical goals directly in micro-management.
+There is also a cognitive approach to RTS AI \cite{SORTS}, using a human-like attention model (with units grouping and vision of a unique screen location) for micro-management.
 
-Craig Reynolds introduced algorithmic steering behaviors to simulate the aggregate motion of swarms using procedural animation\cite{Reynolds_1999}.
-His model describes how a system of multiple agents, each steering according to simple local rules, can produce the collective movement found in flocks of birds.
-With this behaviors, we are able to simulate intelligent squad movements and do tactical moves to flank the opponent in a RTS game\cite{Danielsiek_2008,SynnaeveMicroCig11}, improving the survival of our units.
-
-A good alternative to Flocking is using Reciprocal Velocity Obstacles (RVO)\cite{Berg_2008}, since it is easier to control the behavior and they guarantee to prevent collisions.
-With this algorithm, we can move a large squad crossing another squad and avoid all possible collisions between units.
-This type of techniques are already applied in commercial RTS games like Starcraft 2 or Supreme Commander 2 with great success.
-
-Another popular technique is Potential Fields.
-basically, Potential Fields work like a charged particle moving through a magnetic field.
-The process of tuning and optimizing the parameters of potential field charges can be time reduced applying learning algorithms such as Q-Learning\cite{Liu_2008}.
-Another common issue we have to deal with when using potential fields is the local optima problem that can stuck the unit away from the goal.
-Some examples of the utilities of potential files in RTS games are avoiding obstacles (navigation), avoiding opponent fire, or staying at maximum shooting distance\cite{Hagelback08,Hagelback09}.
-
-There are some interesting uses of reinforcement learning (RL) \cite{Sutton}: \cite{Marthi05} employs concurrent hierarchical Q-learning (units Q-functions are combined at the group level) RL to efficiently control units in a ``one robot with multiple effectors'' fashion.
-\cite{Madeira06} advocates the use of prior domain knowledge to allow faster RL learning and applied their work on a turn-based strategy game.
+There are some interesting uses of reinforcement learning (RL) \cite{Sutton}: Marthi et al.
+\cite{Marthi05} employ concurrent hierarchical Q-learning (units Q-functions are combined at the group level) RL to efficiently control units in a ``one robot with multiple effectors'' fashion.
+Madeira et al.
+\cite{Madeira06} advocate the use of prior domain knowledge to allow faster RL learning and applied their work on a turn-based strategy game.
 In real game setups, the actions spaces to explore is gigantic, it requires to use the structure of the game in a partial program (or a partial Markov decision process) and a shape function (or a heuristic) \cite{Marthi05}.
-There are other was to learn parameters of an underlying model, for instance \cite{GA} used evolutionary learning techniques, but face the same problem of dimensionality.
+There are other way to learn parameters of an underlying model, for instance Ponsen and Spronck \cite{GA} used evolutionary learning techniques, but face the same problem of dimensionality.
 The difficulty to work with multi-scale goals and plans is handled directly by case-based reasoning (CBR), which has been adapted for units behavior with continuous action models \cite{Molineaux08}, an integrated RL/CBR algorithm using continuous models, or with hybrid CBR/RL transfer learning \cite{CBR-RL}.
-Reactive planning (for behavior languages) \cite{WeberCig10}, a decompositional planning similar to hierarchical task networks \cite{HTNPlanning}, allows for plans to be changed at different granularity levels and so for multi-scale (hierarchical) goals integration.
-
-Self-organizing-maps (SOM) learned with evolutionary algorithms provide some way to learn (realistic) parameters for RTS units behavior \cite{teamCompositionRTS}.
-Bayesian programming has been applied to inverse fusion of the sensory inputs of the units \cite{SynnaeveMicroCig11}, which subsumes potential fields, allowing for integration of tactical goals directly in micro-management.
-There are also some cognitive approaches to RTS AI \cite{SORTS}, in which there is a human-like attention model for micro-management.
+Reactive planning \cite{WeberCig10}, a decompositional planning similar to hierarchical task networks \cite{HTNPlanning}, allows for plans to be changed at different granularity levels and so for multi-scale (hierarchical) goals integration of low-level control.
+Synnaeve and Bessi\`{e}re \cite{SynnaeveMicroCig11} achieve hierarchical goals (coming from tactical decisions) integration through the addition of another sensory input corresponding to the goal's objective.
 

### 2012-05-14 survey.tex
Pass over related work

@@
 
+{\color{red}
+TODO merge above and bottom
+}
+
 {\color{blue}
-\begin{LARGE}
-TODO merge above and bottom\\
-\end{LARGE}
 \textbf{Planning}
@@
 \textbf{Learning}\\
-When talking about learning in RTS games, with regard to a particular match against an opponent, one has to differentiate between:
+When talking about learning in RTS games, with regard to a particular match against an opponent (where a match is a sequence of games against the same opponent), one has to differentiate between:
 \begin{itemize}
-\item prior learning: done before the match, like when a bot learns what is likely to happen on a given match-up (and map) from replays, or when a bot optimizes its parameters beforehand through simulated annealing or genetic evolution.
-\item in-game learning: done between the different actions (decisions) taken during a game.
+\item Prior learning: done before the match, like when a bot learns what is likely to happen on a given match-up (and map) from replays, or when a bot optimizes its parameters beforehand through simulated annealing or genetic evolution.
+\item In-game learning: done between the different actions (decisions) taken during a game.
 For instance, a bot can take into account the successes and failures of its attacks (location, type, timings).
-\item in-match learning: done between the different games of match, it encompasses opponent modeling across multiple games and adapting the bots' strategy to its successes and failures.
+\item In-match learning: done between the different games of the match, it encompasses opponent modeling across multiple games and adapting the bots' strategy to its successes and failures.
 \end{itemize}
-All these learning can be supervised (labeled replays, scripted situations) or unsupervised (unlabeled data, exploratory situations).
+Learning in all these situations can be supervised (labeled replays, scripted situations) or unsupervised (unlabeled data, exploratory situations).
 The last two kinds of situations of learning are more inclined towards reinforcement learning because the AI has to deal with the exploration-exploitation duality.
@@
 \subsection{RTS AI Decomposition}
-\begin{figure*}
+\begin{figure}
     \centering
@@
     \label{fig:levels-abstraction}
-\end{figure*}
-As a work on RTS AI has to address several of these aforementioned problems together, we decided to study decompose RTS AI in three tasks, widely known by RTS players: strategy, tactics and units control (micro-management), see Figure~\ref{fig:levels-abstraction}.
-We consider strategy to be everything related to the \textit{technology tree}, being the buildings and the upgrades, and the army composition.
-It is the most intentional level, as a player performs a strategy with future stances (aggressive, defensive, economy, technology) and tactics in mind.
+\end{figure}
+As work on RTS AI has to address several of the aforementioned problems together, in this section, we divide the existing work on RTS AI in three different levels of abstraction, widely known by RTS players: strategy, tactics and units control (micro-management), see Figure~\ref{fig:levels-abstraction}.
+We consider strategy to be everything related to the \textit{technology tree}, buildings, upgrades, and army composition.
+It is the most deliberative level, as a player performs a strategy with future stances (aggressive, defensive, economy, technology) and tactics in mind.
 To change its strategy, we consider that a player need about 3 minutes: they should need to construct a few new buildings and/or research some upgrades and/or produce units changing the army composition.
-Tactics represent the attacks and defenses.
+Tactics represent the attacks and defenses between groups of units.
 For the human player, decisions on tactics are made from spatial (exploiting the terrain) and temporal (army movements) reasoning, constrained on the possible types of attacks by the army composition of the player and their opponent.
-Units control, also called micro-management, is how the player takes action to maximize the efficiency of its units (applying the tactics) in real-time (simultaneous) and adversary conditions.
+Unit control, also called micro-management, is how the player takes action to maximize the efficiency of its individual units (applying the tactics) in real-time (simultaneous) and adversary conditions.
 As shown in Figure~\ref{fig:levels-abstraction}, strategy constrains possible tactics which in turn constrain units control.
@@
 
-In the industry, strategies are often ``hard coded'' with finite state machines (FSM) \cite{FSM_AIGameProgWisdom2003}.
-The idea behind FSM is to decompose an object's behavior into easily manageable states, such as ``attacking'', ``gathering resources'' or ``repairing'' and establish the conditions that trigger transitions between them.
-Hierarchical FSM give higher (strategic) states some control over lower oned (actions).
+In commercial RTS games, strategies are often ``hard coded'' with finite state machines (FSM) \cite{FSM_AIGameProgWisdom2003}.
+The idea behind FSMs is to decompose an object's behavior into easily manageable states, such as ``attacking'', ``gathering resources'' or ``repairing'' and establish the conditions that trigger transitions between them.
+Hierarchical FSM give higher (strategic) states some control over lower ones (actions).
 However, strategic decisions have to be made in a context, not in a vacuum.
@@
 To this end, several approaches have been developed like hierarchical task networks, behavior trees and the comeback of STRIPS \cite{FikesSTRIPS} planning in the industry \cite{orkinGDC_FEAR}.
-Regardless, one cannot but notice that strategy adaptation is not the strong point of industrial RTS AI.
-Strategy adaptation is part of what is called opponent modeling.
-An adapting robot needs to keep track of what the enemy is doing in order to estimate the probability that they will use a specific strategy and adapt its own strategy to it.
+Regardless, one cannot but notice that adaptability is not the strong point of industrial RTS AI.
+In the context of RTS AI strategy adaptation is related to the idea of opponent modeling.
+An adapting AI needs to keep track of what the enemy is doing in order to estimate the probability that they will use a specific strategy and adapt its own strategy accordingly.
 
-In research, the main methods used for strategic decision-making are case-based reasoning (CBR) and planning or plan recognition \cite{LTW,CBR_Planning,OntanonCBR,HTNPlanning,Ramirez}.
+In academic RTS AI research, the main methods used for strategic decision-making have been case-based reasoning (CBR) and planning or plan recognition \cite{LTW,CBR_Planning,OntanonCBR,HTNPlanning,Ramirez}.
 Aha et al.
@@
 Onta\~{n}\'{o}n et al.
-\cite{CBR_Planning} base their real-time case-based planning (CBP) system on a plan dependency graph which is learned from human demonstration.
+\cite{CBR_Planning} base real-time case-based planning (CBP) system on a plan dependency graph which is learned from human demonstration.
 In \cite{OntanonCBR,PlanRetrieval}, they use CBR learned from expert demonstrations on Wargus.
@@
 
-Most commercial games avoid gathering this information by using perfect information all the time (cheating).
-Sometimes they simulate some scouting tasks as Bob Fitch described in his AIIDE 2011 keynote for the WarCraft game series and StarCraft: Broodwar.
-{\color{blue}XXX (perhaps move it above, to the start of subsection}
+One final consideration is that RTS games are typically partially observable.
+For example games like StarCraft implement the ``fog of war'' idea, which basically means that a player can only see the areas of the map close to her own units.
+Areas of the map away from the field of view of individual units are not observable.
+Therefore, in order to play an RTS game, players need to scout the opponent in order to obtain information about the opponent's strategy.
+Most commercial games avoid this problem by using perfect information all the time (cheating).
+In order to make the human player believe the built-in AI of this games does scouting, sometimes they simulate some scouting tasks as Bob Fitch described in his AIIDE 2011 keynote for the WarCraft game series and StarCraft: Broodwar.
+%{\color{blue}XXX (perhaps move it above, to the start of subsection}
 
@@
 Units have different abilities, which leads to different possible tactics.
-Most often, fast or invisible or flying units are not cost-effective in head-to-head fights against slower bulky units.
+For example, very often, fast, invisible or flying units are not cost-effective in head-to-head fights against slower bulky units.
 Taking good tactical decisions is taking advantage of the terrain and positions of the different units (in a timely manner).
@@
 Balla and Fern \cite{UCT} applied upper confidence bounds on trees (UCT: a MCTS algorithm) to tactical assault planning in Wargus.
-On influence maps, \cite{teamCompositionRTS} studied team composition and maneuvering by learning a self-organizing map, while \cite{HagelbackJ08} presented a multi-agent potential field based bot.
-Finally, Avery et al.
-\cite{Avery09} and Smith et al.
-\cite{SmithCIG10} co-evolved influence map trees for spatial (tactical) reasoning in RTS games.
+% (Santi: I commented this one, since it's mentioned above) On influence maps, \cite{teamCompositionRTS} studied team composition and maneuvering by learning a self-organizing map, while \cite{HagelbackJ08} presented a multi-agent potential field based bot.
+Finally, Avery et al. \cite{Avery09} and Smith et al. \cite{SmithCIG10} co-evolved influence map trees for spatial (tactical) reasoning in RTS games.
 
@@
 \cite{Danielsiek_2008} studied intelligent squad movements to do tactical moves to flank the opponent in a RTS game this way.
-An alternative to flocking is to use reciprocal velocity obstacles (RVO) \cite{Berg_2008}, since it is easier to control the behavior and they guarantee to prevent collisions.
+%An alternative to flocking is to use reciprocal velocity obstacles (RVO) \cite{Berg_2008}, since it is easier to control the behavior and they guarantee to prevent collisions.
 With this algorithm, we can move a large squad crossing another squad and avoid all possible collisions between units.
 {\color{blue}Do we put RVO?
-As I don't remember it being used in any game or game AI research.} In recent commercial RTS games like Starcraft 2 or Supreme Commander 2, flocking-like behaviors are inspired of continuum crowds (``flow field'') \cite{Treuille2006}, and a comprehensive review about (grid-based) pathfinding was recently done by Sturtevant \cite{sturtevant2012benchmarks}.
+As I don't remember it being used in any game or game AI research.} (Santi: I agree, let's not mention it)
+In recent commercial RTS games like Starcraft 2 or Supreme Commander 2, flocking-like behaviors are inspired of continuum crowds (``flow field'') \cite{Treuille2006}.
+A comprehensive review about (grid-based) pathfinding was recently done by Sturtevant \cite{sturtevant2012benchmarks}.
 

### 2012-05-14 survey.tex
Incorporated AIUR in the paper

@@
 
-Figure \ref{fig:bot-architecture} shows some representative examples of the integration architectures used by different bots in the AIIDE and CIG 2011 Starcraft AI competitions \cite{url1,url2}: BroodwarQ \cite{???}, Nova \cite{???}, UAlbertaBot \cite{???}, Skynet \cite{???}, SPAR \cite{???}, and BTHAI \cite{???} (see Section \ref{sec:competition} for a comparison of their performance).
+Figure \ref{fig:bot-architecture} shows some representative examples of the integration architectures used by different bots in the AIIDE and CIG 2011 Starcraft AI competitions \cite{url1,url2}: BroodwarQ \cite{???}, Nova \cite{???}, UAlbertaBot \cite{???}, Skynet \cite{???}, SPAR \cite{???}, AIUR \cite{???}, and BTHAI \cite{???} (see Section \ref{sec:competition} for a comparison of their performance).
 Each box represents an individual module with a clearly defined task (only modules with a black background can send actions directly to Starcraft).
@@
 Its high-level module determines the strategy to use, and the tactical decision module divides it into a collection of {\em abstract actions}, that are executed by the lower-level modules.
+\item AIUR: is mainly divide-and-conquer oriented, for macro as well as for micro, with a slight abstraction on macro with a SpendManager deciding how to spend resources among Base, Production and Construction Managers.
+At the beginning of a game, the MoodManager initializes a ``mood'' which will influence both micro and macro.
+For example, micro-management is divided into three independent managers: the {\em Defense Manager}, controlling military units when there is nothing special, the {\em Under Attack Manager}, activated when the opponent is attacking our bases, and the {\em Army Manager}, taking control of units when it is time to attack, following a timing given by the current mood.
 \item BTHAI: uses a two-tier abstraction hierarchy, where a collection of high-level modules command a collection of lower-level agents in charge of each of the units.
@@
 
-Most bots using divide-and-conquer (except for BBQ and UAlbertaBot), assume that each of the modules can act independently and that their actions can be executed without interference.
-BBQ and UAlbertaBot use an arbitrator ({\em Game Commander'} in UAlbertaBot) that makes sure that modules do not send contradictory orders to the same unit.
+Some bots using divide-and-conquer, assume that each of the modules can act independently and that their actions can be executed without interference.
+BBQ, UAlbertaBot and AIUR, however use an arbitrator ({\em Game Commander'} in UAlbertaBot) that makes sure that modules do not send contradictory orders to the same unit.
 However, very little bots handle the problem of how to coordinate resource usage amongst modules, for instance BTHAI uses a first-come-first-serve policy for spending resources, the first module that requests resources is the one that gets them.
 Nova and Skynet are exceptions, and implement some rudimentary prioritization based on the high level strategy.
+Following available resources and timing, AIUR's {\em Spend Manager} orders Base, Production and Construction Managers what they have to build/produce.
+It also orders to start tech research and upgrades.
 
@@
 An interesting exception to that is UAlbertaBot, which uses a search algorithm in the {\em Production Manager} to find near-optimal build orders.
+Another interesting case is AIUR, that uses a {\em Mood Manager} to randomly pick a mood among five (rush, aggressive, defensive, macro, fast expand), which will influence the build order, strategy and tactics, all of them scripted so far.
+This mood can change during the game, following the opponent behavior.
 

### 2012-05-20 survey.tex
Added Florian to acknowledgements: do we want him as an author?
Edited the beginning of the related-work section (intro paragraphs and strategy section)

@@
 
-\subsection{RTS AI Decomposition}
+%\subsection{RTS AI Decomposition}
 \begin{figure}
@@
 \end{figure}
-As work on RTS AI has to address several of the aforementioned problems together, in this section, we divide the existing work on RTS AI in three different levels of abstraction, widely known by RTS players: strategy, tactics and units control (micro-management), see Figure~\ref{fig:levels-abstraction}.
-We consider strategy to be everything related to the \textit{technology tree}, buildings, upgrades, and army composition.
-It is the most deliberative level, as a player performs a strategy with future stances (aggressive, defensive, economy, technology) and tactics in mind.
-To change its strategy, we consider that a player need about 3 minutes: they should need to construct a few new buildings and/or research some upgrades and/or produce units changing the army composition.
-Tactics represent the attacks and defenses between groups of units.
-For the human player, decisions on tactics are made from spatial (exploiting the terrain) and temporal (army movements) reasoning, constrained on the possible types of attacks by the army composition of the player and their opponent.
-Unit control, also called micro-management, is how the player takes action to maximize the efficiency of its individual units (applying the tactics) in real-time (simultaneous) and adversary conditions.
-As shown in Figure~\ref{fig:levels-abstraction}, strategy constrains possible tactics which in turn constrain units control.
-On a prediction/adaptation point of view, the units observation level gives information about the opponent's tactics and strategy.
-Flaws in opponents tactics can also steer our strategy into exploiting them.\\
-{\color{blue}
-perhaps TODO examples?
-}
+
+AI systems that play RTS games need to address most, if not all, the aforementioned problems together.
+Therefore, in order to classify existing work on RTS AI, we will divide the existing work according to three levels of abstraction (division that is widely used by RTS players): strategy, tactics and unit control (micro-management).
+
+Figure~\ref{fig:levels-abstraction} graphically illustrates how strategy, tactics and unit control are three points in a continuum scale where strategy corresponds to decisions making processes that affect long spans of time (several minutes in the case of Starcraft), unit control corresponds to low-level reactive second-by-second individual unit control, and tactics sit in the middle.
+Also, strategic decisions reason about the whole game at once, whereas tactical decisions are localized, and affect only specific groups of units; finally unit control techniques control only individual units.
+Typically, strategic decisions constraint future tactical decisions, which in turn constrain unit control.
+Moreover, information gathered while performing unit control, can cause reconsideration of the tactics being employed; which could trigger further strategical reasoning.
+
+Following this idea, we consider strategy to be everything related to the technology trees, build-order\footnote{The {\em build-order} is the specific sequence in which buildings of different types will be constructed at the beginning of a game, and completely determines the long-term strategy of a player.}, upgrades, and army composition.
+It is the most deliberative level, as a player selected and performs a strategy with future stances (aggressive, defensive, economy, technology) and tactics in mind.
+We consider tactics, everything related to confrontations between groups of units.
+Tactical reasoning involves both spatial (exploiting the terrain) and temporal (army movements) reasoning, constrained on the possible types of attacks by the army composition of the player and their opponent.
+Finally, we will call unit control (also known as micro-management) to how the player controls individual units to maximize their efficiency (when executing tactics) in real-time and adversary conditions.
+
+% Example:
+For example, after starting a game, a player might decide to use a {\em rushing} strategy (which involves quickly building an army and sending it to attack as early as possible in the game); then, when performing the attack use a {\em surrounding} tactic, where the player tries to surround the enemy cutting potential escape routes; finally, while executing the surrounding tactic, the player might decide to use unit control techniques that command individual units to perform repeated {\em attack and flee} movements, to maximize the efficiency of each of the units being used in the attack.
 
@@
 
-In commercial RTS games, strategies are often ``hard coded'' with finite state machines (FSM) \cite{FSM_AIGameProgWisdom2003}.
-The idea behind FSMs is to decompose an object's behavior into easily manageable states, such as ``attacking'', ``gathering resources'' or ``repairing'' and establish the conditions that trigger transitions between them.
-Hierarchical FSM give higher (strategic) states some control over lower ones (actions).
-However, strategic decisions have to be made in a context, not in a vacuum.
-It is hard to encode dynamic, adaptive behaviors in (H)FSM due to the exponential number of transitions (between states) to account for.
-To this end, several approaches have been developed like hierarchical task networks, behavior trees and the comeback of STRIPS \cite{FikesSTRIPS} planning in the industry \cite{orkinGDC_FEAR}.
+Strategic decision making is a hard problem, which has been addressed using many AI paradigms.
+Most of the work in this area falls in one of three categories: hard-coded approaches, planning-based approaches, and machine learning-based approaches.
+Let us cover each of these categories in turn.
+
+Hard-coded approaches have been extensively used in commercial RTS games.
+The most common approaches make use finite state machines (FSM) \cite{FSM_AIGameProgWisdom2003} in order to let the AI author hard-code the strategy that the AI will employ.
+The idea behind FSMs is to decompose the AI behavior into easily manageable states, such as ``attacking'', ``gathering resources'' or ``repairing'' and establish the conditions that trigger transitions between them.
+Commercial approaches also include Hierarchical FSMs, in which FSMS are composed hierarchically.
+These hard-coded approaches have achieved a significant amount of success, and, as we will discuss later, have also been used in many academic RTS AI research systems, as discussed in Section \ref{sec:bot}.
+However, strategic decision making is a hard problem, and these hard-coded approaches struggle to encode dynamic, adaptive behaviors.
+
+% XXX: Santi: I've commented out this paragraph, since it has nothing to do with RTS games.
+There is no commercial RTS game using planning (neither STRIPS nor HTN) that I'm aware of.
+The references below refer to FEAR, which is an FPS game.
+%To this end, several approaches have been developed like hierarchical task networks, behavior trees and the comeback of STRIPS \cite{FikesSTRIPS} planning in the industry \cite{orkinGDC_FEAR}.
 Regardless, one cannot but notice that adaptability is not the strong point of industrial RTS AI.
@@
 
-In academic RTS AI research, the main methods used for strategic decision-making have been case-based reasoning (CBR) and planning or plan recognition \cite{LTW,CBR_Planning,OntanonCBR,HTNPlanning,Ramirez}.
-Aha et al.
-\cite{LTW} used CBR to perform dynamic plan retrieval extracted from domain knowledge in Wargus (Warcraft II clone).
-Onta\~{n}\'{o}n et al.
-\cite{CBR_Planning} base real-time case-based planning (CBP) system on a plan dependency graph which is learned from human demonstration.
-In \cite{OntanonCBR,PlanRetrieval}, they use CBR learned from expert demonstrations on Wargus.
-They update their previous CBP approach by using situation assessment for retrieving the ``right'' plan, and improve the speed of CBP by using a decision tree to select relevant features.
-Hsieh and Sun \cite{HsiehS08} based their work on \cite{LTW} CBR model and used StarCraft replays to construct states and building sequences (``build orders'').
-Schadd et al.
-\cite{SchaddBS07} describe opponent modeling through hierarchically structured models of the opponent behavior and they applied their work to the Spring RTS (Total Annihilation clone).
-Hoang et al.
-\cite{HTNPlanning} use hierarchical task networks (HTN) to model strategies in a first person shooter with the goal to use HTN planners.
-%Madeira et al.
-\cite{Madeira06}
+Approaches using planning techniques have also been explored in the literature.
+For example Onta\~{n}\'{o}n et al.
+\cite{CBR_Planning} explored the use of real-time case-based planning (CBP) in the domain of Wargus (a Warcraft II clone).
+In their work, they used human demonstration to learn plans, that are then composed at run-time in order to form full-fledges strategies to play the game.
+In \cite{PlanRetrieval} they their previous CBP approach by using situation assessment for retrieving the ``right'' plan, and improve the speed of CBP by using a decision tree to select relevant features.
+Hierarchical Task-Network (HTN) planning has also been explored with some success in the context of simpler first-person shooter games.
+Planning approaches offer more adaptivity of the AI strategy compared to hard-coded approaches.
+However, the real-time constraints of RTS games limit the planning approaches that can be applied, being HTN and case-based planning the only ones explored so far.
 
-Weber and Mateas \cite{WeberCig09} proposed ``a data mining approach to strategy prediction'' and performed supervised learning (from buildings features) on labeled StarCraft replays.
+Concerning machine learning-based approaches, Weber and Mateas \cite{WeberCig09} proposed ``a data mining approach to strategy prediction'' and performed supervised learning (from buildings features) on labeled StarCraft replays.
 Dereszynski et al.
@@
 
+Also falling into the machine-learning category, a significant group of researchers has explored case-based reasoning (CBR) \cite{Aamodt94CBR} approaches for strategic decision making.
+For example Aha et al.
+\cite{LTW} used CBR to perform dynamic plan retrieval in the Wargus domain.
+Hsieh and Sun \cite{HsiehS08} based their work on \cite{LTW} CBR model and used StarCraft replays to construct states and building sequences (``build orders'').
+Finally, Schadd et al.
+\cite{SchaddBS07} applied a CBR approach to opponent modeling through hierarchically structured models of the opponent behavior and they applied their work to the Spring RTS (Total Annihilation clone).
+
 One final consideration is that RTS games are typically partially observable.
@@
 Therefore, in order to play an RTS game, players need to scout the opponent in order to obtain information about the opponent's strategy.
-Most commercial games avoid this problem by using perfect information all the time (cheating).
-In order to make the human player believe the built-in AI of this games does scouting, sometimes they simulate some scouting tasks as Bob Fitch described in his AIIDE 2011 keynote for the WarCraft game series and StarCraft: Broodwar.
+Very few of the previous approaches deal with this problem, and use perfect information all the time (in the case of commercial games, most AI implementations cheat, since they have perfect information).
+In commercial games, in order to make the human player believe the built-in AI of this games does scouting, sometimes they simulate some scouting tasks as Bob Fitch described in his AIIDE 2011 keynote for the WarCraft game series and StarCraft: Broodwar.
 %{\color{blue}XXX (perhaps move it above, to the start of subsection}
@@
 
-\section*{Acknowledgments} {\color{blue} This research is partially funded by projects ... and ... }
+\section*{Acknowledgments} {\color{blue} This research is partially funded by projects ...
+and ...
+.
+The authors would like to thank Florian Richoux, author of the AIUR bot, for his helpful comments and help in early drafts of this paper.
+}
 

### 2012-05-20 survey.tex
Changed "units control" to "unit control" in the figure and all over the paper (which is more correct in English)

@@
 
-\subsection{Units Control}
-Units control, also called micro-management in RTS games, aims at maximizing the effectiveness of units, it starts with pathfinding and goes up to simultaneous control of units of different types in complex battles on heterogeneous terrain.
+\subsection{Unit Control}
+Unit control, also called micro-management in RTS games, aims at maximizing the effectiveness of units, it starts with pathfinding and goes up to simultaneous control of units of different types in complex battles on heterogeneous terrain.
 The most common pathfinding algorithm is A*, but its big problem is CPU time and memory consumption, hard to satisfy in a complex, dynamic, real-time environment on multiple units.

### 2012-05-21 survey.tex
Done a pass over the tactical related work section

@@
 
-AI systems that play RTS games need to address most, if not all, the aforementioned problems together.
-Therefore, in order to classify existing work on RTS AI, we will divide the existing work according to three levels of abstraction (division that is widely used by RTS players): strategy, tactics and unit control (micro-management).
+Systems that play RTS games need to address most, if not all, the aforementioned problems together.
+Therefore, it is hard to classify existing work on RTS AI as addressing the different problems above.
+For that reason, in order to classify existing work on RTS AI, we will divide the existing work according to three levels of abstraction (division that is widely used by RTS players): strategy, tactics and unit control (micro-management).
 
@@
 
-Strategic decision making is a hard problem, which has been addressed using many AI paradigms.
-Most of the work in this area falls in one of three categories: hard-coded approaches, planning-based approaches, and machine learning-based approaches.
-Let us cover each of these categories in turn.
+% INTRO:
+Strategic decision making in real-time domains is still an open problem.
+In the context of RTS games is has been addressed using many AI techniques, like hard-coded approaches, planning-based approaches, or machine learning-based approaches.
+Let us cover each of these approaches in turn.
 
+% HARD-CODED:
 Hard-coded approaches have been extensively used in commercial RTS games.
@@
 
+% PLANNING:
 Approaches using planning techniques have also been explored in the literature.
@@
 In their work, they used human demonstration to learn plans, that are then composed at run-time in order to form full-fledges strategies to play the game.
-In \cite{PlanRetrieval} they their previous CBP approach by using situation assessment for retrieving the ``right'' plan, and improve the speed of CBP by using a decision tree to select relevant features.
+In \cite{PlanRetrieval} they improve over  their previous CBP approach by using situation assessment for improving the quality and speed of plan retrieval.
 Hierarchical Task-Network (HTN) planning has also been explored with some success in the context of simpler first-person shooter games.
@@
 However, the real-time constraints of RTS games limit the planning approaches that can be applied, being HTN and case-based planning the only ones explored so far.
+Moreover, none of these approaches addresses any timing or scheduling issues, which are key in RTS games.
 
+% MACHINE LEARNING:
 Concerning machine learning-based approaches, Weber and Mateas \cite{WeberCig09} proposed ``a data mining approach to strategy prediction'' and performed supervised learning (from buildings features) on labeled StarCraft replays.
@@
 
+% CBR:
 Also falling into the machine-learning category, a significant group of researchers has explored case-based reasoning (CBR) \cite{Aamodt94CBR} approaches for strategic decision making.
@@
 
+% SCOUTING:
 One final consideration is that RTS games are typically partially observable.
@@
 In commercial games, in order to make the human player believe the built-in AI of this games does scouting, sometimes they simulate some scouting tasks as Bob Fitch described in his AIIDE 2011 keynote for the WarCraft game series and StarCraft: Broodwar.
-%{\color{blue}XXX (perhaps move it above, to the start of subsection}
+A notable exception is the work of Weber et al.
+\cite{WeberAIIDE11}, who used a particle model with a linear trajectory update to track opponent units under fog of war in StarCraft.
+They also produced tactical goals through reactive planning and goal-driven autonomy \cite{WeberCig10,Weber10}, finding the more relevant goal(s) to spawn in unforeseen situations.
 
@@
 \subsection{Tactics}
-Units have different abilities, which leads to different possible tactics.
-For example, very often, fast, invisible or flying units are not cost-effective in head-to-head fights against slower bulky units.
-Taking good tactical decisions is taking advantage of the terrain and positions of the different units (in a timely manner).
+
+% INTRO:
+Tactical reasoning involves reasoning about the different abilities of the units in a group of units and about the environment (terrain) and positions of the different units in order to gain military advantage in battles.
+For example, it would be a very bad tactical decision to place fast, invisible or flying units (typically expensive) in the first line of fire against slower heavier units, since they will be wiped out fast.
+We will divide the work on tactical reasoning in two parts: terrain analysis and decision making.
+
+% TERRAIN ANALYSIS:
 Terrain analysis supplies the AI with structured information about the map in order to help making decisions.
 This analysis is usually performed off-line, in order to save CPU time during the game.
-Pottinger \cite{Pottinger00} described the \emph{BANG} engine implemented by Ensemble Studios for the game Age of Empires II.
+For example, Pottinger \cite{Pottinger00} described the \emph{BANG} engine implemented by Ensemble Studios for the game Age of Empires II.
 This engine provides terrain analysis functionalities to the game using influence maps and areas with connectivity information.
-Forbus et al.
-\cite{Forbus2002} showed the importance to have qualitative spatial information for wargames, for which they used geometric and pathfinding analysis.
-Hale et al.
-\cite{Hale08} presented a 2D geometric navigation mesh generation method from expanding convex regions from seeds.
+Forbus et al. \cite{Forbus2002} showed the importance to have qualitative spatial information for wargames, for which they used geometric and pathfinding analysis. 
+Hale et al. \cite{Hale08} presented a 2D geometric navigation mesh generation method from expanding convex regions from seeds. 
 Finally, Perkins \cite{Perkins10} applied Voronoi decomposition (then pruning) %\cite{Karavelas04}
@@
 
+% MACHINE LEARNING:
+Concerning tactical decision making, many different approaches have been explored such as machine learning, game tree search and potential fields. Concerning machine learning approaches, 
 Hladky and Bulitko \cite{Hladky2008} benchmarked hidden semi-Markov models (HSMM) and particle filters in first person shooter games (FPS) units tracking.
 They showed that the accuracy of occupancy maps was improved using movement models (learned from the player behavior) in HSMM.
-Hagelb\"{a}ck and Johansson \cite{HagelbackJ08} presented a multi-agent potential fields based bot able to deal with fog of war in Tankbattle.
-Weber et al.
-\cite{WeberAIIDE11} used a particle model with a simple linear trajectory update to track opponent units under fog of war in StarCraft.
-They also produced tactical goals through reactive planning and goal-driven autonomy \cite{WeberCig10,Weber10}, finding the more relevant goal(s) to spawn in unforeseen situations.
 Kabanza et al.
@@
 \cite{CBR-RL} combined CBR and reinforcement learning to enable reuse of tactical plan components.
-Cadena and Garrido \cite{CadenaG11} used fuzzy CBR (fuzzy case matching) for strategic and tactical planning.
+Finally, Cadena and Garrido \cite{CadenaG11} used fuzzy CBR (fuzzy case matching) for strategic and tactical planning.
+
+% GAME TREE SEARCH:
+Game tree search techniques have also been explored for tactical decision making.
 \cite{Chung05} adapted Monte-Carlo tree search (MCTS) to planning in RTS games and applied it to a capture-the-flag mod of Open RTS.
@@
 % (Santi: I commented this one, since it's mentioned above) On influence maps, \cite{teamCompositionRTS} studied team composition and maneuvering by learning a self-organizing map, while \cite{HagelbackJ08} presented a multi-agent potential field based bot.
-Finally, Avery et al. \cite{Avery09} and Smith et al. \cite{SmithCIG10} co-evolved influence map trees for spatial (tactical) reasoning in RTS games.
+
+% POTENTIAL FIELDS:
+Finally, potential fields (or influence maps) have also been found to be a useful technique for tactical decision making.
+Hagelb\"{a}ck and Johansson \cite{HagelbackJ08} presented a multi-agent potential fields based bot able to deal with fog of war in the Tankbattle game.
+Avery et al.
+\cite{Avery09} and Smith et al.
+\cite{SmithCIG10} co-evolved influence map trees for spatial (tactical) reasoning in RTS games.
+Danielsiek et al.
+\cite{Danielsiek_2008} used influence maps to achieve intelligent squad tactical moves to flank the opponent in a RTS game.
+%An alternative to flocking is to use reciprocal velocity obstacles (RVO) \cite{Berg_2008}, since it is easier to control the behavior and they guarantee to prevent collisions.
+With this algorithm, we can move a large squad crossing another squad and avoid all possible collisions between units.
+{\color{blue}Do we put RVO?
+As I don't remember it being used in any game or game AI research.} (Santi: I agree, let's not mention it)
+
+% SCOUTING:
+Additionally, scouting is equally important in tactical decision making as in strategic decision making.
+However, as mentioned earlier, very little work has been done in this respect, being that of Weber et al.
+\cite{WeberAIIDE11} the only exception.
 
 \subsection{Unit Control}
+
 Unit control, also called micro-management in RTS games, aims at maximizing the effectiveness of units, it starts with pathfinding and goes up to simultaneous control of units of different types in complex battles on heterogeneous terrain.
@@
 His model describes how a system of multiple agents, each steering according to simple local rules, can produce the collective movement found in flocks of birds.
-Danielsiek et al.
-\cite{Danielsiek_2008} studied intelligent squad movements to do tactical moves to flank the opponent in a RTS game this way.
-%An alternative to flocking is to use reciprocal velocity obstacles (RVO) \cite{Berg_2008}, since it is easier to control the behavior and they guarantee to prevent collisions.
-With this algorithm, we can move a large squad crossing another squad and avoid all possible collisions between units.
-{\color{blue}Do we put RVO?
-As I don't remember it being used in any game or game AI research.} (Santi: I agree, let's not mention it)
 In recent commercial RTS games like Starcraft 2 or Supreme Commander 2, flocking-like behaviors are inspired of continuum crowds (``flow field'') \cite{Treuille2006}.

### 2012-05-24 survey.tex
some progress in competition section (more to come)

@@
 
-{\color{magenta}
+{\color{violet}
+
+In order to inject new AI code into the StarCraft game, all recent
+tournaments attached to scientific conferences employ the 
+Brood War Application Programming Interface 
+(BWAPI)\footnote{\url{http://code.google.com/p/bwapi/}} which 
+enables replacing the human player interface with a C++ library,
+some auxiliary libraries and the bot code. A consequence of this
+indirect way of integrating new bots is that every machine can
+only run one custom bot, requiring two computers for a 2-player game.
+
+\subsection{AIIDE}
+\label{sec:AIIDE}
+this is left empty for now
 
@@
 These mainly stemmed from the desire to use evolved, largely untested
-maps which proved to look interesting but made the submitted bots crash
-so often that it would have been unjustifiable to announce a winner.
+maps which proved to look interesting but made the submitted bots 
+(and/or the Broodwar Terrain Analyzer (BWTA) provided with the BWapi interface) crash
+so frequently that it would have been unjustifiable to announce a winner.
 
-At the CIG 2011, the tournament was therefore run with a (secret) selection
+At CIG 2011, the tournament was therefore run with a (secret) selection
 of maps used in league play. It was organized by Tobias Mahlmann and Mike
-Preuss and attracted 10 bots.
+Preuss and attracted 10 bots. In addition to the ones discussed in previous
+sections (UAlbertaBot, Skynet, AIUR, Nova, BroodwarBotQ, BTHAI), 
+the set also contained LSAI, Xelnaga, Protoss Beast Jelly, and EvoBot,
+these are shortly described in the following:
+
+\paragraph*{LSAI (Zerg)} utilizes a heavily modified BWSAL to divide management
+ of the units to different modules that communicate via a centralized information
+ module. It works using a simple reactive strategy to try and survive early game
+ attacks and macro up to a larger attack force and maintain map control.
+
+\paragraph*{Xelnaga (Protoss)} is a modification of the AIUR bot that chooses the 
+Dark Templar Opening in order to destroy the enemy base before defenses against
+invisible units are available.
+
+\paragraph*{Protoss Beast Jelly (Protoss)}
+
+\paragraph*{EvoBot (Terran)} employs an evolutionary algorithm for obtaining rational
+ unit combinations and influence map techniques for deciding the strategic locations. Note
+ that this bot was submitted in a very early version, with many of its designed features not
+ yet fully ready.
 
 \begin{table*}[htb]
-\caption{Bracket A results at CIG 2011, 
-qualified for the final round: UAlbertaBot and Skynet.}
+\caption{Results of the first round at CIG 2011, held in two brackets.
+Qualified for the final round: UAlbertaBot and Skynet (from A), Xelnaga
+and BroodwarBotQ (from B, the latter by comparing direct encounters
+with BTHAI of which 6:4 were won).}
 \label{tab:correlations}
@@
 \begin{center}
+% bracket A
 \begin{tabular}{|l|c|c|c|}
 \hline
+\multicolumn{4}{|c|}{Bracket A} \\ \hline
 Crashes & games & bot	& wins\\ \hline
@@
 \end{tabular}
+% bracket B
+\begin{tabular}{|l|c|c|c|}
+\hline
+\multicolumn{4}{|c|}{Bracket B} \\ \hline
+Crashes & games & bot	& wins\\ \hline
+12 &	 40 &	 Xelnaga &	 25\\
+3 &   40 &	 BroodwarBotQ  &  23 (6)\\
+0 &	 40 &	 BTHAI	  &  23 (4)\\
+17 &	 40 &	 Protoss Beast Jelly  & 17\\
+0 &	 40 &	 EvoBot	  &  12\\
+\hline
+\end{tabular}
 \end{center}
@@
 
-Bracket B
+As BroodwarBotQ and BTHAI have the same number of wins, their direct encounter is evaluated which accounted 6 : 4 for the BroodwarBotQ.
+Note that all qualified bots play Protoss.
 
-Crashes	 Games	 Bot	 Wins
-12	 40	 Xelnaga	 25
-3	 40	 BroodwarBotQ	 23
-0	 40	 BTHAI	 23
-17	 40	 Protoss Beast Jelly	 17
-0	 40	 EvoBot	 12
-As BroodwarBotQ and BTHAI have the same number of wins, their direct encounter is evaluated which accounted 6 : 4 for the BroodwarBotQ.
-
-Qualified for the final round: Xelnaga and BroodwarBotQ. Note that all qualified bots play Protoss.
 Maps used for the first round: (2)MatchPoint 1.3, (4)Fighting Spirit 1.3, iCCupdestination 1.1, iCCup gaia, iCCup great barrier reef

### 2012-05-29 survey.tex
Completed my pass over the related work section

@@
 Therefore, it is hard to classify existing work on RTS AI as addressing the different problems above.
-For that reason, in order to classify existing work on RTS AI, we will divide the existing work according to three levels of abstraction (division that is widely used by RTS players): strategy, tactics and unit control (micro-management).
+For that reason, in order to classify existing work on RTS AI, we will divide the existing work according to three levels of abstraction (division that is widely used by RTS players): strategy, tactics and reactive control (micro-management).
 
-Figure~\ref{fig:levels-abstraction} graphically illustrates how strategy, tactics and unit control are three points in a continuum scale where strategy corresponds to decisions making processes that affect long spans of time (several minutes in the case of Starcraft), unit control corresponds to low-level reactive second-by-second individual unit control, and tactics sit in the middle.
-Also, strategic decisions reason about the whole game at once, whereas tactical decisions are localized, and affect only specific groups of units; finally unit control techniques control only individual units.
-Typically, strategic decisions constraint future tactical decisions, which in turn constrain unit control.
-Moreover, information gathered while performing unit control, can cause reconsideration of the tactics being employed; which could trigger further strategical reasoning.
+Figure~\ref{fig:levels-abstraction} graphically illustrates how strategy, tactics and reactive control are three points in a continuum scale where strategy corresponds to decisions making processes that affect long spans of time (several minutes in the case of Starcraft), reactive control corresponds to low-level second-by-second decisions, and tactics sit in the middle.
+Also, strategic decisions reason about the whole game at once, whereas tactical or reactive control decisions are localized, and affect only specific groups of units.
+Typically, strategic decisions constraint future tactical decisions, which in turn condition reactive control.
+Moreover, information gathered while performing reactive control, can cause reconsideration of the tactics being employed; which could trigger further strategical reasoning.
 
@@
 Tactical reasoning involves both spatial (exploiting the terrain) and temporal (army movements) reasoning, constrained on the possible types of attacks by the army composition of the player and their opponent.
-Finally, we will call unit control (also known as micro-management) to how the player controls individual units to maximize their efficiency (when executing tactics) in real-time and adversary conditions.
+Finally, we will call reactive control (also known as micro-management) to how the player controls individual units to maximize their efficiency (when executing tactics) in real-time and adversary conditions.
+The main difference between tactics and reactive control is that tactical reasoning typically involves some sort of planning ahead for some short spans of time, whereas reactive control involves no planning ahead whatsoever.
 
 % Example:
-For example, after starting a game, a player might decide to use a {\em rushing} strategy (which involves quickly building an army and sending it to attack as early as possible in the game); then, when performing the attack use a {\em surrounding} tactic, where the player tries to surround the enemy cutting potential escape routes; finally, while executing the surrounding tactic, the player might decide to use unit control techniques that command individual units to perform repeated {\em attack and flee} movements, to maximize the efficiency of each of the units being used in the attack.
+For example, after starting a game, a player might decide to use a {\em rushing} strategy (which involves quickly building an army and sending it to attack as early as possible in the game); then, when performing the attack use a {\em surrounding} tactic, where the player tries to surround the enemy cutting potential escape routes; finally, while executing the surrounding tactic, the player might decide to use reactive control techniques that command individual units to perform repeated {\em attack and flee} movements, to maximize the efficiency of each of the units being used in the attack.
 
@@
 % MACHINE LEARNING:
-Concerning tactical decision making, many different approaches have been explored such as machine learning, game tree search and potential fields. Concerning machine learning approaches, 
+Concerning tactical decision making, many different approaches have been explored such as machine learning or game tree search. Concerning machine learning approaches, 
 Hladky and Bulitko \cite{Hladky2008} benchmarked hidden semi-Markov models (HSMM) and particle filters in first person shooter games (FPS) units tracking.
@@
 
-% POTENTIAL FIELDS:
-Finally, potential fields (or influence maps) have also been found to be a useful technique for tactical decision making.
-Hagelb\"{a}ck and Johansson \cite{HagelbackJ08} presented a multi-agent potential fields based bot able to deal with fog of war in the Tankbattle game.
-Avery et al.
-\cite{Avery09} and Smith et al.
-\cite{SmithCIG10} co-evolved influence map trees for spatial (tactical) reasoning in RTS games.
-Danielsiek et al.
-\cite{Danielsiek_2008} used influence maps to achieve intelligent squad tactical moves to flank the opponent in a RTS game.
-%An alternative to flocking is to use reciprocal velocity obstacles (RVO) \cite{Berg_2008}, since it is easier to control the behavior and they guarantee to prevent collisions.
-With this algorithm, we can move a large squad crossing another squad and avoid all possible collisions between units.
-{\color{blue}Do we put RVO?
-As I don't remember it being used in any game or game AI research.} (Santi: I agree, let's not mention it)
-
 % SCOUTING:
@@
 
-\subsection{Unit Control}
+\subsection{Reactive Control}
 
-Unit control, also called micro-management in RTS games, aims at maximizing the effectiveness of units, it starts with pathfinding and goes up to simultaneous control of units of different types in complex battles on heterogeneous terrain.
-The most common pathfinding algorithm is A*, but its big problem is CPU time and memory consumption, hard to satisfy in a complex, dynamic, real-time environment on multiple units.
-%To solve this the search space needs to be reduced as much as possible.
-A common technique consists in reducing the search into a navigation graph map.
-Among others, Triangulation Reduction A* computes polygonal triangulations on a grid-based map \cite{Demyen_2006}.
-One of the problems of A* related approaches in RTS games is that moving units block/get in the way of/ each others and the path has to be recomputed often.
-Craig Reynolds \cite{Reynolds_1999} introduced algorithmic steering behaviors, which simulate the aggregate motion of swarms.
-His model describes how a system of multiple agents, each steering according to simple local rules, can produce the collective movement found in flocks of birds.
-In recent commercial RTS games like Starcraft 2 or Supreme Commander 2, flocking-like behaviors are inspired of continuum crowds (``flow field'') \cite{Treuille2006}.
-A comprehensive review about (grid-based) pathfinding was recently done by Sturtevant \cite{sturtevant2012benchmarks}.
+Reactive control, also called micro-management in RTS games, aims at maximizing the effectiveness of units, including simultaneous control of units of different types in complex battles on heterogeneous terrain.
 
-Another popular technique is potential fields (PF) and influence maps.
-Basically, PF work like a charged particle moving through a magnetic field.
-The parameters of PF charges can be learned through Q-Learning \cite{Liu_2008}.
-Another common issue of potential fields is the local optima problem that can stuck the unit away from the goal.
+% POTENTIAL FIELDS:
+Potential fields (or influence maps) have been found to be a useful technique for reactive decision making.
 Some uses of potential fields in RTS games are: avoiding obstacles (navigation), avoiding opponent fire, or staying at maximum shooting distance \cite{Hagelback08,Hagelback09}.
-Self-organizing-maps (SOM) learned with evolutionary algorithms provide some way to learn (realistic) parameters for RTS units behavior \cite{teamCompositionRTS}.
+Hagelb\"{a}ck and Johansson \cite{HagelbackJ08} presented a multi-agent potential fields based bot able to deal with fog of war in the Tankbattle game.
+Avery et al.
+\cite{Avery09} and Smith et al.
+\cite{SmithCIG10} co-evolved influence map trees for spatial reasoning in RTS games.
+Danielsiek et al.
+\cite{Danielsiek_2008} used influence maps to achieve intelligent squad movement to flank the opponent in a RTS game.
+A drawback go potential field-based techniques is the large number of parameters that has to be tuned in order to achieve the desired behavior.
+Approaches for automatically learning such parameters have been explored, for example, using reinforcement Learning \cite{Liu_2008}, or self-organizing-maps (SOM) \cite{teamCompositionRTS}.
+We would like to note that potential fields are a reactive control technique, and as such, they do not perform any form of lookahead.
+As a consequence, these techniques are prone to make units stuck in local optima.
+
+% Machine learning:
+There has been a significant amount of work on using machine learning techniques for the problem of reactive control.
 Bayesian modeling has been applied to inverse fusion of the sensory inputs of the units \cite{SynnaeveMicroCig11}, which subsumes potential fields, allowing for integration of tactical goals directly in micro-management.
-There is also a cognitive approach to RTS AI \cite{SORTS}, using a human-like attention model (with units grouping and vision of a unique screen location) for micro-management.
-
-There are some interesting uses of reinforcement learning (RL) \cite{Sutton}: Marthi et al.
+Additionally, there has been some interesting uses of reinforcement learning (RL) \cite{Sutton}: Marthi et al.
 \cite{Marthi05} employ concurrent hierarchical Q-learning (units Q-functions are combined at the group level) RL to efficiently control units in a ``one robot with multiple effectors'' fashion.
@@
 In real game setups, the actions spaces to explore is gigantic, it requires to use the structure of the game in a partial program (or a partial Markov decision process) and a shape function (or a heuristic) \cite{Marthi05}.
-There are other way to learn parameters of an underlying model, for instance Ponsen and Spronck \cite{GA} used evolutionary learning techniques, but face the same problem of dimensionality.
+Other approaches that aim at learning the parameters of an underlying model have also been explored.
+For example Ponsen and Spronck \cite{GA} used evolutionary learning techniques, but face the same problem of dimensionality.
 The difficulty to work with multi-scale goals and plans is handled directly by case-based reasoning (CBR), which has been adapted for units behavior with continuous action models \cite{Molineaux08}, an integrated RL/CBR algorithm using continuous models, or with hybrid CBR/RL transfer learning \cite{CBR-RL}.
@@
 Synnaeve and Bessi\`{e}re \cite{SynnaeveMicroCig11} achieve hierarchical goals (coming from tactical decisions) integration through the addition of another sensory input corresponding to the goal's objective.
+
+% Other:
+Other research falling into reactive control has been performed in the field of cognitive science, where Wintermute et al.
+\cite{SORTS} have explored human-like attention models (with units grouping and vision of a unique screen location) for micro-management.
+
+% Pathfinding:
+Finally, although pathfinding does not fall under our previous definition of reactive control, we include it in this section, since it is typically performed as a low-level service, not part of either tactical nor strategical reasoning (although there are some exceptions, like the tactical pathfinding of Danielsiek et al.
+\cite{Danielsiek_2008}).
+The most common pathfinding algorithm is A*, but its big problem is CPU time and memory consumption, hard to satisfy in a complex, dynamic, real-time environment with large numbers of units.
+Even if specialized algorithms, such as D*-Lite \cite{KoenigL02} exist, it is most common to use A* combined with a map simplification technique that generates a simpler navigation graph to be used for pathfinding.
+An example of such technique is Triangulation Reduction A*, that computes polygonal triangulations on a grid-based map \cite{Demyen_2006}.
+Considering movement for groups of units, rather then individual units, techniques such as steering of flocking behaviors \cite{Reynolds_1999} can be used on top of a path-finding algorithm in order to make whole groups of units follow a given path.
+In recent commercial RTS games like Starcraft 2 or Supreme Commander 2, flocking-like behaviors are inspired of continuum crowds (``flow field'') \cite{Treuille2006}.
+A comprehensive review about (grid-based) pathfinding was recently done by Sturtevant \cite{sturtevant2012benchmarks}.
 
@@
 
-One interesting aspect of the five bots described above is that, while all of them are reactive at the lower level (unit control), most if not all of them, are scripted at the highest level of abstraction.
+One interesting aspect of the five bots described above is that, while all of them are reactive at the lower level (reactive control), most if not all of them, are scripted at the highest level of abstraction.
 BTHAI reads build and squad formations from a predefined script, Nova's {\em Strategy Manager} is a predefined finite-state machine, BBQ's construction manager reads the build order from a predefined script, and Skynet's {\em BuildOrder Manager} is basically a predefined script.

### 2012-05-30 survey.tex
small addition

@@
 \paragraph*{Protoss Beast Jelly (Protoss)}
+always goes for a 5-gate Zealot rush, supported by an effective harvesting 
+strategy named power-mining (2 probes are assigned to every mineral patch,
+thereby needing 18 probes for 100\% saturation in a normal map, prior
+to expanding). Gas is not mined as it is not needed for constructing Zealots.
 

### 2012-06-30 survey.tex
small correction in related work

@@
 Game tree search techniques have also been explored for tactical decision making.
-\cite{Chung05} adapted Monte-Carlo tree search (MCTS) to planning in RTS games and applied it to a capture-the-flag mod of Open RTS.
+\cite{Chung05} applied Monte-Carlo planning to a capture-the-flag mod of Open RTS.
 Balla and Fern \cite{UCT} applied upper confidence bounds on trees (UCT: a MCTS algorithm) to tactical assault planning in Wargus.
@@
 \cite{Danielsiek_2008} used influence maps to achieve intelligent squad movement to flank the opponent in a RTS game.
-A drawback go potential field-based techniques is the large number of parameters that has to be tuned in order to achieve the desired behavior.
+A drawback for potential field-based techniques is the large number of parameters that has to be tuned in order to achieve the desired behavior.
 Approaches for automatically learning such parameters have been explored, for example, using reinforcement Learning \cite{Liu_2008}, or self-organizing-maps (SOM) \cite{teamCompositionRTS}.

### 2012-07-19 survey.tex
added a preliminary outline of the open challenges section

@@
 {\color{blue}
-Here, describe what is left to be done.
+[This is a preliminary list of open questions, derived from the Skype call on July 19. Some work is still required to turn this into the final list]
+
+There are a set of problems that are either solved in RTS games (path-finding), or for which we have very good solutions (micro management).
+But there are features that are not exhibited by any of the current Starcraft playing bots.
+For example, current bots are incapable of coming up on their own with tactical moves such as unit drops.
+Some bots do drops, but only if this is hard-coded; no bot has the capability of reasoning about all the possible moves available, and discovering that doing a drop is the best thing to do right now.
+This is related to the lack of real-time adversarial planning techniques that scale up to RTS games.
+Similarly, current bots do not exhibit adaptation to opponent strategies.
+Some switch between different build-orders, but do not fully adapt their strategy.
+No bot is capable of observing the opponent and autonomously come up with a good plan (from scratch) to counter the opponent strategy.
+
+In general, this is a list of open problems:
+
+\begin{itemize}
+\item Adaptation to opponent
+\item Learning (from opponent, from traces or from mistakes)
+\item Scalability: this includes:
+\begin{itemize}
+\item a) how to scale up existing algorithms for adversarial domains like game tree search or others, and 
+\item b) multi-scale planning/reasoning, how to integrate control, tactics and strategy together (e.g. how to interface them).
+\end{itemize}
+\item Adversarial real-time planning and scheduling under uncertainty with resources: this basically includes three general unsolved AI problems:
+\begin{itemize}
+\item Adversarial planning under real-time constraints (there is some work on this area to be presented this year in AIIDE)
+\item Adversarial planning under uncertainty 
+\item Planning with resources (there exist planning algorithms that handle resources, like GRT-R, but cannot scale up to the size of problems needed for e.g. Starcraft)
+\end{itemize}
+\item How to exploit and incorporate domain knowledge: we know how to incorporate some aspects of domain knowledge (e.g.
+build orders), but, in general, how to incorporate some forms of domain knowledge into algorithms for RTS games is still an open problem.
+For example, standard techniques to encode strategies for other forms of games, like Behavior Trees, are hard to deploy in RTS games.
+\end{itemize}
 }

### 2012-07-19 survey.tex
competition results of CIG 2011 completed

@@
 At CIG 2011, the tournament was therefore run with a (secret) selection
-of maps used in league play. It was organized by Tobias Mahlmann and Mike
+of maps used in league play, which can be regarded as the most important
+difference to the AIIDE tournament that employed a known list of maps.
+The competition was organized by Tobias Mahlmann and Mike
 Preuss and attracted 10 bots. In addition to the ones discussed in previous
@@
 with BTHAI of which 6:4 were won).}
-\label{tab:correlations}
+\label{tab:cig-first-round}
 \begin{small}
@@
 
-As BroodwarBotQ and BTHAI have the same number of wins, their direct encounter is evaluated which accounted 6 : 4 for the BroodwarBotQ.
-Note that all qualified bots play Protoss.
+\subsubsection{First Round}
+\label{sec:cig-first-round}
 
-Maps used for the first round: (2)MatchPoint 1.3, (4)Fighting Spirit 1.3, iCCupdestination 1.1, iCCup gaia, iCCup great barrier reef
+As the CIG competition games were executed manually due to
+a lack of available software (the AIIDE program was not yet
+available at that time), we separated the ten entries into
+two brackets. In each bracket of 5 bots, a round-robin
+tournament was held with 10 repetitions per pairing, resulting
+in 40 games per bot.
+The 5 maps chosen for the first round were selected from the pool
+of well-known league play maps found on the internet:
+\emph{(2)MatchPoint 1.3}, \emph{(4)Fighting Spirit 1.3}, 
+\emph{iCCupdestination 1.1}, \emph{iCCup gaia}, and 
+\emph{iCCup great barrier reef}. Each bot pairing played
+on every map twice, with switched starting positions.
 
-Final Round
+The two top bots of every bracket qualified
+for the final round. Table~\ref{tab:cig-first-round} summarizes
+the results.
+Note that as BroodwarBotQ and BTHAI have the same number of wins,
+their direct encounter was evaluated which accounted 6:4 for the BroodwarBotQ.
+The bots going into the final were thus UAlbertaBot, Skynet (from bracket A)
+and Xelnaga and BroodwarBotQ (from bracket B). All qualified bots play the
+Protoss faction. Most bots proved pretty stable, only Xelnaga and Protoss 
+Beast Jelly crashed relatively often (each in more than a quarter of the games). 
+Crashing of course resulted in an instant win for the other bot.
+In some cases, neither bot was able to finish the other off completely,
+so that they went into a passive state. We manually ended such games after
+around 15 minutes and assigned victory to the bot that had obtained more
+points as indicated on the end game screen.
 
-The final round was played in a similar mode as each of the first round brackets, using another set of 5 maps (see below), resulting in 30 games for each bot.
+\subsubsection{Final Round}
+\label{sec:cig-final-round}
 
-Crashes	 Games	 Bot	 Wins
-0	 30	 Skynet	 26
-0	 30	 UAlbertaBot	 22
-3	 30	 Xelnaga	 11
-2	 30	 BroodwarBotQ	 1
+The final round was played in a similar mode as each of the
+first round brackets,
+using another set of 5 previously unknown maps:
+\emph{iCCup lost temple 2.4}, \emph{iCCup rush hour 3.1},
+\emph{iCCup swordinthemoon 2.1}, \emph{iCCup yellow 1.1},
+and \emph{La\_Mancha 1.1}. 
+Letting each pairing play on each map twice again with
+switching starting positions resulted in 30 games per bot.
+The final results are displayed in table~\ref{tab:cig-final-round},
+indicating Skynet as winner and UAlbertaBot as runner-up, being
+almost equally strong, and the two other bots as clearly inferior.
 
-Here describe the results obtained in the competition comparing the bots.
+\begin{table}[bth]
+\caption{Results of the CIG 2011 final round.}
+\label{tab:cig-final-round}
+\begin{small}
+\begin{center}
+% bracket A
+\begin{tabular}{|l|c|c|c|}
+\hline
+Crashes & games & bot	& wins\\ \hline
+0 &  30 &	 Skynet	  		&  26\\
+0 &	 30 &	 UAlbertaBot 	&  22\\
+3 &	 30 &	 Xelnaga  		&  11\\
+2 &	 30 &	 BroodwarBotQ  	&  1\\
+\hline
+\end{tabular}
+\end{center}
+\end{small}
+\end{table}
+
 }

### 2012-08-15 survey.tex
Some more work on the open problems section

@@
 \section{Real-Time Strategy Games}\label{sec:rts}
+
+{\color{blue} [Here, a brief overview of RTS games, which are their main characteristics, etc.]}
 
@@
 
+\subsection{Holistic Approaches}
+
+{\color{red} [Santi: some RTS AI techniques attempt to address the problem in a holistic way, for example the Darmok system, is there enough work on this for justifying a whole subsection?
+or should we just spread the few pieces of work on this over the previous sections].}
+
 \section{State of the Art Bots for Starcraft - \color{blue}{Information/Productions}}\label{sec:bot}
@@
 [This is a preliminary list of open questions, derived from the Skype call on July 19. Some work is still required to turn this into the final list]
+}
 
-There are a set of problems that are either solved in RTS games (path-finding), or for which we have very good solutions (micro management).
-But there are features that are not exhibited by any of the current Starcraft playing bots.
-For example, current bots are incapable of coming up on their own with tactical moves such as unit drops.
-Some bots do drops, but only if this is hard-coded; no bot has the capability of reasoning about all the possible moves available, and discovering that doing a drop is the best thing to do right now.
-This is related to the lack of real-time adversarial planning techniques that scale up to RTS games.
+As illustrated in this paper, there is a set of problems in RTS game AI that could be considered mostly solved, of for which we have very good solutions.
+One example of such problems is pathfinding (mostly solved) or micro-management (we have good solutions).
+
+However, there are many other problems for which this is not the case.
+For example, there is no current Starcraft bot that can come up with its own tactical moves, such as ``unit drops''.
+Some bots do drops, but only if this is hard-coded; no bot has the capability of reasoning about the current situation, synthesize a tactical move that involves a ``unit drop'', and determine that this move is the best one in the current situation.
+This is related to the lack of real-time adversarial planning techniques that scale up to the size required for RTS games.
 Similarly, current bots do not exhibit adaptation to opponent strategies.
 Some switch between different build-orders, but do not fully adapt their strategy.
-No bot is capable of observing the opponent and autonomously come up with a good plan (from scratch) to counter the opponent strategy.
+No bot is capable of observing the opponent and autonomously synthesize a good plan from scratch to counter the opponent strategy.
 
-In general, this is a list of open problems:
+Specifically, the list of problems that are currently unsolved, grouped in various categories:
+
+{\color{blue}
 
 \begin{itemize}
-\item Adaptation to opponent
-\item Learning (from opponent, from traces or from mistakes)
-\item Scalability: this includes:
+
+\item Learning and adaptation:
 \begin{itemize}
-\item a) how to scale up existing algorithms for adversarial domains like game tree search or others, and 
-\item b) multi-scale planning/reasoning, how to integrate control, tactics and strategy together (e.g. how to interface them).
+\item Adaptation to opponent strategy.
+\item Learning from experience in RTS games.
+\item Learning from observation (from demonstration, or from observing the opponent) in RTS games.
 \end{itemize}
-\item Adversarial real-time planning and scheduling under uncertainty with resources: this basically includes three general unsolved AI problems:
+
+\item Planning:
 \begin{itemize}
-\item Adversarial planning under real-time constraints (there is some work on this area to be presented this year in AIIDE)
-\item Adversarial planning under uncertainty 
-\item Planning with resources (there exist planning algorithms that handle resources, like GRT-R, but cannot scale up to the size of problems needed for e.g. Starcraft)
+\item Adversarial planning under real-time constraints (there is some work on this area to be presented this year in AIIDE).
+\item Adversarial planning under uncertainty.
+\item Adversarial planning in partially-observable domains.
+\item Adversarial planning with resources (there exist planning algorithms that handle resources, like GRT-R, but cannot scale up to the size of problems needed for e.g. Starcraft).
 \end{itemize}
-\item How to exploit and incorporate domain knowledge: we know how to incorporate some aspects of domain knowledge (e.g.
+
+\item Scalability:
+\begin{itemize}
+\item How to scale up existing algorithms for adversarial domains like game tree search or Monte Carlo search to the very large dimensions of RTS games.
+\end{itemize}
+
+\item Integration:
+\begin{itemize}
+\item Multi-scale planning/reasoning: how to integrate reasoning at multiple levels of abstraction.
+\end{itemize}
+
+\item Domain Knowledge: 
+\begin{itemize}
+\item we know how to incorporate some aspects of domain knowledge (e.g.
 build orders), but, in general, how to incorporate some forms of domain knowledge into algorithms for RTS games is still an open problem.
 For example, standard techniques to encode strategies for other forms of games, like Behavior Trees, are hard to deploy in RTS games.
+\end{itemize}
 \end{itemize}

### 2012-09-20 survey.tex
minor corrections

--- 2012-07-19 survey.tex
@@
 Hard-coded approaches have been extensively used in commercial RTS games.
-The most common approaches make use finite state machines (FSM) \cite{FSM_AIGameProgWisdom2003} in order to let the AI author hard-code the strategy that the AI will employ.
+The most common approaches use finite state machines (FSM) \cite{FSM_AIGameProgWisdom2003} in order to let the AI author hard-code the strategy that the AI will employ.
 The idea behind FSMs is to decompose the AI behavior into easily manageable states, such as ``attacking'', ``gathering resources'' or ``repairing'' and establish the conditions that trigger transitions between them.
@@
 \cite{HMMstrat_RTS_AIIDE11} used HMM to learn the transition probabilities of sequences of constructions and kept the most probables to produce probabilistic behavior models (in StarCraft).
-Synnaeve and Bessi\`{e}re \cite{SynnaeveOpeningCig11} extended \cite{WeberCig09} and used a Bayesian semi-supervised (labeling based on EM clustering considering appropriate features) model to learn from replays and predict openings (early game strategies) from StarCraft replays.
-Then \cite{SynnaeveAIIDE11}, they presented an unsupervised learning Bayesian model for tech-tree prediction, still using replays.
+Synnaeve and Bessi\`{e}re \cite{SynnaeveOpeningCig11} used the dataset of \cite{WeberCig09} and presented a Bayesian semi-supervised model to learn from replays and predict openings (early game strategies) from StarCraft replays.
+The openings are labeled by EM clustering considering appropriate features.
+Then in \cite{SynnaeveAIIDE11}, they presented an unsupervised learning Bayesian model for tech-tree prediction, still using replays.
 
@@
 {\color{blue}
-[This is a preliminary list of open questions, derived from the Skype call on July 19. Some work is still required to turn this into the final list]
-
-There are a set of problems that are either solved in RTS games (path-finding), or for which we have very good solutions (micro management).
-But there are features that are not exhibited by any of the current Starcraft playing bots.
-For example, current bots are incapable of coming up on their own with tactical moves such as unit drops.
-Some bots do drops, but only if this is hard-coded; no bot has the capability of reasoning about all the possible moves available, and discovering that doing a drop is the best thing to do right now.
-This is related to the lack of real-time adversarial planning techniques that scale up to RTS games.
-Similarly, current bots do not exhibit adaptation to opponent strategies.
-Some switch between different build-orders, but do not fully adapt their strategy.
-No bot is capable of observing the opponent and autonomously come up with a good plan (from scratch) to counter the opponent strategy.
-
-In general, this is a list of open problems:
-
-\begin{itemize}
-\item Adaptation to opponent
-\item Learning (from opponent, from traces or from mistakes)
-\item Scalability: this includes:
-\begin{itemize}
-\item a) how to scale up existing algorithms for adversarial domains like game tree search or others, and 
-\item b) multi-scale planning/reasoning, how to integrate control, tactics and strategy together (e.g. how to interface them).
-\end{itemize}
-\item Adversarial real-time planning and scheduling under uncertainty with resources: this basically includes three general unsolved AI problems:
-\begin{itemize}
-\item Adversarial planning under real-time constraints (there is some work on this area to be presented this year in AIIDE)
-\item Adversarial planning under uncertainty 
-\item Planning with resources (there exist planning algorithms that handle resources, like GRT-R, but cannot scale up to the size of problems needed for e.g. Starcraft)
-\end{itemize}
-\item How to exploit and incorporate domain knowledge: we know how to incorporate some aspects of domain knowledge (e.g.
-build orders), but, in general, how to incorporate some forms of domain knowledge into algorithms for RTS games is still an open problem.
-For example, standard techniques to encode strategies for other forms of games, like Behavior Trees, are hard to deploy in RTS games.
-\end{itemize}
+Here, describe what is left to be done.
 }

### 2012-12-11 survey.tex
Updated the challenges section

--- 2012-08-15 survey.tex
@@
 
+{\color{blue}
 \begin{abstract}
@@
 \end{abstract}
+}
 
@@
 \IEEEPARstart{S}{ince} Michael Buro's call for research in RTS games \cite{Buro03rts}, many researchers have answered the call.
-Specially, AI competitions like the Starcraft one have caused many AI techniques to be
+Specially, AI competitions like the ``AIIDE Starcraft AI Competition'' have caused many AI techniques to be
 applied to RTS AI. We will list and classify these approaches, explain their 
@@
 
+{\color{red} make sure to introduce the term ``bot''}
+
 \section{Real-Time Strategy Games}\label{sec:rts}
 
-{\color{blue} [Here, a brief overview of RTS games, which are their main characteristics, etc.]}
+Real-time Strategy (RTS) is a sub-genre of strategy games where players need to build an economy (gathering resources and building a base) and a military power (training units and researching technologies) in order to defeat their opponents (destroying their army and base).
+From a theoretical point of view, the main differences between RTS games and traditional board games such as Chess are:
 
-\subsection{StarCraft {\color{blue}- Situation}}\label{subsec:starcraft}
+\begin{itemize}
+\item They are {\em simultaneous move} games, where more than one player can issue actions at the same time.
+Additionally, these actions are {\em durative}, i.e.
+actions are not instantaneous, but take some amount of time time to complete.
+\item RTS games are ``real-time'', which actually means is that each player has a very small amount of time to decide the next move.
+Compared to Chess, where players have several minutes to decide the next action, in Starcraft, the game executes at 24 iterations per second, which means that players only have 55ms to decide a move, before the game state changes.
+\item Most RTS games are partially observable: players can only see the part of the map that has been explored.
+\item Most RTS games are non-deterministic.
+\item And finally, the complexity of these games, both in terms of state space size and in terms of number of actions available at each decision cycle is very large.
+{\color{red} For example, the state space of Chess is typically estimated to be around $10^{50}$, whereas the state space of Starcraft in a typical map is estimated to be around $10^{10000}$.}
+\end{itemize}
+
+For those reasons, standard techniques used for playing classic board games, such as minimax game tree search, cannot be directly applied to solve RTS games without the definition of some level of abstraction, or some other simplification.
+Interestingly enough, humans seem to be able to deal with the complexity of RTS games, and are still vastly superior to computers in these type of games \cite{burochurchill2012aimagazine}.
+For those reasons, a large spectrum of techniques have been attempted to deal with this domain, as we will describe below.
+The remainder of this section is devoted to describe Starcraft as a research testbed, and on detailing which are the open challenges in RTS game AI.
+
+\subsection{StarCraft}\label{subsec:starcraft}
 
@@
 Starcraft is set in a science-fiction based world where the player must choose one of the three races: Terran, Protoss or Zerg.
-The good work done by the people of Blizzard makes this game one of the most extremely well-balanced RTS games ever created.
+One of the most remarkable aspects of Starcraft is that the three races in Starcraft are extremely well balanced:
+
 \begin{itemize}
-	\item Terrans, human exiled from Earth, provide units that are versatile and flexible giving a balanced option between Protoss and Zergs.
+	\item Terrans, provide units that are versatile and flexible giving a balanced option between Protoss and Zergs.
 	\item Protoss units have lengthy and expensive manufacturing process, but they are strong and resistant. These conditions makes players follow a strategy of quality over quantity.
@@
 
-{\color{blue}TODO: explain the gameplay a little more, how are ppl not familiar to SC able to understand the problems for AI that we raise below?}
+Figure \ref{fig:starcraft} shows a screenshot of Starcraft showing a player playing Terrans.
+In order to win a Starcraft game, players must first gather resources (minerals and gas), which they can spend in creating additional buildings and units.
+As resources become available, players need to allocate them for creating more buildings (which reinforce the economy, and allow players to create units or unlock stronger units) and train attack units.
+Units must be distributed to accomplish different tasks such as: reconnaissance, defense and attack.
+While performing all of those tasks, players also need to strategically understand the geometry of the map at hand, in order to decide where to place new buildings (concentrate in a single area, or expand to different areas) or where to set defensive outposts.
+Finally, when offensive units of two players meet, each player must quickly maneuver each of the units in order to fight the a battle, which requires quick and reactive control of each of the units in order to successfully fight a battle.
 
-\subsection{Challenges in RTS Game AI {\color{blue}- Problems}}\label{subsec:challenges}
+\begin{figure}
+    \centering
+    \includegraphics[width=0.8\columnwidth]{figures/starcraft1.png}
+    \caption{A screenshot of Starcraft {\color{red} is this a good screenshot? can we actually have a screenshot? or is this copyrighted? Do we even want a screenshot?}.}
+    \label{fig:starcraft}
+\end{figure}
 
-{\color{blue}
-This section should be brief, and just give us some basic ideas of which are the challenges in RTS Game AI, so that when we refer to related work, we can use this as a reference point.
-The most important thing of this section is to demonstrate that RTS games are complex.
+{\color{red} A typical Starcraft map is defined as a rectangular grid, typically ranging between 64x64 to 128x128 in size (although the resolution at which units move is finer), and each player can control up to 200 units (not including buildings).
+Moreover, each different race contains between 30 to 35 different types of units and buildings, each one with a significant number of special abilities}.
+All these factors together make Starcraft a significant challenge, in which humans are still much better than computers.
 
-Many years have passed since Buro's call for research (8!!). he identified 6 challenges:
+\subsection{Challenges in RTS Game AI}\label{subsec:challenges}
+
+Many years have passed since Buro's call for research, where he identified the following 6 challenges:
 \begin{itemize}
@@
 
-There has been a lot of work in many, but others have been untouched (e.g.
-Collaboration).
-Additionally, other challenges that are not in the list appeared, and have been worked on, like: how to exploit the existing domain knowledge (strategies, build-orders, replays, etc.), or how to design an architecture that integrates all the modules required for an agent to play an RTS games.
+While there has been a significant work in many, others have been untouched (e.g.
+collaboration).
+Additionally, after the increase attention to the problem of RTS games, further, unforeseen challenges appeared.
+For example, how to exploit the massive amounts of existing domain knowledge (strategies, build-orders, replays, and so on), or how to design an architecture that integrates all the modules required for an agent to play an RTS games.
+In the following subsections, we describe current challenges in RTS Game AI, grouped in different areas.
 
-\begin{itemize}
-\item Task Decomposition (or ``Architecture'')
-\item Integration of Domain Knowledge
-\item Reasoning with Uncertainty (including information gathering)
-\item Opponent Modeling and Adaptation: opponent modeling is key if we have to adapt our strategy.
+%\begin{itemize}
+%\item Task Decomposition (or ``Architecture'')
+%\item Integration of Domain Knowledge
+%\item Reasoning with Uncertainty (including information gathering)
+%\item Opponent Modeling and Adaptation: opponent modeling is key if we have to adapt our strategy.
 As different strategies are dominating each others, forming multiple Nash equilibria, the AI has to be able to infer the intend of its opponent.
-\item Group and Individual Control (``micro''): the task of controlling units efficiently (we can not speak of optimality here due to the huge state space, and of the enemy behavior entails numerous Nash equilibria), focusing fire to diminish enemy's firepower and keeping our units alive the longest, casting defensive and offensive spells and abilities.
-\item Planning and Resource Allocation (``macro'')
-\item Spatial reasoning (``tactics'')
-\end{itemize}
+%\item Group and Individual Control (``micro''): the task of controlling units efficiently (we can not speak of optimality here due to the huge state space, and of the enemy behavior entails numerous Nash equilibria), focusing fire to diminish enemy's firepower and keeping our units alive the longest, casting defensive and offensive spells and abilities.
+%\item Planning and Resource Allocation (``macro'')
+%\item Spatial reasoning (``tactics'')
+%\end{itemize}
 
-Maybe the list above goes better in Section \ref{sec:questions}?
-}
+\subsubsection{Planning}
+{\color{blue}
+TODO (Santi?) just a paragraph on exposing why there is a planning problem in RTS AI (macro and micro).
 
-{\color{red}
-TODO merge above and bottom
-}
-
-{\color{blue}
-\textbf{Planning}
-TODO (Santi?) just a paragraph on exposing why there is a planning problem in RTS AI.
-
-\textbf{Spatial and Temporal Reasoning}
-TODO just a paragraph on exposing why there are temporal and spatial reasoning in RTS AI.
-}
-
-\textbf{Learning}\\
+\paragraph{Learning}
 When talking about learning in RTS games, with regard to a particular match against an opponent (where a match is a sequence of games against the same opponent), one has to differentiate between:
@@
 
-\textbf{Uncertainty}\\
+\subsubsection{Uncertainty}
 In RTS games, there are two main kinds of uncertainty, existing for two different reasons.
@@
 
+\subsubsection{Spatial and Temporal Reasoning}
+TODO just a paragraph on exposing why there are temporal and spatial reasoning in RTS AI.
+
+\subsubsection{Domain Knowledge Exploitation}
+TODO
+
+\subsubsection{Task Decomposition}
+TODO
+}
+
 \section{Existing work on RTS Game AI - \color{blue}{Resolutions}}\label{sec:review}
 
-%\subsection{RTS AI Decomposition}
 \begin{figure}
@@
 
-{\color{blue}
-[This is a preliminary list of open questions, derived from the Skype call on July 19. Some work is still required to turn this into the final list]
-}
-
 As illustrated in this paper, there is a set of problems in RTS game AI that could be considered mostly solved, of for which we have very good solutions.
-One example of such problems is pathfinding (mostly solved) or micro-management (we have good solutions).
+One example of such problems is pathfinding (mostly solved) or low-scale micro-management (for which we have good solutions).
 
@@
 
-Specifically, the list of problems that are currently unsolved, grouped in various categories:
-
-{\color{blue}
+We present here a list of problems that are currently unsolved, grouped in various categories.
 
@@
 
-\item Learning and adaptation:
+\item {\bf Learning and adaptation:}
 \begin{itemize}
-\item Adaptation to opponent strategy.
-\item Learning from experience in RTS games.
-\item Learning from observation (from demonstration, or from observing the opponent) in RTS games.
+\item Adaptation to opponent strategy: observing the opponent strategy, and synthesizing an adequate counter strategy.
+Current bots switch between predefined strategies based on hard-coded preconditions, or based on the performance of each predefined strategy against an opponent in previous games, but no current bot creates new strategies (like Chess or Go playing programs do).
+\item Learning from experience in RTS games: how can we make a bot that improves performance over time?
+some current bots learn which strategy (out of a predefined set of strategies) is best against a given opponent, but how can we devise learning strategies that can perform more general learning?
+\item Learning from observation (from demonstration, or from observing the opponent) in RTS games: how can we learn by observing the game play of other players?
+can we devise algorithms that can automatically extract strategies from observation, and later apply them?
 \end{itemize}
 
-\item Planning:
+\item {\bf Planning:}
 \begin{itemize}
-\item Adversarial planning under real-time constraints (there is some work on this area to be presented this year in AIIDE).
-\item Adversarial planning under uncertainty.
-\item Adversarial planning in partially-observable domains.
-\item Adversarial planning with resources (there exist planning algorithms that handle resources, like GRT-R, but cannot scale up to the size of problems needed for e.g. Starcraft).
+\item Adversarial planning under real-time constraints: although some solutions for small-scale real-time planning have been recently proposed (such as \cite{churchill2012AIIDE}, based on alpha-beta game-tree search), the problem of large-scale adversarial planning under real-time constraints is still open.
+\item Adversarial planning under uncertainty of partially-observable domains: how can we adapt adversarial planning techniques for dealing with uncertainty?
+This problem has been widely studied in the context of simple games such as back-gammon \cite{???}, or Poker \cite{???}.
+However, the techniques developed for those domains do not scale to RTS-game scenarios.
+\item Adversarial planning with resources: similarly, even if there exist planning algorithms that handle resources (like GRT-R \cite{???}), they cannot scale up to the size of problems needed for RTS games like Starcraft.
 \end{itemize}
 
-\item Scalability:
+\item {\bf Integration:}
 \begin{itemize}
-\item How to scale up existing algorithms for adversarial domains like game tree search or Monte Carlo search to the very large dimensions of RTS games.
+\item Multi-scale planning/reasoning: as described in this paper, all the bots developed for the Starcraft AI competitions decompose the problem of playing an RTS game into smaller subproblems, and then solutions for each of those subproblems are integrated in to a common architecture to play the game.
+However, the integration of each of the modules in a unified architecture is still an open problem.
+For example, how can decisions made at high-level modules be integrated with decisions made at lower-level modules?
 \end{itemize}
 
-\item Integration:
+\item {\bf Domain Knowledge: }
 \begin{itemize}
-\item Multi-scale planning/reasoning: how to integrate reasoning at multiple levels of abstraction.
-\end{itemize}
-
-\item Domain Knowledge: 
-\begin{itemize}
-\item we know how to incorporate some aspects of domain knowledge (e.g.
-build orders), but, in general, how to incorporate some forms of domain knowledge into algorithms for RTS games is still an open problem.
+\item We know how to incorporate some aspects of domain knowledge (e.g.
+build orders) into RTS game playing agents.
+But, in general, how to incorporate some forms of domain knowledge into algorithms for RTS games is still an open problem.
 For example, standard techniques to encode strategies for other forms of games, like Behavior Trees, are hard to deploy in RTS games.
+Is it possible to devise techniques that can automatically mine the existing collections of domain knowledge for an RTS game like Starcraft, and incorporate it into the bot?
+An initial exploration of this idea was carried out by Branavan et al.
+\cite{???}
 \end{itemize}
 \end{itemize}
-}
 

### 2013-01-03 survey.tex
bibliography (CIG 2012)

--- 2012-09-20 survey.tex
@@
 Potential fields (or influence maps) have been found to be a useful technique for reactive decision making.
-Some uses of potential fields in RTS games are: avoiding obstacles (navigation), avoiding opponent fire, or staying at maximum shooting distance \cite{Hagelback08,Hagelback09}.
+Some uses of potential fields in RTS games are: avoiding obstacles (navigation), avoiding opponent fire, or staying at maximum shooting distance \cite{Hagelback09}.
+Potential fields have also been combined with A* path-finding to avoid local traps \cite{Hagelback12}.
 Hagelb\"{a}ck and Johansson \cite{HagelbackJ08} presented a multi-agent potential fields based bot able to deal with fog of war in the Tankbattle game.

### 2013-02-03 survey.tex

@@
 The openings are labeled by EM clustering considering appropriate features.
-Then in \cite{SynnaeveAIIDE11}, they presented an unsupervised learning Bayesian model for tech-tree prediction, still using replays.
+Then, in \cite{SynnaeveAIIDE11}, they presented an unsupervised learning Bayesian model for tech-tree prediction, still using replays.
+% Finally \cite{GemineImitative} used the same ``build-order features'' approach to learn to adapt the bot's build order to the opponents strategy in StarCraft 2.
 
@@
 \cite{CBR-RL} combined CBR and reinforcement learning to enable reuse of tactical plan components.
-Finally, Cadena and Garrido \cite{CadenaG11} used fuzzy CBR (fuzzy case matching) for strategic and tactical planning.
+Cadena and Garrido \cite{CadenaG11} used fuzzy CBR (fuzzy case matching) for strategic and tactical planning.
+Finally, \cite{SynnaeveTactics} combined space abstraction into regions from \cite{Perkins10} and tactical-decision making by assigning scores (economical, defenses, etc.) to regions and looking for their correspondences to tactical moves (attacks) in pro-gamers replays.
 
@@
 Bayesian modeling has been applied to inverse fusion of the sensory inputs of the units \cite{SynnaeveMicroCig11}, which subsumes potential fields, allowing for integration of tactical goals directly in micro-management.
-Additionally, there has been some interesting uses of reinforcement learning (RL) \cite{Sutton}: Marthi et al.
+
+Additionally, there has been some interesting uses of reinforcement learning (RL) \cite{Sutton}: 
+%\cite{WenderRL} evaluated the different major RL algorithms for (decentralized) micro-management, which perform all equally.
+Marthi et al.
 \cite{Marthi05} employ concurrent hierarchical Q-learning (units Q-functions are combined at the group level) RL to efficiently control units in a ``one robot with multiple effectors'' fashion.
@@
 In real game setups, the actions spaces to explore is gigantic, it requires to use the structure of the game in a partial program (or a partial Markov decision process) and a shape function (or a heuristic) \cite{Marthi05}.
+Evolutionary optimization by simulating fights can easily adapted to any parameter-dependent micro-management control model, as shown by \cite{OthmanSimu} which optimizes an AIIDE 2010 micro-management competition bot.
+
 Other approaches that aim at learning the parameters of an underlying model have also been explored.

### 2013-02-03 survey.tex
Updated related work with papers from AIIDE 2012

@@
 % Finally \cite{GemineImitative} used the same ``build-order features'' approach to learn to adapt the bot's build order to the opponents strategy in StarCraft 2.
+Finally, evolutionary approaches to determine priorities of high level tasks was explored by Young and Hawes in their QUORUM system \cite{young2012evolutionary}, showing improvement over static priorities.
 
@@
 Potential fields (or influence maps) have been found to be a useful technique for reactive decision making.
-Some uses of potential fields in RTS games are: avoiding obstacles (navigation), avoiding opponent fire, or staying at maximum shooting distance \cite{Hagelback09}.
+Some uses of potential fields in RTS games are: avoiding obstacles (navigation), avoiding opponent fire \cite{uriarte2012kiting}, or staying at maximum shooting distance \cite{Hagelback09}.
 Potential fields have also been combined with A* path-finding to avoid local traps \cite{Hagelback12}.
@@
 Additionally, there has been some interesting uses of reinforcement learning (RL) \cite{Sutton}: 
-%\cite{WenderRL} evaluated the different major RL algorithms for (decentralized) micro-management, which perform all equally.
+Wender and Watson \cite{WenderRL} evaluated the different major RL algorithms for (decentralized) micro-management, which perform all equally.
 Marthi et al.
@@
 \cite{Madeira06} advocate the use of prior domain knowledge to allow faster RL learning and applied their work on a turn-based strategy game.
-In real game setups, the actions spaces to explore is gigantic, it requires to use the structure of the game in a partial program (or a partial Markov decision process) and a shape function (or a heuristic) \cite{Marthi05}.
-Evolutionary optimization by simulating fights can easily adapted to any parameter-dependent micro-management control model, as shown by \cite{OthmanSimu} which optimizes an AIIDE 2010 micro-management competition bot.
+The actions spaces to explore is gigantic for real game setups.
+It requires to use the structure of the game in a partial program (or a partial Markov decision process) and a shape function (or a heuristic) \cite{Marthi05}.
+Another approach is that proposed by Jaide and Mu{\~n}oz-Avila \cite{jaidee2012classq} through learning just one Q-function for each unit type, in order to cut down the search space.
 
@@
 Synnaeve and Bessi\`{e}re \cite{SynnaeveMicroCig11} achieve hierarchical goals (coming from tactical decisions) integration through the addition of another sensory input corresponding to the goal's objective.
+Finally, evolutionary optimization by simulating fights can easily adapted to any parameter-dependent micro-management control model, as shown by \cite{OthmanSimu} which optimizes an AIIDE 2010 micro-management competition bot.
 
@@
 In conclusion, we can see that there are two basic tools that can be used in an integration architecture: abstraction and divide-and-conquer, which are widely used by the existing Starcraft bots.
+For space reasons, we do not include an exhaustive comparison of the architectures of all the participating bots.
+Some other bots have been documented by their authors, such as SCAIL \cite{YoungSCAIL} or QUORUM \cite{young2012evolutionary}.
 

### 2013-02-06 survey.tex
ncorporated DAve's input for the AIIDE competitions

@@
 
-\section{Existing work on RTS Game AI - \color{blue}{Resolutions}}\label{sec:review}
+\section{Existing work on RTS Game AI}\label{sec:review}
 
@@
 Moreover, none of these approaches addresses any timing or scheduling issues, which are key in RTS games.
+On notable exception is the work of Churchill and Buro \cite{churchill2011build}, who used planning in order to construct its economic build-orders taking into account timing constraints of the different actions.
 
@@
 \cite{Chung05} applied Monte-Carlo planning to a capture-the-flag mod of Open RTS.
-Balla and Fern \cite{UCT} applied upper confidence bounds on trees (UCT: a MCTS algorithm) to tactical assault planning in Wargus.
+Balla and Fern \cite{UCT} applied the UCT algorithm (a Monte Carlo Tree Search algorithm) to tactical assault planning in Wargus.
 % (Santi: I commented this one, since it's mentioned above) On influence maps, \cite{teamCompositionRTS} studied team composition and maneuvering by learning a self-organizing map, while \cite{HagelbackJ08} presented a multi-agent potential field based bot.
@@
 
-\section{State of the Art Bots for Starcraft - \color{blue}{Information/Productions}}\label{sec:bot}
+\section{State of the Art Bots for Starcraft}\label{sec:bot}
 
@@
 
-Figure \ref{fig:bot-architecture} shows some representative examples of the integration architectures used by different bots in the AIIDE and CIG 2011 Starcraft AI competitions \cite{url1,url2}: BroodwarQ \cite{???}, Nova \cite{???}, UAlbertaBot \cite{???}, Skynet \cite{???}, SPAR \cite{???}, AIUR \cite{???}, and BTHAI \cite{???} (see Section \ref{sec:competition} for a comparison of their performance).
+Figure \ref{fig:bot-architecture} shows some representative examples of the integration architectures used by different bots in the AIIDE and CIG Starcraft AI competitions (see Section \ref{sec:competition}): BroodwarQ \cite{???}, Nova \cite{???}, UAlbertaBot \cite{???}, Skynet \cite{???}, SPAR \cite{???}, AIUR \cite{???}, and BTHAI \cite{???}.
 Each box represents an individual module with a clearly defined task (only modules with a black background can send actions directly to Starcraft).
@@
 
-{\color{violet}
-
 In order to inject new AI code into the StarCraft game, all recent
@@
 
-\subsection{AIIDE}
-\label{sec:AIIDE}
-this is left empty for now
+The following subsections summarize the results of all the Starcraft AI competitions held at the AIIDE (Artificial Intelligence for Interactive Digital Entertainment) and CIG (Computational Intelligence in Games) conferences during the past years.
+
+\subsection{AIIDE}\label{sec:AIIDE}
+
+Started in 2010, the AIIDE StarCraft AI Competition\footnote{\url{http://www.StarCraftAICompetition.com}} is the most
+well known and longest running StarCraft AI Competition in the world. Each year
+AI bots are submitted by competitors to do battle within the retail version of
+StarCraft: BroodWar, with prizes supplied by Blizzard Entertainment. This
+competition has been made possible by the BWAPI StarCraft programming interface,
+which allows players to write programs which can retrieve game data and issue
+game commands to StarCraft via an easy to use C++ API.
+
+The first competition in 2010 was organized and run by Ben Weber in the Expressive
+Intelligence Studio at University of California , Santa Cruz\footnote{\url{http://eis.ucsc.edu/StarCraftAICompetition}}. 26 total submissions
+were received from around the world. As this was the first year of the competition,
+and little infrastructure had been created, each game of the tournament was run 
+manually on two laptop computers and monitored by Ben to record the results. Also,
+no persistent data was kept for bots to learn about opponents between matches.
+
+The 2010 competition had 4 different tournament categories in which to compete. Tournament 1
+was a flat-terrain unit micromanagement battle consisting of four separate unit
+composition games. Of the six competitors, FreSCBot won the competition with
+Sherbrooke coming in 2nd place. Tournament 2 was another micro-focused game with
+non-trivial terrain. Two competitors submitted for this category, with FreSCBot
+once again coming in 1st by beating Sherbrooke.
+
+Tournament 3 was a tech-limited StarCraft game on a single known map with no fog of war enforced. Players
+were only allowed to choose the Protoss race, with no late game units allowed. 8 bots
+faced off in this double-elimination tournament with mimicbot taking first place over botnik in the final.
+As this was a perfect information variant of StarCraft, mimicbot adopted a strategy of
+"mimics its opponent's build order, gaining an economic advantage whenever possible" which
+worked quite well.
+
+Tournament 4 was the complete game of {\em StarCraft: BroodWar} with fog of war enforced. The tournament
+was run with a random pairing double-elimination format with each match being best of 5 games.
+Competitors could play as any of the three races, with the only limitations in gameplay
+being those which were considered 'cheating' in the StarCraft community. A map pool of
+5 well-known professional maps was announced to competitors in advance, with a random map being chosen for each game.
+
+Results are shown in Table \ref{tab:aiide2010}. The team that won was Overmind\footnote{\url{http://overmind.cs.berkeley.edu}}, from University of California, Berkeley. Using the Zerg race, their
+strategy was to defend early aggression with zergling units while amassing mutalisk units
+which they used to contain and eventually defeat their opponents. The mutalisk is a very fast and agile
+flying unit which is able to attack while moving with no drawback, which makes them quite a powerful
+unit when controlled by a computer, which has super-human input speeds. Overmind used a potential-field
+based micromanagement system to guide their mutalisks, which led them to victory. Krasi0 came in 2nd place
+with a standard defensive Terran opening strategy which transitioned into mech play in the late game.
+
+\begin{table}[bth]
+\caption{Results of the AIIDE 2010 competition.}
+\label{tab:aiide2010}
+\begin{small}
+\begin{center}
+% bracket A
+\begin{tabular}{|l|c|}
+\hline
+Position & bot\\ \hline
+1 & Overmind \\
+2 & Krasi0 \\
+3 & Chronos \\ \hline
+\end{tabular}
+\end{center}
+\end{small}
+\end{table}
+
+In 2011 the competition was hosted by the University of Alberta, with organization by Michael Buro and
+David Churchill\footnote{\url{https://skatgame.net/mburo/sc2011/}}. Due to a lack of entrants in tournament categories 1-3 in the 2010 competition, it was
+decided that only the full game category would be played in the 2011 competition. Another important
+change in the 2011 competition was the introduction of automated tournament-managing software running
+StarCraft games simultaneously on 20 computers, allowing a total of 1170 games to be played in the far less
+time than the 108 games of the 2010 competition. This increase in games played also allowed the tournament
+to switch to a round-robin format, eliminating the ``luck'' factor of the pairings inherent in bracket
+style tournaments. The bot which achieved the highest win percentage over the course of the competition would
+be determined the winner. Also decided was that the competition would become open-source, in an effort not
+only to prevent possible cheating, but to promote healthy competition in future tournaments by giving
+newcomers and easier entry point by basing their design off of previous bots.
+
+In the end, Skynet\footnote{\url{http://code.google.com/p/skynetbot/}} won the competition with its solid Protoss play (results are summarized in Table \ref{tab:aiide2011}).
+The bot executed one of a small set of
+strategies randomly at the start of the match based on the map and the race of the opponent. Skynet would
+then amass a medium to large sized army and expand before moving out to attack. Good use of of Dragoon range
+and kiting micromanagement allowed it to hold off the early aggression of other bots such as UAlbertaBot,
+which came in 2nd.
+
+UAlbertaBot used an early zealot-rush strategy to take advantage of the power of early game Protoss units. 
+It would send out the first zealots which were made and immediately attack the enemy base, using a unit
+counting heuristic to determine whether or retreat or keep pushing. Of note is that UAlbertaBot used an
+online planning algorithm to construct all of its economic build-orders\cite{churchill2011build}, as no hard-coded build orders
+were used.
+
+AIUR also chose Protoss, with a strategy which was in between Skynet and UAlbertaBot in terms of attack
+timings. Of note in these results was that a rock-paper-scissors effect happened among the top 3 finishers.
+Of the 30 rounds, Skynet beat UAlbertaBot 26 times, UAlbertaBot beat AIUR 29 times, and AIUR beat Skynet
+19 times. Another notable result is that Overmind did not choose to compete despite winning the 2010
+competition. After the competition, many bot programmers (including the Overmind team) realized that 
+their 2010 strategy was quite easily defeated by early game rushing strategies, and so they submitted
+a Terran bot instead, called Undermind, which finished in 7th.
+
+\begin{table}[bth]
+\caption{Results of the AIIDE 2011 competition.}
+\label{tab:aiide2011}
+\begin{small}
+\begin{center}
+% bracket A
+\begin{tabular}{|l|c|c|}
+\hline
+Position & bot & win \% \\ \hline
+1 & Skynet & 88.9\% \\
+2 & UAlbertaBot & 79.4\% \\
+3 & AIUR & 70.3\% \\
+4 & ItayUndermind & 65.8\% \\
+5 & EISBot & 60.6\% \\ \hline
+\end{tabular}
+\end{center}
+\end{small}
+\end{table}
+
+After the competition was over, a man vs. machine match was held between the winner (Skynet) and an
+ex-professional StarCraft player named Oriol Vinyals. Oriol was a competitor in the 2001 World Cyber
+Games StarCraft competition, and though he had been out of practice for a few years was still quite a
+good player. The match was arranged to see how well StarCraft AI bots had progressed and to see if
+they could actually beat a decent human opponent.
+
+For the best-of-three match, Oriol chose his races randomly and ended up beating Skynet in a 2-0. 
+In the first match, Oriol played Zerg vs. Skynet's Protoss on Python, a four player map. Oriol chose
+to start with a fast expansion strategy and transition into two base mutalisk production. Skynet chose
+to rush with a few early zealots, which was luckily the best possible choice given Oriol's strategy.
+Skynet's initial attack destroyed Oriol's early defenses, and nearly won the game in the first few
+minutes, however it then proceeded to send zealots to attack one at a time rather than group up its
+units before moving in, which allowed Oriol to catch up. Once Oriol produced his mutalisks, Skynet did
+not produce sufficient air defenses and Oriol quickly destroyed Skynet's base. In the second game,
+Oril played Terran, again on Python. After holding off early Dragoon pressure from Skynet, Oriol
+moved out with a few marines, medics and tanks. Skynet tried to defend with its army of Dragoons,
+however due to poor unit targeting decisions it started to attack useless medics after the marines
+had died, rather than the tanks. Oriol overcame the Dragoon army and was victorious. Later analysis
+of the match concluded that Skynet, while dominant over the other bots, was unable to properly
+adapt and transition into a mid-game strategy in game one once its early pressure failed, and in game
+two made a key blunder in unit targeting which cost it the game. The humans were still in command.
+
+The 2012 competition was also hosted by the University of Alberta, with the major difference from the 2011
+competition being the addition of persistent storage. Bots could now write information to disk during a
+match, and then read the information during other matches, allowing them to adjust strategies based
+on previous results. 6 of the 10 entrants used this feature to aid in strategy selection, including the
+top 4 finishers. More improvements to the tournament environment also meant that a total of 4240 games
+could now be played in the same time period. Results are shown in Table \ref{tab:aiide2012}.
+
+Skynet once again won the competition with its solid Protoss build orders and good Dragoon kiting. AIUR
+and UAlbertaBot switched positions from the previous year to come 2nd and 3rd respectively. Both AIUR
+and UAlbertaBot used data stored from the results of previous games to select a strategy for future matches.
+UAlbertaBot did this using the UCB algorithm, while AIUR used a custom algorithm to achieve similar
+results. The effect of this strategy selection process can be seen Figure \ref{fig:aiide2012} which shows bot
+win percentages over time. While the earlier rounds of the tournament fluctuated wildly in results,
+eventually the results converged to their final values. One of the main reasons for this is due to
+the bots learning which strategies to use as the tournament progressed.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\columnwidth]{figures/aiide2012.png}
+    \caption{Evolution of the win percentage of each bot participating in the AIIDE 2012 competition}
+    \label{fig:aiide2012}
+\end{figure}
+
+\begin{table}[bth]
+\caption{Results of the AIIDE 2012 competition.}
+\label{tab:aiide2012}
+\begin{small}
+\begin{center}
+% bracket A
+\begin{tabular}{|l|c|c|}
+\hline
+Position & bot & win \% \\ \hline
+1 & Skynet & 84.4\% \\
+2 & AIUR & 72.2\% \\
+3 & UAlbertaBot & 68.6\% \\
+4 & BroodwarBotQ & 59.1\% \\
+5 & AdjutantBot & 52.8\% \\ \hline
+\end{tabular}
+\end{center}
+\end{small}
+\end{table}
+
+The 2012 man vs. machine match again used the winner of the competition (Skynet), who played against
+Mike Lange, aka LRM) Bakuryu. At the time of the match, Bakuryu was an A- ranked Zerg player on ICCup,
+and known as one of the best non-Korean Zerg players in the world. Bakuryu was considered much
+stronger than Oriol at the time that the match was played, and the results showed that this was true.
+In the first game of the best-of-three, Bakuryu made Skynet look quite silly by running around inside
+Skynet's base with a small number of zerglings while Skynet's zealots and half of its worked chased
+then in vain. After killing off several probes and buying enough time to set up his expansion, he
+cleaned up Skynet's army with a flying army of mutalisks. In the second game, Bakuryu contained Skynet
+inside its base with a group of zerglings positioned within Skynet's expansion. Skynet then constructed
+several Dark Templar and along with some Dragoons and Zealots attacked into Bakuryu's expansion which
+was heavily defended, and was crushed almost instantly, allowing Bakuryu's zergling force to finish
+off the Protoss base.
+
+In this match it was shown that the true weakness of state of the art StarCraft AI systems was that
+humans are very adept at recognizing scripted behaviors and exploiting them to the fullest. A human
+player in Skynet's position in the first game would have realized he was being taken advantage of and
+adapted his strategy accordingly, however the inability to put the local context (Bakuryu kiting his
+units around his base) into the larger context of the game (that this would delay Skynet until
+reinforcements arrived) and then the lack of strategy change to fix the situation led to an easy
+victory for the human. These problems remain the main challenges in RTS AI today: to both recognize
+the strategy and intent of an opponent's actions, and how to effectively adapt your own strategy to 
+overcome them.
+
+All results, videos, and replays from the AIIDE StarCraft AI Competition can be found in\footnote{\url{http://www.StarCraftAICompetition.com}}
 
@@
 
-}
-
 \section{Open Questions in RTS Game AI}\label{sec:questions}

### 2013-02-14 survey.tex
Fix some typos.

@@
 While performing all of those tasks, players also need to strategically understand the geometry of the map at hand, in order to decide where to place new buildings (concentrate in a single area, or expand to different areas) or where to set defensive outposts.
-Finally, when offensive units of two players meet, each player must quickly maneuver each of the units in order to fight the a battle, which requires quick and reactive control of each of the units in order to successfully fight a battle.
+Finally, when offensive units of two players meet, each player must quickly maneuver each of the units in order to fight a battle, which requires quick and reactive control of each of the units in order to successfully fight the battle.
 
@@
 The idea behind FSMs is to decompose the AI behavior into easily manageable states, such as ``attacking'', ``gathering resources'' or ``repairing'' and establish the conditions that trigger transitions between them.
-Commercial approaches also include Hierarchical FSMs, in which FSMS are composed hierarchically.
+Commercial approaches also include Hierarchical FSMs, in which FSMs are composed hierarchically.
 These hard-coded approaches have achieved a significant amount of success, and, as we will discuss later, have also been used in many academic RTS AI research systems, as discussed in Section \ref{sec:bot}.
@@
 Its high-level module determines the strategy to use, and the tactical decision module divides it into a collection of {\em abstract actions}, that are executed by the lower-level modules.
-\item AIUR: is mainly divide-and-conquer oriented, for macro as well as for micro, with a slight abstraction on macro with a SpendManager deciding how to spend resources among Base, Production and Construction Managers.
+\item AIUR:  is mainly divide-and-conquer oriented, for  macro as well
+  as  for  micro,  with  a  slight  abstraction  on  macro  due  to  a
+  SpendManager deciding how to spend and share resources among Base, Production and Construction Managers.
 At the beginning of a game, the MoodManager initializes a ``mood'' which will influence both micro and macro.
@@
 
-One interesting aspect of the five bots described above is that, while all of them are reactive at the lower level (reactive control), most if not all of them, are scripted at the highest level of abstraction.
+One interesting aspect of the seven bots described above is that, while all of them are reactive at the lower level (reactive control), most if not all of them, are scripted at the highest level of abstraction.
 BTHAI reads build and squad formations from a predefined script, Nova's {\em Strategy Manager} is a predefined finite-state machine, BBQ's construction manager reads the build order from a predefined script, and Skynet's {\em BuildOrder Manager} is basically a predefined script.
@@
 strategies randomly at the start of the match based on the map and the race of the opponent. Skynet would
-then amass a medium to large sized army and expand before moving out to attack. Good use of of Dragoon range
+then amass a medium to large sized army and expand before moving out to attack. Good use of Dragoon range
 and kiting micromanagement allowed it to hold off the early aggression of other bots such as UAlbertaBot,
@@
 a lack of available software (the AIIDE program was not yet
-available at that time), we separated the ten entries into
+available at that time), the organizers separated the ten entries into
 two brackets. In each bracket of 5 bots, a round-robin

### 2013-02-14 survey.tex
Add iCCup ranking in intro, and start II.B.3 spatial and temporal reasoning

@@
 
-{\color{red} A typical Starcraft map is defined as a rectangular grid, typically ranging between 64x64 to 128x128 in size (although the resolution at which units move is finer), and each player can control up to 200 units (not including buildings).
-Moreover, each different race contains between 30 to 35 different types of units and buildings, each one with a significant number of special abilities}.
-All these factors together make Starcraft a significant challenge, in which humans are still much better than computers.
+{\color{red} A typical Starcraft map is defined as a rectangular grid,
+  typically  ranging between 64x64  to 128x128  in size  (although the
+  resolution  at which  units  move  is finer),  and  each player  can
+  control up  to 200 units (not including  buildings).  Moreover, each
+  different race  contains between 30  to 35 different types  of units
+  and  buildings,  each  one  with  a significant  number  of  special
+  abilities}. All these factors  together make Starcraft a significant
+challenge,   in   which   humans    are   still   much   better   than
+computers. Indeed, best Starcraft AIs are ranked between $D$ and $D^-$
+on the iCCup\footnote{http://www.iccup.com/starcraft/} ranking system,
+where average  amateur players are  ranked between $C^+$ and  $B$. For
+comparison, Starcraft professional  players are usually ranked between
+$A^-$ and $A^+$.
 
@@
 \subsubsection{Spatial and Temporal Reasoning}
-TODO just a paragraph on exposing why there are temporal and spatial reasoning in RTS AI.
+% TODO just a paragraph on exposing why there are temporal and spatial
+% reasoning in RTS AI.
+Spatial  reasoning   is  related  to   each  aspect  of   the  terrain
+exploitation.  This  involves  buildings  placing  where  one  has  to
+carefully  consider buildings  position  into its  own  bases to  both
+protect them  by creating  a wall against  invasions and to  avoid bad
+configurations  where large  units could  be stuck.  Base expansion
+also belongs  to spatial reasoning,  in the sense  that the AI  has to
+choose good available emplacements to  build a new base, regarding its
+own  position  and opponent's  bases.  Finally,  spatial reasoning  is
+closely related to  tactics with questions like ``Where  should I place
+my  units for battle?'',  favoring for  instance engagements  when the
+opponent's units are into a  bottleneck. In Starcraft, it is always an
+advantage to  have its units  on high ground  whereas enemy is  on low
+ground, since units  on low ground have no vision  on high ground, and
+firing from low  ground have only 70\% of chance to  hit units on high
+ground.
+
+As well as spatial reasoning, temporal reasoning is related to tactics
+since it implies  to decide about army movements,  like when attacking
+or withdrawing  units from battle.  Strategy also copes  with temporal
+reasoning, mainly  with planning problems:  when researching upgrades,
+when constructing wished buildings,  when switching strategy if needed,
+etc.
 
@@
 \subsubsection{Task Decomposition}
-TODO
-}
+TODO }
 

### 2013-02-14 survey.tex
Start II.B.5 task decomposition

@@
 \subsubsection{Task Decomposition}
-TODO }
+Playing a  RTS game like Starcraft  can be decomposed  into three main
+tasks: 
+\begin{itemize}
+\item Strategy:  corresponds to the  decision making process.  This is
+  the  highest level of  abstraction for  the game  comprehension. RTS
+  games   are   all   about   finding   an   efficient   strategy   or
+  counter-strategy  against a  given opponent.  It concerns  the whole
+  set of  units and buildings  one owns. If  we may allow  ourselves a
+  military analogy, strategy is orders given by generals at the HQ. 
+\item  Tactics: are  the implementation  of the  current  strategy. It
+  implies  army and  building positioning,  movements, timing,  and so
+  on.  Tactics concerns a  group of  units. They  are orders  given by
+  officers on the battlefield.
+\item  Reactive  control:  is  the  implementation  of  tactics.  This
+  consists in   moving,    targeting,   firing,    fleeing   during
+  battle. Reactive control focuses on  a specific unit. It is soldiers
+  applying officers' orders.
+\end{itemize}
+
+\noindent
+Both     strategy      and     tactics     are      usually     called
+\textit{macro-management},  whereas reactive  control  is also  called
+\textit{micro-management}. In addition of the three tasks shown above,
+one could also consider these two tasks that an AI has to deal with:
+\begin{itemize}
+\item Terrain analysis: consists  in the analysis of regions composing
+  the  map,  chokes,  minerals  and  gas emplacements,  low  and  high
+  walkable grounds, islands, etc.
+\item Intelligence gathering:  corresponds to information collected on
+  our  opponent. Because of  the fog  of war,  a non-cheating  AI must
+  regulary send scouts to localize and spy enemy bases, and must see or
+  infer what buildings and technology the enemy has developed in order
+  to adapt its strategy.
+\end{itemize}
+
+\noindent
+The reader can find a good presentation of task decomposition for AIs 
+playing RTS in~\cite{weber2011acs}.  }
 

### 2013-02-15 survey.tex
Details about AIUR, and url for each bot.

@@
 since it implies  to decide about army movements,  like when attacking
-or withdrawing  units from battle.  Strategy also copes  with temporal
-reasoning, mainly  with planning problems:  when researching upgrades,
-when constructing wished buildings,  when switching strategy if needed,
-etc.
+or  withdrawing  units from  battle.   A  part  of temporal  reasoning
+consists  of  threat predictions,  where  one  has  to plan  when  the
+opponent is  likely running an  attack.  Strategy copes  with temporal
+reasoning  as well,  mainly with  planning problems:  when researching
+upgrades, when constructing  wished buildings, when switching strategy
+if needed, etc.
 
@@
 \begin{itemize}
-\item BroodwarBotQ: uses abstraction for micro-management, and divide-and-conquer for macro-management and intelligence gathering.
-To avoid conflicts between modules (since the individual tasks of each of the modules are not completely independent), BBQ uses an arbitrator.
-\item Nova: is similar in design as BroodwarBotQ, and uses abstraction for micro-management, and divide-and-conquer for macro-management.
-The differences are that Nova does not have an arbitrator to resolve conflicts, but has a higher-level module ({\em strategy manager}), which posts information to the blackboard that the rest of modules follow (thus, making use of abstraction).
-\item UAlbertaBot: also uses abstraction in micro-management like the previous two bots.
-But it also uses it in macro-management: as can be seen, the production manager sends commands to the building manager, who is in charge of producing the buildings.
-This bot also uses divide-and-conquer, and tasks like scouting and resource gathering are managed by separate, independent modules.
-\item Skynet: makes extensive use of both abstraction and divide-and-conquer.
-Considering the decision making component of Skynet, we can see a high level module that issues commands to a series of tactics modules.
-The collection of tactic modules queue {\em tasks} (that are analogous to the abstract actions used in SPAR).
-Each different task has a specific low level module that knows how to execute it.
-Thus, Skynet uses a 3 layered abstraction hierarchy, and uses divide-and-conquer in all levels except the highest.
-\item SPAR: only uses abstraction.
-Its high-level module determines the strategy to use, and the tactical decision module divides it into a collection of {\em abstract actions}, that are executed by the lower-level modules.
-\item AIUR:  is mainly divide-and-conquer oriented, for  macro as well
-  as  for  micro,  with  a  slight  abstraction  on  macro  due  to  a
-  SpendManager deciding how to spend and share resources among Base, Production and Construction Managers.
-At the beginning of a game, the MoodManager initializes a ``mood'' which will influence both micro and macro.
-For example, micro-management is divided into three independent managers: the {\em Defense Manager}, controlling military units when there is nothing special, the {\em Under Attack Manager}, activated when the opponent is attacking our bases, and the {\em Army Manager}, taking control of units when it is time to attack, following a timing given by the current mood.
-\item BTHAI: uses a two-tier abstraction hierarchy, where a collection of high-level modules command a collection of lower-level agents in charge of each of the units.
-At the high-level, BTHAI uses divide-and-conquer, having multiple high-level modules issuing commands to the lower-level units.
+\item
+  BroodwarBotQ\footnote{\url{http://github.com/SnippyHolloW/BroodwarBotQ}}:
+  uses  abstraction for  micro-management, and  divide-and-conquer for
+  macro-management  and  intelligence  gathering. To  avoid  conflicts
+  between modules (since  the individual tasks of each  of the modules
+  are not completely independent), BBQ uses an arbitrator.
+\item  Nova\footnote{\url{http://nova.wolfwork.com/}}:  is similar  in
+  design as  BroodwarBotQ, and uses  abstraction for micro-management,
+  and  divide-and-conquer for  macro-management.  The differences  are
+  that Nova does not have  an arbitrator to resolve conflicts, but has
+  a  higher-level   module  ({\em  strategy   manager}),  which  posts
+  information to the blackboard that the rest of modules follow (thus,
+  making use of abstraction).
+\item
+  UAlbertaBot\footnote{\url{http://code.google.com/p/ualbertabot/}}:
+  also  uses abstraction  in  micro-management like  the previous  two
+  bots. But it  also uses it in macro-management: as  can be seen, the
+  production manager sends commands to the building manager, who is in
+  charge   of   producing   the   buildings.  This   bot   also   uses
+  divide-and-conquer, and  tasks like scouting  and resource gathering
+  are managed by separate, independent modules.
+\item       Skynet\footnote{\url{http://code.google.com/p/skynetbot/}}:
+  makes     extensive     use      of     both     abstraction     and
+  divide-and-conquer.  Considering the  decision  making component  of
+  Skynet, we  can see a  high level module  that issues commands  to a
+  series of  tactics modules. The  collection of tactic  modules queue
+  {\em  tasks} (that  are analogous  to the  abstract actions  used in
+  SPAR).  Each different  task has  a specific  low level  module that
+  knows how to  execute it. Thus, Skynet uses  a 3 layered abstraction
+  hierarchy,  and uses  divide-and-conquer  in all  levels except  the
+  highest.
+\item
+  SPAR\footnote{\url{http://www.planiart.usherbrooke.ca/projects/spar/}}:
+  only uses abstraction. Its high-level module determines the strategy
+  to  use,  and  the  tactical  decision  module  divides  it  into  a
+  collection  of {\em  abstract  actions}, that  are  executed by  the
+  lower-level modules.
+\item  AIUR\footnote{\url{http://code.google.com/p/aiurproject/}}:  is
+  mainly divide-and-conquer oriented, for  macro as well as for micro,
+  with a  slight abstraction on  macro due to a  SpendManager deciding
+  how  to  spend  and  share  resources  among  Base,  Production  and
+  Construction Managers.  At the beginning of a  game, the MoodManager
+  initializes  a  ``mood''  which  will  influence  both  tactics  and
+  strategy. Tactics  are divided into three  independent managers: the
+  {\em  Defense Manager},  controlling  military units  when there  is
+  nothing special, the {\em  Under Attack Manager}, activated when the
+  opponent is attacking our bases,  and the {\em Army Manager}, taking
+  control of units when it is time to attack, following a timing given
+  by the current mood.
+\item  BTHAI\footnote{\url{http://code.google.com/p/bthai/}}:  uses  a
+  two-tier  abstraction hierarchy,  where a  collection  of high-level
+  modules command a collection of lower-level agents in charge of each
+  of  the units.  At  the high-level,  BTHAI uses  divide-and-conquer,
+  having   multiple  high-level  modules   issuing  commands   to  the
+  lower-level units.
 \end{itemize}
@@
 
-Some bots using divide-and-conquer, assume that each of the modules can act independently and that their actions can be executed without interference.
-BBQ, UAlbertaBot and AIUR, however use an arbitrator ({\em Game Commander'} in UAlbertaBot) that makes sure that modules do not send contradictory orders to the same unit.
-However, very little bots handle the problem of how to coordinate resource usage amongst modules, for instance BTHAI uses a first-come-first-serve policy for spending resources, the first module that requests resources is the one that gets them.
-Nova and Skynet are exceptions, and implement some rudimentary prioritization based on the high level strategy.
-Following available resources and timing, AIUR's {\em Spend Manager} orders Base, Production and Construction Managers what they have to build/produce.
-It also orders to start tech research and upgrades.
-
-One interesting aspect of the seven bots described above is that, while all of them are reactive at the lower level (reactive control), most if not all of them, are scripted at the highest level of abstraction.
-BTHAI reads build and squad formations from a predefined script, Nova's {\em Strategy Manager} is a predefined finite-state machine, BBQ's construction manager reads the build order from a predefined script, and Skynet's {\em BuildOrder Manager} is basically a predefined script.
-Such scripts describe the strategy that the bots will use, however, such strategy is always fixed.
-One could see this pre-scripting as if each bot defined a ``high-level programming language'' to describe Starcraft strategies, and the bots themselves are just interpreters of such strategy.
-Compared to current approaches for Chess or Go, this scripting seems a rigid and inflexible, but responds to the much higher complexity of the Starcraft game.
-An interesting exception to that is UAlbertaBot, which uses a search algorithm in the {\em Production Manager} to find near-optimal build orders.
-Another interesting case is AIUR, that uses a {\em Mood Manager} to randomly pick a mood among five (rush, aggressive, defensive, macro, fast expand), which will influence the build order, strategy and tactics, all of them scripted so far.
-This mood can change during the game, following the opponent behavior.
+Some bots  using divide-and-conquer, assume  that each of  the modules
+can act independently  and that their actions can  be executed without
+interference.  BBQ,  UAlbertaBot and  AIUR, however use  an arbitrator
+({\em Game Commander'} in UAlbertaBot) that makes sure that modules do
+not send contradictory orders to  the same unit.  However, very little
+bots handle  the problem of  how to coordinate resource  usage amongst
+modules, for  instance BTHAI uses a  first-come-first-serve policy for
+spending resources,  the first module  that requests resources  is the
+one that gets them. Nova and Skynet are exceptions, and implement some
+rudimentary prioritization based on the high level strategy. Following
+available  resources and  timing,  AIUR's {\em  Spend Manager}  orders
+Base,  Production   and  Construction  Managers  what   they  have  to
+build/produce. It also orders to start tech research and upgrades. The
+idea here is  not to allocation resources among  different managers to
+satisfy a need (informally: ``I want  to do that, so I need M minerals
+and G gas''), but to do the  opposite, that is, to find how the AI can
+spend the available money: ``I have right now M minerals and G gas, so
+how can I spend it?''.
+
+One  interesting aspect  of the  seven bots  described above  is that,
+while all of them are  reactive at the lower level (reactive control),
+most  if  not all  of  them,  are scripted  at  the  highest level  of
+abstraction. BTHAI reads build  and squad formations from a predefined
+script,  Nova's {\em  Strategy Manager}  is a  predefined finite-state
+machine,  BBQ's construction  manager  reads the  build  order from  a
+predefined script, and Skynet's  {\em BuildOrder Manager} is basically
+a predefined script. Such scripts  describe the strategy that the bots
+will use, however,  such strategy is always fixed.  One could see this
+pre-scripting  as  if  each  bot defined  a  ``high-level  programming
+language'' to  describe Starcraft strategies, and  the bots themselves
+are just interpreters of such strategy. Compared to current approaches
+for  Chess or Go,  this scripting  seems a  rigid and  inflexible, but
+responds  to the  much higher  complexity  of the  Starcraft game.  An
+interesting  exception to  that is  UAlbertaBot, which  uses  a search
+algorithm in  the {\em Production Manager} to  find near-optimal build
+orders.  Another interesting  case  is  AIUR, that  uses  a {\em  Mood
+  Manager}  to   randomly  pick  a  mood  among   six  (cheese,  rush,
+aggressive, defensive,  macro, fast expand), which  will influence the
+build order,  strategy and tactics. All  of them scripted  so far, but
+build  orders  can been  modified  but  the  Spend Manager,  following
+available resources.  This mood can change during  the game, regarding
+the opponent behavior.
 
@@
 
-\begin{table}[bth]
+\begin{table}[t!]
 \caption{Results of the AIIDE 2010 competition.}
@@
 
-In the end, Skynet\footnote{\url{http://code.google.com/p/skynetbot/}} won the competition with its solid Protoss play (results are summarized in Table \ref{tab:aiide2011}).
-The bot executed one of a small set of
-strategies randomly at the start of the match based on the map and the race of the opponent. Skynet would
-then amass a medium to large sized army and expand before moving out to attack. Good use of Dragoon range
-and kiting micromanagement allowed it to hold off the early aggression of other bots such as UAlbertaBot,
+In the  end, Skynet  won the competition  with its solid  Protoss play
+(results  are  summarized   in  Table  \ref{tab:aiide2011}).  The  bot
+executed one of a small set of strategies randomly at the start of the
+match based on the map and the race of the opponent. Skynet would then
+amass a  medium to large  sized army and  expand before moving  out to
+attack. Good  use of Dragoon range and  kiting micromanagement allowed
+it to hold off the early aggression of other bots such as UAlbertaBot,
 which came in 2nd.
@@
 
-AIUR also chose Protoss, with a strategy which was in between Skynet and UAlbertaBot in terms of attack
-timings. Of note in these results was that a rock-paper-scissors effect happened among the top 3 finishers.
-Of the 30 rounds, Skynet beat UAlbertaBot 26 times, UAlbertaBot beat AIUR 29 times, and AIUR beat Skynet
-19 times. Another notable result is that Overmind did not choose to compete despite winning the 2010
-competition. After the competition, many bot programmers (including the Overmind team) realized that 
-their 2010 strategy was quite easily defeated by early game rushing strategies, and so they submitted
-a Terran bot instead, called Undermind, which finished in 7th.
-
-\begin{table}[bth]
+AIUR also chose  Protoss, with a strategy which  was in between Skynet
+and UAlbertaBot in terms of  attack timings.  At that time, AIUR chose
+one  mood among  five (leading  to slightly  different  strategies and
+tactics) at the  beginning of a game and kept it  until the end. These
+five moods were:
+\begin{itemize}
+\item  Rush:  where  the  bot  tries  early  attacks,  and  have  good
+  probabilities to send  the two or three first  Zealots to harass the
+  opponent.
+\item Aggressive: where  we have less chance to  perform harasses with
+  the first  Zealots, and  the first attack  is usually a  bit delayed
+  with regard to the Rush mood.
+\item Macro: where the AI do not try any early attacks and focus a bit
+  more on its economy before attacking.
+\item Defense: where the AI  ``turtles'' and wait to have a consequent
+  army before running an attack.
+\item  Fast expand:  where the  first building  constructed it  a base
+  expansion, for a very economical-oriented game.
+\end{itemize}
+Notice that  build orders are not  fully hard-coded since  they can be
+altered by AIUR's Spend Manager.
+
+Of  note  in  these  results  was that  a  rock-paper-scissors  effect
+happened among  the top  3 finishers.  Of  the 30 rounds,  Skynet beat
+UAlbertaBot 26  times, UAlbertaBot beat  AIUR 29 times, and  AIUR beat
+Skynet  19 times.   Another notable  result is  that Overmind  did not
+choose  to compete despite  winning the  2010 competition.   After the
+competition,  many  bot  programmers  (including  the  Overmind  team)
+realized that their  2010 strategy was quite easily  defeated by early
+game rushing strategies,  and so they submitted a  Terran bot instead,
+called Undermind, which finished in 7th.
+
+\begin{table}[t!]
 \caption{Results of the AIIDE 2011 competition.}
@@
 
-Skynet once again won the competition with its solid Protoss build orders and good Dragoon kiting. AIUR
-and UAlbertaBot switched positions from the previous year to come 2nd and 3rd respectively. Both AIUR
-and UAlbertaBot used data stored from the results of previous games to select a strategy for future matches.
-UAlbertaBot did this using the UCB algorithm, while AIUR used a custom algorithm to achieve similar
-results. The effect of this strategy selection process can be seen Figure \ref{fig:aiide2012} which shows bot
-win percentages over time. While the earlier rounds of the tournament fluctuated wildly in results,
-eventually the results converged to their final values. One of the main reasons for this is due to
-the bots learning which strategies to use as the tournament progressed.
-
-\begin{figure}
+Skynet once  again won  the competition with  its solid  Protoss build
+orders  and  good  Dragoon  kiting.   AIUR  and  UAlbertaBot  switched
+positions from  the previous  year to come  2nd and  3rd respectively.
+Both  AIUR  and UAlbertaBot  used  data  stored  from the  results  of
+previous games  to select a strategy for  future matches.  UAlbertaBot
+did  this  using  the  UCB   algorithm,  while  AIUR  used  a  uniform
+distribution  to choose  its  mood before  altering this  distribution
+after  some  games  against  the  same  opponent  to  favor  efficient
+strategies, achieving  similar results than  UAlbertaBot. Notice that,
+compared  to AIIDE 2011,  AIUR proposes  a new  mood, \textit{Cheese},
+implementing a  Photon Cannon rush  strategy in order to  surprise the
+opponent and  to finish the game  as soon as possible.   The effect of
+this strategy selection process can be seen Figure \ref{fig:aiide2012}
+which shows bot  win percentages over time.  While  the earlier rounds
+of the tournament fluctuated wildly in results, eventually the results
+converged to their  final values. One of the main  reasons for this is
+due to  the bots  learning which strategies  to use as  the tournament
+progressed.
+
+\begin{figure}[t!]
     \centering
-    \includegraphics[width=\columnwidth]{figures/aiide2012.png}
+    \includegraphics[width=\columnwidth]{figures/aiide2012_v2}
     \caption{Evolution of the win percentage of each bot participating in the AIIDE 2012 competition}
@@
 
-\begin{table}[bth]
+\begin{table}[t!]
 \caption{Results of the AIIDE 2012 competition.}
@@
 
-\begin{table*}[htb]
+\begin{table*}[t!]
 \caption{Results of the first round at CIG 2011, held in two brackets.
@@
 
-\begin{table}[bth]
+\begin{table}[t!]
 \caption{Results of the CIG 2011 final round.}

### 2013-02-15 survey.tex
Minor mod about reactive control (in II.B.5 and AIUR's description).

@@
 \item  Reactive  control:  is  the  implementation  of  tactics.  This
-  consists in   moving,    targeting,   firing,    fleeing   during
-  battle. Reactive control focuses on  a specific unit. It is soldiers
-  applying officers' orders.
+  consists   in  moving,   targeting,  firing,   fleeing,  hit-and-run
+  techniques (aka ``kiting'')  during battle. Reactive control focuses
+  on a specific unit. It is soldiers applying officers' orders. 
 \end{itemize}
@@
   control of units when it is time to attack, following a timing given
-  by the current mood.
+  by the  current mood. This bot  does not manage however  any kind of
+  reactive controls so far.
 \item  BTHAI\footnote{\url{http://code.google.com/p/bthai/}}:  uses  a
@@
 One  interesting aspect  of the  seven bots  described above  is that,
-while all of them are  reactive at the lower level (reactive control),
-most  if  not all  of  them,  are scripted  at  the  highest level  of
-abstraction. BTHAI reads build  and squad formations from a predefined
-script,  Nova's {\em  Strategy Manager}  is a  predefined finite-state
-machine,  BBQ's construction  manager  reads the  build  order from  a
-predefined script, and Skynet's  {\em BuildOrder Manager} is basically
-a predefined script. Such scripts  describe the strategy that the bots
-will use, however,  such strategy is always fixed.  One could see this
-pre-scripting  as  if  each  bot defined  a  ``high-level  programming
-language'' to  describe Starcraft strategies, and  the bots themselves
-are just interpreters of such strategy. Compared to current approaches
-for  Chess or Go,  this scripting  seems a  rigid and  inflexible, but
-responds  to the  much higher  complexity  of the  Starcraft game.  An
-interesting  exception to  that is  UAlbertaBot, which  uses  a search
-algorithm in  the {\em Production Manager} to  find near-optimal build
-orders.  Another interesting  case  is  AIUR, that  uses  a {\em  Mood
-  Manager}  to   randomly  pick  a  mood  among   six  (cheese,  rush,
-aggressive, defensive,  macro, fast expand), which  will influence the
-build order,  strategy and tactics. All  of them scripted  so far, but
-build  orders  can been  modified  but  the  Spend Manager,  following
-available resources.  This mood can change during  the game, regarding
-the opponent behavior.
+while all  of them  - except AIUR  - are  reactive at the  lower level
+(reactive  control), most  if not  all of  them, are  scripted  at the
+highest level  of abstraction. BTHAI reads build  and squad formations
+from  a  predefined  script,   Nova's  {\em  Strategy  Manager}  is  a
+predefined finite-state machine,  BBQ's construction manager reads the
+build  order from a  predefined script,  and Skynet's  {\em BuildOrder
+  Manager} is basically a predefined script. Such scripts describe the
+strategy  that the  bots will  use, however,  such strategy  is always
+fixed.   One could see  this pre-scripting  as if  each bot  defined a
+``high-level programming language''  to describe Starcraft strategies,
+and   the   bots   themselves    are   just   interpreters   of   such
+strategy.  Compared  to  current  approaches  for Chess  or  Go,  this
+scripting  seems a  rigid and  inflexible,  but responds  to the  much
+higher complexity of the  Starcraft game.  An interesting exception to
+that  is  UAlbertaBot, which  uses  a  search  algorithm in  the  {\em
+  Production  Manager}  to find  near-optimal  build orders.   Another
+interesting case is  AIUR, that uses a {\em  Mood Manager} to randomly
+pick  a mood among  six (cheese,  rush, aggressive,  defensive, macro,
+fast  expand), which  will  influence the  build  order, strategy  and
+tactics.  All of  them  scripted so  far,  but build  orders can  been
+modified but  the Spend Manager, following  available resources.  This
+mood can change during the game, regarding the opponent behavior.
 

### 2013-02-20 survey.tex
Did a complete pass over Section 2, now we have a complete first draft on that.

@@
 
-Real-time Strategy (RTS) is a sub-genre of strategy games where players need to build an economy (gathering resources and building a base) and a military power (training units and researching technologies) in order to defeat their opponents (destroying their army and base).
+Real-time Strategy (RTS) is a sub-genre of strategy games where players need to build an economy (gathering resources and building a base) and military power (training units and researching technologies) in order to defeat their opponents (destroying their army and base).
 From a theoretical point of view, the main differences between RTS games and traditional board games such as Chess are:
@@
 
-For those reasons, standard techniques used for playing classic board games, such as minimax game tree search, cannot be directly applied to solve RTS games without the definition of some level of abstraction, or some other simplification.
+For those reasons, standard techniques used for playing classic board games, such as alpha-beta search, cannot be directly applied to solve RTS games without the definition of some level of abstraction, or some other simplification.
 Interestingly enough, humans seem to be able to deal with the complexity of RTS games, and are still vastly superior to computers in these type of games \cite{burochurchill2012aimagazine}.
 For those reasons, a large spectrum of techniques have been attempted to deal with this domain, as we will describe below.
-The remainder of this section is devoted to describe Starcraft as a research testbed, and on detailing which are the open challenges in RTS game AI.
+The remainder of this section is devoted to describe Starcraft as a research testbed, and on detailing the open challenges in RTS game AI.
 
@@
 
-Starcraft: Brood War is a popular RTS game released in 1998 by Blizzard Entertainment.
+{\em Starcraft: Brood War} is an immensely popular RTS game released in 1998 by Blizzard Entertainment.
 Starcraft is set in a science-fiction based world where the player must choose one of the three races: Terran, Protoss or Zerg.
@@
 As resources become available, players need to allocate them for creating more buildings (which reinforce the economy, and allow players to create units or unlock stronger units) and train attack units.
-Units must be distributed to accomplish different tasks such as: reconnaissance, defense and attack.
+Units must be distributed to accomplish different tasks such as reconnaissance, defense and attack.
 While performing all of those tasks, players also need to strategically understand the geometry of the map at hand, in order to decide where to place new buildings (concentrate in a single area, or expand to different areas) or where to set defensive outposts.
@@
 
-{\color{red} A typical Starcraft map is defined as a rectangular grid,
+A typical Starcraft map is defined as a rectangular grid,
 typically  ranging between 64x64  to 128x128  in size  (although the
@@
 and  buildings,  each  one  with  a significant  number  of  special
-  abilities}. All these factors  together make Starcraft a significant
+abilities. All these factors  together make Starcraft a significant
 challenge,   in   which   humans    are   still   much   better   than
@@
 
-Many years have passed since Buro's call for research, where he identified the following 6 challenges:
+Early research in AI for RTS games \cite{Buro03rts} identified the following sic challenges: 
 \begin{itemize}
@@
 \item Spatial and temporal reasoning
-\item Collaboration
-\item Opponent modeling, Learning
+\item Collaboration (between multiple AIs)
+\item Opponent modeling and learning
 \item Adversarial real-time planning
@@
 collaboration).
-Additionally, after the increase attention to the problem of RTS games, further, unforeseen challenges appeared.
-For example, how to exploit the massive amounts of existing domain knowledge (strategies, build-orders, replays, and so on), or how to design an architecture that integrates all the modules required for an agent to play an RTS games.
-In the following subsections, we describe current challenges in RTS Game AI, grouped in different areas.
+Moreover, recent research in this area has identified several additional research challenges, such as how to exploit the massive amounts of existing domain knowledge (strategies, build-orders, replays, and so on).
+Below, we describe current challenges in RTS Game AI, grouped in six main different areas.
 
@@
 \subsubsection{Planning}
-{\color{blue}
-TODO (Santi?) just a paragraph on exposing why there is a planning problem in RTS AI (macro and micro).
-
-\paragraph{Learning}
-When talking about learning in RTS games, with regard to a particular match against an opponent (where a match is a sequence of games against the same opponent), one has to differentiate between:
-\begin{itemize}
-\item Prior learning: done before the match, like when a bot learns what is likely to happen on a given match-up (and map) from replays, or when a bot optimizes its parameters beforehand through simulated annealing or genetic evolution.
-\item In-game learning: done between the different actions (decisions) taken during a game.
-For instance, a bot can take into account the successes and failures of its attacks (location, type, timings).
-\item In-match learning: done between the different games of the match, it encompasses opponent modeling across multiple games and adapting the bots' strategy to its successes and failures.
-\end{itemize}
-Learning in all these situations can be supervised (labeled replays, scripted situations) or unsupervised (unlabeled data, exploratory situations).
-The last two kinds of situations of learning are more inclined towards reinforcement learning because the AI has to deal with the exploration-exploitation duality.
-%prior learning: supervised \cite{WeberCIG09} \cite{SynnaeveCIG11}, unsupervised: \cite{SynnaeveAIIDE11}
+As mentioned above, the size of the state space in RTS games is vastly larger than that of traditional board-games such as Chess or Go.
+Additionally, the number of actions that can be executed at given instant of time is also much larger.
+Thus, standard adversarial planning approaches, such as game tree search are not directly applicable.
+As we elaborate later, planning in RTS games can be seen as having multiple levels of abstraction: at a higher level, players need long-term planning capabilities, in order to develop a strong economy in the game; at a low level, individual units need to be moved in coordination to fight battles taking into account the terrain and the opponent.
+Techniques that can address this large planning problems by either sampling, or hierarchical decomposition do not yet exist.
+
+\subsubsection{Learning}
+Given the difficulties in playing RTS games by directly using adversarial planning techniques, many research groups have turned attention to learning techniques.
+We can distinguish three types of learning problems in RTS games:
+\begin{itemize}
+\item {\em Prior learning}: how can we exploit available data, such as existing replays, or information about specific maps for learning appropriate strategies before hand?
+A significant amount of work has gone in this direction.%, such as \cite{WeberCIG09,SynnaeveCIG11,SynnaeveAIIDE11,OntanonMSR10}.
+\item {\em In-game learning}: how can bots can deploy online learning techniques that allow them to improve their game play while playing a game?
+these techniques might include reinforcement learning techniques, but also opponent modeling.
+The main problem again is the fact that the state space is too large.
+\item {\em Inter-game learning}: what can be learned from one game that can be used to increase the chances of victory in the next game?
+Some work has used simple game-theoretical solutions to select amongst a pool of predefined strategies, but the general problem remains unanswered.
+\end{itemize}
 
@@
 \subsubsection{Uncertainty}
-In RTS games, there are two main kinds of uncertainty, existing for two different reasons.
-First, there is incompleteness of information, due to the fog of war, which leads to some kind of ``extensional''
-%``extensive'' 
-uncertainty, that can be lowered by good scouting, and knowledge representation (to infer what is possible given what has been seen).
-Secondly, there is the fact that we will never be able to read the opponent's mind, the ``intentional''
-%or ``intensive'' 
-uncertainty that we cannot really be sure to scratch off, even with the best scouting possible.
-For intentional uncertainty, the AI, as the human player, can only build a sensible model of what the opponent is likely to do, to think, or to think knowing what we think he thinks.
-Both these kinds of uncertainty can be studied through probabilities, but only for extensional uncertainty can $P(Hidden | Seen)$ be \textit{directly} linked to data.
-%now for the clumsy part:
-One could argue that the extensional data is the only thing that matters because it is the only thing that has an influence on the game, but in reality, for very highly skilled players, discovering the intentions of the enemy is the key to winning.
+Adversarial planning under uncertainty in domains of the size of RTS games is still an unsolved challenge.
+In RTS games, there are two main kinds of uncertainty.
+First, the game is partially observable, and players cannot observe the whole game map (like in Chess), but need to scout in order to see what the opponent is doing.
+This type of uncertainty can be lowered by good scouting, and knowledge representation (to infer what is possible given what has been seen).
+Second, there is also uncertainty arising from the fact that the games are adversarial, and a player cannot predict the actions that the opponent(s) will execute.
+For this type of uncertainty, the AI, as the human player, can only build a sensible model of what the opponent is likely to do.
+% [Santi: I commented out this part, since I'm not sure it's correct.
+The second type of uncertainty can also be linked to data.
+For example, by analyzing player replays, one could learn "plan recognition" probability tables, which can be used to infer the strategy of the opponent, and thus, reduce the uncertainty of the opponent's future actions]
+%Both these kinds of uncertainty can be studied through probabilities, but only the first type, $P(Hidden | Seen)$, can be {\em easily} linked to data. %now for the clumsy part:
+%One could argue that the extensional data is the only thing that matters because it is the only thing that has an influence on the game, but in reality, for very highly skilled players, discovering the intentions of the enemy is the key to winning.
 It allows for a compact understanding of the game state and development, while also giving additional information about the future.
@@
 Spatial  reasoning   is  related  to   each  aspect  of   the  terrain
-exploitation.  This  involves  buildings  placing  where  one  has  to
+exploitation.  Spatial reasoning is involved in tasks such as building placing, or base expansion. In the former, the player needs to
 carefully  consider buildings  position  into its  own  bases to  both
 protect them  by creating  a wall against  invasions and to  avoid bad
-configurations  where large  units could  be stuck.  Base expansion
-also belongs  to spatial reasoning,  in the sense  that the AI  has to
+configurations  where large  units could  be stuck. In base expansion, the player has to
 choose good available emplacements to  build a new base, regarding its
 own  position  and opponent's  bases.  Finally,  spatial reasoning  is
-closely related to  tactics with questions like ``Where  should I place
-my  units for battle?'',  favoring for  instance engagements  when the
-opponent's units are into a  bottleneck. In Starcraft, it is always an
-advantage to  have its units  on high ground  whereas enemy is  on low
-ground, since units  on low ground have no vision  on high ground, and
-firing from low  ground have only 70\% of chance to  hit units on high
-ground.
-
-As well as spatial reasoning, temporal reasoning is related to tactics
-since it implies  to decide about army movements,  like when attacking
-or  withdrawing  units from  battle.   A  part  of temporal  reasoning
-consists  of  threat predictions,  where  one  has  to plan  when  the
-opponent is  likely running an  attack.  Strategy copes  with temporal
-reasoning  as well,  mainly with  planning problems:  when researching
-upgrades, when constructing  wished buildings, when switching strategy
-if needed, etc.
+key to tactical reasoning: players need to decide where to place units for battle, favoring, for  instance, engagements  when the opponent's units are into a  bottleneck.
+%In Starcraft, it is always an advantage to  have its units  on high ground  whereas enemy is  on low ground, since units  on low ground have no vision  on high ground, and firing from low  ground have only 70\% of chance to  hit units on high ground.
+
+Analogously, temporal reasoning is key in tactical or strategic reasoning.
+For example, timing attacks and retreats to gain advantage.
+At a higher strategic level, players need to reason about when to perform long-impact economic actions such as upgrades, building construction, strategy switching, etc.
+all taking into account that the effects of these actions are not immediate, but longer term.
 
 \subsubsection{Domain Knowledge Exploitation}
-TODO
+In traditional board-games such as Chess, researchers have exploited the large amounts of existing domain knowledge to create good evaluation functions to be used by alpha-beta search algorithms, extensive opening books, or end-game tables.
+In the case of RTS games, it is still unclear the ways in which the significantly large amount of domain knowledge (in the forms or strategy guides, replays, etc.) can be exploited by bots.
+Most work in this area has focused on two main directions: on the one hand, researchers are finding ways in which to hard-code existing strategies into bots, so that bots only need to decide which strategies to deploy, instead of having to solve the complete problem of deciding which actions to execute by each individual unit at each time step.
+One the other hand, large datasets of replays have been created \cite{synnaeve2012dataset}, from where strategies, trends or plans have been tried to learn.
+However, Starcraft games are quite complex, and how to automatically learn from such datasets is still an open problem.
 
 \subsubsection{Task Decomposition}
-Playing a  RTS game like Starcraft  can be decomposed  into three main
-tasks: 
-\begin{itemize}
-\item Strategy:  corresponds to the  decision making process.  This is
+For all the previous reasons, most existing approaches to play games like Starcraft work by decomposing the problem of playing a RTS game into a collection of smaller problems, to be solved independently.
+Specifically, a common subdivision is:
+\begin{itemize}
+\item {\em Strategy}: often called {\em macro-management}, corresponds to the high-level decision making process.  This is
   the  highest level of  abstraction for  the game  comprehension. RTS
@@
   counter-strategy  against a  given opponent.  It concerns  the whole
-  set of  units and buildings  one owns. If  we may allow  ourselves a
-  military analogy, strategy is orders given by generals at the HQ. 
-\item  Tactics: are  the implementation  of the  current  strategy. It
+  set of  units and buildings  one owns. %If  we may allow  ourselves a  military analogy, strategy is orders given by generals at the HQ. 
+\item  {\em Tactics}: are  the implementation  of the  current  strategy. It
   implies  army and  building positioning,  movements, timing,  and so
-  on.  Tactics concerns a  group of  units. They  are orders  given by
-  officers on the battlefield.
-\item  Reactive  control:  is  the  implementation  of  tactics.  This
+  on.  Tactics concerns a  group of  units. %They  are orders  given by officers on the battlefield.
+\item  {\em Reactive  control}: often called, {\em micro-management} is  the  implementation  of  tactics.  This
   consists   in  moving,   targeting,  firing,   fleeing,  hit-and-run
   techniques (aka ``kiting'')  during battle. Reactive control focuses
-  on a specific unit. It is soldiers applying officers' orders. 
-\end{itemize}
-
-\noindent
-Both     strategy      and     tactics     are      usually     called
-\textit{macro-management},  whereas reactive  control  is also  called
-\textit{micro-management}. In addition of the three tasks shown above,
-one could also consider these two tasks that an AI has to deal with:
-\begin{itemize}
-\item Terrain analysis: consists  in the analysis of regions composing
-  the  map,  chokes,  minerals  and  gas emplacements,  low  and  high
+  on a specific unit. %It is soldiers applying officers' orders. 
+%\end{itemize}
+%
+%\noindent
+%Both     strategy      and     tactics     are      usually     called
+%\textit{macro-management},  whereas reactive  control  is also  called
+%\textit{micro-management}. In addition of the three tasks shown above,
+%one could also consider these two tasks that an AI has to deal with:
+%\begin{itemize}
+\item {\em Terrain analysis}: consists  in the analysis of regions composing
+  the  map:  choke-points,  minerals  and  gas emplacements,  low  and  high
   walkable grounds, islands, etc.
-\item Intelligence gathering:  corresponds to information collected on
-  our  opponent. Because of  the fog  of war,  a non-cheating  AI must
-  regulary send scouts to localize and spy enemy bases, and must see or
-  infer what buildings and technology the enemy has developed in order
-  to adapt its strategy.
-\end{itemize}
-
-\noindent
+\item {\em Intelligence gathering}:  corresponds to information collected about
+  the  opponent. Because of  the fog-of-war,  players must
+  regularly send scouts to localize and spy enemy bases.%, and must see or  infer what buildings and technology the enemy has developed in order  to adapt its strategy.
+\end{itemize}
+
 The reader can find a  good presentation of task decomposition for AIs
-playing RTS in~\cite{weber2011acs}.  }
+playing RTS in~\cite{weber2011acs}.
+Although the previous task decomposition is common, a significant challenge is on designing architectures so that the individual AI techniques that address each of those tasks can communicate and effectively work together, resolving conflicts, prioritizing resources between them, etc.
 

### 2013-02-20 survey.tex
Updated game state sizes with DAve's information

@@
 \item And finally, the complexity of these games, both in terms of state space size and in terms of number of actions available at each decision cycle is very large.
-{\color{red} For example, the state space of Chess is typically estimated to be around $10^{50}$, whereas the state space of Starcraft in a typical map is estimated to be around $10^{10000}$.}
+For example, the state space of Chess is typically estimated to be around $10^{50}$, heads up no-limit texas holdem poker around $10^{80}$, and Go around $10^{170}$.
+In comparison, the state space of Starcraft in a typical map is estimated to be at least {\color{red} $10^{10000}$.}
 \end{itemize}

### 2013-02-22 survey.tex
Comment at line 252 in the tex: add chance to hit from low to high ground? which % is right?
CIG competition update 1

@@
 own  position  and opponent's  bases.  Finally,  spatial reasoning  is
-key to tactical reasoning: players need to decide where to place units for battle, favoring, for  instance, engagements  when the opponent's units are into a  bottleneck.
+key to tactical reasoning: players need to decide where to place units
+for battle,  favoring, for  instance, engagements when  the opponent's
+units are into a bottleneck.
+
+% [Flo:  70% to hit high  ground units according to  Ben Weber's paper
+% [3], and 53.125% according  to BWAPI website. Who's right? Should we
+% add this in the paper?]
 %In Starcraft, it is always an advantage to  have its units  on high ground  whereas enemy is  on low ground, since units  on low ground have no vision  on high ground, and firing from low  ground have only 70\% of chance to  hit units on high ground.
@@
 
-\subsection{CIG 2011}
+\subsection{CIG}
 \label{sec:cig2011}
@@
 % bracket A
-\begin{tabular}{|l|c|c|c|}
+\begin{tabular}{|l|c|c|c|r|}
 \hline
-\multicolumn{4}{|c|}{Bracket A} \\ \hline
-Crashes & games & bot	& wins\\ \hline
-0 &	 40 &	 UAlbertaBot &	 33\\
-1 &   40 &	 Skynet	  &  31\\
-2 &	 40 &	 AIUR	  &  24\\
-1 &	 40 &	 Nova	  &  8\\
-0 &	 40 &	 LSAI	  &  4\\
+\multicolumn{5}{|c|}{Bracket A} \\ \hline
+Position & crashes & games & bot	& win\%\\ \hline
+A1 & 0 &	 40 &	 UAlbertaBot & 82.5\% \\
+A2 & 1 &  	 40 &	 Skynet	  &  77.5\% \\
+A3 & 2 &	 40 &	 AIUR	  &  60.0\% \\
+A4 & 1 &	 40 &	 Nova	  &  20.0\% \\
+A5 & 0 &	 40 &	 LSAI	  &  10.0\%\\
 \hline
@@
 % bracket B
-\begin{tabular}{|l|c|c|c|}
+\begin{tabular}{|l|c|c|c|r|}
 \hline
-\multicolumn{4}{|c|}{Bracket B} \\ \hline
-Crashes & games & bot	& wins\\ \hline
-12 &	 40 &	 Xelnaga &	 25\\
-3 &   40 &	 BroodwarBotQ  &  23 (6)\\
-0 &	 40 &	 BTHAI	  &  23 (4)\\
-17 &	 40 &	 Protoss Beast Jelly  & 17\\
-0 &	 40 &	 EvoBot	  &  12\\
+\multicolumn{5}{|c|}{Bracket B} \\ \hline
+Position & crashes & games & bot	& win\%\\ \hline
+B1 & 12 &	 40 &	 Xelnaga &	 62.5\%\\
+B2 & 3 &   40 &	 BroodwarBotQ  &  57.5\%\\
+B3 & 0 &	 40 &	 BTHAI	  &  57.5\%\\
+B4 & 17 &	 40 &	 Protoss Beast Jelly  & 42.5\%\\
+B5 & 0 &	 40 &	 EvoBot	  &  30.0\%\\
 \hline
@@
 % bracket A
-\begin{tabular}{|l|c|c|c|}
+\begin{tabular}{|l|c|c|c|r|}
 \hline
-Crashes & games & bot	& wins\\ \hline
-0 &  30 &	 Skynet	  		&  26\\
-0 &	 30 &	 UAlbertaBot 	&  22\\
-3 &	 30 &	 Xelnaga  		&  11\\
-2 &	 30 &	 BroodwarBotQ  	&  1\\
+Position & crashes & games & bot	& win\%\\ \hline
+1 & 0 &  30 &	 Skynet	  		&  86.7\%\\
+2 & 0 &	 30 &	 UAlbertaBot 	&  73.3\%\\
+3 & 3 &	 30 &	 Xelnaga  		&  36.7\%\\
+4 & 2 &	 30 &	 BroodwarBotQ  	&  3.3\%\\
 \hline
@@
 \end{table}
+
+For CIG 2012, the AIIDE tournament software was employed,
+leading to a total of $4050$ games played in $90$ rounds
+of round robin. As 6 different maps were used, this means
+that each bot played every other on every map 15 times.
 

### 2013-02-23 survey.tex
updated text CIG 2012

@@
 
-The 2012 man vs. machine match again used the winner of the competition (Skynet), who played against
-Mike Lange, aka LRM) Bakuryu. At the time of the match, Bakuryu was an A- ranked Zerg player on ICCup,
-and known as one of the best non-Korean Zerg players in the world. Bakuryu was considered much
-stronger than Oriol at the time that the match was played, and the results showed that this was true.
-In the first game of the best-of-three, Bakuryu made Skynet look quite silly by running around inside
-Skynet's base with a small number of zerglings while Skynet's zealots and half of its worked chased
-then in vain. After killing off several probes and buying enough time to set up his expansion, he
-cleaned up Skynet's army with a flying army of mutalisks. In the second game, Bakuryu contained Skynet
-inside its base with a group of zerglings positioned within Skynet's expansion. Skynet then constructed
-several Dark Templar and along with some Dragoons and Zealots attacked into Bakuryu's expansion which
-was heavily defended, and was crushed almost instantly, allowing Bakuryu's zergling force to finish
-off the Protoss base.
-
-In this match it was shown that the true weakness of state of the art StarCraft AI systems was that
-humans are very adept at recognizing scripted behaviors and exploiting them to the fullest. A human
-player in Skynet's position in the first game would have realized he was being taken advantage of and
-adapted his strategy accordingly, however the inability to put the local context (Bakuryu kiting his
-units around his base) into the larger context of the game (that this would delay Skynet until
-reinforcements arrived) and then the lack of strategy change to fix the situation led to an easy
-victory for the human. These problems remain the main challenges in RTS AI today: to both recognize
-the strategy and intent of an opponent's actions, and how to effectively adapt your own strategy to 
-overcome them.
-
-All results, videos, and replays from the AIIDE StarCraft AI Competition can be found in\footnote{\url{http://www.StarCraftAICompetition.com}}
-
-\subsection{CIG}
-\label{sec:cig2011}
-
-An initial attempt to run a StarCraft tournament at the Computational
-Intelligence in Games conference 2010 suffered from technical problems.
-These mainly stemmed from the desire to use evolved, largely untested
-maps which proved to look interesting but made the submitted bots 
-(and/or the Broodwar Terrain Analyzer (BWTA) provided with the BWapi interface) crash
-so frequently that it would have been unjustifiable to announce a winner.
-
-At CIG 2011, the tournament was therefore run with a (secret) selection
-of maps used in league play, which can be regarded as the most important
-difference to the AIIDE tournament that employed a known list of maps.
-The competition was organized by Tobias Mahlmann and Mike
-Preuss and attracted 10 bots. In addition to the ones discussed in previous
-sections (UAlbertaBot, Skynet, AIUR, Nova, BroodwarBotQ, BTHAI), 
-the set also contained LSAI, Xelnaga, Protoss Beast Jelly, and EvoBot,
-these are shortly described in the following:
-
-\paragraph*{LSAI (Zerg)} utilizes a heavily modified BWSAL to divide management
- of the units to different modules that communicate via a centralized information
- module. It works using a simple reactive strategy to try and survive early game
- attacks and macro up to a larger attack force and maintain map control.
-
-\paragraph*{Xelnaga (Protoss)} is a modification of the AIUR bot that chooses the 
-Dark Templar Opening in order to destroy the enemy base before defenses against
-invisible units are available.
-
-\paragraph*{Protoss Beast Jelly (Protoss)}
-always goes for a 5-gate Zealot rush, supported by an effective harvesting 
-strategy named power-mining (2 probes are assigned to every mineral patch,
-thereby needing 18 probes for 100\% saturation in a normal map, prior
-to expanding). Gas is not mined as it is not needed for constructing Zealots.
-
-\paragraph*{EvoBot (Terran)} employs an evolutionary algorithm for obtaining rational
- unit combinations and influence map techniques for deciding the strategic locations. Note
- that this bot was submitted in a very early version, with many of its designed features not
- yet fully ready.
-
-\begin{table*}[t!]
+\begin{table*}[t]
 \caption{Results of the first round at CIG 2011, held in two brackets.
@@
 
-\subsubsection{First Round}
-\label{sec:cig-first-round}
-
-As the CIG competition games were executed manually due to
-a lack of available software (the AIIDE program was not yet
-available at that time), the organizers separated the ten entries into
-two brackets. In each bracket of 5 bots, a round-robin
-tournament was held with 10 repetitions per pairing, resulting
-in 40 games per bot.
-The 5 maps chosen for the first round were selected from the pool
-of well-known league play maps found on the internet:
-\emph{(2)MatchPoint 1.3}, \emph{(4)Fighting Spirit 1.3}, 
-\emph{iCCupdestination 1.1}, \emph{iCCup gaia}, and 
-\emph{iCCup great barrier reef}. Each bot pairing played
-on every map twice, with switched starting positions.
-
-The two top bots of every bracket qualified
-for the final round. Table~\ref{tab:cig-first-round} summarizes
-the results.
-Note that as BroodwarBotQ and BTHAI have the same number of wins,
-their direct encounter was evaluated which accounted 6:4 for the BroodwarBotQ.
-The bots going into the final were thus UAlbertaBot, Skynet (from bracket A)
-and Xelnaga and BroodwarBotQ (from bracket B). All qualified bots play the
-Protoss faction. Most bots proved pretty stable, only Xelnaga and Protoss 
-Beast Jelly crashed relatively often (each in more than a quarter of the games). 
-Crashing of course resulted in an instant win for the other bot.
-In some cases, neither bot was able to finish the other off completely,
-so that they went into a passive state. We manually ended such games after
-around 15 minutes and assigned victory to the bot that had obtained more
-points as indicated on the end game screen.
-
-\subsubsection{Final Round}
-\label{sec:cig-final-round}
-
-The final round was played in a similar mode as each of the
-first round brackets,
-using another set of 5 previously unknown maps:
-\emph{iCCup lost temple 2.4}, \emph{iCCup rush hour 3.1},
-\emph{iCCup swordinthemoon 2.1}, \emph{iCCup yellow 1.1},
-and \emph{La\_Mancha 1.1}. 
-Letting each pairing play on each map twice again with
-switching starting positions resulted in 30 games per bot.
-The final results are displayed in table~\ref{tab:cig-final-round},
-indicating Skynet as winner and UAlbertaBot as runner-up, being
-almost equally strong, and the two other bots as clearly inferior.
-
-\begin{table}[t!]
+\begin{table}[b]
 \caption{Results of the CIG 2011 final round.}
@@
 
+The 2012 man vs. machine match again used the winner of the competition (Skynet), who played against
+Mike Lange, aka LRM) Bakuryu. At the time of the match, Bakuryu was an A- ranked Zerg player on ICCup,
+and known as one of the best non-Korean Zerg players in the world. Bakuryu was considered much
+stronger than Oriol at the time that the match was played, and the results showed that this was true.
+In the first game of the best-of-three, Bakuryu made Skynet look quite silly by running around inside
+Skynet's base with a small number of zerglings while Skynet's zealots and half of its worked chased
+then in vain. After killing off several probes and buying enough time to set up his expansion, he
+cleaned up Skynet's army with a flying army of mutalisks. In the second game, Bakuryu contained Skynet
+inside its base with a group of zerglings positioned within Skynet's expansion. Skynet then constructed
+several Dark Templar and along with some Dragoons and Zealots attacked into Bakuryu's expansion which
+was heavily defended, and was crushed almost instantly, allowing Bakuryu's zergling force to finish
+off the Protoss base.
+
+In this match it was shown that the true weakness of state of the art StarCraft AI systems was that
+humans are very adept at recognizing scripted behaviors and exploiting them to the fullest. A human
+player in Skynet's position in the first game would have realized he was being taken advantage of and
+adapted his strategy accordingly, however the inability to put the local context (Bakuryu kiting his
+units around his base) into the larger context of the game (that this would delay Skynet until
+reinforcements arrived) and then the lack of strategy change to fix the situation led to an easy
+victory for the human. These problems remain the main challenges in RTS AI today: to both recognize
+the strategy and intent of an opponent's actions, and how to effectively adapt your own strategy to 
+overcome them.
+
+All results, videos, and replays from the AIIDE StarCraft AI Competition can be found in\footnote{\url{http://www.StarCraftAICompetition.com}}
+
+\subsection{CIG}
+\label{sec:cig2011}
+
+An initial attempt to run a StarCraft tournament at the Computational
+Intelligence in Games conference 2010 suffered from technical problems.
+These mainly stemmed from the desire to use evolved, largely untested
+maps which proved to look interesting but made the submitted bots 
+(and/or the Broodwar Terrain Analyzer (BWTA) provided with the BWapi interface) crash
+so frequently that it would have been unjustifiable to announce a winner.
+
+At CIG 2011, the tournament was therefore run with a (secret) selection
+of maps used in league play, which can be regarded as the most important
+difference to the AIIDE tournament that employed a known list of maps.
+The competition was organized by Tobias Mahlmann and Mike
+Preuss and attracted 10 bots. In addition to the ones discussed in previous
+sections (UAlbertaBot, Skynet, AIUR, Nova, BroodwarBotQ, BTHAI), 
+the set also contained LSAI, Xelnaga, Protoss Beast Jelly, and EvoBot,
+these are shortly described in the following:
+
+\paragraph*{LSAI (Zerg)} utilizes a heavily modified BWSAL to divide management
+ of the units to different modules that communicate via a centralized information
+ module. It works using a simple reactive strategy to try and survive early game
+ attacks and macro up to a larger attack force and maintain map control.
+
+\paragraph*{Xelnaga (Protoss)} is a modification of the AIUR bot that chooses the 
+Dark Templar Opening in order to destroy the enemy base before defenses against
+invisible units are available.
+
+\paragraph*{Protoss Beast Jelly (Protoss)}
+always goes for a 5-gate Zealot rush, supported by an effective harvesting 
+strategy named power-mining (2 probes are assigned to every mineral patch,
+thereby needing 18 probes for 100\% saturation in a normal map, prior
+to expanding). Gas is not mined as it is not needed for constructing Zealots.
+
+\paragraph*{EvoBot (Terran)} employs an evolutionary algorithm for obtaining rational
+ unit combinations and influence map techniques for deciding the strategic locations. Note
+ that this bot was submitted in a very early version, with many of its designed features not
+ yet fully ready.
+
+\begin{figure}[t]
+	%\vspace*{-1ex}
+    \centering
+    \includegraphics[width=0.95\columnwidth]{figures/cig2012-ResultsRound90.pdf}
+	%\vspace*{-1ex}
+    \caption{Evolution of the win percentage of each bot participating in the
+CIG 2012 competition}
+    \label{fig:cig2012-results}
+\end{figure}
+
+\begin{table}[b]
+\caption{Results of the CIG 2012 competition.}
+\label{tab:cig2012}
+\begin{small}
+\begin{center}
+% bracket A
+\begin{tabular}{|l|c|c|}
+\hline
+Position & bot & win \% \\ \hline
+1 & Skynet & 78.3\% \\
+2 & UAlbertaBot & 65.2\% \\
+3 & AIUR & 60.4\% \\
+4 & Adjutant & 58.6\% \\
+5 & Nova & 52.4\% \\ \hline
+\end{tabular}
+\end{center}
+\end{small}
+\end{table}
+
+\subsubsection{First Round}
+\label{sec:cig-first-round}
+
+As the CIG competition games were executed manually due to
+a lack of available software (the AIIDE program was not yet
+available at that time), the organizers separated the ten entries into
+two brackets. In each bracket of 5 bots, a round-robin
+tournament was held with 10 repetitions per pairing, resulting
+in 40 games per bot.
+The 5 maps chosen for the first round were selected from the pool
+of well-known league play maps found on the internet:
+\emph{(2)MatchPoint 1.3}, \emph{(4)Fighting Spirit 1.3}, 
+\emph{iCCupdestination 1.1}, \emph{iCCup gaia}, and 
+\emph{iCCup great barrier reef}. Each bot pairing played
+on every map twice, with switched starting positions.
+
+The two top bots of every bracket qualified
+for the final round. Table~\ref{tab:cig-first-round} summarizes
+the results.
+Note that as BroodwarBotQ and BTHAI have the same number of wins,
+their direct encounter was evaluated which accounted 6:4 for the BroodwarBotQ.
+The bots going into the final were thus UAlbertaBot, Skynet (from bracket A)
+and Xelnaga and BroodwarBotQ (from bracket B). All qualified bots play the
+Protoss faction. Most bots proved pretty stable, only Xelnaga and Protoss 
+Beast Jelly crashed relatively often (each in more than a quarter of the games). 
+Crashing of course resulted in an instant win for the other bot.
+In some cases, neither bot was able to finish the other off completely,
+so that they went into a passive state. We manually ended such games after
+around 15 minutes and assigned victory to the bot that had obtained more
+points as indicated on the end game screen.
+
+\begin{figure*}[t]
+	%\vspace*{-2ex}
+    \centering
+    \includegraphics[width=0.34\textwidth]{figures/vstable3.pdf}
+    \includegraphics[width=0.3\textwidth]{figures/vstable6.pdf}
+    \includegraphics[width=0.3\textwidth]{figures/vstable.pdf}
+	%\vspace*{-4ex}
+    \caption{Win percentages of CIG 2012 competition, from left
+    to right: 3-player maps only, 6-player maps only, all maps.
+    Read from line to column, bot in row wins given fraction of
+    games against bot in column. For some bots, we find interesting
+    differences, e.g. Xelnaga gets worse on 6-player maps, UAlbertaBot
+    gets better. Only Xelnaga can reliably beat Skynet, but only on 
+    3-player maps}
+    \label{fig:cig2012-mapresults}
+\end{figure*}
+
+\subsubsection{Final Round}
+\label{sec:cig-final-round}
+
+The final round was played in a similar mode as each of the
+first round brackets,
+using another set of 5 previously unknown maps:
+\emph{iCCup lost temple 2.4}, \emph{iCCup rush hour 3.1},
+\emph{iCCup swordinthemoon 2.1}, \emph{iCCup yellow 1.1},
+and \emph{La\_Mancha 1.1}. 
+Letting each pairing play on each map twice again with
+switching starting positions resulted in 30 games per bot.
+The final results are displayed in table~\ref{tab:cig-final-round},
+indicating Skynet as winner and UAlbertaBot as runner-up, being
+almost equally strong, and the two other bots as clearly inferior.
+The competition setup, documentation and results can be found
+in\footnote{\url{http://ls11-www.cs.tu-dortmund.de/rts-competition/starcraft-cig2011}}.
+
 For CIG 2012, the AIIDE tournament software was employed,
@@
 that each bot played every other on every map 15 times. 
+As in the AIIDE competition, writing to and reading from
+a bot specific directory was enabled, however, due to 
+technical reasons, this feature was constrained to the
+computer (of 6) the game was actually run on. We can 
+therefore assume that this feature was of minor use for
+the CIG competition. The only other difference to the
+AIIDE competition was that the used maps were not 
+made available to the competitors in advance.
+
+These maps came in two flavors, namely 3 3-player maps:
+\emph{Athena-II}, \emph{Neo Moon Glaive}, \emph{Tears of the Moon},
+and 3 6-player maps:
+\emph{Legacy}, \emph{River of Light}, and 
+\emph{The Huntress 1.1}. 
+We shall note that some bots consistently crashed on one
+of the originally considered maps which has thus been replaced.
+This is surprising as all maps are well known league play maps
+or have been provided with the StarCraft Broodwar distribution
+itself.
+
+\ref{tab:cig2012}
+
+\ref{fig:cig2012-results}
+\ref{fig:cig2012-mapresults}
+
+map ‘size‘ has an effect!
+ Nova against BBQ: good on 6-player, bad on 3-player
+ SCAIL and Adjutant better on 6-player, BBQ worse on 	6-player
+ UAlbertaBot better on 6-player (due to scouting?)
+ Nova consistently good against UAlbertaBot
+ Xelnaga only bot to beat Skynet, but only on 3-player 	maps
+
+Here are the 2 replays shown in the presentation: Xelnaga surprising Skynet with a targetted DTS attack on its probes, and Nova successfully luring UAlbertaBot's probes into its own base.
+
+things (many) bots can‘t do:
+ handle islands
+ play zerg
+ play maps of different sizes (!!!)
+ play against random race
+ finish off beaten opponent
+ adapt to opponent strategy
+ cooperate (no teamplay)
+
+Setup and results for the CIG 2012 competition are 
+here\footnote{\url{http://ls11-www.cs.tu-dortmund.de/rts-competition/starcraft-cig2012}}.
 

### 2013-02-23 survey.tex
boing

@@
 itself.
-
-\ref{tab:cig2012}
-
-\ref{fig:cig2012-results}
-\ref{fig:cig2012-mapresults}
-
-map ‘size‘ has an effect!
- Nova against BBQ: good on 6-player, bad on 3-player
- SCAIL and Adjutant better on 6-player, BBQ worse on 	6-player
- UAlbertaBot better on 6-player (due to scouting?)
- Nova consistently good against UAlbertaBot
- Xelnaga only bot to beat Skynet, but only on 3-player 	maps
-
-Here are the 2 replays shown in the presentation: Xelnaga surprising Skynet with a targetted DTS attack on its probes, and Nova successfully luring UAlbertaBot's probes into its own base.
-
-things (many) bots can‘t do:
- handle islands
- play zerg
- play maps of different sizes (!!!)
- play against random race
- finish off beaten opponent
- adapt to opponent strategy
- cooperate (no teamplay)
-
-Setup and results for the CIG 2012 competition are 
+Setup, replays and results for the CIG 2012 competition can be found
 here\footnote{\url{http://ls11-www.cs.tu-dortmund.de/rts-competition/starcraft-cig2012}}.
+
+The overall results are displayed in table~\ref{tab:cig2012},
+and the win rate evolution over time in figure~\ref{fig:cig2012-results}.
+These are quite consistent with the results of the AIIDE 2012 competition,
+so that we can conclude that the best bots are not very dependent on 
+knowing the maps beforehand. However, the bot vs. bot win rates as
+displayed in figure~\ref{fig:cig2012-mapresults} show some interesting
+trends. On the maps with more possible start points, some bots do 
+better than others, namely SCAIL, Adjutant, Nova, and UAlbertaBot,
+the latter probably due to its very efficient scouting routine.
+Some bots however suffer from the increased uncertainty about
+the enemies position, namely Xelnaga and the BroodwarBotQ.
+
+As already observed before in the previously described competitions,
+there are also bots who consistently beat top ranked bots but have
+severe problems against lower ranked bots. E.g., Xelnaga is especially
+strong against Skynet on the 3-player maps (about 70\% wins). Reviewing
+the replays led to the assumption that Xelnaga usually tries to attack
+Skynet's probes with a dark templar strategy, and often succeeds. 
+Nova does very well against the UAlbertaBot, and the replays show
+that it sometimes succeeds to lure the probes into its own base, where
+they get killed, leading to severe resource problems. However, we cannot
+tell how often this happens as this would require to review every single
+replay between the 2 bots. Summarizing, most bots seem to have improved,
+which becomes clear if the nearly unchanged BTHAI bot is taken as a baseline.
+In 2011, it won more than half of its qualifying games, in 2012 it came
+out last with around 20\% wins. However, designing a bot in order to 
+beat a top bot (as for Xelnaga with Skynet) leads to a very restricted
+strategy that often leads to failure if playing against different bots.
+Note that in the direct encounter between Xelnaga and AIUR, its ancestor,
+Xelnaga looses consistently.
+
+Nevertheless, from the observations we made during the tournament,
+we can draw the conclusion that the available bots are still 
+very constrained. No bot in the competition played the Zerg race,
+which is surprising as the AIIDE 2010 winner (Overmind) did so.
+Presumably, implementing a good Zerg strategy is more demanding
+than for the Protoss or Terran races. Many bots consistently crashed
+when playing against a random race built-in bot for testing, and
+also did so when the map size was changed from $128\times 128$ to 
+any other. Furthermore, every single bot sometimes failed to finish
+off an already beaten opponent, such that the game had to be stopped
+after a previously determined maximum time. It also seems that most of
+the current bots are not very good at adapting their strategy to the
+one of their opponent during a game, or at least (via the read/write
+procedure of game information) within a series of games.
 

### 2013-07-29 reviewers-responses.tex
First version of the responses to reviewers

\documentclass{article}

\usepackage{cite,graphicx}
\usepackage[usenames,dvipsnames]{xcolor}

\usepackage{graphicx}
\usepackage[cmex10]{amsmath}
\usepackage{algorithmic}
\usepackage{array}
\usepackage{mdwmath}
\usepackage{mdwtab}
\usepackage{eqparbox}
%\usepackage[tight,footnotesize]{subfigure}
%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
%\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{fixltx2e}
\usepackage{stfloats}
\usepackage{url}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{A Survey of Real-Time Strategy Game AI\\ Research and Competition in StarCraft \\ Responses to Reviewers Comments}

\author{Santiago~Onta\~{n}\'{o}n\thanks{Santiago~Onta\~{n}\'{o}n is with the Computer Science Department at Drexel University, Philadelphia, PA, USA.},
        Gabriel~Synnaeve\thanks{Gabriel~Synnaeve is with the Laboratory of Cognitive Science and Psycholinguistics (LSCP) of ENS Ulm in Paris, France.},
        Alberto~Uriarte\thanks{Alberto Uriarte is with the Computer Science Department at Drexel University, Philadelphia, PA, USA.},\\
        Florian~Richoux\thanks{Florian~Richoux is with the Nantes Atlantic Computer Science Laboratory, France.},
        David~Churchill\thanks{David~Churchill is with the Computing Science Department of the University of Alberta, Edmonton, Canada.},
        Mike~Preuss\thanks{Mike~Preuss is with the Department of Computer Science of Technische Universit{\"a}t Dortmund, Germany.}}

\maketitle

We'd like to start by thanking the reviewers for the constructive comments and suggestions, which have definitively helped in making the paper stronger. Below, we include responses to the specific comments by each of the reviewers.

\section*{Reviewer 1}

{\bf Reviewer Comment}: {\em There are several spelling and grammar errors that needs to be fixed.}

{\bf Answer}: We have revised the paper accordingly. Thanks for the suggestion.

\vspace{0.3cm}
\noindent {\bf Reviewer Comment}: {\em First in the introduction, I think the ORTS competition that preceded the Starcraft competition should be mentioned.}

{\bf Answer}: Indeed, we have included a mention to it in the introduction.

\vspace{0.3cm}
\noindent {\bf Reviewer Comment}: {\em On page 2 the authors state that ``... 30 to 35 different types of units and buildings, each one with a significant number of special abilities.''. This is not entirely true. There are both units (for example Zealots) and buildings (for example Pylons) that does not have any special abilities. I would prefer to use the word ``most'' instead of ``each one''.}

{\bf Answer}: Fixed.

\vspace{0.3cm}
\noindent {\bf Reviewer Comment}: {\em On page 5: "Most research in RTS game AI assumes perfect information all the time.". As the authors state later in the paper, most of the mentioned bot competitions use fog-of-war and therefore have imperfect information for the AI.}

{\bf Answer}: Thanks for noticing, that sentence was indeed problematic, we have changed the wording to reflect what we intended to sa

\vspace{0.3cm}
\noindent {\bf Reviewer Comment}: {\em On page 6: ``Potential fields (or influence maps) have been...''. There are significant differences between these two techniques and they are not the same, as the reader might think by reading that sentence.}

{\bf Answer}: Thanks for pointing this out, we have corrected this in all the places in the paper were the same confusion appeared.

\vspace{0.3cm}
\noindent {\bf Reviewer Comment}: {\em Chapter ``III. Existing work on RTS game AI'' is a bit confusing. As a reader I want to know what work has been done in the RTS game AI area, but in several places (for example in page 5) other very different game genres (FPS games) are mentioned. I think these related work should either be removed, or the authors should state more clearly why this particular related work in FPS games is relevant to RTS game AI.}

{\bf Answer}: We have clarified that this piece of work in particular is relevant for RTS games.

\vspace{0.3cm}
\noindent {\bf Reviewer Comment}: {\em Table I, II and III. The tables show the top bots in each competition, but it is nowhere mentioned that they only show the top bots and that other bots participated as well. For example Table I shows three bots while a total of 17 bots participated in the tournament. Table VI only shows the top bots as well, but in this case it was mentioned in the text that more bots participated.}

{\bf Answer}: Agreed, and fixed.

\section*{Reviewer 2}

{\bf Reviewer Comment}: {\em Page 2. Please make your estimates a little better. I've played SC and SC2 and it is not that I think you are off, I am just really curious and want to get it as close and as supportable as possible.  The paper would be much improved if, for example, you looked at average estimates for b and d from real games? Novices versus experts. Specifically, when estimating pow(b, d),

\begin{enumerate}
\item For d: Should you also consider buildings? Should you look at
early, mid, and late game value for d? From say 30 pro and 30 novice games?

\item For b: What is the range? SC is well defined, you should be able to
come up with a more justifiable estimate.
\end{enumerate}

Please do spend a little more time on coming up with more refined estimates and supporting them with evidence. This will be most helpful in defining the role of RTS games in CI and AI research.
}

{\bf Answer}: Indeed coming up with accurate estimates is very important to determining the actual complexity of games like StarCraft, and compare it with other tasks in AI and CI. The original estimate in our paper was intended to be a rough lower-bound (which was already tremendously large), just to give the reader an idea of the difference between RTS games and games like Chess or Go. We have improved our estimate (we carefully assessed the number of actions of each unit available in StarCraft for the Terran race. With that analysis, we came up with a more accurate estimate. This is still a lower-bound, and indeed buildings could be added to the equation, making $b$ even larger. But again, our intention is just to provide a lower bound, which shows that StarCraft is many, many, orders of magnitude larger than any game solved to date.

However, we would like to point out that our intention is just to provide a theoretical measure of complexity, which only depends on the rules of the game, and not on the specific players (whether they are novices or experts). The play-style of the player affects the measure of complexity, at most, in the number of units they control at once. But everything else is just determined by the rules of the game. However, an exact analysis of the branching factor $b$ that actually occurs in specific instances of games, like those player by humans, seems like a very interesting experiment, which we will indeed consider for our future work. Thanks for the suggestion!

\vspace{0.3cm}
\noindent {\bf Reviewer Comment}: {\em Page 4. There is at least one other way you can subdivide the tasks to get high quality, RTS bots, and I would refer you to some of Chris Miles' work in RTS games (and also the first thesis on RTS games: \url{http://www.cse.unr.edu/~miles/papers/index.htm#dissertation} ). Strangely hard to find with google}

{\bf Answer}: Agreed. Thanks for pointing to this piece of work, which we had missed. We have added a few references in different parts of the paper where the work of Chris Miles was relevant.

\vspace{0.3cm}
\noindent {\bf Reviewer Comment}: {\em Fig 2. Caption seems rushed, please fix the obvious grammer issues. Nice Figure. }

{\bf Answer}: Fixed.

\vspace{0.3cm}
\noindent {\bf Reviewer Comment}: {\em Column 2: line 13: Should be: constrain}

{\bf Answer}: Fixed.

\vspace{0.3cm}
\noindent {\bf Reviewer Comment}: {\em Column 2: Please rewrite: Finally, we will call reactive control to how the player controls individual units to maximize their efﬁciency (when executing tactics) in real-time.}

{\bf Answer}: Done.

\vspace{0.3cm}
\noindent {\bf Reviewer Comment}: {\em Column 2: lines 34 - 43, You might want to point out the relationship with micro and macro here?}

{\bf Answer}: Indeed, that was missing. Done.

\vspace{0.3cm}
\noindent {\bf Reviewer Comment}: {\em Page 5: Column 1: line 27: ``being'' should come after ``planning''}

{\bf Answer}: Fixed.

\vspace{0.3cm}
\noindent {\bf Reviewer Comment}: {\em line 50 or thereabouts: Please describe Miles' approach to strategic planning as well}

{\bf Answer}: Done.

\vspace{0.3cm}
\noindent {\bf Reviewer Comment}: {\em Page 6: Column 1: line 20: remove the "is" at the beginning of the sentence}

{\bf Answer}: Done.

\vspace{0.3cm}
\noindent {\bf Reviewer Comment}: {\em Page 9+: There are several small grammatical errors that can be easily fixed. I am not going to point them all out.}

{\bf Answer}: Thanks, we have proofread the complete paper for typos and grammar.

\end{document}

### 2014-08-22 ECGG_RTSProblems/ECGG_RTSProblems.tex
Added files to start working on the ECGG papers

\documentclass{llncs}

\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{color}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{mathabx}
%\usepackage{hyperref}
%\usepackage{subfigure}
%\usepackage[colorinlistoftodos, textwidth=3.2cm, shadow]{todonotes}

\usepackage{algorithm}
\usepackage{fixltx2e}
\usepackage{algpseudocode}

\usepackage{multirow}

% To use Call inside another Call (algorithms)
\MakeRobust{\Call}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Title
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{RTS AI Problems and Techniques
%\thanks{Supported by ...}
}

\author{Santiago~Onta\~{n}\'{o}n\inst{1} \and 
		Gabriel~Synnaeve\inst{2} \and
		Alberto~Uriarte\inst{1} \and
		Florian~Richoux\inst{3} \and
		David~Churchill\inst{4} \and
		Mike~Preuss\inst{5}
		}

\institute{
	Computer Science Department at Drexel University, Philadelphia, PA, USA. \\
	\email{\{santi,albertouri\}@cs.drexel.edu}
\and
	Cognitive Science and Psycholinguistics (LSCP) of ENS Ulm, Paris, France. \\
	\email{gabriel.synnaeve@gmail.com}
\and
	Nantes Atlantic Computer Science Laboratory (LINA) of the Universit{\'e} de Nantes, France. \\
	\email{florian.richoux@univ-nantes.fr}
\and
	Computing Science Department of the University of Alberta, Edmonton, Canada. \\
	\email{cdavid@cs.ualberta.ca}
\and
	Department of Computer Science of Technische Universit{\"a}t Dortmund, Germany.
	\email{mike.preuss@cs.tu-dortmund.de}
}

\begin{document}

\maketitle

\section{Definition}

{\color{blue} to-do}

\bibliographystyle{splncs}
\bibliography{references}

\end{document}

### 2014-08-22 ECGG_StarCraftCompetitions/ECGG_StarCraftCompetitions.tex
Added files to start working on the ECGG papers

--- 2014-08-22 ECGG_RTSProblems/ECGG_RTSProblems.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{RTS AI Problems and Techniques
+\title{StarCraft Bots and Competitions
 %\thanks{Supported by ...}
@@
 
-\author{Santiago~Onta\~{n}\'{o}n\inst{1} \and 
-		Gabriel~Synnaeve\inst{2} \and
-		Alberto~Uriarte\inst{1} \and
+\author{David~Churchill\inst{1} \and
+		Mike~Preuss\inst{2}		
 		Florian~Richoux\inst{3} \and
-		David~Churchill\inst{4} \and
-		Mike~Preuss\inst{5}
+		Gabriel~Synnaeve\inst{4} \and
+		Alberto~Uriarte\inst{5} \and
+		Santiago~Onta\~{n}\'{o}n\inst{5} \and 
 		}
@@
 \institute{
-	Computer Science Department at Drexel University, Philadelphia, PA, USA. \\
-	\email{\{santi,albertouri\}@cs.drexel.edu}
-\and
-	Cognitive Science and Psycholinguistics (LSCP) of ENS Ulm, Paris, France. \\
-	\email{gabriel.synnaeve@gmail.com}
-\and
-	Nantes Atlantic Computer Science Laboratory (LINA) of the Universit{\'e} de Nantes, France. \\
-	\email{florian.richoux@univ-nantes.fr}
-\and
 	Computing Science Department of the University of Alberta, Edmonton, Canada. \\
@@
 	\email{mike.preuss@cs.tu-dortmund.de}
+\and
+	Nantes Atlantic Computer Science Laboratory (LINA) of the Universit{\'e} de Nantes, France. \\
+	\email{florian.richoux@univ-nantes.fr}
+\and
+	Cognitive Science and Psycholinguistics (LSCP) of ENS Ulm, Paris, France. \\
+	\email{gabriel.synnaeve@gmail.com}
+\and
+	Computer Science Department at Drexel University, Philadelphia, PA, USA. \\
+	\email{\{santi,albertouri\}@cs.drexel.edu}
 }

### 2014-12-08 ECGG_RTSProblems/ECGG_RTSProblems.tex
Copy/Paste Intro + Section 2 into the paper. Must shorten everything (the paper should be between 6 and 10 pages).

@@
 \and
-	Nantes Atlantic Computer Science Laboratory (LINA) of the Universit{\'e} de Nantes, France. \\
+	Nantes Atlantic Computer Science Laboratory (LINA), Univ. Nantes, France.\\
 	\email{florian.richoux@univ-nantes.fr}
@@
 
-\section{Definition}
+\begin{abstract}
+  This is the abstract.
+\end{abstract}
 
-{\color{blue} to-do}
+\section{Introduction}\label{sec:intro}
+The  field of  Real-Time Strategy  (RTS) game  Artificial Intelligence
+(AI) has advanced significantly since Michael Buro's call for research
+in this area \cite{Buro03rts}. Specially, competitions like the ``ORTS
+RTS  Game AI  Competition''  (held  from 2006  to  2009), the  ``AIIDE
+StarCraft AI Competition'' (held since  2010), and the ``CIG StarCraft
+RTS AI Competition'' (held since  2011) have motivated the exploration
+of many  AI approaches  in the  context of  RTS AI.  We will  list and
+classify these approaches, explain  their strengths and weaknesses and
+conclude on what is left to achieve human-level RTS AI.
+
+Complex  dynamic  environments,  where neither  perfect  nor  complete
+information  about the  current state  or  about the  dynamics of  the
+environment are available, pose  significant challenges for artificial
+intelligence. Road traffic, finance, or weather forecasts are examples
+of such large, complex, real-life  dynamic environments. RTS games can
+be seen  as a simplification  of one such real-life  environment, with
+simpler dynamics in a finite and smaller world, although still complex
+enough to  study some  of the key  interesting problems  like decision
+making under  uncertainty or  real-time adversarial  planning. Finding
+efficient techniques for tackling these problems on RTS games can thus
+benefit other  AI disciplines and  application domains, and  also have
+concrete and direct applications in the ever growing industry of video
+games.
+
+% Some other arguments for the motivation: 
+% 1. It's fun 
+% 2. Togelius' argument: Some problems  are very similar than the ones
+% in  robotics (decision  making, pathfinding,  ...).  However  RTS AI
+% does  not have  the same  limitations for  running experiments:  few
+% hardware constraints (we just need  a computer), we can have control
+% on the time flow, we can run many experiments in parallel, ...
+
+This article aims to  provide an overview on what is  the state of the
+art  in  RTS AI,  with  a  particular emphasis  on  the  work done  in
+StarCraft.  It is organized as follows: 
+{\color{blue} TODO}
+
+% Section~\ref{sec:rts}
+% introduces RTS games, in particular the game StarCraft, and their main
+% AI challenges.  Section~\ref{sec:review} reviews  the existing work on
+% tackling  these   challenges  in  RTS   games.   Section~\ref{sec:bot}
+% analyzes  several current  state of  the art  RTS game  playing agents
+% (called  {\em  bots}),  selected   from  the  participants  to  annual
+% StarCraft  AI  competitions.   Section~\ref{sec:competition}  presents
+% results of  the recent annual competitions  held at the AIIDE  and CIG
+% conferences  and  a  StarCraft bot  game  ladder\footnote{An  extended
+%   tournament,   which   can    potentially   go   on   indefinitely.}.
+% Section~\ref{sec:questions}  compiles  open   questions  in  RTS  game
+% AI. Finally, the paper concludes on discussions and perspectives.
+
+\section{Real-Time Strategy Games}\label{sec:rts}
+
+Real-Time Strategy is a sub-genre of strategy games where players need
+to  build an  economy (gathering  resources and  building a  base) and
+military power (training units  and researching technologies) in order
+to defeat  their opponents  (destroying their army  and base).  From a
+theoretical point of view, the  main differences between RTS games and
+traditional board games such as Chess or Go are:
+
+\begin{itemize}
+\item  They are  {\em simultaneous  move} games,  where more  than one
+  player  can issue  actions  at the  same  time. Additionally,  these
+  actions are {\em durative}, i.e.  actions are not instantaneous, but
+  take some amount of time to complete.
+\item RTS games  are ``real-time'', which actually means  is that each
+  player  has  a  very  small  amount  of  time  to  decide  the  next
+  move. Compared to  Chess, where players may have  several minutes to
+  decide the next action, in StarCraft, the game executes at 24 frames
+  per second, which means that players  can act as fast as every 42ms,
+  before the game state changes.
+\item Most  RTS games are  partially observable: players can  only see
+  the part of the  map that has been explored. This  is referred to as
+  the {\em fog-of-war}.
+\item Most RTS games are non-deterministic. Some actions have a chance
+  of success.
+\item And  finally, the complexity  of these  games, both in  terms of
+  state space size and in terms of number of actions available at each
+  decision cycle is very large. For  example, the state space of Chess
+  is typically  estimated to  be around  $10^{50}$, heads  up no-limit
+  Texas holdem  poker around $10^{80}$,  and Go around  $10^{170}$. In
+  comparison,  the  state space  of  StarCraft  in  a typical  map  is
+  estimated to be  many orders of magnitude larger than  any of those,
+  as discussed in the next section.
+\end{itemize}
+
+For those reasons, standard techniques  used for playing classic board
+games, such as  game tree search, cannot be directly  applied to solve
+RTS games without the definition of some level of abstraction, or some
+other simplification. Interestingly enough, humans  seem to be able to
+deal with the  complexity of RTS games, and are  still vastly superior
+to       computers      in       these       types      of       games
+\cite{burochurchill2012aimagazine}.   For  those   reasons,  a   large
+spectrum of techniques  have been attempted to deal  with this domain,
+as we will describe below. The remainder of this section is devoted to
+describe StarCraft  as a research  testbed, and on detailing  the open
+challenges in RTS game AI.
+
+\subsection{StarCraft}\label{subsec:StarCraft}
+
+{\em StarCraft: Brood  War} is an immensely popular  RTS game released
+in   1998  by   Blizzard  Entertainment.   StarCraft  is   set  in   a
+science-fiction based universe where the player must choose one of the
+three  races: Terran,  Protoss or  Zerg.  One of  the most  remarkable
+aspects  of StarCraft  is  that  the three  races  are extremely  well
+balanced:
+
+\begin{itemize}
+\item Terrans provide  units that are versatile and  flexible giving a
+  balanced option between Protoss and Zergs.
+\item  Protoss   units  have   lengthy  and   expensive  manufacturing
+  processes, but they are strong and resistant.  These conditions make
+  players follow a strategy of quality over quantity.
+\item Zergs, the insectoid race, units are cheap and weak. They can be
+  produced fast, encouraging players to overwhelm their opponents with
+  sheer numbers.
+\end{itemize}
+
+% What about copyright problems with this screenshot?
+% Figure~\ref{fig:StarCraft} shows  a screenshot of StarCraft  showing a
+% player playing the Terran race. 
+In order to win a StarCraft  game, players must first gather resources
+(minerals  and Vespene  gas). As  resources become  available, players
+need to allocate them for creating more buildings (which reinforce the
+economy, and allow players to  create units or unlock stronger units),
+research  new technologies  (in order  to  use new  unit abilities  or
+improve the units)  and train attack units. Units  must be distributed
+to  accomplish different  tasks  such as  reconnaissance, defense  and
+attack.  While  performing all  of those tasks,  players also  need to
+strategically understand the geometry of the  map at hand, in order to
+decide where to place new buildings  (concentrate in a single area, or
+expand  to  different  areas)  or where  to  set  defensive  outposts.
+Finally, when  offensive units of  two players meet, each  player must
+quickly maneuver each  of the units in order to  fight a battle, which
+requires quick and reactive control of each of the units.
+
+% \begin{figure}
+%     \centering
+%     \includegraphics[width=0.8\columnwidth]{figures/starcraft1.png}
+%     \caption{A screenshot of \emph{StarCraft: Brood War}.}
+%     \label{fig:StarCraft}
+% \end{figure}
+
+A typical  StarCraft map is defined  as a rectangular grid,  where the
+$width \times  height$ of  the map  is measured in  the number  of $32
+\times 32$ squares of pixels, also  known as build tiles. However, the
+resolution of  walkable areas is  in squares  of $8 \times  8$ pixels,
+also known as  walk tiles. The typical dimensions for  maps range from
+$64  \times 64$  to  $256 \times  256$ build  tiles.  Each player  can
+control  up to  200 units  (plus  an unlimited  number of  buildings).
+Moreover,  each different  race contains  between 30  to 35  different
+types of units  and buildings, most of them with  a significant number
+of  special abilities.  All these  factors together  make StarCraft  a
+significant  challenge, in  which humans  are still  much better  than
+computers.      For     instance,      in     the      game     ladder
+iCCup\footnote{http://www.iccup.com/StarCraft/} where users are ranked
+by their current point totals ($E$ being the lowest possible rank, and
+$A^+$  and  $Olympic$ being  the  second  highest and  highest  ranks,
+respectively), the best  StarCraft AI bots are ranked  between $D$ and
+$D^+$,  where average  amateur players  are ranked  between $C^+$  and
+$B$. For comparison, StarCraft professional players are usually ranked
+between $A^-$ and $A^+$.
+
+From a theoretical point of view,  the state space of a StarCraft game
+for a given map is enormous.  For example, consider a $128 \times 128$
+map. At any given moment there might be between 50 to 400 units in the
+map,  each of  which might  have a  complex internal  state (remaining
+energy  and hit-points,  action  being executed,  etc.). This  quickly
+leads to an immense number of  possible states (way beyond the size of
+smaller games, such as Chess or Go). For example, just considering the
+location of  each unit (with  $128 \times 128$ possible  positions per
+unit),  and 400  units, gives  us  an initial  number of  $16384^{400}
+\approx 10^{1685}$. If we add the  other factors playing a role in the
+game, we obtain even larger numbers.
+
+Another way to measure the complexity of the game is by looking at the
+branching factor,  $b$, and the  depth of  the game, $d$,  as proposed
+in~\cite{Gaby}, with  a total game  complexity of $b^d$. In  Chess, $b
+\approx 35$  and $d \approx  80$. In more  complex games, like  Go, $b
+\approx  30$ to  $300$, and  $d  \approx 150$  to $200$.  In order  to
+determine the  branching factor in StarCraft  when an AI plays  it, we
+must have in mind, that the  AI can issue actions simultaneously to as
+many  units in  the  game as  desired. Thus,  considering  that, in  a
+typical game, a player controls between 50 to 200 units, the branching
+factor  would be  between $u^{50}$  and  $u^{200}$, where  $u$ is  the
+average number of actions each unit can execute.  Estimating the value
+of $u$ is not easy, since the  number of actions a unit can execute is
+highly  dependent   on  the  context.   Let  us  make   the  following
+assumptions: 1) at most 16 enemy units  will be in range of a friendly
+unit (larger values  are possible, but unlikely), 2) when  an AI plays
+StarCraft, it only makes sense to  consider movement in the 8 cardinal
+directions per unit  (instead of assuming that the player  can issue a
+``move'' command to anywhere in the map  at any point in time), 3) for
+``build'' actions,  we consider that  SCVs (Terran worker  units) only
+build in their  current location (otherwise, if they need  to move, we
+consider  that  as  first  issuing  a  ``move''  action,  and  then  a
+``build''), and  4) let's  consider only the  {\em Terran}  race. With
+those assumptions,  units in  StarCraft can  execute between  1 (units
+like ``Supply  Depots'', whose only  action is  to be ``idle'')  to 43
+actions (Terran ``Ghosts''), with typical values around 20 to 30. Now,
+if we have in mind that actions have cool-down times, and thus not all
+units can  execute all of  the actions at every  frame, we can  take a
+conservative estimation  of about $10$  possible actions per  unit per
+game frame. This results in  a conservative estimate for the branching
+factor  between $b  \in  [10^{50},10^{200}]$,  only considering  units
+(ignoring the actions buildings can execute).
+% Assuming  a conservative  value for  $u$  of about  $30$ (moving  to
+% nearby positions, attacking units in range, using special abilities,
+% etc.), we  obtain a  branching factor of  $b$ between  $30^{50}$ and
+% $30^{200}$.
+Now, to  compute $d$, we simply  consider the fact that  typical games
+last for  about 25  minutes, which  results in  $d \approx  36000$ (25
+minutes $\times$ 60 seconds $\times$ 24 frames per second).
+
+% In StarCraft,  humans can issue only  one action at a  time (this is
+% because of  the GUI  used by  the game,  computer players  can issue
+% several  actions, to  different  units, at  any  given time),  thus,
+% assuming a  player controls between 50  to 200 units, and  that each
+% unit can execute  between 4 to 20  actions (moving in each  of the 4
+% cardinal  directions,   attacking  units   in  range,   use  special
+% abilities, etc.), we estimate the branching factor for human players
+% to be around $b \approx 200  - 4000$. Good players can execute about
+% 300 actions per minute, and the typical length of a game is about 25
+% minutes. Thus, $d \approx 7500$.
+
+% However, this  is a weak indication  of the complexity of  the game,
+% since many  of those states  would never be  reached in a  game. For
+% that  reason, it  is  more useful  to  just think  in  terms of  the
+% branching factor  and depth  of a  game, when a  human plays  it. In
+% Chess,  the branching  factor is  around  35 and  the depth,  around
+% 80. In Go, the  branching factor is between 30 to  300 and the depth
+% 150 to 200. In StarCraft, humans can execute about ...
+
+% From Gabriel's dissertation:
+% ------------------------------------
+%How does the state of possible actions grow? To measure this, we
+%used a measure from perfect information zero-sum games (as Checkers,
+%Chess and Go): the branching factor* b and the depth d of a typical
+%game. The complexity of a game (for taking a decision) is proportional
+%to bd.
+%[...]
+%In RTS games, b = 200 is a lower bound (in StarCraft we may have
+%between 50 to 400 units to control), and very good amateurs and
+%professional players perform more than 300 actions per minute. As
+%StarCraft is a simultaneous move, multi-units game, a strict number
+%for b would be |actions||units| (actions account for atomic moves and
+%abilities), thus for StarCraft b would be around 30^{60}. Strictly
+%speaking, for a StarCraft game, d = 24  game_seconds (24 game frames
+%per second), with a game duration of 25 minutes, this gives d =
+%36,000, thus b^d Å 30^{60^{36000}} .
+
+\subsection{Challenges in RTS Game AI}\label{subsec:challenges}
+
+Early research  in AI  for RTS  games \cite{Buro03rts}  identified the
+following six challenges:
+\begin{itemize}
+\item Resource management
+\item Decision making under uncertainty
+\item Spatial and temporal reasoning
+\item Collaboration (between multiple AIs)
+\item Opponent modeling and learning
+\item Adversarial real-time planning
+\end{itemize}
+
+While there  has been  a significant  work in  many, others  have been
+untouched (e.g. collaboration). Moreover, recent research in this area
+has identified several additional research  challenges, such as how to
+exploit the massive amounts  of existing domain knowledge (strategies,
+build-orders,  replays,  and  so   on).  Below,  we  describe  current
+challenges in RTS Game AI, grouped in six main different areas.
+
+%\begin{itemize}
+%\item Task Decomposition (or ``Architecture'')
+%\item Integration of Domain Knowledge
+%\item Reasoning with Uncertainty (including information gathering)
+%\item Opponent Modeling  and Adaptation: opponent modeling  is key if
+%  we  have  to  adapt  our  strategy.  As  different  strategies  are
+%  dominating each  others, forming  multiple Nash equilibria,  the AI
+%  has to be able to infer the intend of its opponent.
+%\item  Group   and  Individual  Control  (``micro''):   the  task  of
+%  controlling units efficiently (we can  not speak of optimality here
+%  due to  the huge  state space,  and of  the enemy  behavior entails
+%  numerous  Nash  equilibria),  focusing  fire  to  diminish  enemy's
+%  firepower  and  keeping  our   units  alive  the  longest,  casting
+%  defensive and offensive spells and abilities.
+%\item Planning and Resource Allocation (``macro'')
+%\item Spatial reasoning (``tactics'')
+%\end{itemize}
+
+\subsubsection{Planning}
+As mentioned above, the  size of the state space in  RTS games is much
+larger  than  that  of  traditional  board  games  such  as  Chess  or
+Go. Additionally,  the number  of actions  that can  be executed  at a
+given instant of time is  also much larger. Thus, standard adversarial
+planning  approaches,  such  as  game tree  search  are  not  directly
+applicable. As we  elaborate later, planning in RTS games  can be seen
+as having multiple  levels of abstraction: at a  higher level, players
+need long-term  planning capabilities,  in order  to develop  a strong
+economy in the game; at a low level, individual units need to be moved
+in coordination to  fight battles taking into account  the terrain and
+the  opponent.  Techniques  that  can  address  these  large  planning
+problems by either sampling, or  hierarchical decomposition do not yet
+exist.
+
+\subsubsection{Learning}
+Given  the  difficulties  in  playing  RTS  games  by  directly  using
+adversarial  planning techniques,  many  research  groups have  turned
+attention to  learning techniques. We  can distinguish three  types of
+learning problems in RTS games:
+\begin{itemize}
+\item {\em Prior learning}: How can we exploit available data, such as
+  existing replays,  or information  about specific maps  for learning
+  appropriate strategies before hand? A significant amount of work has
+  gone in this
+  direction.%, such as \cite{WeberCIG09,SynnaeveCIG11,SynnaeveAIIDE11,OntanonMSR10}.
+\item  {\em In-game  learning}: How  can bots  deploy online  learning
+  techniques that allow them to  improve their game play while playing
+  a  game?  These  techniques  might  include  reinforcement  learning
+  techniques, but  also opponent modeling.  The main problem  again is
+  the fact  that the state  space is too large  and the fact  that RTS
+  games are partially observable.
+\item {\em  Inter-game learning}:  What can be  learned from  one game
+  that can  be used  to increase  the chances of  victory in  the next
+  game? Some work has used simple game-theoretical solutions to select
+  amongst a  pool of  predefined strategies,  but the  general problem
+  remains unsolved.
+\end{itemize}
+
+%%% I  think there  are two  important aspects  of uncertainty  in RTS
+%%% games:
+%%% \begin{itemize}
+%%% \item First  is trying to model  uncertainty and try to  reduce it
+%%%   (information gathering,  and representation of  knowledge), this
+%%%   includes scouting,  and maintaining good estimates  of locations
+%%%   and number of units of the enemy, etc.
+%%% \item Second  is actually using the  uncertain information. Taking
+%%%   decisions  like  army strength  estimation  (like  to attack  or
+%%%   retreat).
+%%% \end{itemize}
+
+\subsubsection{Uncertainty}
+Adversarial planning under  uncertainty in domains of the  size of RTS
+games is  still an unsolved  challenge.  In  RTS games, there  are two
+main kinds  of uncertainty. First,  the game is  partially observable,
+and players  cannot observe the  whole game  map (like in  Chess), but
+need to scout in order to see what the opponent is doing. This type of
+uncertainty   can  be   lowered  by   good  scouting,   and  knowledge
+representation  (to  infer  what  is  possible  given  what  has  been
+seen). Second,  there is also  uncertainty arising from the  fact that
+the games  are adversarial,  and a player  cannot predict  the actions
+that the opponent(s)  will execute. For this type  of uncertainty, the
+AI, as the human  player, can only build a sensible  model of what the
+opponent is likely to do.
+% [Santi:  I  commented  out  this  part,  since  I'm  not  sure  it's
+% correct.  The second  type  of  uncertainty can  also  be linked  to
+% data.  For example,  by analyzing  player replays,  one could  learn
+% "plan recognition"  probability tables, which  can be used  to infer
+% the strategy  of the opponent,  and thus, reduce the  uncertainty of
+% the opponent's future  actions] Both these kinds  of uncertainty can
+% be studied through probabilities, but only the first type, $P(Hidden
+% | Seen)$, can be {\em easily} linked to
+% data. %now for the clumsy part:
+% One could  argue that the  extensional data  is the only  thing that
+% matters because  it is the only  thing that has an  influence on the
+% game, but in  reality, for very highly  skilled players, discovering
+% the intentions of the  enemy is the key to winning.  It allows for a
+% compact understanding of the game  state and development, while also
+% giving additional information about the  future. Whereas a player is
+% doing action  A to  follow it  up by action  B or  C depends  on its
+% intention,  while the  information  about action  A  is visible  and
+% information about  B or  C is  hidden, only  intention allows  for a
+% choice.
+
+\subsubsection{Spatial and Temporal Reasoning}
+% TODO just a paragraph on exposing why there are temporal and spatial
+% reasoning in RTS AI.
+Spatial reasoning is  related to each aspect  of terrain exploitation.
+It  is  involved   in  tasks  such  as  building   placement  or  base
+expansion.  In the  former,  the player  needs  to carefully  consider
+building  positioning into  its  own  bases to  both  protect them  by
+creating  a wall  against invasions  and to  avoid bad  configurations
+where large units could be stuck. In base expansion, the player has to
+choose good available locations to build a new base, regarding its own
+position and  opponent's bases. Finally,  spatial reasoning is  key to
+tactical reasoning:  players need to  decide where to place  units for
+battle, favoring, for instance,  engagements when the opponent's units
+are lead into a bottleneck.
+
+% [Flo:  70% to hit high  ground units according to  Ben Weber's paper
+% [3], and 53.125% according  to BWAPI website. Who's right? Should we
+% add this in the paper?]
+
+% [Alberto: I  guess Weber's number comes  from StarCraft walkthroughs
+% like                                                           this:
+% http://www.supercheats.com/pc/walkthroughs/starcraftbroodwar-walkthrough01.txt,
+% there is also a chance to hit in even grounds (like
+% 98%). But since we cannot see the original code all of these numbers are approximations. In any case we can keep explaining the concept without numbers ;)]
+Another example of spatial reasoning in StarCraft is that it is always
+an advantage to  have own units on  high ground while the  enemy is on
+low ground,  since units on  low ground have  no vision onto  the high
+ground. %, and firing from low ground have a chance to hit units on high ground.
+%MP: took out last half sentence as it seems to be rather confusing.
+
+Analogously,  temporal  reasoning  is  key in  tactical  or  strategic
+reasoning.  For  example,  timing  attacks and  retreats  to  gain  an
+advantage. At a  higher strategic level, players need  to reason about
+when to  perform long-term impact  economic actions such  as upgrades,
+building  construction,  strategy  switching,  etc.  all  taking  into
+account  that the  effects of  these  actions are  not immediate,  but
+longer term.
+
+\subsubsection{Domain Knowledge Exploitation}
+In traditional board  games such as Chess,  researchers have exploited
+the  large  amounts  of  existing  domain  knowledge  to  create  good
+evaluation  functions  to be  used  by  alpha-beta search  algorithms,
+extensive opening books, or end-game tables. In the case of RTS games,
+it  is still  unclear how  the  significantly large  amount of  domain
+knowledge  (in the  forms or  strategy guides,  replays, etc.)  can be
+exploited by  bots. Most  work in  this area has  focused on  two main
+directions: on the one hand, researchers  are finding ways in which to
+hard-code existing  strategies into  bots, so that  bots only  need to
+decide  which strategies  to deploy,  instead of  having to  solve the
+complete  problem  of  deciding  which  actions  to  execute  by  each
+individual unit at each time step.  One the other hand, large datasets
+of  replays have  been created  \cite{WeberCig09,synnaeve2012dataset},
+from  where   strategies,  trends   or  plans   have  been   tried  to
+learn.  However,  StarCraft  games  are  quite  complex,  and  how  to
+automatically learn from such datasets is still an open problem.
+
+\subsubsection{Task Decomposition}
+For all the  previous reasons, most existing approaches  to play games
+as StarCraft  work by decomposing the  problem of playing an  RTS game
+into   a    collection   of    smaller   problems,   to    be   solved
+independently. Specifically, a common subdivision is:
+\begin{itemize}
+\item {\em  Strategy}: corresponds  to the high-level  decision making
+  process.  This is  the highest  level  of abstraction  for the  game
+  comprehension.  Finding an  efficient  strategy or  counter-strategy
+  against a given opponent is key in RTS games, and concerns the whole
+  set of units a player
+  owns. %If we may allow ourselves a military analogy, strategy is orders given by generals at the HQ.
+\item   {\em  Tactics}:   are  the   implementation  of   the  current
+  strategy.  It  implies  army and  building  positioning,  movements,
+  timing, and so on. Tactics concerns a group of
+  units. %They are orders given by officers on the battlefield.
+\item {\em Reactive  control}: is the implementation  of tactics. This
+  consists   in  moving,   targeting,  firing,   fleeing,  hit-and-run
+  techniques  (also  knows  as  ``kiting'')  during  battle.  Reactive
+  control focuses on a specific
+  unit. %It is soldiers applying officers' orders.
+%\end{itemize}
+%
+%\noindent
+%Both  strategy  and  tactics  are  usually  called
+%\textit{macro-management}, whereas reactive control is also called
+%\textit{micro-management}. In addition of the three tasks shown above,
+%one could also consider these two tasks that an AI has to deal with:
+%\begin{itemize}
+\item  {\em Terrain  analysis}: consists  in the  analysis of  regions
+  composing the map: choke-points,  minerals and gas emplacements, low
+  and high walkable grounds, islands, etc.
+\item  {\em   Intelligence  gathering}:  corresponds   to  information
+  collected  about the  opponent. Because  of the  fog-of-war, players
+  must regularly send scouts to localize and spy enemy
+  bases.%, and must see or infer what buildings and technology the enemy has developed in order to adapt its strategy.
+\end{itemize}
+
+In comparison, when humans play StarCraft, they typically divide their
+decision  making in  a  very different  way.  The StarCraft  community
+typically talks about two tasks:
+\begin{itemize}
+\item  {\em  Micro}: is  the  ability  to control  units  individually
+  (roughly corresponding to {\em Reactive  Control} above, and part of
+  {\em Tactics}). A good \emph{micro} player usually keeps their units
+  alive over a longer period of time.
+\item {\em  Macro}: is the ability  to produce units and  to expand at
+  the  appropriate times  to  keep your  production  of units  flowing
+  (roughly corresponding to everything  but {\em Reactive Control} and
+  part of {\em Tactics} above). A good \emph{macro} player usually has
+  the larger army.
+\end{itemize}
+
+The reader can find a good  presentation of task decomposition for AIs
+playing  RTS   in~\cite{weber2011acs}.  Although  the   previous  task
+decomposition  is  common, a  significant  challenge  is on  designing
+architectures so that  the individual AI techniques  that address each
+of  those  tasks  can   communicate  and  effectively  work  together,
+resolving conflicts, prioritizing resources between them, etc. Section
+\ref{sec:bot}  provides an  overview of  the task  decompositions that
+state-of-the-art bots use.  Moreover, we would like to  point out that
+the task decomposition  above is not the only  possible approach. Some
+systems, such  as IMAI  \cite{miles2007co}, divide gameplay  into much
+smaller  tasks, which  are then  assigned resources  depending on  the
+expected benefits of achieving each task.
 

### 2015-04-22 ECGG_StarCraftCompetitions/ECGG_StarCraftCompetitions.tex
Init SC Competitions paper

@@
 \usepackage{algpseudocode}
+\usepackage{adjustbox}
 
@@
 
-\section{Definition}
+\section{State of the Art Bots for StarCraft}\label{sec:bot}
 
-{\color{blue} to-do}
+Thanks to the recent organization of international game AI competitions focused around the popular StarCraft game (see Section \ref{sec:competition}), several groups have been working on integrating many of the techniques described in the previous section into complete ``bots'', capable of playing complete StarCraft games.
+In this section we will overview some of the currently available top bots.
+
+%\subsection{RTS Bot Architectures}\label{sec:integration}
+
+Playing an RTS game involves dealing with all the problems described above.
+A few approaches, like CAT \cite{LTW}, Darmok \cite{OntanonMSR10} or ALisp \cite{Marthi05} try to deal with the problem in a monolithic manner, by using a single AI technique.
+However, none of those systems aims at achieving near human performance.
+In order to achieve human-level performance, RTS AI designers use a lot of domain knowledge in order to divide the task of playing the game into a collection of sub-problems, which can be dealt-with using individual AI techniques.
+
+\begin{figure}[h!]
+  \begin{adjustbox}{addcode={\begin{minipage}{\width}}{\caption{%
+    %\centering
+            Architecture  of 7  StarCraft bots  obtained by  analyzing
+            their  source code.  Modules  with  black background  sent
+            commands  directly to  StarCraft, dashed  arrows represent
+            data flow, and solid arrows represent control.
+          }\end{minipage}},rotate=90,center}
+    \includegraphics[width=0.935\textheight]{figures/figure-bot-architectures-wide.pdf}
+    % \caption{Architecture of 7  StarCraft bots obtained by analyzing
+    % their  source code.  Modules with  black background  sent commands
+    % directly  to StarCraft,  dashed  arrows represent  data flow,  and
+    % solid arrows represent control.}
+  \end{adjustbox}
+    \label{fig:bot-architecture}
+\end{figure}
+
+Figure \ref{fig:bot-architecture} shows some representative examples of the architectures used by different bots in the AIIDE and CIG StarCraft AI competitions (see Section \ref{sec:competition}): BroodwarBotQ \cite{SynnaeveMicroCig11}, Nova \cite{uriarte2012kiting}, UAlbertaBot \cite{churchill2011build}, Skynet, SPAR, AIUR, and BTHAI \cite{Hagelback12}.
+Each box represents an individual module with a clearly defined task (only modules with a black background can send actions directly to StarCraft).
+Dashed arrows represent data flow, and solid arrows represent control (when a module can command another module to perform some task).
+For example, we can see how SPAR is divided in two sets of modules: {\em intelligence} and {\em decision making}.
+Intelligence in SPAR has three modules dedicated to analyze the current situation of the game.
+Decision making in SPAR is done through four hierarchically organized modules, with the higher-level module ({\em strategic decision}) issuing commands to the next module ({\em tactical decision}), which sends commands to the next module ({\em action implementation}), and so on.
+Only the two lower-level modules can send actions directly to StarCraft.
+
+On the other hand, bots such as Nova or BroodwarBotQ (BBQ) only use a hierarchical organization for {\em combat} (controlling the attack units), but use a decentralized organization for the rest of the bot.
+In Nova and BBQ, there is a collection of modules that control different aspects of the game (workers, production, construction, etc.).
+These modules can all send actions directly to StarCraft.
+In Nova those modules coordinate mostly through writing data in a shared blackboard, and in BBQ they coordinate only when they have to use a shared resource (unit) by means of an arbitrator: a bidding market and broker for settling units control, military and civilian groups/task forces bid for units proportionally to their usefulness and the task importance.
+
+By analyzing the structure of these bots, we can see that there are two main tools being used in these integration architectures:
+
+\begin{itemize}
+\item {\em Abstraction}: complex tasks can be formulated at different levels of abstraction.
+For example, playing an RTS game can be seen as issuing individual low-level actions to each of the units in the game, or at a higher level, it can be seen as deploying a specific strategy (e.g.
+a ``BBS strategy'', or a ``Reaver Drop'' strategy).
+Some bots, reason at multiple levels of abstraction at the same time, making the task of playing StarCraft simpler.
+Assuming that each module in the architecture of a bot has a goal and determines some actions to achieve that goal, the actions determined by higher-level modules are considered as the goals of the lower level modules.
+In this way, each module can focus on reasoning at only one level of abstraction, thus, making the problem easier.
+
+\item {\em Divide-and-conquer}: playing a complex RTS, such as StarCraft, requires performing many conceptually different tasks, such as gathering resources, attacking, placing buildings, etc.
+Assuming each of these tasks can be performed relatively independently and without interference, we can have one module focusing on each of the tasks independently, thus making the problem easier.
+\end{itemize}
+
+If we imagine the different tasks to perform in a complex RTS game in a two-dimensional plane, where the vertical axis represents abstraction, and the horizontal axis represents the different aspects of the game (combat, resource gathering, etc.), abstraction can be seen as dividing the space with horizontal lines, whereas divide-and-conquer divides the space using vertical lines.
+
+Different bots use different combinations of these two tools. Looking back at Figure \ref{fig:bot-architecture}, we can see the following use of abstraction and divide-in-conquer in the bots:
+
+\begin{itemize}
+\item
+  BroodwarBotQ\footnote{\url{http://github.com/SnippyHolloW/BroodwarBotQ}}:
+  uses  abstraction for  {\em combat}, and  divide-and-conquer for
+  {\em economy}  and  {\em intelligence  gathering}. To  avoid  conflicts
+  between modules (since  the individual tasks of each  of the modules
+  are not completely independent), BBQ uses an arbitrator.
+\item  Nova\footnote{\url{http://nova.wolfwork.com/}}:  is similar  in
+  design as  BroodwarBotQ, and uses  abstraction for {\em combat},
+  and  divide-and-conquer for  {\em economy}.  The differences  are
+  that Nova does not have  an arbitrator to resolve conflicts, but has
+  a  higher-level   module  ({\em  strategy   manager}),  which  posts
+  information to the blackboard that the rest of modules follow (thus,
+  making use of abstraction).
+\item
+  UAlbertaBot\footnote{\url{http://code.google.com/p/ualbertabot/}}:
+  also  uses abstraction  in  {\em combat} like  the previous  two
+  bots. But it  also uses it in {\em economy}: as  can be seen, the
+  production manager sends commands to the building manager, who is in
+  charge   of   producing   the   buildings.  This   bot   also   uses
+  divide-and-conquer, and  tasks like scouting  and resource gathering
+  are managed by separate, independent modules.
+\item       Skynet\footnote{\url{http://code.google.com/p/skynetbot/}}:
+  makes     extensive     use      of     both     abstraction     and
+  divide-and-conquer.  We  can see a  high level module  that issues commands  to a
+  series of  tactics modules. The  collection of tactic  modules queue
+  {\em  tasks} (that  are analogous  to the  abstract actions  used in
+  SPAR).  Each different  task has  a specific  low level  module that
+  knows how to  execute it. Thus, Skynet uses  a 3 layered abstraction
+  hierarchy,  and uses  divide-and-conquer  in all  levels except  the
+  highest.
+\item
+  SPAR\footnote{\url{http://www.planiart.usherbrooke.ca/projects/spar/}}:
+  only uses abstraction. Its high-level module determines the strategy
+  to  use,  and  the  tactical  decision  module  divides  it  into  a
+  collection  of {\em  abstract  actions}, that  are  executed by  the
+  lower-level modules.
+\item  AIUR\footnote{\url{http://code.google.com/p/aiurproject/}}:  is
+  mainly divide-and-conquer oriented,
+  with a  slight abstraction on  {\em economy} due to a  SpendManager deciding
+  how  to  spend  and  share  resources  among  Base,  Production  and
+  Construction Managers.  At the beginning of a  game, the MoodManager
+  initializes  a  ``mood''  which  will  influence  both  tactics  and
+  strategy. {\em Combat} is divided into three  independent managers: the
+  {\em  Defense Manager},  controlling  military units  when there  is
+  nothing special, the {\em  Under Attack Manager}, activated when the
+  opponent is attacking our bases,  and the {\em Army Manager}, taking
+  control of units when it is time to attack, following a timing given
+  by the  current mood. This bot  does not manage however  any kind of
+  reactive controls so far.
+\item  BTHAI\footnote{\url{http://code.google.com/p/bthai/}}:  uses  a
+  two-tier  abstraction hierarchy,  where a  collection  of high-level
+  modules command a collection of lower-level agents in charge of each
+  of  the units.  At  the high-level,  BTHAI uses  divide-and-conquer,
+  having   multiple  high-level  modules   issuing  commands   to  the
+  lower-level units.
+\end{itemize}
+
+Additionally, except for BTHAI, all other agents use divide-and-conquer at a higher-level bot design and divide all the modules into two or three categories: {\em intelligence gathering} and {\em decision making} (sometimes divided into {\em combat} and {\em economy}).
+
+Some bots  using divide-and-conquer, assume  that each of  the modules
+can act independently  and that their actions can  be executed without
+interference.  BBQ,  UAlbertaBot and  AIUR, however use  an arbitrator
+({\em Game Commander'} in UAlbertaBot) that makes sure that modules do
+not send contradictory orders to  the same unit.  However, very little
+bots handle  the problem of  how to coordinate resource  usage amongst
+modules, for  instance BTHAI uses a  first-come-first-serve policy for
+spending resources,  the first module  that requests resources  is the
+one that gets them. Nova and Skynet are exceptions, and implement some
+rudimentary prioritization based on the high level strategy. Following
+available  resources and  timing,  AIUR's {\em  Spend Manager}  orders
+Base,  Production   and  Construction  Managers  what   they  have  to
+build/produce. It also orders to start tech research and upgrades. The
+idea here is not to let the different managers allocate the resources they want, but to do the opposite, that is: finding how the AI can
+spend the available money.
+
+One  interesting aspect  of the  seven bots  described above  is that,
+while all  of them (except AIUR) are  reactive at the  lower level
+(reactive  control), most  if not  all of  them, are  scripted  at the
+highest level  of abstraction. BTHAI reads build  and squad formations
+from  a  predefined  script,   Nova's  {\em  Strategy  Manager}  is  a
+predefined finite-state machine,  BBQ's construction manager reads the
+build  order from a  predefined script,  and Skynet's  {\em BuildOrder
+  Manager} is basically a predefined script. Such scripts describe the
+strategy  that the  bots will  use, however,  such strategy  is always
+fixed.   One could see  this pre-scripting  as if  each bot  defined a
+``high-level programming language''  to describe StarCraft strategies,
+and   the   bots   themselves    are   just   interpreters   of   such
+strategy.  Compared  to  current  approaches  for Chess  or  Go,  this
+scripting  seems a  rigid and  inflexible,  but responds  to the  much
+higher complexity of the  StarCraft game.  An interesting exception to
+that  is  UAlbertaBot, which  uses  a  search  algorithm in  the  {\em
+  Production  Manager}  to find  near-optimal  build orders.   Another
+interesting case is  AIUR, that uses a {\em  Mood Manager} to randomly
+pick  a mood among  six (cheese,  rush, aggressive,  defensive, macro,
+fast  expand), which  will  influence the  build  order, strategy  and
+tactics.
+% All of  them  scripted so  far,  but build  orders can  been modified but  the Spend Manager, following  available resources.
+This mood can change during the game, regarding the opponent behavior.
+
+In conclusion, we can see that there are two basic tools that can be used in an integration architecture: abstraction and divide-and-conquer, which are widely used by the existing StarCraft bots.
+For space reasons, we do not include an exhaustive comparison of the architectures of all the participating bots.
+Some other bots have been documented by their authors, such as SCAIL \cite{YoungSCAIL} or QUORUM \cite{young2012evolutionary}.
+%This section has focused on discussing the architecture of existing StarCraft bots. 
+Let us now focus on their performance.
+
+%\subsection{Individual AI Techniques}\label{sec:techniques}
+
+%{\color{red} TODO}
+
+\section{Recent StarCraft AI Competitions}\label{sec:competition}
+
+This section reviews the results of the recent international competitions on AI for StarCraft.
+These competitions, typically co-located with scientific conferences, have been possible thanks to the existence of the Brood War Application Programming Interface
+(BWAPI)\footnote{\url{http://code.google.com/p/bwapi/}}, which 
+enables replacing the human player interface with C++ code.
+%A consequence of this indirect way of integrating new bots is that every machine can only run one custom bot, requiring two computers for a 2-player game.
+The following subsections summarize the results of all the StarCraft AI competitions held at the AIIDE (Artificial Intelligence for Interactive Digital Entertainment) and CIG (Computational Intelligence in Games) conferences during the past years.
+Additionally we analyze the statistics from the StarCraft Bot Ladder, where the best bots play against each other continuously over time.
+
+\subsection{AIIDE}\label{sec:AIIDE}
+
+Started in 2010, the AIIDE StarCraft AI Competition\footnote{\url{http://www.StarCraftAICompetition.com}} is the most
+well known and longest running StarCraft AI Competition in the world. Each year,
+AI bots are submitted by competitors to do battle within the retail version of
+StarCraft: Brood War, with prizes supplied by Blizzard Entertainment.
+%This competition has been made possible by the BWAPI StarCraft programming interface, which allows players to write programs which can retrieve game data and issue game commands to StarCraft via an easy to use C++ API.
+
+The first competition in 2010 was organized and run by Ben Weber in the Expressive
+Intelligence Studio at University of California, Santa Cruz\footnote{\url{http://eis.ucsc.edu/StarCraftAICompetition}}. 26 total submissions
+were received from around the world. As this was the first year of the competition,
+and little infrastructure had been created, each game of the tournament was run 
+manually on two laptop computers and monitored by hand to record the results. Also,
+no persistent data was kept for bots to learn about opponents between matches.
+
+The 2010 competition had 4 different tournament categories in which to compete. Tournament 1
+was a flat-terrain unit micro-management battle consisting of four separate unit
+composition games. Of the six competitors, FreSCBot won the competition with
+Sherbrooke coming in 2nd place. Tournament 2 was another micro-focused game with
+non-trivial terrain. Two competitors submitted for this category, with FreSCBot
+once again coming in 1st by beating Sherbrooke.
+
+Tournament 3 was a tech-limited StarCraft game on a single known map with no fog-of-war enforced. Players
+were only allowed to choose the Protoss race, with no late game units allowed. 8 bots
+faced off in this double-elimination tournament with MimicBot taking first place over Botnik in the final.
+As this was a perfect information variant of StarCraft, MimicBot adopted a strategy of
+``mimic its opponent's build order, gaining an economic advantage whenever possible'' which
+worked quite well.
+
+Tournament 4 was the complete game of {\em StarCraft: Brood War} with fog-of-war enforced. The tournament
+was run with a random pairing double-elimination format with each match being best of 5 games.
+Competitors could play as any of the three races, with the only limitations in gameplay
+being those that were considered ``cheating'' in the StarCraft community. A map pool of
+5 well-known professional maps were announced to competitors in advance, with a random map being chosen for each game.
+
+Results are shown in Table \ref{tab:aiide2010}. The team that won was Overmind\footnote{\url{http://overmind.cs.berkeley.edu}}, from University of California, Berkeley. Using the Zerg race, their
+strategy was to defend early aggression with zergling units while amassing mutalisk units,
+which they used to contain and eventually defeat their opponents. The mutalisk is a very fast and agile
+flying unit which is able to attack while moving with no drawback, which makes them quite a powerful
+unit when controlled by a computer. Overmind used a potential-field
+based micro-management system to guide their mutalisks, which led them to victory. Krasi0 came in 2nd place
+with a standard defensive Terran opening strategy that transitioned into ``mech'' play in the late game.
+
+\begin{table}[!t]
+\caption{Ranking of the three best bots of the AIIDE 2010 competition}
+\label{tab:aiide2010}
+\centering
+\begin{tabular}{|c|l|}
+\hline
+{\bfseries Position} & {\bfseries Bot}\\
+\hline
+1 & Overmind \\
+2 & Krasi0 \\
+3 & Chronos \\
+\hline
+\end{tabular}
+\end{table}
+
+In 2011 the University of Alberta hosted the competition, with organization by Michael Buro and
+David Churchill\footnote{\url{https://skatgame.net/mburo/sc2011/}}. Due to a lack of entrants in tournament categories 1-3 in the 2010 competition, it was
+decided that only the full game category would be played in the 2011 competition. Another important
+change in the 2011 competition was the introduction of automated tournament-managing software running
+StarCraft games simultaneously on 20 computers, allowing a total of 1170 games to be played in far less
+time than the 108 games of the 2010 competition. This increase in games played also allowed the tournament
+to switch to a round-robin format, eliminating the ``luck'' factor of the pairings inherent in bracket
+style tournaments. The bot that achieved the highest win percentage over the course of the competition would
+be determined the winner. Also, the competition became open-source, in an effort not
+only to prevent possible cheating, but to promote healthy competition in future tournaments by giving
+newcomers and easier entry point by basing their design off of previous bots.
+
+In the  end, Skynet  won the competition  with its solid  Protoss play
+(results  are  summarized   in  Table  \ref{tab:aiide2011}).  The  bot
+executed one of a small set of strategies randomly at the start of the
+match based on the map and the race of the opponent. Skynet would then
+amass a  medium to large  sized army and  expand before moving  out to
+attack. Good  use of Dragoon (powerful ranged ground unit with clumsy movement) range and  kiting micro-management allowed
+it to hold off the early aggression of other bots such as UAlbertaBot,
+which came in 2nd.
+
+\begin{table}[!b]
+\caption{Results of the five best bots of the AIIDE 2011 competition}
+\label{tab:aiide2011}
+\centering
+\begin{tabular}{|c|l|r|}
+\hline
+{\bfseries Position} & {\bfseries Bot} & {\bfseries Win \%} \\
+\hline
+1 & Skynet & 88.9\% \\
+2 & UAlbertaBot & 79.4\% \\
+3 & AIUR & 70.3\% \\
+4 & ItayUndermind & 65.8\% \\
+5 & EISBot & 60.6\% \\
+\hline
+\end{tabular}
+\end{table}
+
+UAlbertaBot used an early zealot-rush strategy to take advantage of the power of early game Protoss units. 
+It would send out the first zealots that were made and immediately attack the enemy base, using a unit
+counting heuristic to determine whether or retreat or keep pushing. Of note is that UAlbertaBot used an
+online planning algorithm to construct all of its economic build-orders\cite{churchill2011build}, as no hard-coded build orders
+were used.
+
+AIUR also chose  Protoss, with a strategy that was in between Skynet
+and UAlbertaBot in terms of  attack timings.  At that time, AIUR chose
+one  mood among  five (leading  to slightly  different  strategies and
+tactics) at the  beginning of a game and kept it  until the end. These
+five moods were:
+\begin{itemize}
+\item  Rush:  where  the  bot  tries  early  attacks,  and  have  good
+  probabilities to send the two or three first Zealots (basic contact attack ground unit) to harass the
+  opponent.
+\item Aggressive: where  we have less chance to  perform harasses with
+  the first Zealots, and the first attack is usually a bit delayed
+  with regard to the Rush mood.
+\item Macro: where the AI do not try any early attacks and focus a bit
+  more on its economy before attacking.
+\item Defense: where the AI  ``turtles'' and wait to have a consequent
+  army before running an attack.
+\item  Fast expand:  where the  first building  constructed it  a base
+  expansion, for a very economical-oriented game.
+\end{itemize}
+Notice that  build orders are not  fully hard-coded since  they can be
+altered by AIUR's Spend Manager.
+
+Of  note  in  these  results  was that  a  rock-paper-scissors  effect
+happened among  the top  3 finishers.  Of  the 30 rounds,  Skynet beat
+UAlbertaBot 26  times, UAlbertaBot beat  AIUR 29 times, and  AIUR beat
+Skynet  19 times.   Another notable  result is  that Overmind  did not
+choose  to compete despite  winning the  2010 competition.   After the
+competition,  many  bot  programmers  (including  the  Overmind  team)
+realized that their  2010 strategy was quite easily  defeated by early
+game rushing strategies,  and so they submitted a  Terran bot instead,
+called Undermind, which finished in 7th.
+
+After the competition was over, a man vs. machine match was held between the winner (Skynet) and an
+ex-professional StarCraft player named Oriol Vinyals. Oriol was a competitor in the 2001 World Cyber
+Games StarCraft competition, and though he had been out of practice for a few years was still quite a
+good player. The match was arranged to see how well StarCraft AI bots had progressed and to see if
+they could actually beat a decent human opponent.
+
+\begin{figure}[t!]
+    \centering
+    \includegraphics[width=\columnwidth]{figures/aiide2012_v2}
+    \caption{Evolution of the win percentage of each bot participating in the AIIDE 2012 competition}
+    \label{fig:aiide2012}
+\end{figure}
+
+For the best-of-three match, Oriol chose his races randomly and ended up beating Skynet in a 2-0. 
+In the first match, Oriol played Zerg vs. Skynet's Protoss on Python, a four player map. Oriol chose
+to start with a fast expansion strategy and transition into two base mutalisk production. Skynet chose
+to rush with a few early zealots, which was luckily the best possible choice given Oriol's strategy.
+Skynet's initial attack destroyed Oriol's early defenses, and nearly won the game in the first few
+minutes, however it then proceeded to send zealots to attack one at a time rather than group up its
+units before moving in, which allowed Oriol to catch up. Once Oriol produced his mutalisks, Skynet did
+not produce sufficient air defenses and Oriol quickly destroyed Skynet's base. In the second game,
+Oril played Terran, again on Python. After holding off early Dragoon pressure from Skynet, Oriol
+moved out with a few marines, medics and tanks. Skynet tried to defend with its army of Dragoons,
+however due to poor unit targeting decisions it started to attack useless medics after the marines
+had died, rather than the tanks. Oriol overcame the Dragoon army and was victorious. Later analysis
+of the match concluded that Skynet, while dominant over the other bots, was unable to properly
+adapt and transition into a mid-game strategy in game one once its early pressure failed, and in game
+two made a key blunder in unit targeting which cost it the game. Humans were still in command.
+
+\begin{table}[!t]
+\caption{Results of the five best bots of the AIIDE 2012 competition}
+\label{tab:aiide2012}
+\centering
+\begin{tabular}{|c|l|r|}
+\hline
+{\bfseries Position} & {\bfseries Bot} & {\bfseries Win \%} \\
+\hline
+1 & Skynet & 84.4\% \\
+2 & AIUR & 72.2\% \\
+3 & UAlbertaBot & 68.6\% \\
+4 & BroodwarBotQ & 59.1\% \\
+5 & AdjutantBot & 52.8\% \\
+\hline
+\end{tabular}
+\end{table}
+
+The University of Alberta also hosted the 2012 competition, with the major difference from the 2011
+competition being the addition of persistent storage. Bots could now write information to disk during a
+match, and then read the information during other matches, allowing them to adjust strategies based
+on previous results. 6 of the 10 entrants used this feature to aid in strategy selection, including the
+top 4 finishers. More improvements to the tournament environment also meant that a total of 4240 games
+could now be played in the same time period. Results are shown in Table \ref{tab:aiide2012}.
+
+Skynet once  again won  the competition with  its solid  Protoss build
+orders  and  good  Dragoon  kiting.   AIUR  and  UAlbertaBot  switched
+positions from  the previous  year to come  2nd and  3rd respectively.
+Both  AIUR  and UAlbertaBot  used  data  stored  from the  results  of
+previous games  to select a strategy for  future matches.  UAlbertaBot
+did  this  using  the  UCB  \cite{auer2002finite} algorithm,  while  AIUR  used  a  uniform
+distribution  to choose  its  mood before  altering this  distribution
+after  some  games  against  the  same  opponent  to  favor  efficient
+strategies, achieving  similar results than  UAlbertaBot. Notice that,
+compared  to AIIDE 2011,  AIUR proposes  a new  mood, \textit{Cheese},
+implementing a  Photon Cannon rush  strategy in order to  surprise the
+opponent and  to finish the game  as soon as possible.   The effect of
+this strategy selection process can be seen Figure \ref{fig:aiide2012}
+which shows bot  win percentages over time.  While  the earlier rounds
+of the tournament fluctuated wildly in results, eventually the results
+converged to their  final values. One of the main  reasons for this is
+due to  the bots  learning which strategies  to use as  the tournament
+progressed.
+
+The 2012 man vs. machine match again used the winner of the competition (Skynet), who played against
+Mike Lange, also known as Bakuryu. At the time of the match, Bakuryu was an A- ranked Zerg player on ICCup,
+and known as one of the best non-Korean Zerg players in the world. Bakuryu was considered much
+stronger than Oriol at the time that the match was played, and the results showed that this was true.
+In the first game of the best-of-three, Bakuryu made Skynet look quite silly by running around inside
+Skynet's base with a small number of zerglings while Skynet's zealots and half of its worked chased
+then in vain. After killing off several probes and buying enough time to set up his expansion, he
+cleaned up Skynet's army with a flying army of mutalisks. In the second game, Bakuryu contained Skynet
+inside its base with a group of zerglings positioned within Skynet's expansion. Skynet then constructed
+several Dark Templar and along with some Dragoons and Zealots attacked into Bakuryu's expansion which
+was heavily defended, and was crushed almost instantly, allowing Bakuryu's zergling force to finish
+off the Protoss base.
+
+In this match it was shown that the true weakness of state of the art StarCraft AI systems was that
+humans are very adept at recognizing scripted behaviors and exploiting them to the fullest. A human
+player in Skynet's position in the first game would have realized he was being taken advantage of and
+adapted his strategy accordingly, however the inability to put the local context (Bakuryu kiting his
+units around his base) into the larger context of the game (that this would delay Skynet until
+reinforcements arrived) and then the lack of strategy change to fix the situation led to an easy
+victory for the human. These problems remain as some of the main challenges in RTS AI today: to both recognize
+the strategy and intent of an opponent's actions, and how to effectively adapt your own strategy to 
+overcome them.
+
+All results, videos, and replays from the AIIDE StarCraft AI Competition can be found in \url{http://www.StarCraftAICompetition.com}.
+% \footnote{\url{http://www.StarCraftAICompetition.com}}.
+
+\begin{table*}[!t]
+\caption{Results of the first round at CIG 2011, held in two brackets.
+Qualified for the final round: UAlbertaBot and Skynet (from A), Xelnaga
+and BroodwarBotQ (from B, the latter by comparing direct encounters
+with BTHAI of which 6:4 were won)}
+\label{tab:cig-first-round}
+\centering
+% bracket A
+\begin{tabular}{|c|c|c|l|r|}
+\hline
+\multicolumn{5}{|c|}{Bracket A} \\ \hline
+{\bfseries Position} & {\bfseries Crashes} & {\bfseries Games} & {\bfseries Bot} & {\bfseries Win \%} \\
+\hline
+A1 & 0 &   40 &  UAlbertaBot & 82.5\% \\
+A2 & 1 &     40 &  Skynet   &  77.5\% \\
+A3 & 2 &   40 &  AIUR   &  60.0\% \\
+A4 & 1 &   40 &  Nova   &  20.0\% \\
+A5 & 0 &   40 &  LSAI   &  10.0\%\\
+\hline
+\end{tabular}
+% bracket B
+\begin{tabular}{|c|c|c|l|r|}
+\hline
+\multicolumn{5}{|c|}{Bracket B} \\ \hline
+{\bfseries Position} & {\bfseries Crashes} & {\bfseries Games} & {\bfseries Bot} & {\bfseries Win \%} \\
+\hline
+B1 & 12 &  40 &  Xelnaga &   62.5\%\\
+B2 & 3 &   40 &  BroodwarBotQ  &  57.5\%\\
+B3 & 0 &   40 &  BTHAI    &  57.5\%\\
+B4 & 17 &  40 &  Protoss Beast Jelly  & 42.5\%\\
+B5 & 0 &   40 &  EvoBot   &  30.0\%\\
+\hline
+\end{tabular}
+\end{table*}
+
+\subsection{CIG}
+\label{sec:cig2011}
+
+An initial attempt to run a StarCraft tournament at the Computational
+Intelligence in Games conference (CIG 2010) suffered from technical problems.
+These mainly stemmed from the desire to use evolved, largely untested
+maps which proved to look interesting but made the submitted bots and the Brood War Terrain Analyzer (BWTA) provided with the BWAPI interface crash
+so frequently that it would have been unjustifiable to announce a winner.
+
+At CIG 2011, the tournament was therefore run with a (secret) selection
+of maps used in league play, which can be regarded as the most important
+difference to the AIIDE tournament that employed a known list of maps.
+The competition was organized by Tobias Mahlmann and Mike
+Preuss and attracted 10 bots. In addition to the ones discussed in previous
+sections (UAlbertaBot, Skynet, AIUR, Nova, BroodwarBotQ, BTHAI), 
+the set also contained LSAI, Xelnaga, Protoss Beast Jelly, and EvoBot,
+these are shortly described in the following:
+
+\paragraph*{LSAI (Zerg)} utilizes a heavily modified BWSAL\footnote{\url{https://code.google.com/p/bwsal/}} to divide management
+ of the units to different modules that communicate via a centralized information
+ module. It works using a simple reactive strategy to try and survive early game
+ attacks and macro up to a larger attack force and maintain map control.
+
+\paragraph*{Xelnaga (Protoss)} is a modification of the AIUR bot that chooses the 
+Dark Templar Opening in order to destroy the enemy base before defenses against
+invisible units are available.
+
+\paragraph*{Protoss Beast Jelly (Protoss)}
+always goes for a 5-gate Zealot rush, supported by an effective harvesting 
+strategy named power-mining (2 probes are assigned to every mineral patch,
+thereby needing 18 probes for 100\% saturation in a normal map, prior
+to expanding). Gas is not mined as it is not needed for constructing Zealots.
+
+\paragraph*{EvoBot (Terran)} employs an evolutionary algorithm for obtaining rational
+ unit combinations and influence map techniques for deciding the strategic locations. Note
+ that this bot was submitted in a very early version, with many of its designed features not
+ yet fully ready.
+
+\subsubsection{First Round}
+\label{sec:cig-first-round}
+
+\begin{figure}[t]
+	%\vspace*{-1ex}
+    \centering
+    \includegraphics[width=\columnwidth]{figures/cig2012-ResultsRound90.pdf}
+	%\vspace*{-1ex}
+    \caption{Evolution of the win percentage of each bot participating in the
+CIG 2012 competition}
+    \label{fig:cig2012-results}
+\end{figure}
+
+\begin{figure*}[b]
+	%\vspace*{-2ex}
+    \centering
+    \includegraphics[width=0.32\textwidth]{figures/vstable3.pdf}
+    \includegraphics[width=0.32\textwidth]{figures/vstable6.pdf}
+    \includegraphics[width=0.32\textwidth]{figures/vstable.pdf}
+	%\vspace*{-4ex}
+    \caption{Win percentages of CIG 2012 competition, from left
+    to right: 3-player maps only, 6-player maps only, all maps.
+    Read from line to column, bot in row wins given fraction of
+    games against bot in column. For some bots, we find interesting
+    differences, e.g. Xelnaga gets worse on 6-player maps, UAlbertaBot
+    gets better. Only Xelnaga can reliably beat Skynet, but only on 
+    3-player maps}
+    \label{fig:cig2012-mapresults}
+\end{figure*}
+
+As the CIG competition games were executed manually due to
+a lack of available software (the AIIDE program was not yet
+available at that time), the organizers separated the ten entries into
+two brackets. In each bracket of 5 bots, a round-robin
+tournament was held with 10 repetitions per pairing, resulting
+in 40 games per bot.
+The 5 maps chosen for the first round were selected from the pool
+of well-known league play maps found on the Internet:
+\emph{(2)MatchPoint 1.3}, \emph{(4)Fighting Spirit 1.3}, 
+\emph{iCCup Destination 1.1}, \emph{iCCup Gaia}, and 
+\emph{iCCup Great Barrier Reef}. Each bot pairing played
+on every map twice, with switched starting positions.
+
+The two top bots of every bracket qualified
+for the final round. Table~\ref{tab:cig-first-round} summarizes
+the results.
+Note that as BroodwarBotQ and BTHAI have the same number of wins,
+their direct encounter was evaluated which accounted 6:4 for the BroodwarBotQ.
+The bots going into the final were thus UAlbertaBot, Skynet (from bracket A)
+and Xelnaga and BroodwarBotQ (from bracket B). All qualified bots play the
+Protoss faction. Most bots proved pretty stable, only Xelnaga and Protoss 
+Beast Jelly crashed relatively often (each in more than a quarter of the games). 
+Crashing of course resulted in an instant win for the other bot.
+In some cases, neither bot was able to finish the other off completely,
+so that they went into a passive state. We manually ended such games after
+around 15 minutes and assigned victory to the bot that had obtained more
+points as indicated on the end game screen.
+
+\begin{table}[!b]
+\caption{Results of the CIG 2011 competition}
+\label{tab:cig-final-round}
+\centering
+\begin{tabular}{|c|c|c|l|r|}
+\hline
+{\bfseries Position} & {\bfseries Crashes} & {\bfseries Games} & {\bfseries Bot} & {\bfseries Win \%} \\
+\hline
+1 & 0 &  30 &  Skynet       &  86.7\%\\
+2 & 0 &  30 &  UAlbertaBot  &  73.3\%\\
+3 & 3 &  30 &  Xelnaga      &  36.7\%\\
+4 & 2 &  30 &  BroodwarBotQ   &  3.3\%\\
+\hline
+\end{tabular}
+\end{table}
+
+\subsubsection{Final Round}
+\label{sec:cig-final-round}
+
+The final round was played in a similar mode as each of the
+first round brackets,
+using another set of 5 previously unknown maps:
+\emph{iCCup lost temple 2.4}, \emph{iCCup rush hour 3.1},
+\emph{iCCup swordinthemoon 2.1}, \emph{iCCup yellow 1.1},
+and \emph{La\_Mancha 1.1}. 
+Letting each pairing play on each map twice again with
+switching starting positions resulted in 30 games per bot.
+The final results are displayed in table~\ref{tab:cig-final-round},
+indicating Skynet as winner and UAlbertaBot as runner-up, being
+almost equally strong, and the two other bots as clearly inferior.
+The competition setup, documentation and results can be found
+in\footnote{\url{http://ls11-www.cs.tu-dortmund.de/rts-competition/StarCraft-cig2011}}.
+
+For CIG 2012, the AIIDE tournament software was employed,
+leading to a total of $4050$ games played in $90$ rounds
+of round robin. As 6 different maps were used, this means
+that each bot played every other on every map 15 times. 
+As in the AIIDE competition, writing to and reading from
+a bot specific directory was enabled, however, due to 
+technical reasons, this feature was constrained to the
+computer (of 6) the game was actually run on. We can 
+therefore assume that this feature was of minor use for
+the CIG competition. The only other difference to the
+AIIDE competition was that the used maps were not 
+made available to the competitors in advance.
+
+These maps came in two flavors, namely three 3-player maps:
+\emph{Athena-II}, \emph{Neo Moon Glaive}, \emph{Tears of the Moon},
+and three 6-player maps:
+\emph{Legacy}, \emph{River of Light}, and 
+\emph{The Huntress 1.1}. 
+We shall note that some bots consistently crashed on one
+of the originally considered maps which has thus been replaced.
+This is surprising as all maps are well known league play maps
+or have been provided with the StarCraft Brood War distribution
+itself.
+Setup, replays and results for the CIG 2012 competition can be found
+here\footnote{\url{http://ls11-www.cs.tu-dortmund.de/rts-competition/StarCraft-cig2012}}.
+
+\begin{table}[t]
+\caption{Results of the CIG 2012 competition.}
+\label{tab:cig2012}
+\centering
+\begin{tabular}{|c|l|r|}
+\hline
+{\bfseries Position} & {\bfseries Bot} & {\bfseries Win \%} \\
+\hline
+1 & Skynet & 78.3\% \\
+2 & UAlbertaBot & 65.2\% \\
+3 & AIUR & 60.4\% \\
+4 & Adjutant & 58.6\% \\
+5 & Nova & 52.4\% \\ 
+\hline
+\end{tabular}
+\end{table}
+
+\begin{figure*}[tb]
+    \centering
+    \includegraphics[width=\textwidth]{figures/botLadder1.pdf}
+    \caption{Bot's Elo Rating from February 1, 2012 to June 20, 2012}
+    \label{fig:botLadder1}
+\end{figure*}
+
+\begin{figure*}[tb]
+    \centering
+    \includegraphics[width=\textwidth]{figures/botLadder2.pdf}
+    \caption{Bot's Elo Rating from September 1, 2012 to January 10, 2013}
+    \label{fig:botLadder2}
+\end{figure*}
+
+The overall results are displayed in table~\ref{tab:cig2012},
+and the win rate evolution over time in figure~\ref{fig:cig2012-results}.
+These are quite consistent with the results of the AIIDE 2012 competition,
+so that we can conclude that the best bots are not very dependent on 
+knowing the maps beforehand. However, the bot vs. bot win rates as
+displayed in figure~\ref{fig:cig2012-mapresults} show some interesting
+trends. On the maps with more possible start points, some bots do 
+better than others, namely SCAIL, Adjutant, Nova, and UAlbertaBot,
+the latter probably due to its very efficient scouting routine.
+Some bots however suffer from the increased uncertainty about
+the enemies' position, namely Xelnaga and BroodwarBotQ.
+
+As already observed before in the previously described competitions,
+there are also bots who consistently beat top ranked bots but have
+severe problems against lower ranked bots. E.g., Xelnaga is especially
+strong against Skynet on the 3-player maps (about 70\% wins). Reviewing
+the replays led to the assumption that Xelnaga usually tries to attack
+Skynet's probes with a dark templar strategy, and often succeeds. 
+Nova does very well against the UAlbertaBot, and the replays show
+that it sometimes succeeds to lure the probes into its own base, where
+they get killed, leading to severe resource problems. However, we cannot
+tell how often this happens as this would require to review every single
+replay between the 2 bots. Summarizing, most bots seem to have improved,
+which becomes clear if the nearly unchanged BTHAI bot is taken as a baseline.
+In 2011, it won more than half of its qualifying games, in 2012 it came
+out last with around 20\% wins. However, designing a bot in order to 
+beat a top bot (as for Xelnaga with Skynet) leads to a very restricted
+strategy that often leads to failure if playing against different bots.
+Note that in the direct encounter between Xelnaga and AIUR, its ancestor,
+Xelnaga looses consistently.
+
+Nevertheless, from the observations we made during the tournament,
+we can draw the conclusion that the available bots are still 
+very constrained. No bot in the competition played the Zerg race,
+which is surprising as the AIIDE 2010 winner (Overmind) did so.
+Presumably, implementing a good Zerg strategy is more demanding
+than implementing one for the Protoss or Terran races. Many bots consistently crashed
+when playing against a random race built-in bot for testing, and
+also did so when the map size was changed from $128\times 128$ to 
+any other. Furthermore, every single bot sometimes failed to finish
+off an already beaten opponent, such that the game had to be stopped
+after a previously determined maximum time. It also seems that most of
+the current bots are not very good at adapting their strategy to the
+one of their opponent during a game, or at least (via the read/write
+procedure of game information) within a series of games.
+
+\subsection{StarCraft Bot Ladder}\label{sec:ladder}
+
+The StarCraft Bot Ladder is a website\footnote{\url{http://bots-stats.krasi0.com}} where bot versus bot matches are automatized, and are running all the time.
+This ladder is a great resource for creating data sets (all the game replays are available) and statistics.
+For bot ranking, the ladder uses an Elo rating system suitable for calculating the \emph{relative skill level} of a bot in two-player games.
+In the Elo system each player has a numerical rating that gets incremented or decremented some points after each game.
+The amount of points depends on the difference in the ratings of the players.
+A player will gain more points by beating a higher-rated player than by beating a lower-rated player.
+This kind of rating system is widely used in games like chess.
+This bot ladder compiles different versions of bots from the main worldwide competitions (like AIIDE, CIG or SSCAI\footnote{\url{http://sscaitournament.com}}), even some independent or ``under construction'' bots.
+Therefore, it is a very good resource to test the performance of new bots against the current state of the art in StarCraft bots before participating in the official competitions.
+
+Figure \ref{fig:botLadder1} shows the Elo rating of the bots in the ladder during the first half year of 2012.
+The ranking is practically equal than the AIIDE 2011 competition, showing the lack of adaptability of the current bots.
+We can notice these more extremely in Figure \ref{fig:botLadder2} for the second half year of 2012.
+During this period new bots were introduced in the ladder.
+We can observe how the first version of KillerBot made a huge impact on Skynet ranking and finally, the second version of KillerBot quickly became the best bot for more than one month (again we can see how the rest of the bots aren't able to adapt and the ranking doesn't change so much).
+And finally, in January, the Ximp bot appears with a new strategy that overcomes the rest of the bots.
+Both KillerBot and Ximp use hard-coded strategies without any kind of adaptation capabilities.
+However, they implement strategies that no other bot has a counter for, and thus manage to win a very large percentage of games.
+This points out, once again, that one of the major open challenges in RTS game AI is how achieving adaptive strategies, that can recognize the opponent's intentions, and select an adequate response.
+
+\subsection{Adaptation Analysis}
+
+One of the major conclusions from the results of the StarCraft competitions is the lack of adaptation of bots.
+Some switch between different build-orders, but do not fully adapt their strategy.
+No bot is capable of observing the opponent and autonomously synthesize a good plan from scratch to counter the opponent strategy.
+In this section we analyzed this claim quantitatively using the tools of \cite{SynnaeveOpeningCig11}.
+We analyzed the replays from the 2011 and 2012 AIIDE competitions (shown in Figures \ref{fig:aiide2011-botopenings} and \ref{fig:aiide2012-botopenings} respectively).
+We analyzed the way bots choose their openings depending on which other bot they are playing against.
+Given that there is no dominant strategy in StarCraft, and that it is necessary to see what the opponent is doing in order to determine the best opening, we would expect bots to change their openings depending on which other bot they are playing (since each bot uses a different strategy, or set of strategies).
+Using clustering, we identified the most common openings in all the replays from the 2011 and 2012 competitions (each one shown with a different color in the figures).
+No data is shown for the ItayUnvermind bot, since its opening did not match significantly with any of the ones used in our study (extracted from humans pro-gamers).
+
+\begin{figure*}[tb]
+    \centering
+    \includegraphics[width=\textwidth]{figures/botopenings2011.png}
+    \caption{Distribution of different openings performed by the different bots participating in the AIIDE 2011 competition.
+For each bot match-up, the colors show the proportion of times that the column bot used a particular opening against the row bot.}
+    \label{fig:aiide2011-botopenings}
+\end{figure*}
+
+\begin{figure*}[tb]
+    \centering
+    \includegraphics[width=\textwidth]{figures/botopenings2012.png}
+    \caption{Distribution of different openings performed by the different bots participating in the AIIDE 2012 competition.
+For each bot match-up, the colors show the proportion of times that the column bot used a particular opening against the row bot.}
+    \label{fig:aiide2012-botopenings}
+\end{figure*}
+
+Specifically, we identified the following openings (for a better comprehension of strategies, buildings or units of StarCraft, we refer the reader to Teamliquid's wiki\footnote{\url{http://wiki.teamliquid.net/starcraft/Category:Strategies}}):
+\begin{itemize}
+  \item Protoss openings:
+  \begin{itemize}
+    \item \emph{two\_gates}: Build two \emph{Gateways} and keep training \emph{Zealots} (basic contact attack ground unit), this is the quickest way to apply pressure.
+    \item \emph{fast\_dt}: Produce \emph{Dark Templars} (technologically advanced stealth ground unit) as soon as possible, sacrificing early game power for a technological advance, hard-countered by detectors technology.
+    \item \emph{templar}: Train \emph{High Templars} (technologically advanced zone attack unit) as fast as possible, same as above, less deadly but is less easily countered.
+    \item \emph{speedzeal}: Train \emph{Zealots} and research attack and speed upgrades as soon as possible, some early game power transitioning into late game tech.
+    \item \emph{corsair}: Produce \emph{Corsairs} (air-air flying unit) as soon as possible and then transition into training \emph{Dark Templars} (safe from Zerg's flying detectors thanks to \emph{Corsairs}), \emph{Reavers} (ground artillery unit) or \emph{Dragoons}.
+Weak early game.
+    \item \emph{nony}: Build three \emph{Gateways} and massive training of \emph{Dragoons}. Slower than \emph{two\_gates} but still some early game (ranged) power.
+    \item \emph{reaver\_drop}: Train \emph{Reavers} as soon as possible to be able to do drops (air transport of artillery units).
+  \end{itemize}
+  \item Terran openings:
+  \begin{itemize}
+    \item \emph{bio}: Produce a large army of \emph{Marines} (basic ranged ground unit) and \emph{Medics} (can heal biological units). Quickest way to apply pressure.
+    \item \emph{rax\_fe}: Take the closest ``natural expansion'' as soon as possible. This provides a big economic boost in the mid game by sacrificing some early game power.
+    \item \emph{two\_facto}: Build two \emph{Factories} and keep producing \emph{Tanks} (ground artillery unit). Vulnerable while building up to it and then very powerful on ground.
+    \item \emph{vultures}: Produce mainly \emph{Vultures} (fast ground ranged unit, excels against small units) and research mines.
+Quicker to reach (technologically) and build than tanks, can transition into tanks.
+    \item \emph{air}: Produce \emph{Wraiths} (ranged flying units) as soon as possible for an air attack. Vulnerable to anti-air openings or quick rushes.
+    \item \emph{drop}: Train \emph{Dropships} (flying transports) as soon as possible to be able to do (mostly tanks or marines) drops, leveraging efficient tactics.
+  \end{itemize}
+  \item Zerg openings:
+  \begin{itemize}
+    \item \emph{speedlings}: Train \emph{Zerlings} (basic cheap, fast, ground contact attack unit) and research speed upgrade as soon as possible. Quickest way to apply pressure.
+    \item \emph{fast\_mutas}: Produce mainly \emph{Mutalisks} (ranged flying units). Vulnerable in the early game while gathering gas and researching the technology.
+    \item \emph{mutas}: Expand two times for a stronger economy before massive training of \emph{Mutalisks}. Slower but more powerful build-up than above.
+    \item \emph{lurkers}: Train \emph{Lurkers} (ground, stealth artillery unit) as soon as possible to benefit from their (advanced technology) zone attack and cloak ability. 
+    \item \emph{hydras}: Massive production of \emph{Hydralisks} (ground ranged unit). Much quicker to reach technologically than \emph{Lurkers} and can transition into them.
+  \end{itemize}
+\end{itemize}
+
+As the figures show, the top three ranked bots in the competition (Skynet, Aiur and UalbertaBot) do not change their strategy at all depending on their opponent.
+For example, the Skynet bot (both in 2011 and 2012), always uses the same opening (\emph{two\_gates}), except when playing a Terran opponent, when it uses \emph{nony}.
+This reflects the trend that the performance of bots is still more dependent on carefully handcrafted and non-adaptive behaviors, than on on-line decision making procedures.
+This is so, since most of the problems that need to be solved in order to implement such procedures are still open.
+
+% {\color{blue} to-do}
 

### 2015-04-24 ECGG_StarCraftCompetitions/ECGG_StarCraftCompetitions.tex
Start remove some text, add subsubsections for each competition year

@@
 
+\subsubsection{AIIDE 2010}
+
 Started in 2010, the AIIDE StarCraft AI Competition\footnote{\url{http://www.StarCraftAICompetition.com}} is the most
@@
 
+\subsubsection{AIIDE 2011}
+
 In 2011 the University of Alberta hosted the competition, with organization by Michael Buro and
@@
 tactics) at the  beginning of a game and kept it  until the end. These
-five moods were:
-\begin{itemize}
-\item  Rush:  where  the  bot  tries  early  attacks,  and  have  good
+five moods were: 1. Rush: the bot tries early attacks, and have good
   probabilities to send the two or three first Zealots (basic contact attack ground unit) to harass the
-  opponent.
-\item Aggressive: where  we have less chance to  perform harasses with
+  opponent. 2. Aggressive: it has less chance to perform harasses with
   the first Zealots, and the first attack is usually a bit delayed
-  with regard to the Rush mood.
-\item Macro: where the AI do not try any early attacks and focus a bit
-  more on its economy before attacking.
-\item Defense: where the AI  ``turtles'' and wait to have a consequent
-  army before running an attack.
-\item  Fast expand:  where the  first building  constructed it  a base
-  expansion, for a very economical-oriented game.
-\end{itemize}
-Notice that  build orders are not  fully hard-coded since  they can be
+  with regard to the Rush mood. 3. Macro: the AI do not try any early attacks and focus a bit
+  more on its economy before attacking. 4. Defense: AIUR  ``turtles'' and wait to have a consequent
+  army before running an attack. 5. Fast expand:  the  first building  constructed it  a base
+  expansion, for a very economical-oriented game. Notice that build orders are not fully hard-coded since they can be
 altered by AIUR's Spend Manager.
@@
 
-After the competition was over, a man vs. machine match was held between the winner (Skynet) and an
-ex-professional StarCraft player named Oriol Vinyals. Oriol was a competitor in the 2001 World Cyber
-Games StarCraft competition, and though he had been out of practice for a few years was still quite a
-good player. The match was arranged to see how well StarCraft AI bots had progressed and to see if
-they could actually beat a decent human opponent.
+% After the competition was over, a man vs. machine match was held between the winner (Skynet) and an
+% ex-professional StarCraft player named Oriol Vinyals. Oriol was a competitor in the 2001 World Cyber
+% Games StarCraft competition, and though he had been out of practice for a few years was still quite a
+% good player. The match was arranged to see how well StarCraft AI bots had progressed and to see if
+% they could actually beat a decent human opponent.
 
@@
 
-For the best-of-three match, Oriol chose his races randomly and ended up beating Skynet in a 2-0. 
-In the first match, Oriol played Zerg vs. Skynet's Protoss on Python, a four player map. Oriol chose
-to start with a fast expansion strategy and transition into two base mutalisk production. Skynet chose
-to rush with a few early zealots, which was luckily the best possible choice given Oriol's strategy.
-Skynet's initial attack destroyed Oriol's early defenses, and nearly won the game in the first few
-minutes, however it then proceeded to send zealots to attack one at a time rather than group up its
-units before moving in, which allowed Oriol to catch up. Once Oriol produced his mutalisks, Skynet did
-not produce sufficient air defenses and Oriol quickly destroyed Skynet's base. In the second game,
-Oril played Terran, again on Python. After holding off early Dragoon pressure from Skynet, Oriol
-moved out with a few marines, medics and tanks. Skynet tried to defend with its army of Dragoons,
-however due to poor unit targeting decisions it started to attack useless medics after the marines
-had died, rather than the tanks. Oriol overcame the Dragoon army and was victorious. Later analysis
-of the match concluded that Skynet, while dominant over the other bots, was unable to properly
-adapt and transition into a mid-game strategy in game one once its early pressure failed, and in game
-two made a key blunder in unit targeting which cost it the game. Humans were still in command.
+% For the best-of-three match, Oriol chose his races randomly and ended up beating Skynet in a 2-0. 
+% In the first match, Oriol played Zerg vs. Skynet's Protoss on Python, a four player map. Oriol chose
+% to start with a fast expansion strategy and transition into two base mutalisk production. Skynet chose
+% to rush with a few early zealots, which was luckily the best possible choice given Oriol's strategy.
+% Skynet's initial attack destroyed Oriol's early defenses, and nearly won the game in the first few
+% minutes, however it then proceeded to send zealots to attack one at a time rather than group up its
+% units before moving in, which allowed Oriol to catch up. Once Oriol produced his mutalisks, Skynet did
+% not produce sufficient air defenses and Oriol quickly destroyed Skynet's base. In the second game,
+% Oril played Terran, again on Python. After holding off early Dragoon pressure from Skynet, Oriol
+% moved out with a few marines, medics and tanks. Skynet tried to defend with its army of Dragoons,
+% however due to poor unit targeting decisions it started to attack useless medics after the marines
+% had died, rather than the tanks. Oriol overcame the Dragoon army and was victorious. Later analysis
+% of the match concluded that Skynet, while dominant over the other bots, was unable to properly
+% adapt and transition into a mid-game strategy in game one once its early pressure failed, and in game
+% two made a key blunder in unit targeting which cost it the game. Humans were still in command.
+
+\subsubsection{AIIDE 2012}
 
@@
 
-The 2012 man vs. machine match again used the winner of the competition (Skynet), who played against
-Mike Lange, also known as Bakuryu. At the time of the match, Bakuryu was an A- ranked Zerg player on ICCup,
-and known as one of the best non-Korean Zerg players in the world. Bakuryu was considered much
-stronger than Oriol at the time that the match was played, and the results showed that this was true.
-In the first game of the best-of-three, Bakuryu made Skynet look quite silly by running around inside
-Skynet's base with a small number of zerglings while Skynet's zealots and half of its worked chased
-then in vain. After killing off several probes and buying enough time to set up his expansion, he
-cleaned up Skynet's army with a flying army of mutalisks. In the second game, Bakuryu contained Skynet
-inside its base with a group of zerglings positioned within Skynet's expansion. Skynet then constructed
-several Dark Templar and along with some Dragoons and Zealots attacked into Bakuryu's expansion which
-was heavily defended, and was crushed almost instantly, allowing Bakuryu's zergling force to finish
-off the Protoss base.
+\subsubsection{AIIDE 2013}
 
-In this match it was shown that the true weakness of state of the art StarCraft AI systems was that
-humans are very adept at recognizing scripted behaviors and exploiting them to the fullest. A human
-player in Skynet's position in the first game would have realized he was being taken advantage of and
-adapted his strategy accordingly, however the inability to put the local context (Bakuryu kiting his
-units around his base) into the larger context of the game (that this would delay Skynet until
-reinforcements arrived) and then the lack of strategy change to fix the situation led to an easy
-victory for the human. These problems remain as some of the main challenges in RTS AI today: to both recognize
-the strategy and intent of an opponent's actions, and how to effectively adapt your own strategy to 
-overcome them.
+\begin{table}[!t]
+\caption{Results of the five best bots of the AIIDE 2013 competition}
+\label{tab:aiide2013}
+\centering
+\begin{tabular}{|c|l|r|}
+\hline
+{\bfseries Position} & {\bfseries Bot} & {\bfseries Win \%} \\
+\hline
+1 & UAlbertaBot & 84.49\% \\
+2 & Skynet & 66.26\% \\
+3 & AIUR & 58.51\% \\
+4 & Ximp & 56.57\% \\
+5 & ICEStarCraft & 48.53\% \\
+\hline
+\end{tabular}
+\end{table}
 
-All results, videos, and replays from the AIIDE StarCraft AI Competition can be found in \url{http://www.StarCraftAICompetition.com}.
+\begin{figure}[t!]
+    \centering
+    \includegraphics[width=\columnwidth]{figures/aiide2013}
+    \caption{Evolution of the win percentage of each bot participating in the AIIDE 2013 competition}
+    \label{fig:aiide2013}
+\end{figure}
+
+{\color{blue} AIIDE 2013: Skynet and AIUR learning, BBQ opponent strategy
+  learning,  UAB  branch-and-bound  for   build  order  during  games,
+  SparCraft for combat simulations}
+
+\subsubsection{AIIDE 2014}
+
+\begin{table}[!t]
+\caption{Results of the five best bots of the AIIDE 2014 competition}
+\label{tab:aiide2014}
+\centering
+\begin{tabular}{|c|l|r|}
+\hline
+{\bfseries Position} & {\bfseries Bot} & {\bfseries Win \%} \\
+\hline
+1 & ICEBot & 85.86\% \\
+2 & Ximp & 84.64\% \\
+3 & LetaBot & 82.09\% \\
+4 & AIUR & 70.94\% \\
+5 & Skynet & 68.74\% \\
+\hline
+\end{tabular}
+\end{table}
+
+{\color{blue} AIIDE 2014: ???}
+
+% The 2012 man vs. machine match again used the winner of the competition (Skynet), who played against
+% Mike Lange, also known as Bakuryu. At the time of the match, Bakuryu was an A- ranked Zerg player on ICCup,
+% and known as one of the best non-Korean Zerg players in the world. Bakuryu was considered much
+% stronger than Oriol at the time that the match was played, and the results showed that this was true.
+% In the first game of the best-of-three, Bakuryu made Skynet look quite silly by running around inside
+% Skynet's base with a small number of zerglings while Skynet's zealots and half of its worked chased
+% then in vain. After killing off several probes and buying enough time to set up his expansion, he
+% cleaned up Skynet's army with a flying army of mutalisks. In the second game, Bakuryu contained Skynet
+% inside its base with a group of zerglings positioned within Skynet's expansion. Skynet then constructed
+% several Dark Templar and along with some Dragoons and Zealots attacked into Bakuryu's expansion which
+% was heavily defended, and was crushed almost instantly, allowing Bakuryu's zergling force to finish
+% off the Protoss base.
+
+% In this match it was shown that the true weakness of state of the art StarCraft AI systems was that
+% humans are very adept at recognizing scripted behaviors and exploiting them to the fullest. A human
+% player in Skynet's position in the first game would have realized he was being taken advantage of and
+% adapted his strategy accordingly, however the inability to put the local context (Bakuryu kiting his
+% units around his base) into the larger context of the game (that this would delay Skynet until
+% reinforcements arrived) and then the lack of strategy change to fix the situation led to an easy
+% victory for the human. These problems remain as some of the main challenges in RTS AI today: to both recognize
+% the strategy and intent of an opponent's actions, and how to effectively adapt your own strategy to 
+% overcome them.
+
+% All results, videos, and replays from the AIIDE StarCraft AI Competition can be found in \url{http://www.StarCraftAICompetition.com}.
 % \footnote{\url{http://www.StarCraftAICompetition.com}}.
@@
 so frequently that it would have been unjustifiable to announce a winner.
+
+\subsubsection{CIG 2011}
 
@@
 
+\subsubsection{CIG 2012}
+
 For CIG 2012, the AIIDE tournament software was employed,
@@
 
-\begin{figure*}[tb]
-    \centering
-    \includegraphics[width=\textwidth]{figures/botLadder1.pdf}
-    \caption{Bot's Elo Rating from February 1, 2012 to June 20, 2012}
-    \label{fig:botLadder1}
-\end{figure*}
+% \begin{figure*}[tb]
+%     \centering
+%     \includegraphics[width=\textwidth]{figures/botLadder1.pdf}
+%     \caption{Bot's Elo Rating from February 1, 2012 to June 20, 2012}
+%     \label{fig:botLadder1}
+% \end{figure*}
 
-\begin{figure*}[tb]
-    \centering
-    \includegraphics[width=\textwidth]{figures/botLadder2.pdf}
-    \caption{Bot's Elo Rating from September 1, 2012 to January 10, 2013}
-    \label{fig:botLadder2}
-\end{figure*}
+% \begin{figure*}[tb]
+%     \centering
+%     \includegraphics[width=\textwidth]{figures/botLadder2.pdf}
+%     \caption{Bot's Elo Rating from September 1, 2012 to January 10, 2013}
+%     \label{fig:botLadder2}
+% \end{figure*}
 
@@
 
-\subsection{StarCraft Bot Ladder}\label{sec:ladder}
+\subsubsection{CIG 2013}
 
-The StarCraft Bot Ladder is a website\footnote{\url{http://bots-stats.krasi0.com}} where bot versus bot matches are automatized, and are running all the time.
+\subsubsection{CIG 2014}
+
+% \subsection{StarCraft Bot Ladder}\label{sec:ladder}
+
+% The StarCraft Bot Ladder is a website\footnote{\url{http://bots-stats.krasi0.com}} where bot versus bot matches are automatized, and are running all the time.
 This ladder is a great resource for creating data sets (all the game replays are available) and statistics.
@@
 
-Figure \ref{fig:botLadder1} shows the Elo rating of the bots in the ladder during the first half year of 2012.
+% Figure \ref{fig:botLadder1} shows the Elo rating of the bots in the ladder during the first half year of 2012.
 The ranking is practically equal than the AIIDE 2011 competition, showing the lack of adaptability of the current bots.
@@
 
-\subsection{Adaptation Analysis}
+% \subsection{Adaptation Analysis}
 
-One of the major conclusions from the results of the StarCraft competitions is the lack of adaptation of bots.
+% One of the major conclusions from the results of the StarCraft competitions is the lack of adaptation of bots.
 Some switch between different build-orders, but do not fully adapt their strategy.
@@
 
-\begin{figure*}[tb]
-    \centering
-    \includegraphics[width=\textwidth]{figures/botopenings2011.png}
-    \caption{Distribution of different openings performed by the different bots participating in the AIIDE 2011 competition.
+% \begin{figure*}[tb]
+%     \centering
+%     \includegraphics[width=\textwidth]{figures/botopenings2011.png}
+%     \caption{Distribution of different openings performed by the different bots participating in the AIIDE 2011 competition.
 For each bot match-up, the colors show the proportion of times that the column bot used a particular opening against the row bot.}
-    \label{fig:aiide2011-botopenings}
-\end{figure*}
+%     \label{fig:aiide2011-botopenings}
+% \end{figure*}
 
-\begin{figure*}[tb]
-    \centering
-    \includegraphics[width=\textwidth]{figures/botopenings2012.png}
-    \caption{Distribution of different openings performed by the different bots participating in the AIIDE 2012 competition.
+% \begin{figure*}[tb]
+%     \centering
+%     \includegraphics[width=\textwidth]{figures/botopenings2012.png}
+%     \caption{Distribution of different openings performed by the different bots participating in the AIIDE 2012 competition.
 For each bot match-up, the colors show the proportion of times that the column bot used a particular opening against the row bot.}
-    \label{fig:aiide2012-botopenings}
-\end{figure*}
+%     \label{fig:aiide2012-botopenings}
+% \end{figure*}
 
-Specifically, we identified the following openings (for a better comprehension of strategies, buildings or units of StarCraft, we refer the reader to Teamliquid's wiki\footnote{\url{http://wiki.teamliquid.net/starcraft/Category:Strategies}}):
-\begin{itemize}
-  \item Protoss openings:
-  \begin{itemize}
-    \item \emph{two\_gates}: Build two \emph{Gateways} and keep training \emph{Zealots} (basic contact attack ground unit), this is the quickest way to apply pressure.
-    \item \emph{fast\_dt}: Produce \emph{Dark Templars} (technologically advanced stealth ground unit) as soon as possible, sacrificing early game power for a technological advance, hard-countered by detectors technology.
-    \item \emph{templar}: Train \emph{High Templars} (technologically advanced zone attack unit) as fast as possible, same as above, less deadly but is less easily countered.
-    \item \emph{speedzeal}: Train \emph{Zealots} and research attack and speed upgrades as soon as possible, some early game power transitioning into late game tech.
-    \item \emph{corsair}: Produce \emph{Corsairs} (air-air flying unit) as soon as possible and then transition into training \emph{Dark Templars} (safe from Zerg's flying detectors thanks to \emph{Corsairs}), \emph{Reavers} (ground artillery unit) or \emph{Dragoons}.
+% Specifically, we identified the following openings (for a better comprehension of strategies, buildings or units of StarCraft, we refer the reader to Teamliquid's wiki\footnote{\url{http://wiki.teamliquid.net/starcraft/Category:Strategies}}):
+% \begin{itemize}
+%   \item Protoss openings:
+%   \begin{itemize}
+%     \item \emph{two\_gates}: Build two \emph{Gateways} and keep training \emph{Zealots} (basic contact attack ground unit), this is the quickest way to apply pressure.
+%     \item \emph{fast\_dt}: Produce \emph{Dark Templars} (technologically advanced stealth ground unit) as soon as possible, sacrificing early game power for a technological advance, hard-countered by detectors technology.
+%     \item \emph{templar}: Train \emph{High Templars} (technologically advanced zone attack unit) as fast as possible, same as above, less deadly but is less easily countered.
+%     \item \emph{speedzeal}: Train \emph{Zealots} and research attack and speed upgrades as soon as possible, some early game power transitioning into late game tech.
+%     \item \emph{corsair}: Produce \emph{Corsairs} (air-air flying unit) as soon as possible and then transition into training \emph{Dark Templars} (safe from Zerg's flying detectors thanks to \emph{Corsairs}), \emph{Reavers} (ground artillery unit) or \emph{Dragoons}.
 Weak early game.
-    \item \emph{nony}: Build three \emph{Gateways} and massive training of \emph{Dragoons}. Slower than \emph{two\_gates} but still some early game (ranged) power.
-    \item \emph{reaver\_drop}: Train \emph{Reavers} as soon as possible to be able to do drops (air transport of artillery units).
-  \end{itemize}
-  \item Terran openings:
-  \begin{itemize}
-    \item \emph{bio}: Produce a large army of \emph{Marines} (basic ranged ground unit) and \emph{Medics} (can heal biological units). Quickest way to apply pressure.
-    \item \emph{rax\_fe}: Take the closest ``natural expansion'' as soon as possible. This provides a big economic boost in the mid game by sacrificing some early game power.
-    \item \emph{two\_facto}: Build two \emph{Factories} and keep producing \emph{Tanks} (ground artillery unit). Vulnerable while building up to it and then very powerful on ground.
-    \item \emph{vultures}: Produce mainly \emph{Vultures} (fast ground ranged unit, excels against small units) and research mines.
+%     \item \emph{nony}: Build three \emph{Gateways} and massive training of \emph{Dragoons}. Slower than \emph{two\_gates} but still some early game (ranged) power.
+%     \item \emph{reaver\_drop}: Train \emph{Reavers} as soon as possible to be able to do drops (air transport of artillery units).
+%   \end{itemize}
+%   \item Terran openings:
+%   \begin{itemize}
+%     \item \emph{bio}: Produce a large army of \emph{Marines} (basic ranged ground unit) and \emph{Medics} (can heal biological units). Quickest way to apply pressure.
+%     \item \emph{rax\_fe}: Take the closest ``natural expansion'' as soon as possible. This provides a big economic boost in the mid game by sacrificing some early game power.
+%     \item \emph{two\_facto}: Build two \emph{Factories} and keep producing \emph{Tanks} (ground artillery unit). Vulnerable while building up to it and then very powerful on ground.
+%     \item \emph{vultures}: Produce mainly \emph{Vultures} (fast ground ranged unit, excels against small units) and research mines.
 Quicker to reach (technologically) and build than tanks, can transition into tanks.
-    \item \emph{air}: Produce \emph{Wraiths} (ranged flying units) as soon as possible for an air attack. Vulnerable to anti-air openings or quick rushes.
-    \item \emph{drop}: Train \emph{Dropships} (flying transports) as soon as possible to be able to do (mostly tanks or marines) drops, leveraging efficient tactics.
-  \end{itemize}
-  \item Zerg openings:
-  \begin{itemize}
-    \item \emph{speedlings}: Train \emph{Zerlings} (basic cheap, fast, ground contact attack unit) and research speed upgrade as soon as possible. Quickest way to apply pressure.
-    \item \emph{fast\_mutas}: Produce mainly \emph{Mutalisks} (ranged flying units). Vulnerable in the early game while gathering gas and researching the technology.
-    \item \emph{mutas}: Expand two times for a stronger economy before massive training of \emph{Mutalisks}. Slower but more powerful build-up than above.
-    \item \emph{lurkers}: Train \emph{Lurkers} (ground, stealth artillery unit) as soon as possible to benefit from their (advanced technology) zone attack and cloak ability. 
-    \item \emph{hydras}: Massive production of \emph{Hydralisks} (ground ranged unit). Much quicker to reach technologically than \emph{Lurkers} and can transition into them.
-  \end{itemize}
-\end{itemize}
+%     \item \emph{air}: Produce \emph{Wraiths} (ranged flying units) as soon as possible for an air attack. Vulnerable to anti-air openings or quick rushes.
+%     \item \emph{drop}: Train \emph{Dropships} (flying transports) as soon as possible to be able to do (mostly tanks or marines) drops, leveraging efficient tactics.
+%   \end{itemize}
+%   \item Zerg openings:
+%   \begin{itemize}
+%     \item \emph{speedlings}: Train \emph{Zerlings} (basic cheap, fast, ground contact attack unit) and research speed upgrade as soon as possible. Quickest way to apply pressure.
+%     \item \emph{fast\_mutas}: Produce mainly \emph{Mutalisks} (ranged flying units). Vulnerable in the early game while gathering gas and researching the technology.
+%     \item \emph{mutas}: Expand two times for a stronger economy before massive training of \emph{Mutalisks}. Slower but more powerful build-up than above.
+%     \item \emph{lurkers}: Train \emph{Lurkers} (ground, stealth artillery unit) as soon as possible to benefit from their (advanced technology) zone attack and cloak ability. 
+%     \item \emph{hydras}: Massive production of \emph{Hydralisks} (ground ranged unit). Much quicker to reach technologically than \emph{Lurkers} and can transition into them.
+%   \end{itemize}
+% \end{itemize}
 
-As the figures show, the top three ranked bots in the competition (Skynet, Aiur and UalbertaBot) do not change their strategy at all depending on their opponent.
+% As the figures show, the top three ranked bots in the competition (Skynet, Aiur and UalbertaBot) do not change their strategy at all depending on their opponent.
 For example, the Skynet bot (both in 2011 and 2012), always uses the same opening (\emph{two\_gates}), except when playing a Terran opponent, when it uses \emph{nony}.

### 2015-12-09 ECGG_StarCraftCompetitions/ECGG_StarCraftCompetitions.tex
added aiide details

@@
 \item
-  UAlbertaBot\footnote{\url{http://code.google.com/p/ualbertabot/}}:
+  UAlbertaBot\footnote{\url{http://github.com/davechurchill/ualbertabot/}}:
   also  uses abstraction  in  {\em combat} like  the previous  two
@@
 \hline
-{\bfseries Position} & {\bfseries Bot}\\
+{\bfseries Rank} & {\bfseries Bot}\\
 \hline
@@
 change in the 2011 competition was the introduction of automated tournament-managing software running
-StarCraft games simultaneously on 20 computers, allowing a total of 1170 games to be played in far less
-time than the 108 games of the 2010 competition. This increase in games played also allowed the tournament
+StarCraft games simultaneously on 20 computers, allowing a total of 1170 games to be played in the five days that the tournament ran. 
+This increase in games played also allowed the tournament
 to switch to a round-robin format, eliminating the ``luck'' factor of the pairings inherent in bracket
 style tournaments. The bot that achieved the highest win percentage over the course of the competition would
-be determined the winner. Also, the competition became open-source, in an effort not
-only to prevent possible cheating, but to promote healthy competition in future tournaments by giving
+be determined the winner. Also, the competition became open-source, in an effort 
+to prevent possible cheating and to promote healthy competition in future tournaments by giving
 newcomers and easier entry point by basing their design off of previous bots.
@@
 \centering
-\begin{tabular}{|c|l|r|}
+\begin{tabular}{|c|c|c|}
 \hline
-{\bfseries Position} & {\bfseries Bot} & {\bfseries Win \%} \\
+{\bfseries Rank} & {\bfseries Bot} & {\bfseries Win \%} \\
 \hline
@@
 The University of Alberta also hosted the 2012 competition, with the major difference from the 2011
-competition being the addition of persistent storage. Bots could now write information to disk during a
+competition being the addition of file reading and writing for learning. Bots could now write information to disk during a
 match, and then read the information during other matches, allowing them to adjust strategies based
@@
 
+A total of 8 bots competed in the 2013 AIIDE competition with many of the same names from the 2012 competition, and some key updates were made to existing bots for 2013 which shook up the results from the previous years.
+UAlbertaBot took 1st place with a dominant 84.49\% win rate, with Skynet in 2nd with 72.77\%.
+The major addition to UAlbertaBot was a combat simulation package called SparCraft.
+UAlbertaBot was re-engineered so that it now always attacked from the first combat unit created, using the SparCraft system to simulate whether or not the bot could a fight against known enemy units, retreating automatically if it thought it could not.
+This addition, combined with some additional bug fixes led to the victory.
+Skynet and Aiur both implemented strategy learning in 2013, which is evident from \ref{fig:aiide2013} in which both Skynet and Aiur's win percentage over time climb dramatically from the first few rounds of the tournament to the later rounds.
+Ximp unfortunately crashed all games on the Fortress map and lost 10\% of its games for free as a result, which meant it could have possibly came in 2nd place if not for all those crashes.
+
 \begin{table}[!t]
@@
 \centering
-\begin{tabular}{|c|l|r|}
+\begin{tabular}{|c|c|c|}
 \hline
-{\bfseries Position} & {\bfseries Bot} & {\bfseries Win \%} \\
+{\bfseries Rank} & {\bfseries Bot} & {\bfseries Win \%} \\
 \hline
 1 & UAlbertaBot & 84.49\% \\
-2 & Skynet & 66.26\% \\
+2 & Skynet & 72.77\% \\
 3 & AIUR & 58.51\% \\
@@
 
-{\color{blue} AIIDE 2013: Skynet and AIUR learning, BBQ opponent strategy
-  learning,  UAB  branch-and-bound  for   build  order  during  games,
-  SparCraft for combat simulations}
+\subsubsection{AIIDE 2014}
 
-\subsubsection{AIIDE 2014}
+The 2014 AIIDE competition saw 18 total entrants, over double the number from the 2013 competition.
+This dramatic increase in numbers was partially because of a large advertising effort by competition organizers, combined with the inclusion of all the 2013 competition entrants (if they had not re-submitted new versions).
+UAlbertaBot and Skynet who finished 1st and 2nd the year before were not updated in 2014 and so their previous versions were re-submitted.
+Along with the larger registration numbers came a lot of new Terran bots, a welcome change from the Protoss-dominated fields of previous years.
+The tournament managing software also underwent some major updates in 2014, allowing for faster game scheduling and the easy pausing and resuming of tournaments, resulting in 10251 total games being played, which was almost double the amount of 5579 from the previous year.
+
+The results of the 2014 competition were dramatically different from previous years.
+Since 2011 the top 3 bots in the AIIDE competition had been some permutation of Skynet, UAlbertaBot, and Aiur, and in 2014 the top 3 finishers were all completely different.
+In first place was IceBot with a win rate of 85.85\%, which had a complete overhaul in terms of strategy and tactics from the previous year.
+IceBot had several strategies which it could implement, most starting with an early bunker defense and transitioning into attacking once a set of heuristics conditions had been met.
+IceBot's source code was originally based on Aiur's modular design, and changed to play the Terran race.
+In a close 2nd place finish was Ximp - which had fixed the crashes and logic errors which had plagued it the year before.
+Ximp implemented a solid forge-first fast expansion strategy which transitioned into late-game carrier play.
+By building a large amount of early photon cannons in its expansion, Ximp was easily able to stop most of the early rushing bots, and then holding them off until its carriers cleaned up in the late game.
+3rd place was taken by LetaBot, a new Terran bot whose source code was written on top of the 2012 version of UAlbertaBot and changed to play the Terran race.
+LetaBot implemented a number of strategies including a 'cheesy' but strong bunker rush strategy which killed many bots very quickly, as well as a Wraith (mid-game flying unit) strategy.
 
@@
 
-{\color{blue} AIIDE 2014: ???}
+\subsubsection{AIIDE 2015}
+
+The 2015 AIIDE competition was the largest competition to date with 23 total bots competing.
+The competition environment was also changed so that the tournament was now run on virtual machines instea dof physical machines.
+Doing this enabled the tournament to run for a full two weeks instead of the normal one week, resulting in a total of 20788 games being played - 90 full round robins between each bot pairing.
+2015 saw the most even distribution of race selection ever in a Starcraft AI competition, with 5 new Zerg submissions along with 7 Protoss bots and 9 Terran bots.
+The first ever Random race entry was also submitted as UAlbertaBot had been updated to play all three of the available races.
+When playing Random race, Starcraft randomly assigns one of Protoss, Terran, or Zerg after the game has started, with your opponent not knowing which race you are until they have scouted you on the map.
+This provides a significant advantage as the enemy's initial build order must now have a period of uncertainty during which is has to guess what race it is playing against.
+
+The results of the 2015 competition were extremely close, with 1st / 2nd place as well as 3rd / 4th place being separated by less than 1\% win rate.
+1st place was won by Tscmoo, a new Zerg bot which played nearly a dozen different strategies, learning which one to choose over time via the file I/O system.
+In a close second place was ZZZKbot, which implement a 4-pool Zergling rush strategy every game.
+Despite the relatively simple strategy, most bots did not have proper defense capabilities and lost in very short games.
+In 3rd place was Overkill, another Zerg bot which had several different strategies including Mutalisks and Hydralisks, which learned over time as well.
+In a close 4th place was UAlbertaBot, which played Random and implemented three main rushing strategies - one for each race.
+An interesting result from this tournament was that despite UAlbertaBot coming 4th place, it finished with a winning rate greater than 50\% vs each other bot.
+Aiur came in 5th place and was a clear demonstration of how learning over time can dramatically improve results in a tournament, going from 63\% win rate early in the competition to a final win rate of over 73\%.
+
+\begin{table}[!t]
+\caption{Results of the five best bots of the AIIDE 2015 competition}
+\label{tab:aiide2014}
+\centering
+\begin{tabular}{|c|c|c|}
+\hline
+{\bfseries Rank} & {\bfseries Bot} & {\bfseries Win \%} \\
+\hline
+1 & Tscmoo & 88.52\% \\
+2 & ZZZKBot & 87.84\% \\
+3 & Overkill & 80.69\% \\
+4 & UAlbertaBot & 80.20\% \\
+5 & Aiur & 73.02\% \\
+\hline
+\end{tabular}
+\end{table}
 
@@
 % bracket A
-\begin{tabular}{|c|c|c|l|r|}
+\begin{tabular}{|c|c|c|c|c|}
 \hline
 \multicolumn{5}{|c|}{Bracket A} \\ \hline
-{\bfseries Position} & {\bfseries Crashes} & {\bfseries Games} & {\bfseries Bot} & {\bfseries Win \%} \\
+{\bfseries Rank} & {\bfseries Crashes} & {\bfseries Games} & {\bfseries Bot} & {\bfseries Win \%} \\
 \hline
@@
 % bracket B
-\begin{tabular}{|c|c|c|l|r|}
+\begin{tabular}{|c|c|c|c|c|}
 \hline
 \multicolumn{5}{|c|}{Bracket B} \\ \hline
-{\bfseries Position} & {\bfseries Crashes} & {\bfseries Games} & {\bfseries Bot} & {\bfseries Win \%} \\
+{\bfseries Rank} & {\bfseries Crashes} & {\bfseries Games} & {\bfseries Bot} & {\bfseries Win \%} \\
 \hline
@@
 These mainly stemmed from the desire to use evolved, largely untested
-maps which proved to look interesting but made the submitted bots and the Brood War Terrain Analyzer (BWTA) provided with the BWAPI interface crash
+maps which initially looked interesting but made the submitted bots and the Brood War Terrain Analyzer (BWTA) provided with the BWAPI interface crash
 so frequently that it would have been unjustifiable to announce a winner.
@@
 \centering
-\begin{tabular}{|c|c|c|l|r|}
+\begin{tabular}{|c|c|c|c|c|}
 \hline
-{\bfseries Position} & {\bfseries Crashes} & {\bfseries Games} & {\bfseries Bot} & {\bfseries Win \%} \\
+{\bfseries Rank} & {\bfseries Crashes} & {\bfseries Games} & {\bfseries Bot} & {\bfseries Win \%} \\
 \hline

### 2015-12-10 ECGG_StarCraftCompetitions/ECGG_StarCraftCompetitions.tex
cig 2014 and 2015

@@
 The following subsections summarize the results of all the StarCraft AI competitions held at the AIIDE (Artificial Intelligence for Interactive Digital Entertainment) and CIG (Computational Intelligence in Games) conferences during the past years.
+Since 2011, the CIG and AIIDE competitions have typically been held in August / September in each year, as the conferences are scheduled quite close to each other.
+As a consequence of this, many of the same bots get entered into both competitions.
+
 Additionally we analyze the statistics from the StarCraft Bot Ladder, where the best bots play against each other continuously over time.
@@
 
+Mike will fill this out?
+
+\begin{table}[t]
+\caption{Results of the CIG 2013 competition.}
+\label{tab:cig2013}
+\centering
+\begin{tabular}{|c|c|c|}
+\hline
+{\bfseries Rank} & {\bfseries Bot} & {\bfseries Win \%} \\
+\hline
+1 & Skynet & 91.1\% \\
+2 & UAlbertaBot & 67.4\% \\
+3 & AIUR & 54.9\% \\
+4 & Xelnaga & 53.6\% \\
+5 & Adjutant & 42.4\% \\ 
+\hline
+\end{tabular}
+\end{table}
+
 \subsubsection{CIG 2014}
+The CIG 2014 competition was organized by Kyung-Joong Kim, Ho-Chul Cho, and In-Seok Oh of Sejong University.
+For 2014, the CIG competition used a total of 20 different maps which were unknown to the competitors before the competition started, which was by far the most maps ever used in a Starcraft AI competition and presented a challenge to many of the bots who entered.
+A total of 4680 games were played which meant that each bot played each other bot 60 times, or 3 times per map.
+The CIG 2014 competition was held just a few weeks before the AIIDE 2014 competition and so many of the entrants to the competition were identical, and the results definitely showed this.
+The top 4 bots were all the same as the 2014 AIIDE competition with IceBot coming 1st, Ximp in 2nd, LetaBot in 3rd and Aiur in 4th place.
+For descriptions of these bots please refer to the AIIDE 2014 competition description as none of the top-finishing bots had any significant changes made.
+
+\begin{table}[t]
+\caption{Results of the CIG 2014 competition.}
+\label{tab:cig2014}
+\centering
+\begin{tabular}{|c|c|c|}
+\hline
+{\bfseries Rank} & {\bfseries Bot} & {\bfseries Win \%} \\
+\hline
+1 & IceBot & 83.1\% \\
+2 & Ximp & 78.1\% \\
+3 & LetaBot & 68.5\% \\
+4 & Aiur & 66.1\% \\
+5 & UAlbertaBot & 60.0\% \\ 
+\hline
+\end{tabular}
+\end{table}
+
+\subsubsection{CIG 2015}
+There were some significant rule changes to the CIG 2015 competition, which was once organized by mebers from Sejong University.
+The most significant rule change was that entrants no longer had to be open source, in an attempt to attract more competitors who may not want to open source their bots.
+The second rule change was that a single competitor could submit multiple entries to the competition - which caused some controversy among registrants since this introduces the possibility of collusion between entries.
+Thankfully no such collusion was detected during the competition.
+The 2015 competition was also not run quite as long as previous competitions with only 2730 games being played in total, or 30 between each bot pairing, 6 between each bot on each of the 5 chosen maps.
+There were several new Zerg entries to the CIG 2015 competition which ended up finishing in the top three positions, with results very similar to those of the AIIDE 2015 competition.
+ZZZBot took 1st place, tscmoo-Z (a Zerg bot written by tscmoo) came 2nd, and Overkill came 3rd.
+For descriptions of these bots and their strategies used please refer to the AIIDE 2015 competition section as the strategies remained largely unchanged between the CIG and AIIDE competitions.
+
+\begin{table}[t]
+\caption{Results of the CIG 2015 competition.}
+\label{tab:cig2015}
+\centering
+\begin{tabular}{|c|c|c|}
+\hline
+{\bfseries Rank} & {\bfseries Bot} & {\bfseries Win \%} \\
+\hline
+1 & ZZZBot & 81.03\% \\
+2 & tscmoo-Z & 73.59\% \\
+3 & Overkill & 62.05\% \\
+4 & LetaBot & 61.54\% \\
+5 & Ximp & 60.26\% \\ 
+\hline
+\end{tabular}
+\end{table}
 

### 2015-12-11 ECGG_StarCraftCompetitions/ECGG_StarCraftCompetitions.tex
removed the * from some tables and figures, since this is a single column document
fixed a typo in the authors line, and a duplicated label in a table

@@
 \author{David~Churchill\inst{1} \and
-		Mike~Preuss\inst{2}		
+		Mike~Preuss\inst{2} \and
 		Florian~Richoux\inst{3} \and
@@
 		Alberto~Uriarte\inst{5} \and
-		Santiago~Onta\~{n}\'{o}n\inst{5} \and 
+		Santiago~Onta\~{n}\'{o}n\inst{5}
 		}
@@
 \caption{Results of the five best bots of the AIIDE 2015 competition}
-\label{tab:aiide2014}
+\label{tab:aiide2015}
 \centering
@@
 
-\begin{table*}[!t]
+\begin{table}[!t]
 \caption{Results of the first round at CIG 2011, held in two brackets.
@@
 \end{tabular}
-\end{table*}
+\end{table}
 
@@
 
-\begin{figure*}[b]
+\begin{figure}[b]
 	%\vspace*{-2ex}
@@
     \label{fig:cig2012-mapresults}
-\end{figure*}
+\end{figure}
 

### 2015-12-11 ECGG_StarCraftCompetitions/ECGG_StarCraftCompetitions.tex
updated CIG competitions text

@@
 \and
-	Department of Computer Science of Technische Universit{\"a}t Dortmund, Germany.
-	\email{mike.preuss@cs.tu-dortmund.de}
+	Information Systems and Statistics, Westf. Wilhelmsuniversit{\"a}t M{\"u}nster, Germany.
+	\email{mike.preuss@uni-muenster.de}
 \and
@@
 
-\subsubsection{CIG 2013}
-
-Mike will fill this out?
+\subsubsection{CIG 2013} The CIG 2013 competition was once again
+organized by Mike Preuss (TU Dortmund) and Tobias Mahlmann (ITU Copenhagen),
+adding Antonio Mora Garc{\'i}a from the Universidad de Granada as
+third team member. Whereas the rules and even the participants
+to this competition were almost identical to the AIIDE 2013
+competition setup, several technical changes were made in comparison
+to 2012\footnote{\url{http://ls11-www.cs.tu-dortmund.de/rts-competition/StarCraft-cig2013}}.
+A new competition software was implemented by Tobias but
+not completely finished in time, so that the read/write function
+that enables learning between games had to be disabled again. 
+This is probably the most important difference to the AIIDE competition,
+however, as the result provided in table~\ref{tab:cig2013} shows, the effect is
+limited. Due to the late availability of the competition software,
+only 32 rounds of random robin were played, making 896 games
+altogether, and 224 games by each of the 8 submitted bots.
+
+As the map set has also been changed to the 10 standard
+maps used by the AIIDE competition (without letting the competitors
+know about this in advance as usual for the CIG competitions), 
+the competitions were otherwise very similar. We presume that 
+disabling the learning was a disadvantage for the UAlbertaBot who
+won the AIIDE competition and was only runner-up to the Skynet bot
+here. The 3rd place went to AIUR as in 2012, followed by 
+Xelnaga and Adjutant. A direct comparison to the results of
+CIG 2012 (table~\ref{tab:cig2012}) shows that bot evolution 
+between these two competitions was obviously limited. It also
+shows that some bots as the UAlbertaBot make good use of 
+online learning, whereas others, as Skynet, do not profit from
+it that much.
 

### 2015-12-16 ECGG_StarCraftCompetitions/ECGG_StarCraftCompetitions.tex
Added a SSCAIT section.

@@
 		Alberto~Uriarte\inst{5} \and
-		Santiago~Onta\~{n}\'{o}n\inst{5}
+		Santiago~Onta\~{n}\'{o}n\inst{5} \and
+		Michal~\v{C}ertick\'{y}\inst{6}
 		}
@@
 	\email{\{santi,albertouri\}@cs.drexel.edu}
+\and
+	Agent Technology Center at Czech Technical University in Prague,
+	Czech Republic.
+	\\
+	\email{certicky@agents.fel.cvut.cz}
 }
@@
 
-This section reviews the results of the recent international competitions on AI for StarCraft.
-These competitions, typically co-located with scientific conferences, have been possible thanks to the existence of the Brood War Application Programming Interface
+This section reviews the results of three recent international competitions on AI for StarCraft. These competitions
+%, typically co-located with scientific conferences, 
+have been possible thanks to the existence of the Brood War Application Programming Interface 
 (BWAPI)\footnote{\url{http://code.google.com/p/bwapi/}}, which 
@@
 %A consequence of this indirect way of integrating new bots is that every machine can only run one custom bot, requiring two computers for a 2-player game.
-The following subsections summarize the results of all the StarCraft AI competitions held at the AIIDE (Artificial Intelligence for Interactive Digital Entertainment) and CIG (Computational Intelligence in Games) conferences during the past years.
-Since 2011, the CIG and AIIDE competitions have typically been held in August / September in each year, as the conferences are scheduled quite close to each other.
-As a consequence of this, many of the same bots get entered into both competitions.
-
-Additionally we analyze the statistics from the StarCraft Bot Ladder, where the best bots play against each other continuously over time.
+The following subsections summarize the results of StarCraft AI competitions co-located with two scientific conferences -- AIIDE (Artificial Intelligence for Interactive Digital Entertainment) and CIG (Computational Intelligence in Games) -- and one stand\-alone competition called SSCAIT (Student StarCraft AI Tournament), during the past years.
+Since 2011, the CIG and AIIDE competitions have typically been held in August / September in each year, as the conferences are scheduled quite close to each other, and SSCAIT have been held in December / January.
+As a consequence of this, AIIDE and CIG competitions share a big portion of the entrants.
+
+% the section with the Ladder seems to be commented out, so this sentence should be as well
+%Additionally we analyze the statistics from the StarCraft Bot Ladder, where the best bots play against each other continuously over time.
 
@@
 \end{table}
+
+\subsection{SSCAIT}\label{sec:SSCAIT}
+Student StarCraft AI Tournament\footnote{\url{http://sscaitournament.com/}} (SSCAIT) started in 2011 at Comenius University in Bratislava, Slovakia and immediately became the competition with the highest numbers of participants.
+There are three fundamental differences between SSCAIT and the remaining two conferences:
+\begin{enumerate}
+  \item SSCAIT is an online-only event. Unlike AIIDE or CIG, it is not co-located with a scientific conference or any other real-world event.
+  \item There are two phases of SSCAIT each year: a {\em competitive} phase, lasting for up to three weeks and a {\em sandbox} phase which runs for approximately eleven months each year.
+In other words, SSCAIT is live at all times with only a few short interruptions.
+  \item All the games are publicly {\em streamed live} 24 hours a day.
+The stream uses a custom observer script \cite{mattsson2015automatic} designed specifically to improve the viewer experience and creates a continuous feedback loop for the participants who can watch their bots play against the others.
+\end{enumerate}
+
+\subsubsection{SSCAIT 2011}
+Taking place in winter of 2011, the first SSCAIT tournament was organized by Michal \v{C}ertick\'{y} as part of the {\em ``Introduction to Artificial Intelligence''} course at the Department of Applied Informatics, Comenius University in Bratislava, Slovakia.
+It was inspired by the media coverage of AIIDE StarCraft AI competition held in 2010, but over the following years it diverged from AIIDE competition's format significantly.
+
+The first year of SSCAIT hosted 50 participants, all of whom were students of Comenius University at the time.
+The participants were allowed to select any race and use all the units and technologies in complete 1 vs.
+1 games of StarCraft.
+Vast majority of the bots were implemented to execute hard-coded strategies and behaviour.
+Participants were divided into 10 groups of 5 people, and 16 of them advanced into a double elimination bracket.
+Final match of the elimination bracket was won by R.
+Danielis with a 2-1 score against M.
+Piovar\v{c}i (both playing as Zerg).
+
+2011 was the only year when the games were not streamed live. However, the replays were published in real time and the games from the elimination bracket were recorded and uploaded to YouTube.
+
+\subsubsection{SSCAIT 2012-2015}
+In 2012, the SSCAIT became a truly international competition with 52 participants from numerous universities including 
+UC Berkeley (USA), University of Alberta (CAN), Washington State University (USA), University of Nantes (FRA), University of Grenoble (FRA), New University of Lisbon (POT), Gdansk University of Technology (POL), Ritsumeikan University (JAP), University of Bielefeld (GER), Sofia University (BUL), Comenius University (SVK), University of \v{Z}ilina (SVK) and Technical University in Ko\v{s}ice (SVK), as well as a number of non-student participants.
+
+The format of SSCAIT changed significantly in 2012. The tournament was divided into two phases:
+
+\textbf{{Sandbox phase}} is played non-stop during first 11 months of the year.
+All the games are streamed via a video streaming service, such as Twitch.tv or Hitbox.tv.
+The stream is meant to attract new entrants and to help current participants improve their bots before the second phase begins.
+The participants can watch their bots play against randomly selected other bots.
+They are allowed to upload new versions of the bots via a web interface at any time during this phase.
+The  sandbox phase has no winners.
+However, the bots may use this long phase to collect useful data about their opponent's behaviour.
+
+\textbf{{Competitive phase}} is played for up to three weeks (depending on the number of participants) at the end of the year.
+New bot versions cannot be uploaded during this phase and the matches are not scheduled randomly any more.
+This phase has two divisions with different rules:
+	\begin{itemize}
+      \item {\em Student division:} Only the bots created by a single participant, who is currently a student, are considered ``student'' bots.
+Other bots are tagged as ``mixed-division'' bots.
+Each student bot plays a single game against every other bot and collects points for the wins.
+Three student participants with the highest score are considered winners and receive  certificates of achievement (and in years 2012 and 2013 also financial prizes collected via crowdfunding campaigns).
+Ties are broken using the results of the mutual game and (if necessary) additional games are scheduled.
+The student division was created so that the students stand a chance of winning the certificates in the presence of more experienced, non-student participants and team-created entries.
+      \item {\em Mixed division:} Eight or sixteen bots with the highest win rate among all the participants (including student and non-student entries) are selected for the additional ``mixed division" elimination bracket.
+This part of the tournament is typically recorded with a commentary and uploaded to YouTube, concluding each SSCAIT season since 2012.
+    \end{itemize}
+
+The results of SSCAIT seasons 2012-2015 follow:
+\begin{itemize}
+  \item {\bf 2012:} 52 participants. Student division winners: 
+  	\begin{enumerate}
+  	  \item Matej Isten\'{i}k's bot
+  	  \item Marcin Bartnicki's bot
+  	  \item UAlbertaBot
+  	\end{enumerate}
+  Mixed division winner of the elimination bracket of 8 bots: ICEBot, after defeating Marcin Bartnicki's bot. 
+  \item {\bf 2013:} 50 participants. Student division winner after 1190 round-robin games: 
+    \begin{enumerate}
+  	  \item Ximp
+  	  \item W.O.P.R.
+  	  \item UAlbertaBot
+  	\end{enumerate}
+  Mixed division winner of the elimination bracket of 8 bots: Krasi0, after defeating ICEBot.
+  \item {\bf 2014:} 42 participants. Student division winner after 861 round-robin games: 
+    \begin{enumerate}
+  	  \item LetaBot
+  	  \item W.O.P.R.
+  	  \item UAlbertaBot, with the same win rate as W.O.P.R., but with a loss in a mutual game
+  	\end{enumerate}
+  Mixed division winner of the elimination bracket of 8 bots: LetaBot, after defeating Ximp.
+  \item {\bf 2015:} At the time of writing this, the SSCAIT 2015 is still in a sandbox phase.
+45 bots have registered for the competitive phase so far, including two bots playing as ``Random'' race.
+The season will be concluded by an elimination bracket featuring 16 bots in 2015.
+\end{itemize}
+
+From a technical perspective, the tournament was run manually in 2011 and by a collection of custom automated scripts in 2012.
+Since the second half of 2013, SSCAIT has been using the automated tournament managing software developed at University of Alberta, which was modified to suit the needs of SSCAIT format and technical environment.
+In contrast to AIIDE and CIG competitions, there is only one game running in SSCAIT at any given time, so that each game can be streamed live (approximately 150 games are played every day).
+The live stream makes use of a custom made observer script by Mattsson et al.
+\cite{mattsson2015automatic} which implements a smooth, prioritized camera movement during the game and a third-party libraries that allow the game to be displayed in high definition screen resolution.
+
+Thanks to the tournament managing software, the game rules and I/O procedures have been synchronized with AIIDE and CIG competitions, so the participants are able to submit the same version of their bot to all three of them.
+In addition to bots written in C++, SSCAIT supports bots using Java interfaces to BWAPI: BWMirror and JNIBWAPI.
 

### 2015-12-16 ECGG_StarCraftCompetitions/ECGG_StarCraftCompetitions.tex
read the paper beginning to end fixing typos, updating URLs (google code urls were old), and adding a quick and dirty conclusions section

@@
 
-Thanks to the recent organization of international game AI competitions focused around the popular StarCraft game (see Section \ref{sec:competition}), several groups have been working on integrating many of the techniques described in the previous section into complete ``bots'', capable of playing complete StarCraft games.
-In this section we will overview some of the currently available top bots.
+Thanks to the recent organization of international game AI competitions focused around the popular StarCraft game, several groups have been working on integrating many of the techniques developed for RTS game AI into complete ``bots'', capable of playing complete StarCraft games.
+In this chapter we will overview some of the currently available top bots, and their results of recent competitions.
 
@@
 
-Playing an RTS game involves dealing with all the problems described above.
+Playing an RTS game involves dealing with a wide variety of problems, ranging from miro-management problems such as unit control, to macro-management problems such as resource allocation.
 A few approaches, like CAT \cite{LTW}, Darmok \cite{OntanonMSR10} or ALisp \cite{Marthi05} try to deal with the problem in a monolithic manner, by using a single AI technique.
@@
   lower-level modules.
-\item  AIUR\footnote{\url{http://code.google.com/p/aiurproject/}}:  is
+\item  AIUR\footnote{\url{https://github.com/AIUR-group/AIUR}}:  is
   mainly divide-and-conquer oriented,
@@
   reactive controls so far.
-\item  BTHAI\footnote{\url{http://code.google.com/p/bthai/}}:  uses  a
+\item  BTHAI\footnote{\url{https://github.com/jhagelback/opprimobot}}:  uses  a
   two-tier  abstraction hierarchy,  where a  collection  of high-level
@@
 For space reasons, we do not include an exhaustive comparison of the architectures of all the participating bots.
-Some other bots have been documented by their authors, such as SCAIL \cite{YoungSCAIL} or QUORUM \cite{young2012evolutionary}.
+Some more bots than the ones described here have been documented by their authors, such as SCAIL \cite{YoungSCAIL} or QUORUM \cite{young2012evolutionary}.
 %This section has focused on discussing the architecture of existing StarCraft bots. 
-Let us now focus on their performance.
+Let us now focus on their performance in recent competitions.
 
@@
 have been possible thanks to the existence of the Brood War Application Programming Interface 
-(BWAPI)\footnote{\url{http://code.google.com/p/bwapi/}}, which 
+(BWAPI)\footnote{\url{https://github.com/bwapi/bwapi}}, which 
 enables replacing the human player interface with C++ code.
@@
 
-\subsubsection{AIIDE 2010}
+\subsubsection{AIIDE 2010.}
 
@@
 
-\subsubsection{AIIDE 2011}
+\subsubsection{AIIDE 2011.}
 
@@
 
-\subsubsection{AIIDE 2012}
+\subsubsection{AIIDE 2012.}
 
@@
 
-\subsubsection{AIIDE 2013}
+\subsubsection{AIIDE 2013.}
 
@@
 
-\subsubsection{AIIDE 2014}
+\subsubsection{AIIDE 2014.}
 
@@
 
-\subsubsection{AIIDE 2015}
+\subsubsection{AIIDE 2015.}
 
 The 2015 AIIDE competition was the largest competition to date with 23 total bots competing.
-The competition environment was also changed so that the tournament was now run on virtual machines instea dof physical machines.
+The competition environment was also changed so that the tournament was now run on virtual machines instead of physical machines.
 Doing this enabled the tournament to run for a full two weeks instead of the normal one week, resulting in a total of 20788 games being played - 90 full round robins between each bot pairing.
@@
 
-\subsubsection{CIG 2011}
+\subsubsection{CIG 2011.}
 
@@
 
-\subsubsection{CIG 2012}
+\subsubsection{CIG 2012.}
 
@@
 
-\subsubsection{CIG 2013} The CIG 2013 competition was once again
+\subsubsection{CIG 2013.} The CIG 2013 competition was once again
 organized by Mike Preuss (TU Dortmund) and Tobias Mahlmann (ITU Copenhagen),
@@
 
-\subsubsection{CIG 2014}
+\subsubsection{CIG 2014.}
 The CIG 2014 competition was organized by Kyung-Joong Kim, Ho-Chul Cho, and In-Seok Oh of Sejong University.
@@
 
-\subsubsection{CIG 2015}
+\subsubsection{CIG 2015.}
 There were some significant rule changes to the CIG 2015 competition, which was once organized by mebers from Sejong University.
@@
 
-\subsubsection{SSCAIT 2011}
+\subsubsection{SSCAIT 2011.}
 Taking place in winter of 2011, the first SSCAIT tournament was organized by Michal \v{C}ertick\'{y} as part of the {\em ``Introduction to Artificial Intelligence''} course at the Department of Applied Informatics, Comenius University in Bratislava, Slovakia.
@@
 1 games of StarCraft.
-Vast majority of the bots were implemented to execute hard-coded strategies and behaviour.
+Vast majority of the bots were implemented to execute hard-coded strategies and behavior.
 Participants were divided into 10 groups of 5 people, and 16 of them advanced into a double elimination bracket.
@@
 
-\subsubsection{SSCAIT 2012-2015}
+\subsubsection{SSCAIT 2012-2015.}
 In 2012, the SSCAIT became a truly international competition with 52 participants from numerous universities including 
@@
 The  sandbox phase has no winners.
-However, the bots may use this long phase to collect useful data about their opponent's behaviour.
+However, the bots may use this long phase to collect useful data about their opponent's behavior.
 
@@
   	  \item W.O.P.R.
-  	  \item UAlbertaBot, with the same win rate as W.O.P.R., but with a loss in a mutual game
+  	  \item UAlbertaBot, with the same win rate as W.O.P.R., but with a loss in a mutual game.
   	\end{enumerate}
@@
 In addition to bots written in C++, SSCAIT supports bots using Java interfaces to BWAPI: BWMirror and JNIBWAPI.
+
+\section{Conclusions}
+
+Competitions have historically been an important drive in artificial intelligence, with challenge tasks such as {\em Chess}, {\em Go}, autonomous vehicle driving, among others having resulted in many algorithmic and theoretical breakthroughs.
+For that reason, we believe RTS AI competitions are playing and will continue to play an important role in the field by both motivating new research and also being an incentive to attract new students and researchers to work on some of the key open challenges of RTS AI.
+
+This chapter has provided a summary of the results of the most three most important StarCraft AI competitions, organized during the past six years, as well as a summary of the architectures used by some of the bots that regularly participate in these competitions.
+Moreover, as evidenced by the poor results that the top bots in these competitions achieve when playing against good human players, a significant amount of open questions remain about how to design AI systems that can handle real-time adversarial domains such as StarCraft.
+Hopefully, these competitions can play an important role in motivating research programs that can ultimately answer these questions.
 

### 2015-12-16 ECGG_StarCraftCompetitions/ECGG_StarCraftCompetitions.tex
small fix to numbers + rewording sscai intro

@@
 change in the 2011 competition was the introduction of automated tournament-managing software running
-StarCraft games simultaneously on 20 computers, allowing a total of 1170 games to be played in the five days that the tournament ran. 
+StarCraft games simultaneously on 20 computers, allowing a total of 2340 games to be played in the five days that the tournament ran. 
 This increase in games played also allowed the tournament
@@
 on previous results. 6 of the 10 entrants used this feature to aid in strategy selection, including the
-top 4 finishers. More improvements to the tournament environment also meant that a total of 4240 games
+top 4 finishers. More improvements to the tournament environment also meant that a total of 8279 games
 could now be played in the same time period. Results are shown in Table \ref{tab:aiide2012}.
@@
 A total of 8 bots competed in the 2013 AIIDE competition with many of the same names from the 2012 competition, and some key updates were made to existing bots for 2013 which shook up the results from the previous years.
+A total of 5597 games were played during the 5 day tournament.
 UAlbertaBot took 1st place with a dominant 84.49\% win rate, with Skynet in 2nd with 72.77\%.
@@
 \subsection{SSCAIT}\label{sec:SSCAIT}
-Student StarCraft AI Tournament\footnote{\url{http://sscaitournament.com/}} (SSCAIT) started in 2011 at Comenius University in Bratislava, Slovakia and immediately became the competition with the highest numbers of participants.
+The Student StarCraft AI Tournament\footnote{\url{http://sscaitournament.com/}} (SSCAIT) started in 2011 at Comenius University in Bratislava, Slovakia and has been well known for being the Starcraft AI competition with the highest number of total participants.
+Started as a part of a course in artificial intellignece at Comenius University, each SSCAI tourament includes several dozen student submissions from this course in addition to submissions from across the globe.
 There are three fundamental differences between SSCAIT and the remaining two conferences:

### 2015-12-16 ECGG_StarCraftCompetitions/ECGG_StarCraftCompetitions.tex
Very small change in SSCAIT section intro.

@@
 The Student StarCraft AI Tournament\footnote{\url{http://sscaitournament.com/}} (SSCAIT) started in 2011 at Comenius University in Bratislava, Slovakia and has been well known for being the Starcraft AI competition with the highest number of total participants.
-Started as a part of a course in artificial intellignece at Comenius University, each SSCAI tourament includes several dozen student submissions from this course in addition to submissions from across the globe.
+Started as a part of a course in artificial intellignece at Comenius University, initial SSCAIT seasons included several dozen student submissions from this course in addition to submissions from across the globe.
 There are three fundamental differences between SSCAIT and the remaining two conferences:

### 2015-12-16 ECGG_StarCraftCompetitions/ECGG_StarCraftCompetitions.tex
Fix typos

@@
 interference.  BBQ,  UAlbertaBot and  AIUR, however use  an arbitrator
-({\em Game Commander'} in UAlbertaBot) that makes sure that modules do
+({\em Game Commander} in UAlbertaBot) that makes sure that modules do
 not send contradictory orders to  the same unit.  However, very little
@@
 This addition, combined with some additional bug fixes led to the victory.
-Skynet and Aiur both implemented strategy learning in 2013, which is evident from \ref{fig:aiide2013} in which both Skynet and Aiur's win percentage over time climb dramatically from the first few rounds of the tournament to the later rounds.
+Skynet and AIUR both implemented strategy learning in 2013, which is evident from \ref{fig:aiide2013} in which both Skynet and AIUR's win percentage over time climb dramatically from the first few rounds of the tournament to the later rounds.
 Ximp unfortunately crashed all games on the Fortress map and lost 10\% of its games for free as a result, which meant it could have possibly came in 2nd place if not for all those crashes.
@@
 The results of the 2014 competition were dramatically different from previous years.
-Since 2011 the top 3 bots in the AIIDE competition had been some permutation of Skynet, UAlbertaBot, and Aiur, and in 2014 the top 3 finishers were all completely different.
+Since 2011 the top 3 bots in the AIIDE competition had been some permutation of Skynet, UAlbertaBot, and AIUR, and in 2014 the top 3 finishers were all completely different.
 In first place was IceBot with a win rate of 85.85\%, which had a complete overhaul in terms of strategy and tactics from the previous year.
 IceBot had several strategies which it could implement, most starting with an early bunker defense and transitioning into attacking once a set of heuristics conditions had been met.
-IceBot's source code was originally based on Aiur's modular design, and changed to play the Terran race.
+IceBot's source code was originally based on AIUR's modular design, and changed to play the Terran race.
 In a close 2nd place finish was Ximp - which had fixed the crashes and logic errors which had plagued it the year before.
@@
 
-The 2015 AIIDE competition was the largest competition to date with 23 total bots competing.
+The 2015 AIIDE competition was  the largest competition to date hosted
+by a conference, with 23 total bots competing.
 The competition environment was also changed so that the tournament was now run on virtual machines instead of physical machines.
@@
 An interesting result from this tournament was that despite UAlbertaBot coming 4th place, it finished with a winning rate greater than 50\% vs each other bot.
-Aiur came in 5th place and was a clear demonstration of how learning over time can dramatically improve results in a tournament, going from 63\% win rate early in the competition to a final win rate of over 73\%.
+AIUR came in 5th place and was a clear demonstration of how learning over time can dramatically improve results in a tournament, going from 63\% win rate early in the competition to a final win rate of over 73\%.
 
@@
 4 & UAlbertaBot & 80.20\% \\
-5 & Aiur & 73.02\% \\
+5 & AIUR & 73.02\% \\
 \hline
@@
 and Xelnaga and BroodwarBotQ (from bracket B). All qualified bots play the
-Protoss faction. Most bots proved pretty stable, only Xelnaga and Protoss 
+Protoss faction. Most bots proved fairly stable, only Xelnaga and Protoss 
 Beast Jelly crashed relatively often (each in more than a quarter of the games). 
@@
 In some cases, neither bot was able to finish the other off completely,
-so that they went into a passive state. We manually ended such games after
-around 15 minutes and assigned victory to the bot that had obtained more
+so that they went into a passive state. Such games were manually ended after
+around 15 minutes and victory were assigned to the bot that had obtained more
 points as indicated on the end game screen.
@@
 The competition setup, documentation and results can be found
-in\footnote{\url{http://ls11-www.cs.tu-dortmund.de/rts-competition/StarCraft-cig2011}}.
+in the 2011 competition web page\footnote{\url{http://ls11-www.cs.tu-dortmund.de/rts-competition/StarCraft-cig2011}}.
 
@@
 The CIG 2014 competition was held just a few weeks before the AIIDE 2014 competition and so many of the entrants to the competition were identical, and the results definitely showed this.
-The top 4 bots were all the same as the 2014 AIIDE competition with IceBot coming 1st, Ximp in 2nd, LetaBot in 3rd and Aiur in 4th place.
+The top 4 bots were all the same as the 2014 AIIDE competition with IceBot coming 1st, Ximp in 2nd, LetaBot in 3rd and AIUR in 4th place.
 For descriptions of these bots please refer to the AIIDE 2014 competition description as none of the top-finishing bots had any significant changes made.
@@
 3 & LetaBot & 68.5\% \\
-4 & Aiur & 66.1\% \\
+4 & AIUR & 66.1\% \\
 5 & UAlbertaBot & 60.0\% \\ 
@@
 \subsubsection{CIG 2015.}
-There were some significant rule changes to the CIG 2015 competition, which was once organized by mebers from Sejong University.
+There were some significant rule changes to the CIG 2015 competition, which was once organized by members from Sejong University.
 The most significant rule change was that entrants no longer had to be open source, in an attempt to attract more competitors who may not want to open source their bots.
@@
 
-% As the figures show, the top three ranked bots in the competition (Skynet, Aiur and UalbertaBot) do not change their strategy at all depending on their opponent.
+% As the figures show, the top three ranked bots in the competition (Skynet, AIUR and UalbertaBot) do not change their strategy at all depending on their opponent.
 For example, the Skynet bot (both in 2011 and 2012), always uses the same opening (\emph{two\_gates}), except when playing a Terran opponent, when it uses \emph{nony}.

### 2015-12-16 ECGG_StarCraftCompetitions/ECGG_StarCraftCompetitions.tex
Add number of entrans and map pool for AIIDE 2011

@@
 David Churchill\footnote{\url{https://skatgame.net/mburo/sc2011/}}. Due to a lack of entrants in tournament categories 1-3 in the 2010 competition, it was
-decided that only the full game category would be played in the 2011 competition. Another important
+decided that only  the full game category would be  played in the 2011
+competition, with 13 entrants. Another important
 change in the 2011 competition was the introduction of automated tournament-managing software running
@@
 to prevent possible cheating and to promote healthy competition in future tournaments by giving
-newcomers and easier entry point by basing their design off of previous bots.
+newcomers  and  easier entry  point  by  basing  their design  off  of
+previous bots. All  maps used by the competition were  known long time
+by advance, so participants can test their bot on them.  These 10 maps
+remains the same for all AIIDE  competition, from 2011 until the last
+one  at the  time this  chapter  has been  written in  2015. Maps  are
+\emph{(2)Benzene},    \emph{(2)Destination}, \emph{(2)HeartbreakRidge},  \emph{(3)Aztec},  \emph{(3)TauCross},  \emph{(4)Andromeda},
+\emph{(4)CircuitBreaker}, \emph{(4)EmpireoftheSun}, \emph{(4)Fortress}
+and \emph{(4)Python}. The number between  parenthesis is the number of
+possible start points.
 

### 2015-12-16 ECGG_StarCraftCompetitions/ECGG_StarCraftCompetitions.tex
sscait section reformatted

@@
 \subsubsection{SSCAIT 2011.}
-Taking place in winter of 2011, the first SSCAIT tournament was organized by Michal \v{C}ertick\'{y} as part of the {\em ``Introduction to Artificial Intelligence''} course at the Department of Applied Informatics, Comenius University in Bratislava, Slovakia.
+%Taking place in winter of 2011, 
+The first SSCAIT tournament was organized by Michal \v{C}ertick\'{y} as part of the {\em ``Introduction to Artificial Intelligence''} course at the Department of Applied Informatics, Comenius University in Bratislava, Slovakia.
 It was inspired by the media coverage of AIIDE StarCraft AI competition held in 2010, but over the following years it diverged from AIIDE competition's format significantly.
@@
 2011 was the only year when the games were not streamed live. However, the replays were published in real time and the games from the elimination bracket were recorded and uploaded to YouTube.
+
+\begin{table}[!t]
+\caption{Ranking of the best bots of the SSCAIT 2012 competition (52 participants)}
+\label{tab:sscait2012}
+\centering
+\begin{tabular}{|c|c|c|c|}
+\hline
+\multicolumn{2}{|c|}{Student division} & \multicolumn{2}{|c|}{Mixed division}\\
+{\bfseries Rank} & {\bfseries Bot} & {\bfseries Rank} & {\bfseries Bot} \\
+\hline
+1 & Matej Isten\'{i}k's bot & 1 & ICEBot \\
+2 & Marcin Bartnicki's bot  & 2 & Marcin Bartnicki's bot \\
+3 & UAlbertaBot 			&   &  \\
+\hline
+\end{tabular}
+\end{table}
 
@@
 
-The results of SSCAIT seasons 2012-2015 follow:
-\begin{itemize}
-  \item {\bf 2012:} 52 participants. Student division winners: 
-  	\begin{enumerate}
-  	  \item Matej Isten\'{i}k's bot
-  	  \item Marcin Bartnicki's bot
-  	  \item UAlbertaBot
-  	\end{enumerate}
-  Mixed division winner of the elimination bracket of 8 bots: ICEBot, after defeating Marcin Bartnicki's bot. 
-  \item {\bf 2013:} 50 participants. Student division winner after 1190 round-robin games: 
-    \begin{enumerate}
-  	  \item Ximp
-  	  \item W.O.P.R.
-  	  \item UAlbertaBot
-  	\end{enumerate}
-  Mixed division winner of the elimination bracket of 8 bots: Krasi0, after defeating ICEBot.
-  \item {\bf 2014:} 42 participants. Student division winner after 861 round-robin games: 
-    \begin{enumerate}
-  	  \item LetaBot
-  	  \item W.O.P.R.
-  	  \item UAlbertaBot, with the same win rate as W.O.P.R., but with a loss in a mutual game.
-  	\end{enumerate}
-  Mixed division winner of the elimination bracket of 8 bots: LetaBot, after defeating Ximp.
-  \item {\bf 2015:} At the time of writing this, the SSCAIT 2015 is still in a sandbox phase.
+\begin{table}[!t]
+\caption{Ranking of the best bots of the SSCAIT 2013 competition (50 participants).
+The student division winner was determined by means of 1190 round-robin games}
+\label{tab:sscait2013}
+\centering
+\begin{tabular}{|c|c|c|c|}
+\hline
+\multicolumn{2}{|c|}{Student division} & \multicolumn{2}{|c|}{Mixed division}\\
+{\bfseries Rank} & {\bfseries Bot} & {\bfseries Rank} & {\bfseries Bot} \\
+\hline
+1 & Ximp & 1 & Krasi0 \\
+2 & W.O.P.R.  & 2 & ICEBot \\
+3 & UAlbertaBot 			&   &  \\
+\hline
+\end{tabular}
+\end{table}
+
+%The results of SSCAIT seasons 2012-2015 follow:
+%\begin{itemize}
+%  \item {\bf 2012:} 52 participants. Student division winners: 
+%  	\begin{enumerate}
+%  	  \item Matej Isten\'{i}k's bot
+%  	  \item Marcin Bartnicki's bot
+%  	  \item UAlbertaBot
+%  	\end{enumerate}
+%  Mixed division winner of the elimination bracket of 8 bots: ICEBot, after defeating Marcin Bartnicki's bot. 
+%  \item {\bf 2013:} 50 participants. Student division winner after 1190 round-robin games: 
+%    \begin{enumerate}
+%  	  \item Ximp
+%  	  \item W.O.P.R.
+%  	  \item UAlbertaBot
+%  	\end{enumerate}
+%  Mixed division winner of the elimination bracket of 8 bots: Krasi0, after defeating ICEBot.
+%  \item {\bf 2014:} 42 participants. Student division winner after 861 round-robin games: 
+%    \begin{enumerate}
+%  	  \item LetaBot
+%  	  \item W.O.P.R.
+%  	  \item UAlbertaBot, with the same win rate as W.O.P.R., but with a loss in a mutual game.
+%  	\end{enumerate}
+%  Mixed division winner of the elimination bracket of 8 bots: LetaBot, after defeating Ximp.
+  %\item {\bf 2015:}
+
+The results of SSCAIT seasons 2012-2014 are provided in the tables~\ref{tab:sscait2012}
+to \ref{tab:sscait2014}.  
+  At the time of writing this, the SSCAIT 2015 is still in a sandbox phase.
 45 bots have registered for the competitive phase so far, including two bots playing as ``Random'' race.
 The season will be concluded by an elimination bracket featuring 16 bots in 2015.
-\end{itemize}
+%\end{itemize}
 
@@
 In addition to bots written in C++, SSCAIT supports bots using Java interfaces to BWAPI: BWMirror and JNIBWAPI.
+
+\begin{table}[!t]
+\caption{Ranking of the best bots of the SSCAIT 2014 competition (42 participants).
+The student division winner was determined by means of 861 round-robin games}
+\label{tab:sscait2014}
+\centering
+\begin{tabular}{|c|c|c|c|}
+\hline
+\multicolumn{2}{|c|}{Student division} & \multicolumn{2}{|c|}{Mixed division}\\
+{\bfseries Rank} & {\bfseries Bot} & {\bfseries Rank} & {\bfseries Bot} \\
+\hline
+1 & LetaBot & 1 & LetaBot \\
+2 & W.O.P.R.  & 2 & Ximp \\
+3 & UAlbertaBot 			&   &  \\
+\hline
+\end{tabular}
+\end{table}
 

### 2015-12-17 ECGG_StarCraftCompetitions/ECGG_StarCraftCompetitions.tex
some final little fixes

@@
 \hline
-\multicolumn{2}{|c|}{Student division} & \multicolumn{2}{|c|}{Mixed division}\\
+\multicolumn{2}{|c|}{Student division} & \multicolumn{2}{|c|}{Mixed division}\\ \hline
 {\bfseries Rank} & {\bfseries Bot} & {\bfseries Rank} & {\bfseries Bot} \\

### 2016-02-11 ECGG_StarCraftCompetitions/ECGG_StarCraftCompetitions.tex
Add Synonyms, Definition and Cross-References sections

@@
 
-\section{State of the Art Bots for StarCraft}\label{sec:bot}
+\section*{Synonyms}
+
+Real-Time Strategy games; RTS games; StarCraft; Artificial Intelligence; AI; Game
+AI; Competition
+
+\section*{Definition}
+
+{\em Real-Time Strategy} (RTS) games  is a sub-genre of strategy games
+where  players  need to  build  an  economy (gathering  resources  and
+building a  base) and military  power (training units  and researching
+technologies)  in order  to defeat  their opponents  (destroying their
+army and base).  Artificial Intelligence  (AI) problems related to RTS
+games deal  with the  behavior of an  artificial player.   Since 2010,
+many international competitions  have been organized to  match AIs, or
+bots,  playing to  the RTS  game  StarCraft. This  chapter presents  a
+review of all  major international competitions from  2010 until 2015,
+and details some competing StarCraft bots.
+
+\section*{State of the Art Bots for StarCraft}\label{sec:bot}
 
@@
 
-%\subsection{RTS Bot Architectures}\label{sec:integration}
+%\subsection*{RTS Bot Architectures}\label{sec:integration}
 
@@
 
-%\subsection{Individual AI Techniques}\label{sec:techniques}
+%\subsection*{Individual AI Techniques}\label{sec:techniques}
 
@@
 
-\section{Recent StarCraft AI Competitions}\label{sec:competition}
+\section*{Recent StarCraft AI Competitions}\label{sec:competition}
 
@@
 
-\subsection{AIIDE}\label{sec:AIIDE}
-
-\subsubsection{AIIDE 2010.}
+\subsection*{AIIDE}\label{sec:AIIDE}
+
+\subsubsection*{AIIDE 2010.}
 
@@
 
-\subsubsection{AIIDE 2011.}
+\subsubsection*{AIIDE 2011.}
 
@@
 
-\subsubsection{AIIDE 2012.}
+\subsubsection*{AIIDE 2012.}
 
@@
 
-\subsubsection{AIIDE 2013.}
+\subsubsection*{AIIDE 2013.}
 
@@
 
-\subsubsection{AIIDE 2014.}
+\subsubsection*{AIIDE 2014.}
 
@@
 
-\subsubsection{AIIDE 2015.}
+\subsubsection*{AIIDE 2015.}
 
@@
 
-\subsection{CIG}
+\subsection*{CIG}
 \label{sec:cig2011}
@@
 
-\subsubsection{CIG 2011.}
+\subsubsection*{CIG 2011.}
 
@@
 
-\subsubsection{First Round}
+\subsubsection*{First Round}
 \label{sec:cig-first-round}
@@
 
-\subsubsection{Final Round}
+\subsubsection*{Final Round}
 \label{sec:cig-final-round}
@@
 
-\subsubsection{CIG 2012.}
+\subsubsection*{CIG 2012.}
 
@@
 
-\subsubsection{CIG 2013.} The CIG 2013 competition was once again
+\subsubsection*{CIG 2013.} The CIG 2013 competition was once again
 organized by Mike Preuss (TU Dortmund) and Tobias Mahlmann (ITU Copenhagen),
@@
 
-\subsubsection{CIG 2014.}
+\subsubsection*{CIG 2014.}
 The CIG 2014 competition was organized by Kyung-Joong Kim, Ho-Chul Cho, and In-Seok Oh of Sejong University.
@@
 
-\subsubsection{CIG 2015.}
+\subsubsection*{CIG 2015.}
 There were some significant rule changes to the CIG 2015 competition, which was once organized by members from Sejong University.
@@
 
-\subsection{SSCAIT}\label{sec:SSCAIT}
+\subsection*{SSCAIT}\label{sec:SSCAIT}
 The Student StarCraft AI Tournament\footnote{\url{http://sscaitournament.com/}} (SSCAIT) started in 2011 at Comenius University in Bratislava, Slovakia and has been well known for being the Starcraft AI competition with the highest number of total participants.
@@
 
-\subsubsection{SSCAIT 2011.}
+\subsubsection*{SSCAIT 2011.}
 %Taking place in winter of 2011, 
@@
 
-\subsubsection{SSCAIT 2012-2015.}
+\subsubsection*{SSCAIT 2012-2015.}
 In 2012, the SSCAIT became a truly international competition with 52 participants from numerous universities including 
@@
 
-\section{Conclusions}
+\section*{Conclusions}
 
@@
 
-% \subsection{StarCraft Bot Ladder}\label{sec:ladder}
+\section*{Cross References}
+In the Springer Encyclopedia of Computer Graphics an Games volume 1:
+\begin{itemize}
+\item  Santiago Ontañón,  Gabriel Synnaeve,  Alberto Uriarte,  Florian
+  Richoux, David Churchill and Mike Preuss, {\it RTS AI: Problems and Techniques}.
+\item Kazuki Yoshizoe and Martin Mueller, {\it Computer Go}.
+\end{itemize}
+
+% \subsection*{StarCraft Bot Ladder}\label{sec:ladder}
 
@@
 
-% \subsection{Adaptation Analysis}
+% \subsection*{Adaptation Analysis}
 
